---
title: "Pesticide Poisoning Surveillance in South Africa"
subtitle: "A Decision Analysis of Short-Term Policy Options for Improved Coordination and Response"
author: "D2P Project"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme:
      light: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(gt)
library(flextable)
library(ggplot2)
library(scales)

tp <- readRDS("tables_and_plots_v2.Rds")
```

# Executive Summary {.unnumbered}

::: {.callout-important appearance="simple"}
## The core question

South Africa already has the legal basis to act on pesticide poisoning — the constraint is **weak execution across existing systems**. Which combination of short-term institutional actions will most efficiently improve hotspot detection and response, and at what cost?
:::

This report evaluates **five policy options** (plus the status quo) for strengthening pesticide poisoning surveillance in South Africa. Each option targets a specific system failure identified through fishbone analysis and can be implemented **without new legislation**.

**Key findings under base-case assumptions:**

- The **status quo** detects and responds to approximately `r round(tp$base_case$effectiveness[tp$base_case$strategy_code == "SQ"], 1)` of an estimated 10 annual poisoning hotspots
- The **lowest-cost improvements** (Options 3 and 4) cost under R65,000/year and can be launched within weeks
- The **Full Package** (all four options) addresses `r round(tp$base_case$effectiveness[tp$base_case$strategy_code == "PKG"], 1)` hotspots — a `r round((tp$base_case$effectiveness[tp$base_case$strategy_code == "PKG"] / tp$base_case$effectiveness[tp$base_case$strategy_code == "SQ"] - 1) * 100)`% improvement over the status quo
- All parameters are **placeholders** — the model is designed to be updated as real data become available from the surveillance pilot

::: {.callout-tip appearance="simple"}
## How to use this document
This report serves three audiences:

1. **Decision-makers** (MAC, NDoH): Read Sections 2–3 for plain-language option comparisons
2. **Technical team** (NICD, NHLS analysts): Read Sections 4–7 for model details and sensitivity analysis
3. **Model developers**: Read Section 8 for instructions on updating parameters and extending the model
:::

# The Problem: Why Act Now?

## Burden of Pesticide Poisoning

From 2020 to 2024, three independent data sources paint a consistent picture:

| Indicator | Value | Source |
|:---|:---|:---|
| NMC notifications (total, 5 years) | `r format(tp$epi_context$nmc_total_2020_2024, big.mark = ",")` | NICD NMC system |
| Average annual NMC notifications | ~`r tp$epi_context$nmc_annual_avg` | NICD NMC system |
| NHLS BChE tests per year | ~`r format(tp$epi_context$bche_tests_per_year, big.mark = ",")` | NHLS laboratory data |
| Severe BChE inhibitions per year | ~`r format(tp$epi_context$bche_severe_per_year, big.mark = ",")` | NHLS laboratory data |
| PIH clinician calls (2019) | `r format(tp$epi_context$pih_calls_2019, big.mark = ",")` | Poisons Information Helpline |
| Seasonal peak | `r tp$epi_context$peak_months` | All three sources |

: Table: Pesticide poisoning burden in South Africa {.striped}

**Who is affected?**

- **Infants and toddlers (< 6 years)**: `r tp$epi_context$accidental_under6_pct`% of exposures are **accidental** — from environmental contact, contaminated food, or pesticides stored unsafely at home
- **Adolescent females (15–19 years)**: The highest age-standardised rate (ASR `r tp$epi_context$highest_asr_bche` per 100,000 for BChE cases), with `r tp$epi_context$selfharm_13_19_pct`% classified as **intentional self-harm**
- **Adult males (30+)**: Occupational exposures in agricultural settings

::: {.callout-warning}
## Under-counting is substantial
The ~690 NMC notifications per year almost certainly under-represent the true burden. BChE data alone show ~2,000 severe inhibitions annually. Evaluations of the NMC system describe incomplete reporting, incorrect data, and discrepancies between notification and laboratory records. The real number of poisonings may be **3–5 times higher** than what is currently captured.
:::

## Root Causes: What Can Be Fixed Now?

The policy brief's fishbone analysis identifies five institutional failures — none require new laws to address:

```{r}
#| label: fig-fishbone
#| fig-cap: "Root causes of surveillance failure and the policy options that address them"
#| fig-width: 10
#| fig-height: 5

fishbone <- tribble(
  ~cause, ~option, ~x, ~y,
  "Fragmented data\n(no single picture)", "Option 1", 1, 2,
  "Weak EHP feedback\n(investigations not returned)", "Option 2", 2, 1,
  "Unclear technical rules\n(codes/tests not mapped)", "Option 3", 3, 2,
  "Low awareness &\nvariable data use", "Option 4", 4, 1,
  "No standing hotspot\nproduct for joint action", "Full Package", 5, 2
)

ggplot(fishbone, aes(x = x, y = y)) +
  geom_segment(aes(xend = x, yend = 0), linewidth = 1.2, colour = "grey60") +
  geom_label(aes(label = cause), size = 2.8, fill = "#fff3e0",
             label.padding = unit(0.3, "lines")) +
  geom_label(aes(y = -0.3, label = option), size = 2.5, fontface = "bold",
             fill = "#e3f2fd", label.padding = unit(0.2, "lines")) +
  geom_hline(yintercept = 0, linewidth = 2, colour = "#1a237e") +
  annotate("text", x = 5.8, y = 0, label = "Surveillance\nFailure",
           fontface = "bold", size = 4, colour = "#b71c1c") +
  ylim(-0.8, 2.8) + xlim(0.3, 6.2) +
  theme_void() +
  theme(plot.margin = margin(5, 10, 5, 10))
```

# Policy Options: What Is on the Table? {#sec-options}

```{r}
#| label: tbl-profiles
#| tbl-cap: "Policy option profiles: what each option does, the problem it solves, and how quickly it can deliver"
tp$tbl_profiles
```

## Option 1 — Integrated Data-Sharing and Hotspot Pilot {#sec-opt1}

::: {.callout-note appearance="minimal"}
**In one sentence:** Combine NMC notifications, cholinesterase lab results, forensic signals and EHP field reports into a single analytic product that shows **where poisoning clusters are happening**.
:::

**What changes for stakeholders:**

- **For NICD analysts**: A routine (initially manual) process to pull data from NMC, NHLS, forensic pathways, and EHP systems into one dataset
- **For district health managers**: A periodic hotspot map showing which sub-districts have clustering events — enabling targeted response rather than waiting for individual case reports
- **For MAC members**: The first evidence product that demonstrates whether integrated surveillance can work before investing in automation

**What this requires:**

- A data-sharing SOP or MoU with each data custodian (NHLS for BChE/FCL, forensic pathology services)
- One dedicated analyst to run routine extracts and produce outputs
- Named leads at each custodian organisation

**Key assumptions in the model:**

| Parameter | Value | What it means |
|:---|:---:|:---|
| P(data access via SOP) | 0.70 | 70% chance the data-sharing agreements are established |
| P(data usable once accessed) | 0.65 | 65% chance extracted data is complete enough to analyse |
| P(hotspot detected) | 0.60 | Given usable data, 60% chance of identifying a real cluster |
| P(response triggered) | 0.65 | Given detection, 65% chance of follow-up action |
| Setup cost | R120,000 | SOP development, data mapping, initial analyst time |
| Annual operating | R70,000 | Ongoing analyst time, routine extracts and reports |

## Option 2 — Standardised EHP Investigation Reporting {#sec-opt2}

::: {.callout-note appearance="minimal"}
**In one sentence:** Give every Environmental Health Practitioner a **single short template** to report pesticide investigations, and require findings to be **fed back** to the surveillance team.
:::

**What changes for stakeholders:**

- **For EHPs**: One clear template replaces variable, person-dependent reporting — reduces ambiguity about what to collect
- **For the surveillance team**: EHP investigation data (source identification, exposure context, packaging information) flows back centrally for the first time
- **For district managers**: Visibility into what actions EHPs are taking and what they are finding

**What this requires:**

- A one-page investigation template (aligned with the 2022 NDoH guideline)
- A circular or service standard requiring submission back to the surveillance team
- Brief training on template completion

**Key assumptions in the model:**

| Parameter | Value | What it means |
|:---|:---:|:---|
| P(template adopted) | 0.75 | 75% of districts adopt the template |
| P(reporting complete) | 0.60 | 60% of investigations completed and fed back |
| P(hotspot detected from EHP data) | 0.50 | 50% detection rate — lower because EHP data alone covers fewer signals |
| P(response triggered) | 0.70 | 70% — higher because EHP data provides actionable source information |
| Setup cost | R80,000 | Template design, training, rollout |
| Annual operating | R60,000 | Quality assurance, template management |

## Option 3 — Technical Code/Test-Set Crosswalk {#sec-opt3}

::: {.callout-note appearance="minimal"}
**In one sentence:** A short technical exercise to create a **reference document** mapping lab test codes, ICD codes, and NMC case definitions so the system consistently identifies poisoning cases across platforms.
:::

**What changes for stakeholders:**

- **For laboratory staff and NMC focal persons**: Clear rules about which test results should trigger a poisoning NMC — reducing the current reliance on individual knowledge
- **For NICD**: Improved sensitivity and specificity of case identification, and a foundation for future automated queries
- **For the surveillance system**: Captures the full range of pesticide classes, not just organophosphates

**What this requires:**

- A 2-day working group with representatives from NICD, NHLS, and forensic pathology
- Documentation of the mapping and dissemination to relevant facilities

**Key assumptions in the model:**

| Parameter | Value | What it means |
|:---|:---:|:---|
| P(exercise completed) | 0.90 | Very high — this is a small, bounded technical task |
| P(rules adopted by NHLS/NMC) | 0.80 | High — institutional buy-in from participants |
| P(detection improvement alone) | 0.45 | Modest — crosswalk alone doesn't create new data flows |
| P(response triggered) | 0.55 | Modest — knowledge of correct codes doesn't directly trigger responses |
| Setup cost | R40,000 | Working group logistics, documentation |
| Annual operating | R15,000 | Periodic updates as codes change |

::: {.callout-tip}
## Highest feasibility, lowest cost
Option 3 scores "Very High" feasibility in the policy brief and has the lowest setup and operating costs. It is also an **enabling condition** for Options 1 and 2 — without agreed technical rules, integrated data and EHP reporting will have inconsistent case identification.
:::

## Option 4 — Notification and Data-Use Activation Package {#sec-opt4}

::: {.callout-note appearance="minimal"}
**In one sentence:** Use the next MAC meeting to **approve the pilot**, use D2P interviews to identify what outputs decision-makers need, and hold a mortality workshop to improve data interpretation.
:::

**What changes for stakeholders:**

- **For MAC members**: A formal decision point — approve the surveillance package, nominate a lead coordinator
- **For D2P researchers**: Structured interviews to understand what information format and frequency decision-makers actually need
- **For all stakeholders**: A mortality data analysis workshop to build shared interpretation skills

**What this requires:**

- Preparation of a MAC agenda item and decision memo
- D2P interview protocol design and execution
- Workshop logistics (can be combined with existing meetings)

**Key assumptions in the model:**

| Parameter | Value | What it means |
|:---|:---:|:---|
| P(awareness raised via MAC) | 0.85 | Very high — the platform already exists |
| P(D2P outputs actually used) | 0.50 | Uncertain — depends on output relevance and format |
| P(detection improvement alone) | 0.35 | Weakest standalone — activation without data infrastructure has limited effect |
| P(response triggered) | 0.60 | Moderate — awareness helps but doesn't replace data |
| Setup cost | R25,000 | MAC preparation, D2P design |
| Annual operating | R20,000 | Communication materials, follow-up meetings |

## Full Package — All Four Options Combined {#sec-pkg}

::: {.callout-note appearance="minimal"}
**In one sentence:** Implement all four options as a **coordinated 6-month package**, using the MAC meeting as the launch mechanism and the proof-of-concept integrated product as the centrepiece.
:::

The policy brief recommends this approach because the options are **complementary, not competing**:

- Option 3 (Crosswalk) creates the **technical rules** needed by Option 1
- Option 2 (EHP) provides the **field investigation data** that makes Option 1 actionable
- Option 4 (Activation) provides the **institutional mandate** and **demand signal** for all other options
- Option 1 (POC) produces the **integrated surveillance product** that demonstrates value

**Key assumptions in the model:**

| Parameter | Value | What it means |
|:---|:---:|:---|
| P(package implemented) | 0.65 | Coordination overhead reduces probability compared to individual options |
| P(surveillance improved) | 0.75 | Synergies between options amplify the effect |
| P(hotspot detected) | 0.75 | Highest detection — multiple data streams feeding one product |
| P(response triggered) | 0.75 | Highest response — clear institutional mandate |
| Setup cost | R180,000 | All four setup costs plus coordination |
| Annual operating | R100,000 | All operating costs plus coordinator time |

# Decision-Maker Summary {#sec-summary}

```{r}
#| label: tbl-stakeholder
#| tbl-cap: "Plain-language comparison of all options for non-technical decision-makers"
tp$tbl_stakeholder
```

::: {.callout-important}
## What the green rows mean
Options highlighted in green are estimated to be **cost-effective** — meaning the additional health gains justify the additional cost, measured against South Africa's cost-effectiveness threshold (R48,474 per DALY averted; Edoka & Stacey, 2020). However, all estimates use placeholder parameters and should be interpreted as indicative rather than definitive.
:::

```{r}
#| label: fig-frontier
#| fig-cap: "Cost-effectiveness frontier: strategies on the dashed line offer the best value; those above it are 'dominated' (more expensive for less benefit)"
#| fig-width: 9
#| fig-height: 6
tp$fig_frontier
```

::: {.callout-tip}
## Reading the frontier
The dashed line connects strategies that offer the **best trade-off** between cost and effectiveness. Strategies above this line are "dominated" — another strategy achieves the same or better results at lower cost. This helps identify which individual options are worth pursuing if the full package is not feasible.
:::

# Technical Model Details {#sec-model}

## Decision Tree Structure

```{r}
#| label: fig-tree
#| fig-cap: "Decision tree with six strategies. Each branch shows the probability pathway and expected outcomes."
#| fig-width: 11
#| fig-height: 7
tp$fig_tree
```

## Model Parameters

```{r}
#| label: tbl-params
#| tbl-cap: "All model input parameters with distributions for probabilistic sensitivity analysis"
tp$tbl_params
```

::: {.callout-tip}
## How to update parameters
Edit [`amua_import_parameters_v2.csv`](amua_import_parameters_v2.csv):

- **`Expression`** column → point estimate (used in base-case)
- **`Notes`** column → PSA distribution specification (e.g., `dist=beta(80,20)`)

Then run `Rscript wrangling_v2.r` and re-render this report.
:::

## Base-Case Results

```{r}
#| label: tbl-basecase
#| tbl-cap: "Base-case results: expected hotspots addressed and costs per strategy"
tp$tbl_basecase
```

## Health Outcomes (DALY Estimates)

```{r}
#| label: tbl-dalys
#| tbl-cap: "Estimated health gains assuming 30% case reduction per hotspot response"
tp$tbl_dalys
```

::: {.callout-caution}
## DALY estimates are illustrative
The conversion from hotspots to DALYs makes several assumptions:

- Each hotspot contains ~69 cases (690 annual cases ÷ 10 hotspots)
- A successful response prevents 30% of cases in that hotspot
- Each prevented case averts 2.5 DALYs (placeholder)
- Case fatality rate among hospitalised cases: 3%

**These should be refined** using data from the prospective surveillance study (concept note target: N = `r tp$epi_context$sample_size_required` patients) and the mortality data workshop.
:::

## Budget Impact Analysis

```{r}
#| label: tbl-bia
#| tbl-cap: "3-year budget impact (discounted at 5% p.a.)"
tp$tbl_bia
```

```{r}
#| label: fig-bia
#| fig-cap: "Total 3-year cost by strategy"
#| fig-width: 9
#| fig-height: 5
tp$fig_bia
```

::: {.callout-note}
## Budget context
The surveillance concept note estimates R`r format(tp$epi_context$surveillance_budget_low / 1e6, digits = 2)`M – R`r format(tp$epi_context$surveillance_budget_high / 1e6, digits = 2)`M for the full prospective surveillance study. The policy options modelled here are **much less expensive** because they use existing data and infrastructure rather than establishing new data collection.
:::

# Sensitivity Analysis {#sec-sensitivity}

## Scenario Analysis

```{r}
#| label: fig-scenarios
#| fig-cap: "How results change under optimistic and conservative assumptions"
#| fig-width: 10
#| fig-height: 7
tp$fig_scenarios
```

```{r}
#| label: tbl-scenarios
#| tbl-cap: "Results under base-case, optimistic (+20% probabilities, −15% costs) and conservative (−20% probabilities, +20% costs) assumptions"
tp$tbl_scenarios
```

::: {.callout-note}
## What the scenarios tell us
- **Optimistic**: If institutional buy-in is strong and costs come in low, all options become clearly cost-effective
- **Conservative**: Even with reduced effectiveness and higher costs, the Full Package still addresses more hotspots than any individual option
- **The key insight**: The ranking of strategies is stable across scenarios — the Full Package is consistently the most effective
:::

## Tornado Diagram (One-Way Sensitivity)

```{r}
#| label: fig-tornado
#| fig-cap: "Which parameters most influence the Full Package ICER (±25% variation)"
#| fig-width: 10
#| fig-height: 6
tp$fig_tornado
```

::: {.callout-tip}
## Priority parameters for investigation
Focus data collection on the parameters with the **widest bars** — these have the most influence on whether the Full Package represents good value. Specifically:

1. **`n_hotspots_baseline`** — How many actionable hotspots are there per year? The pilot POC will answer this directly
2. **`p_pkg_implemented`** — Will all four options actually be implemented together? This depends on MAC endorsement and coordinator capacity
3. **Cost parameters** — What are the actual analyst, EHP, and working group costs? These can be estimated from existing government pay scales
:::

## Probabilistic Sensitivity Analysis

```{r}
#| label: tbl-psa
#| tbl-cap: "PSA summary: mean results with 95% credible intervals from 1,000 simulations"
tp$tbl_psa
```

```{r}
#| label: fig-ce-scatter
#| fig-cap: "Cost-effectiveness plane showing parameter uncertainty across 1,000 simulations"
#| fig-width: 10
#| fig-height: 7
tp$fig_ce_scatter
```

```{r}
#| label: fig-ceac
#| fig-cap: "Probability each strategy is optimal at varying willingness-to-pay thresholds"
#| fig-width: 10
#| fig-height: 6
tp$fig_ceac
```

::: {.callout-note}
## Reading the CEAC
The vertical dotted line shows South Africa's cost-effectiveness threshold. At this threshold:

- If the **Full Package** line is above 50%, there is more than a coin-flip chance it represents the best value
- The point where the **Status Quo** line drops below 50% marks the WTP threshold above which doing nothing is no longer the cheapest optimal strategy
:::

## Strategy Ranking

```{r}
#| label: tbl-ranking
#| tbl-cap: "How often each strategy ranks first across 1,000 PSA iterations"
tp$tbl_ranking
```

# Implementation Roadmap {#sec-implementation}

```{r}
#| label: fig-timeline
#| fig-cap: "24-week implementation timeline showing practical steps by policy option"
#| fig-width: 11
#| fig-height: 7
tp$fig_timeline
```

::: {.callout-important}
## Critical path
The first **12 weeks** are the proof-of-concept phase:

1. **Weeks 1–2**: MAC approval and coordinator nomination (Option 4)
2. **Weeks 2–5**: Dataset mapping and crosswalk exercise (Options 1 & 3) — these run in parallel
3. **Weeks 4–9**: EHP template design and rollout (Option 2)
4. **Weeks 8–12**: Integrated POC begins producing hotspot outputs (Option 1)

**Week 12 milestone**: First hotspot map published. This is the evidence that the package works — and the basis for a scale-up decision.
:::

## Feedback Loop: Using Pilot Data to Refine the Model

The model parameters are placeholders. The implementation process generates the data to replace them:

| Implementation step | Model parameter updated | How |
|:---|:---|:---|
| Integrated POC outputs | `n_hotspots_baseline` | Count actual hotspots identified |
| EHP template submissions | `p_ehp_reporting_completeness` | Track submission rate |
| Crosswalk adoption | `p_xwalk_adopted` | Survey facilities on uptake |
| D2P interviews | `p_act_data_used` | Ask decision-makers if they use the outputs |
| Response tracking | `p_*_response_triggered` | Record whether detected hotspots are acted on |
| Budget actuals | All cost parameters | Replace estimates with actual expenditure |
| Prospective surveillance | `dalys_per_case`, `p_case_fatality` | Clinical data from pilot sites |

::: {.callout-tip}
## Updating the model
After the pilot:

1. Update values in [`amua_import_parameters_v2.csv`](amua_import_parameters_v2.csv)
2. Run `Rscript wrangling_v2.r` in the terminal
3. Re-render this report: `quarto render analysis_report_v2.qmd`
4. Compare the updated results with the current placeholder-based estimates

The model is designed to be a **living decision tool**, not a one-off analysis.
:::

# Extending the Model {#sec-extending}

::: {.callout-tip collapse="true"}
## Adding a new strategy
1. Add parameters to `amua_import_parameters_v2.csv` (probability chain + costs)
2. Add a `make_row()` call in `evaluate_tree()` in `methods_v2.r`
3. Re-run `wrangling_v2.r` — all tables, figures and sensitivity analyses update automatically
:::

::: {.callout-tip collapse="true"}
## Switching to DALY-based ICER
The model already calculates DALYs (Table 4). To make this the primary outcome:

1. Adjust `dalys_per_case`, `p_case_fatality`, `yld_per_survivor`, `yll_per_death` in the CSV
2. The ICER per DALY can then be directly compared to the SA threshold of R48,474
:::

::: {.callout-tip collapse="true"}
## Connecting to Amua
The CSV is formatted for direct import into [Amua](https://github.com/zward/Amua). You can:

- Build the decision tree graphically
- Validate the structure against this R model
- Use Amua's built-in PSA and then compare results
:::

::: {.callout-tip collapse="true"}
## Multi-year Markov extension
For a dynamic model capturing system learning over time:

1. Define health states: No surveillance ↔ Partial ↔ Full integration
2. Transition probabilities from implementation success rates
3. Discounting costs and health outcomes at 5% p.a. (SA Treasury rate)
4. This would be appropriate once pilot data are available (post-week 24)
:::

# Limitations

1. **All parameters are placeholders** — derived from policy documents and expert assumptions, not empirical data
2. **Hotspot-level outcomes** — the primary metric is "hotspots addressed", not individual cases prevented; DALY conversions are illustrative
3. **Single-period model** — does not capture system learning, capacity building, or scale-up effects over time
4. **No equity analysis** — does not weight for differential burden across provinces, age groups, or sexes
5. **Independence assumption** — each chance node is treated as conditionally independent; real-world implementation involves correlated successes and failures
6. **No dynamic feedback** — improved detection should increase reporting which should improve detection; this virtuous cycle is not modelled

Despite these limitations, **delaying action until a perfect model exists would be a mistake**. The current model provides a structured framework for comparing options and identifying which assumptions matter most — enabling focused data collection during the pilot phase.

# References

1. Mathee A, Stephen C, Nzenze SA, Chandu L. New case definitions and thresholds for environmental NMCs in SA. *SAMJ* 2025;115(11):e2565
2. Prinsloo M, Roomaney RA, Neethling I, et al. Tracking poison ingestion deaths for SA. *SAMJ* 2025;115(5):e2993
3. NDoH. Guideline for investigation and control of human chemical poisoning cases. Pretoria: NDoH; 2022
4. Rother HA. Improving poisoning diagnosis and surveillance of street pesticides. *SAMJ* 2012;102(6):485-488
5. Kabanda S, Rother HA. Evaluating a SA mobile application for healthcare professionals. *BMC Med Inform Decis Mak* 2019;19:40
6. Balme KH, et al. Pesticide poisonings at a tertiary children's hospital in SA. *Clin Toxicol* 2010;48(9):928-934
7. FAO/WHO. Guidelines on developing a reporting system for health and environmental incidents from pesticide exposure. 2009
8. London L, Bailie R. Challenges for improving surveillance for pesticide poisoning. *Int J Epidemiol* 2001;30(3):564-570
9. Edoka IP, Stacey NK. Estimating a cost-effectiveness threshold for SA. *Health Policy Plan* 2020;35(5):546-555
10. Buckley NA, et al. Case fatality of agricultural pesticides after self-poisoning in Sri Lanka. *Lancet Glob Health* 2021;9(6):e854-62

```{r}
#| label: session-info
sessionInfo()
```
