---
title: "Parameters Hub"
subtitle: "Canonical model parameters — single source of truth"
author: "D2P Project Team"
date-modified: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show parsing code"
---

::: {.callout-important icon=false}
## Editing parameters

**The single source of truth is the CSV file [`amua_import_parameters_v4.csv`](amua_import_parameters_v4.csv) at the project root.**

To update or verify a parameter:

1. Open `amua_import_parameters_v4.csv`.
2. Edit the relevant row. Each row has four logical fields packed into the `Notes` column: short note · `dist=<distribution>(<params>)` · source citation.
3. Save and re-render the site (`./publish.sh "msg"`). Every place this parameter is used — the [Decision Model](posts/decision_model.qmd), the [Costing Analysis](posts/costing_analysis.qmd), the three policy briefs ([Surveillance](posts/surveillance_policy.qmd) · [Coordination](posts/coordination_policy.qmd) · [Terbufos](posts/terbufos_policy.qmd)), and this hub — updates automatically.
4. If you change a *value* (not just a note), record the rationale in [`parameter_appraisal.qmd`](parameter_appraisal.qmd) under a dated entry.

**Naming convention.** `n_*` = count; `p_*` / `cfr_*` / `d_*` / `eta_*` = probability; `c_*` / `C_*` = cost (lowercase per-unit; uppercase aggregate); `los_*` = length of stay; `yll_*` / `yld_*` = years lost / disabled.

**v4 schema.** The CSV now has four columns: `Name, Expression, Notes, Group`. The `Group` column (e.g. `detect.S1_PIH_MVD`, `costs.S2_BChE`, `terb.burden`) encodes the domain and option each parameter belongs to — useful for programmatic filtering. Rows with `Group = costs.deprecated` or `detect.deprecated` are backward-compatibility aliases; **always use the canonical v4 name** (e.g. `c_S1_PIH_MVD_op`, `d_S2_BChE_autonotify`). Section header rows (`# === ... ===`) have an empty Group field.
:::

```{r}
#| label: load-params
#| message: false
#| warning: false

library(tidyverse)
library(gt)

raw <- readr::read_csv(
  "amua_import_parameters_v4.csv",
  show_col_types = FALSE,
  comment = ""
)

# Separate section header rows (start with "#") from parameter rows
params <- raw |>
  filter(!is.na(Name)) |>
  mutate(
    is_section = str_starts(Name, "#"),
    section = if_else(is_section,
                      str_remove_all(Name, "^#\\s*=*\\s*|\\s*=*\\s*$"),
                      NA_character_)
  ) |>
  fill(section, .direction = "down") |>
  filter(!is_section) |>
  separate(
    Notes,
    into = c("Note", "Distribution", "Source"),
    sep = "\\s*\\|\\s*",
    fill = "right",
    extra = "merge"
  ) |>
  mutate(
    Distribution = str_remove(Distribution, "^dist="),
    section = str_to_title(str_to_lower(section))
  ) |>
  select(Section = section, Parameter = Name, Value = Expression, Note, Distribution, Source)

section_table <- function(df, section_regex) {
  df |>
    filter(str_detect(Section, regex(section_regex, ignore_case = TRUE))) |>
    select(-Section) |>
    gt() |>
    cols_label(
      Parameter = "Name",
      Value = "Value",
      Note = "Description",
      Distribution = "Distribution",
      Source = "Source / status"
    ) |>
    tab_options(table.font.size = "small") |>
    opt_row_striping()
}
```

## 1. Burden parameters

Annual case counts across the three data streams (treated as mutually-exclusive), the true-burden anchor, and the overlap-correction parameter for sensitivity analysis.

```{r}
#| label: tbl-burden
#| tbl-cap: "Burden parameters — annual case counts and true-burden anchor."
section_table(params, "burden parameters")
```

**Interpretation.** Three administrative streams capture different severities. NMC (~1,013) counts clinically-notified cases (2023); NHLS (~1,779 severe inhibitions from ~10,626 BChE tests, 2023) counts severe laboratory cases; PIH (~1,158) counts clinician-initiated calls (2023). Each stream is independently sourced; the sum (~3,950) is the 2023 observed lower bound. The StatsSA-implied true annual burden is ~8,112 (mid-CFR scenario, `n_true_burden_from_statssa_mid`).

## 2. Case fatality rate

```{r}
#| label: tbl-cfr
#| tbl-cap: "Case fatality rate — stratified by severity, with headline range for advocacy."
section_table(params, "case fatality rate")
```

**Stratified vs. headline.** Applying the stratified CFR row-by-row gives **~830 deaths/year**; applying the flat headline 12.5 % gives **~1,460 deaths/year**; the Stats SA Vital Registration triangulation using the broader X40-X49 category previously gave **~960–2,620 deaths/year**. The headline figure is now anchored at **1,014 deaths/year** via **StatsSA MACOD 2022, Table 4.12, pesticide-specific UCOD codes (X48 accidental + X68 intentional self-poisoning + X87 assault + Y18 undetermined intent)** — CSV row `n_deaths_statssa_macod`, mirrored to `n_deaths_headline`. This is the **canonical truth source** for deaths and is updated as new StatsSA releases land; it supersedes both the v1 advocacy range of 1,500–1,800 and the v2/v3 X40-X49 all-noxious-substance figure of 2,620 (which was not pesticide-specific). 2023 pesticide-specific data has not yet been released, so 2022 is used as an interim anchor. The full v2 → v3 rationale is in [`parameter_appraisal.qmd`](parameter_appraisal.qmd).

## 3. Per-case costs

UPFS 2024 tariffs, drug costs, and case-mix-weighted blended morbidity cost.

```{r}
#| label: tbl-per-case-costs
#| tbl-cap: "Per-case cost components and the blended headline used in the decision model."
section_table(params, "per-case costs")
```

## 4. Mortality cost

Human-capital base case plus DALY-threshold and VSL alternatives shown for sensitivity.

```{r}
#| label: tbl-mortality-cost
section_table(params, "mortality cost")
```

## 5. Total economic burden

```{r}
#| label: tbl-burden-total
section_table(params, "total economic burden")
```

The two totals (conservative ~R1.34B vs. headline ~R1.61B) bracket the plausible range. Advocacy products use the headline; academic outputs report both.

## 6. Response efficacy ($\eta$)

```{r}
#| label: tbl-eta
section_table(params, "response efficacy")
```

## 7. Detection probabilities by option

```{r}
#| label: tbl-detection
#| tbl-cap: "Per-option detection probability — all active rows derived on the per-case scale from NICD + NHLS + PIH 2023 stream counts ÷ StatsSA MACOD mid-burden (8,112, 2022 pesticide-specific anchor). v1 hotspot-cluster rows retained as DEPRECATED for back-compat only. Full audit trail in the [Surveillance brief §6](posts/surveillance_policy.qmd#sec-decision-tree)."
section_table(params, "detection probabilities")
```

## 8. Option costs

```{r}
#| label: tbl-option-costs
#| tbl-cap: "Setup and operating costs for each surveillance option."
section_table(params, "option costs")
```

## 8a. Surveillance costing build-up — audit detail

The decision-tree scalars in §8 (e.g. `c_full_package_narrow_operating`) are the **conclusion**. The line-by-line **workings** — every FTE %, every salary anchor, every system line item — live in the editable workbook [`data/surveillance_costing.xlsx`](data/surveillance_costing.xlsx). The three tables below are pulled directly from that workbook at render time; **edit the .xlsx, re-render, and these tables (and the surveillance brief §7) update together**.

```{r}
#| label: costing-load
#| message: false
#| warning: false

library(readxl)

cost_path  <- "data/surveillance_costing.xlsx"
personnel  <- read_excel(cost_path, sheet = "personnel")
system_lvl <- read_excel(cost_path, sheet = "system")
allocation <- read_excel(cost_path, sheet = "allocation")

zar_fmt <- function(x) {
  scales::label_currency(prefix = "R", big.mark = ",", accuracy = 1)(x)
}
```

### 8a.1 Personnel — salary × FTE % allocated to pesticide surveillance

```{r}
#| label: tbl-cost-personnel-hub
#| tbl-cap: "Personnel costs: each role's annual cost-to-employer (CTE) × % of time on pesticide surveillance. Salary anchors: DPSA L10–11 2024, NHLS Patterson D/E 2024, DoH OSD scales 2024."

personnel |>
  mutate(
    `Annual CTE`       = zar_fmt(annual_cte_zar),
    `FTE %`            = sprintf("%.0f%%", fte_pct),
    `Allocated cost`   = zar_fmt(annual_cost_zar)
  ) |>
  select(Stream = stream, Role = role,
         `Annual CTE`, `FTE %`, `Allocated cost`,
         Rationale = rationale, Source = source) |>
  gt() |>
  grand_summary_rows(
    columns = `Allocated cost`,
    fns = list(Total ~ zar_fmt(sum(personnel$annual_cost_zar)))
  ) |>
  opt_row_striping() |>
  tab_options(table.font.size = "small")
```

### 8a.2 System — platforms, assays, courier, hosting

```{r}
#| label: tbl-cost-system-hub
#| tbl-cap: "System costs: unit cost × annual volume. BChE reagent and courier scale with assay volume; platform/hosting/telephony are fixed annual."

system_lvl |>
  mutate(
    `Unit cost`    = zar_fmt(unit_cost_zar),
    Units          = format(units_per_yr, big.mark = ",", scientific = FALSE),
    `Annual cost`  = zar_fmt(annual_cost_zar)
  ) |>
  select(Stream = stream, Item = item,
         `Unit cost`, Units, `Annual cost`,
         Rationale = rationale, Source = source) |>
  gt() |>
  grand_summary_rows(
    columns = `Annual cost`,
    fns = list(Total ~ zar_fmt(sum(system_lvl$annual_cost_zar)))
  ) |>
  opt_row_striping() |>
  tab_options(table.font.size = "small")
```

### 8a.3 Allocation keys — pesticide share of each stream's total throughput

```{r}
#| label: tbl-cost-allocation-hub
#| tbl-cap: "Allocation keys: pesticide cases ÷ total stream throughput. The BChE stream is 100% pesticide-relevant by construction (assay is ordered when OP/carbamate poisoning is suspected); NMC and PIH carry broad notification mandates so only a fraction is pesticide-attributable."

allocation |>
  mutate(
    `Pesticide cases / yr` = format(pesticide_count_yr, big.mark = ",", scientific = FALSE),
    `Total stream / yr`    = format(total_count_yr,    big.mark = ",", scientific = FALSE),
    `Allocation %`         = sprintf("%.2f%%", allocation_pct)
  ) |>
  select(Stream = stream, Basis = basis,
         `Pesticide cases / yr`, `Total stream / yr`, `Allocation %`,
         Rationale = rationale, Source = source) |>
  gt() |>
  opt_row_striping() |>
  tab_options(table.font.size = "small")
```

### 8a.4 Reconciliation — xlsx build-up vs. CSV scalars

This block proves the workbook and the CSV agree. If the discrepancy column is non-zero, **the model and the audit trail have drifted** — either the xlsx needs re-saving with current line items, or the CSV scalar needs updating to match.

```{r}
#| label: tbl-cost-reconciliation
#| tbl-cap: "Reconciliation: xlsx line-item build-up vs. CSV decision-tree scalars. The xlsx is the wider system view (NMC + PIH + BChE personnel + system, allocated to pesticide share); c_status_quo_program in the CSV is the NICD-only slice that already appears in the decision tree as the status-quo arm cost."

xlsx_total <- sum(personnel$annual_cost_zar) + sum(system_lvl$annual_cost_zar)

by_stream <- bind_rows(
  personnel  |> select(stream, annual_cost_zar) |> mutate(layer = "Personnel"),
  system_lvl |> select(stream, annual_cost_zar) |> mutate(layer = "System")
) |>
  group_by(stream) |>
  summarise(total = sum(annual_cost_zar), .groups = "drop")

csv_status_quo <- as.numeric(
  raw |> filter(Name == "c_status_quo_program") |> pull(Expression)
)
csv_pkg_narrow <- as.numeric(
  raw |> filter(Name == "c_full_package_narrow_operating") |> pull(Expression)
)

tibble::tibble(
  Item = c(
    paste0("xlsx — NMC stream (personnel + system)"),
    paste0("xlsx — PIH stream (personnel + system)"),
    paste0("xlsx — BChE stream (personnel + system)"),
    paste0("xlsx — Cross-cutting (data eng, biostats, warehouse)"),
    "xlsx — Grand total (status-quo, allocated to pesticide)",
    "CSV — c_status_quo_program (NICD slice only, used in tree)",
    "CSV — c_full_package_narrow_operating (MVP marginal uplift)"
  ),
  Value = c(
    by_stream |> filter(stream == "NMC")   |> pull(total) |> sum(),
    by_stream |> filter(stream == "PIH")   |> pull(total) |> sum(),
    by_stream |> filter(stream == "BChE")  |> pull(total) |> sum(),
    by_stream |> filter(stream == "Cross") |> pull(total) |> sum(),
    xlsx_total,
    csv_status_quo,
    csv_pkg_narrow
  )
) |>
  mutate(`Annual cost (ZAR)` = zar_fmt(Value)) |>
  select(Item, `Annual cost (ZAR)`) |>
  gt() |>
  tab_row_group(label = "Build-up (workings — data/surveillance_costing.xlsx)",
                rows = 1:5) |>
  tab_row_group(label = "Conclusion (decision-tree scalars — CSV)",
                rows = 6:7) |>
  row_group_order(groups = c("Build-up (workings — data/surveillance_costing.xlsx)",
                             "Conclusion (decision-tree scalars — CSV)")) |>
  opt_row_striping() |>
  tab_options(table.font.size = "small")
```

**How to read this.** The xlsx grand total is the **whole** annual cost of the existing surveillance system (every salary slice + every platform + every assay) allocated to pesticide work. The CSV's `c_status_quo_program` is the **NICD-only slice** that already sits in the decision tree's status-quo arm — the wider build-up is *not* double-counted there because PIH and NHLS are funded outside the NICD vote. The CSV's `c_full_package_narrow_operating` is the **marginal uplift** the MVP adds on top of the status-quo. Auditors can therefore check three independent things:

1. **xlsx grand total** matches the surveillance brief §7 headline ("status-quo R1,853,956/yr");
2. **xlsx NMC stream** is consistent with the CSV `c_status_quo_program` order-of-magnitude (the NICD slice within NMC + cross-cutting);
3. **CSV `c_full_package_narrow_operating`** is the figure the decision tree actually uses to compute incremental cost per detected case.

## 9. Whole-system induced demand

```{r}
#| label: tbl-induced
section_table(params, "whole-system")
```

## 10. Option 3b — sentinel toxicology

Research-surveillance investment, separately funded.

```{r}
#| label: tbl-opt6
section_table(params, "sentinel toxicology|option 6")
```

## 11. DALY monetisation

```{r}
#| label: tbl-daly
section_table(params, "daly monetisation")
```

## 12. Net monetary benefit (NMB) framework

```{r}
#| label: tbl-nmb
section_table(params, "net monetary benefit")
```

---

## Quality and verification status

```{r}
#| label: tbl-verification
#| tbl-cap: "Verification status of every parameter, derived from the Source field."

params |>
  mutate(
    Verification = case_when(
      str_detect(Source, regex("NICD|NHLS|StatsSA|Stats SA|UPFS|EML|Edoka|Buckley", ignore_case = TRUE)) ~ "Sourced (named institution / peer-reviewed)",
      str_detect(Source, regex("Derived", ignore_case = TRUE)) ~ "Derived from other parameters",
      str_detect(Source, regex("Sensitivity|Sum", ignore_case = TRUE)) ~ "Structural / sensitivity bound",
      str_detect(Source, regex("Decision model|Concept Note|Costing analysis|Internal", ignore_case = TRUE)) ~ "Internal model assumption",
      str_detect(Source, regex("Expert elicitation", ignore_case = TRUE)) ~ "Expert elicitation (to be re-estimated from pilot)",
      str_detect(Source, regex("Estimated", ignore_case = TRUE)) ~ "Estimated (requires verification)",
      TRUE ~ "Other / unspecified"
    )
  ) |>
  count(Verification, name = "Parameters") |>
  arrange(desc(Parameters)) |>
  gt() |>
  tab_header(title = "Where each parameter's authority comes from") |>
  opt_row_striping()
```

**Reading the verification table.** "Sourced" parameters are tied to a named institution or peer-reviewed paper. "Expert elicitation" parameters are placeholders awaiting pilot data. "Derived" parameters are calculated from others and only need to be re-derived when a primary input changes.

## Open items — to be retired into the CSV

The following items are referenced in the policy briefs but do **not yet have a row in `amua_import_parameters_v3.csv`** and should be added once the underlying data is collected:

| ID | Parameter | Used in | Owner |
|:---|:----------|:--------|:------|
| TBF-1 | Terbufos annual SA market volume (kg active ingredient) | [Terbufos brief](posts/terbufos_policy.qmd) | DALRRD Pesticide Registrar |
| TBF-2 | Terbufos attributable fraction of severe pesticide-poisoning cases | [Terbufos brief](posts/terbufos_policy.qmd) | NICD + NHLS toxicology (S3) |
| TBF-3 | SA-specific yield impact under leading substitutes (maize) | [Terbufos brief](posts/terbufos_policy.qmd) | ARC + grain industry |
| TBF-4 | Comparative hazard profile of leading substitutes | [Terbufos brief](posts/terbufos_policy.qmd) | DALRRD registration |
| COORD-1 | Per-node coordination transition probabilities Q1–Q7 | [Coordination brief](posts/coordination_policy.qmd) | NICD + DALRRD audit |
| COORD-2 | Joint loop-closure probability under C1–C5 package | [Coordination brief](posts/coordination_policy.qmd) | Pilot evaluation |
| SURV-1 | Live-patient chemical-toxicology sensitivity / specificity | [Surveillance brief](posts/surveillance_policy.qmd) | NHLS clinical chemistry |

Add each item as a new CSV row when the data becomes available, then remove it from this table.
