---
title: "Parameters Hub"
subtitle: "Canonical model parameters — single source of truth"
author: "D2P Project Team"
date-modified: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show parsing code"
---

::: {.callout-important icon=false}
## Editing parameters

**The single source of truth is the CSV file [`amua_import_parameters_v3.csv`](amua_import_parameters_v3.csv) at the project root.**

To update or verify a parameter:

1. Open `amua_import_parameters_v3.csv`.
2. Edit the relevant row. Each row has four logical fields packed into the `Notes` column: short note · `dist=<distribution>(<params>)` · source citation.
3. Save and re-render the site (`./publish.sh "msg"`). Every place this parameter is used — the [Decision Model](posts/decision_model.qmd), the [Costing Analysis](posts/costing_analysis.qmd), the three policy briefs ([Surveillance](posts/surveillance_policy.qmd) · [Coordination](posts/coordination_policy.qmd) · [Terbufos](posts/terbufos_policy.qmd)), and this hub — updates automatically.
4. If you change a *value* (not just a note), record the rationale in [`parameter_appraisal.qmd`](parameter_appraisal.qmd) under a dated entry.

**Naming convention.** `n_*` = count; `p_*` / `cfr_*` / `d_*` / `eta_*` = probability; `c_*` / `C_*` = cost (lowercase per-unit; uppercase aggregate); `los_*` = length of stay; `yll_*` / `yld_*` = years lost / disabled.
:::

```{r}
#| label: load-params
#| message: false
#| warning: false

library(tidyverse)
library(gt)

raw <- readr::read_csv(
  "amua_import_parameters_v3.csv",
  show_col_types = FALSE,
  comment = ""
)

# Separate section header rows (start with "#") from parameter rows
params <- raw |>
  filter(!is.na(Name)) |>
  mutate(
    is_section = str_starts(Name, "#"),
    section = if_else(is_section,
                      str_remove_all(Name, "^#\\s*=*\\s*|\\s*=*\\s*$"),
                      NA_character_)
  ) |>
  fill(section, .direction = "down") |>
  filter(!is_section) |>
  separate(
    Notes,
    into = c("Note", "Distribution", "Source"),
    sep = "\\s*\\|\\s*",
    fill = "right",
    extra = "merge"
  ) |>
  mutate(
    Distribution = str_remove(Distribution, "^dist="),
    section = str_to_title(str_to_lower(section))
  ) |>
  select(Section = section, Parameter = Name, Value = Expression, Note, Distribution, Source)

section_table <- function(df, section_regex) {
  df |>
    filter(str_detect(Section, regex(section_regex, ignore_case = TRUE))) |>
    select(-Section) |>
    gt() |>
    cols_label(
      Parameter = "Name",
      Value = "Value",
      Note = "Description",
      Distribution = "Distribution",
      Source = "Source / status"
    ) |>
    tab_options(table.font.size = "small") |>
    opt_row_striping()
}
```

## 1. Burden parameters

Annual case counts across the three data streams (treated as mutually-exclusive), the true-burden anchor, and the overlap-correction parameter for sensitivity analysis.

```{r}
#| label: tbl-burden
#| tbl-cap: "Burden parameters — annual case counts and true-burden anchor."
section_table(params, "burden parameters")
```

**Interpretation.** Three administrative streams capture different severities. NMC (~1,013) counts clinically-notified cases (2023); NHLS (~1,779 severe inhibitions from ~10,626 BChE tests, 2023) counts severe laboratory cases; PIH (~1,158) counts clinician-initiated calls (2023). Each stream is independently sourced; the sum (~3,950) is the 2023 observed lower bound. The StatsSA-implied true annual burden is ~20,960 (mid-CFR scenario, `n_true_burden_from_statssa_mid`).

## 2. Case fatality rate

```{r}
#| label: tbl-cfr
#| tbl-cap: "Case fatality rate — stratified by severity, with headline range for advocacy."
section_table(params, "case fatality rate")
```

**Stratified vs. headline.** Applying the stratified CFR row-by-row gives **~830 deaths/year**; applying the flat headline 12.5 % gives **~1,460 deaths/year**; the Stats SA Vital Registration triangulation gives **~960–2,620 deaths/year**. The headline figure is now anchored at **2,620 deaths/year** via **StatsSA MACOD 2023, Table 4.12 (X40-X49: Accidental poisoning by and exposure to noxious substance)** — CSV row `n_deaths_statssa_macod`, mirrored to `n_deaths_headline`. This is the **canonical truth source** for deaths and is updated as new StatsSA releases land; it supersedes the v1 advocacy range of 1,500–1,800. The full v2 → v3 rationale is in [`parameter_appraisal.qmd`](parameter_appraisal.qmd).

## 3. Per-case costs

UPFS 2024 tariffs, drug costs, and case-mix-weighted blended morbidity cost.

```{r}
#| label: tbl-per-case-costs
#| tbl-cap: "Per-case cost components and the blended headline used in the decision model."
section_table(params, "per-case costs")
```

## 4. Mortality cost

Human-capital base case plus DALY-threshold and VSL alternatives shown for sensitivity.

```{r}
#| label: tbl-mortality-cost
section_table(params, "mortality cost")
```

## 5. Total economic burden

```{r}
#| label: tbl-burden-total
section_table(params, "total economic burden")
```

The two totals (conservative ~R1.34B vs. headline ~R2.34B) bracket the plausible range. Advocacy products use the headline; academic outputs report both.

## 6. Response efficacy ($\eta$)

```{r}
#| label: tbl-eta
section_table(params, "response efficacy")
```

## 7. Detection probabilities by option

```{r}
#| label: tbl-detection
#| tbl-cap: "Per-option detection probability — currently expert-elicited, to be updated from pilot data."
section_table(params, "detection probabilities")
```

## 8. Option costs

```{r}
#| label: tbl-option-costs
#| tbl-cap: "Setup and operating costs for each surveillance option."
section_table(params, "option costs")
```

## 9. Whole-system induced demand

```{r}
#| label: tbl-induced
section_table(params, "whole-system")
```

## 10. Option 3b — sentinel toxicology

Research-surveillance investment, separately funded.

```{r}
#| label: tbl-opt6
section_table(params, "sentinel toxicology|option 6")
```

## 11. DALY monetisation

```{r}
#| label: tbl-daly
section_table(params, "daly monetisation")
```

## 12. Net monetary benefit (NMB) framework

```{r}
#| label: tbl-nmb
section_table(params, "net monetary benefit")
```

---

## Quality and verification status

```{r}
#| label: tbl-verification
#| tbl-cap: "Verification status of every parameter, derived from the Source field."

params |>
  mutate(
    Verification = case_when(
      str_detect(Source, regex("NICD|NHLS|StatsSA|Stats SA|UPFS|EML|Edoka|Buckley", ignore_case = TRUE)) ~ "Sourced (named institution / peer-reviewed)",
      str_detect(Source, regex("Derived", ignore_case = TRUE)) ~ "Derived from other parameters",
      str_detect(Source, regex("Sensitivity|Sum", ignore_case = TRUE)) ~ "Structural / sensitivity bound",
      str_detect(Source, regex("Decision model|Concept Note|Costing analysis|Internal", ignore_case = TRUE)) ~ "Internal model assumption",
      str_detect(Source, regex("Expert elicitation", ignore_case = TRUE)) ~ "Expert elicitation (to be re-estimated from pilot)",
      str_detect(Source, regex("Estimated", ignore_case = TRUE)) ~ "Estimated (requires verification)",
      TRUE ~ "Other / unspecified"
    )
  ) |>
  count(Verification, name = "Parameters") |>
  arrange(desc(Parameters)) |>
  gt() |>
  tab_header(title = "Where each parameter's authority comes from") |>
  opt_row_striping()
```

**Reading the verification table.** "Sourced" parameters are tied to a named institution or peer-reviewed paper. "Expert elicitation" parameters are placeholders awaiting pilot data. "Derived" parameters are calculated from others and only need to be re-derived when a primary input changes.

## Open items — to be retired into the CSV

The following items are referenced in the policy briefs but do **not yet have a row in `amua_import_parameters_v3.csv`** and should be added once the underlying data is collected:

| ID | Parameter | Used in | Owner |
|:---|:----------|:--------|:------|
| TBF-1 | Terbufos annual SA market volume (kg active ingredient) | [Terbufos brief](posts/terbufos_policy.qmd) | DALRRD Pesticide Registrar |
| TBF-2 | Terbufos attributable fraction of severe pesticide-poisoning cases | [Terbufos brief](posts/terbufos_policy.qmd) | NICD + NHLS toxicology (S3) |
| TBF-3 | SA-specific yield impact under leading substitutes (maize) | [Terbufos brief](posts/terbufos_policy.qmd) | ARC + grain industry |
| TBF-4 | Comparative hazard profile of leading substitutes | [Terbufos brief](posts/terbufos_policy.qmd) | DALRRD registration |
| COORD-1 | Per-node coordination transition probabilities Q1–Q7 | [Coordination brief](posts/coordination_policy.qmd) | NICD + DALRRD audit |
| COORD-2 | Joint loop-closure probability under C1–C5 package | [Coordination brief](posts/coordination_policy.qmd) | Pilot evaluation |
| SURV-1 | Live-patient chemical-toxicology sensitivity / specificity | [Surveillance brief](posts/surveillance_policy.qmd) | NHLS clinical chemistry |

Add each item as a new CSV row when the data becomes available, then remove it from this table.
