---
title: "Technical Appendix B — Coordination Model"
subtitle: "Closed-form joint-probability model for the NICD ↔ EHP ↔ DALRRD response loop"
author: "D2P Project"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

::: {.callout-note}
## Where this appendix fits

This appendix is the **formal quantitative model** behind the [Coordination brief](posts/coordination_policy.qmd). It treats the response loop as a chain of conditionally independent steps, gives the closed-form for the joint probability that the loop fully closes, and pulls the per-node parameters from the canonical CSV (`p_loop_closed_*`, `c_C1…C5_op`).

**Sister appendices:** [A — Surveillance decision analysis](analysis_report_v2.qmd) · [C — Terbufos bracketed estimate](technical_appendix_terbufos.qmd) · [D — Sentinel toxicology](posts/surveillance_concept_note.qmd).
:::

```{r}
#| label: setup
library(tidyverse)
library(gt)
library(scales)

raw <- readr::read_csv("amua_import_parameters_v3.csv", show_col_types = FALSE)
pick <- function(name) {
  v <- raw |> dplyr::filter(Name == name) |> dplyr::pull(Expression)
  if (length(v) == 0) NA_real_ else as.numeric(v[1])
}
zar <- function(x) {
  n <- suppressWarnings(as.numeric(x))
  if (is.na(n)) return(x)
  if (n >= 1e6) sprintf("R%.2f million", n / 1e6)
  else sprintf("R%s", format(round(n), big.mark = ",", scientific = FALSE))
}
```

# 1. The chain

A pesticide-poisoning case must traverse eight conditionally independent steps for the response loop to close:

1. Clinician suspects and notifies (NMC)
2. EHP receives SMS and opens investigation
3. EHP completes field investigation in <14 days
4. EHP outcome is recorded back on NMC
5. NICD reviews the notification and routes it to investigation
6. NICD refers to DALRRD via structured channel
7. DALRRD acts (inspection / withdrawal / formulation review)
8. Action and outcome are fed back to clinician + NICD

Let $p_i$ denote the probability that step $i$ succeeds, conditional on step $i-1$. The joint probability the loop closes is:

$$
P(\text{loop closed}) \;=\; \prod_{i=1}^{8} p_i
$$

Under status quo, several $p_i$ are near zero (especially steps 4, 6, 8 — none recorded back on NMC), giving an aggregate $P(\text{loop closed}) \approx$ **`r sprintf("%.1f%%", 100 * pick("p_loop_closed_status_quo"))`** (CSV: `p_loop_closed_status_quo`).

# 2. Scenarios

```{r}
#| label: tbl-scenarios
tibble::tribble(
  ~Scenario,                     ~`P(loop closed)`,                           ~`Annual operating cost`,
  "Status quo",                  pick("p_loop_closed_status_quo"),            0,
  "C3 only (NICD→DALRRD feed)",  pick("p_loop_closed_C3_only"),               pick("c_C3_intel_feed_op"),
  "Coordination Full Package",   pick("p_loop_closed_FP"),                    pick("c_coord_FP_op")
) |>
  dplyr::mutate(
    `P(loop closed)` = scales::percent(`P(loop closed)`, accuracy = 0.1),
    `Annual operating cost` = sapply(`Annual operating cost`, zar)
  ) |>
  gt() |>
  tab_header(title = "Three operating points",
             subtitle = "All values pulled from amua_import_parameters_v3.csv") |>
  opt_row_striping()
```

The **Coordination Full Package** is the sum of five no-new-legislation sub-options (C1 NMC closure-of-loop field; C2 structured DALRRD referral form; C3 NICD→DALRRD intelligence feed; C4 clinician auto-feedback; C5 joint KPI dashboard). It moves $P(\text{loop closed})$ from `r sprintf("%.1f%%", 100 * pick("p_loop_closed_status_quo"))` to `r sprintf("%.0f%%", 100 * pick("p_loop_closed_FP"))` — a **~`r round(pick("p_loop_closed_FP") / pick("p_loop_closed_status_quo"))`× lift** for `r zar(pick("c_coord_FP_op"))`/year.

# 3. Per-node cost table

```{r}
#| label: tbl-c-options
tibble::tribble(
  ~Code, ~Component,                                  ~`Operating cost`,
  "C1",  "NMC closure-of-loop field",                  pick("c_C1_NMC_closure_op"),
  "C2",  "Structured DALRRD referral form",            pick("c_C2_referral_form_op"),
  "C3",  "NICD→DALRRD intelligence feed",              pick("c_C3_intel_feed_op"),
  "C4",  "Clinician auto-feedback loop",               pick("c_C4_clinician_fb_op"),
  "C5",  "Joint KPI dashboard",                        pick("c_C5_KPI_dashboard_op"),
  "—",   "Full Package (sum)",                         pick("c_coord_FP_op")
) |>
  dplyr::mutate(`Operating cost` = sapply(`Operating cost`, zar)) |>
  gt() |>
  tab_header(title = "Coordination sub-options",
             subtitle = "Per-year operating cost; all under R150k aggregate") |>
  opt_row_striping()
```

# 4. One-way sensitivity

The joint product structure means $P(\text{loop closed})$ is bounded above by the weakest link. The two weakest links under status quo are step 4 (EHP outcome recorded back on NMC) and step 8 (feedback to clinician). C1 directly targets step 4; C4 directly targets step 8. Their combined effect is reflected in `p_loop_closed_FP`.

A reasonable sensitivity range, holding all other $p_i$ fixed at Full Package values, is:

- **Lower:** $P(\text{loop closed}) = 0.50$ (if C4 clinician feedback adoption is poor)
- **Mid:**   $P(\text{loop closed}) = 0.70$ (`p_loop_closed_FP`)
- **Upper:** $P(\text{loop closed}) = 0.85$ (if KPI dashboard creates accountability pressure)

# 5. Limitations

- The eight steps are treated as **conditionally independent**. In practice steps 6–8 are correlated (a poorly-functioning NICD→DALRRD channel hurts both).
- Per-node probabilities are **placeholder Beta priors**. The Coordination pilot is designed to produce posterior estimates within 12 months.
- The model assumes the surveillance signal is already in place. The Detect investments are covered in [Technical Appendix A](analysis_report_v2.qmd) and [Technical Appendix D](posts/surveillance_concept_note.qmd).

---

::: {.footer-cta}
**Brief:** [Coordination policy brief](posts/coordination_policy.qmd) · **Sister appendices:** [A](analysis_report_v2.qmd) · [C](technical_appendix_terbufos.qmd) · [D](posts/surveillance_concept_note.qmd)
:::
