---
title: "Technical Appendix C — Terbufos Bracketed Deaths-Averted Estimate"
subtitle: "Closed-form attributable-fraction model for the T2 restriction option"
author: "D2P Project"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

::: {.callout-note}
## Where this appendix fits

This appendix is the **formal quantitative basis** for the bracketed 200–400 deaths-averted-per-year estimate cited in the [Terbufos brief](posts/terbufos_policy.qmd). It is deliberately a closed-form attributable-fraction calculation — not a decision tree — because the upstream uncertainty is dominated by the attributable fraction itself.

**Sister appendices:** [A — Surveillance decision analysis](analysis_report_v2.qmd) · [B — Coordination model](technical_appendix_coordination.qmd) · [D — Sentinel toxicology](posts/surveillance_concept_note.qmd).
:::

```{r}
#| label: setup
library(tidyverse)
library(gt)

raw <- readr::read_csv("amua_import_parameters_v3.csv", show_col_types = FALSE)
pick <- function(name) {
  v <- raw |> dplyr::filter(Name == name) |> dplyr::pull(Expression)
  if (length(v) == 0) NA_real_ else as.numeric(v[1])
}
zar <- function(x) {
  n <- suppressWarnings(as.numeric(x))
  if (is.na(n)) return(x)
  if (n >= 1e6) sprintf("R%.2f million", n / 1e6)
  else sprintf("R%s", format(round(n), big.mark = ",", scientific = FALSE))
}

f_lo  <- pick("frac_terbufos_attributable_lo")
f_mid <- pick("frac_terbufos_attributable_mid")
f_hi  <- pick("frac_terbufos_attributable_hi")
N     <- pick("n_deaths_headline")
```

# 1. The formula

Annual deaths averted by removing terbufos from informal channels:

$$
\widehat{D}_{T2} \;=\; f_\text{attr} \times \varphi_\text{retired} \times N_\text{deaths}
$$

where

- $f_\text{attr}$ = terbufos attributable fraction of fatal SA pesticide poisonings (CSV: `frac_terbufos_attributable_{lo,mid,hi}` = `r f_lo`, `r f_mid`, `r f_hi`)
- $\varphi_\text{retired}$ = fraction of terbufos exposure successfully retired by enforcement (substitution-adjusted; placeholder 0.6 working, 0.4–0.8 sensitivity)
- $N_\text{deaths}$ = headline annual deaths (CSV: `n_deaths_headline` = `r format(N, big.mark = ",")`)

# 2. Bracketed estimate

```{r}
#| label: tbl-bracket
phi_grid <- c(lo = 0.4, mid = 0.6, hi = 0.8)

tibble::tibble(
  Scenario = c("Lower bound (conservative)", "Working value", "Upper bound (optimistic)"),
  `f_attr` = c(f_lo, f_mid, f_hi),
  `φ_retired` = c(0.4, 0.6, 0.8),
  `N_deaths` = N
) |>
  dplyr::mutate(
    `Deaths averted/yr` = round(`f_attr` * `φ_retired` * `N_deaths`)
  ) |>
  gt() |>
  fmt_percent(columns = c(`f_attr`, `φ_retired`), decimals = 0) |>
  fmt_number(columns = c(`N_deaths`, `Deaths averted/yr`), decimals = 0) |>
  tab_header(
    title = "Bracketed deaths averted per year, Option T2 (restriction + enforcement)",
    subtitle = "Working bracket cited in the Terbufos brief: ~200–400/yr"
  ) |>
  opt_row_striping()
```

Working value: $\widehat{D}_{T2}^{\text{mid}} = `r f_mid` \times 0.6 \times `r N` = `r round(f_mid * 0.6 * N)`$ deaths/yr.

Honest bracket: `r round(f_lo * 0.4 * N)` (lower) — `r round(f_hi * 0.8 * N)` (upper).

# 3. Cost side

```{r}
#| label: tbl-cost
tibble::tibble(
  Scenario   = c("Lower", "Working", "Upper"),
  `Enforcement cost/yr` = c(pick("c_T2_RUP_enforcement_lo"),
                            pick("c_T2_RUP_enforcement_mid"),
                            pick("c_T2_RUP_enforcement_hi"))
) |>
  dplyr::mutate(`Enforcement cost/yr` = sapply(`Enforcement cost/yr`, zar)) |>
  gt() |>
  tab_header(title = "Option T2 enforcement cost",
             subtitle = "CSV: c_T2_RUP_enforcement_{lo,mid,hi} — US EPA RUP cost scaled to SA") |>
  opt_row_striping()
```

# 4. Cost per death averted

Working value: cost/death-averted = `r zar(pick("c_T2_RUP_enforcement_mid") / round(f_mid * 0.6 * N))`.

This is well below any reasonable willingness-to-pay anchor for SA (e.g. 1× GDP/capita per DALY ≈ R85,000–R130,000), so **T2 dominates "do nothing" even at the lower bound of deaths averted.**

# 5. International anchors

- **Sri Lanka:** Single-agent withdrawals of class I OPs (1995–2011) reduced national suicide rate by ~50% with no compensatory substitution to other methods (Gunnell et al. 2017).
- **Bangladesh:** Similar withdrawal pattern, similar mortality reduction.
- **South Africa:** Davies et al. 2023 — terbufos dominates paediatric fatal pesticide poisonings in Soweto.

The Sri Lanka and Bangladesh natural experiments are why we treat $f_\text{attr} \in [0.20, 0.35]$ as defensible without a new SA-specific autopsy study.

# 6. Maize yield impact under a terbufos ban {#sec-yield-impact}

This section underpins the brief paragraph in §7 of the [Terbufos brief](posts/terbufos_policy.qmd). The headline conclusion — **<5 % yield impact, concentrated in non-Bt smallholder maize** — rests on three multiplicative factors.

## 6.1 Closed form

$$
\Delta Y \;=\; s_{\text{non-Bt}} \times u_{\text{terbufos}} \times \delta_{\text{substitute}}
$$

where

- $s_{\text{non-Bt}}$ = share of SA maize area not under Bt-trait varieties (~15 %; the Bt trait covers the dominant stalk-borer use case biologically).
- $u_{\text{terbufos}}$ = share of that non-Bt area where terbufos is the *current* soil-insect / nematode treatment (≤50 %; the rest already uses substitutes or no treatment).
- $\delta_{\text{substitute}}$ = residual per-hectare yield gap when terbufos is replaced by the leading registered substitute (chlorantraniliprole, thiamethoxam, imidacloprid, or biological control). International field-trial evidence: 0–10 %, working value ~5 %.

Working value: $\Delta Y \approx 0.15 \times 0.50 \times 0.05 = 0.375 \%$ of national maize output. Plausible upper bound at $s=0.20$, $u=0.70$, $\delta=0.10$: $\Delta Y \approx 1.4 \%$. The brief's "<5 %" headline is comfortably conservative.

## 6.2 Why the cost is bounded

1. **Bt-trait penetration.** SA commercial maize is ~85 % Bt — the dominant stalk-borer use case is retired biologically before any chemical intervention. This is the single largest reason the SA yield exposure is small relative to the equivalent EU 2009 case.
2. **Substitute acute-hazard profile.** Substitutes are 100–1,000× less acutely toxic on rat oral LD50 (chlorantraniliprole U, thiamethoxam II, imidacloprid II vs terbufos Ia). The acute-poisoning burden retired by withdrawal is therefore the **full** terbufos attributable share, not a redistribution to a similar-hazard substitute.
3. **EU 2009 natural experiment.** Terbufos was withdrawn in the EU more than 15 years ago. No documented maize-yield collapse in any producer state. SA's exposure is *smaller* than the EU's was because of Bt penetration.

## 6.3 Input-cost delta

Substitutes cost roughly 2–5× per hectare vs commodity granular OPs. Maize gross output is ~R20,000/ha. The substitute-cost delta is therefore single-digit-% of gross output — meaningful for smallholder margins but not for the national maize balance sheet.

## 6.4 Smallholder transition support

The remaining concentration of impact on non-Bt smallholder maize is addressed by a two-season DALRRD grower-advisory + targeted subsidy on substitute inputs. Cost is bounded by the size of the non-Bt smallholder pool and the substitute-cost delta, and is small relative to the household-borne mortality cost retired by the ban (~R1.4–2.6B/yr at the working bracket).

## 6.5 Sensitivity

| Driver | Plausible range | Effect on $\Delta Y$ |
|:-------|:----------------|:--------------------:|
| Bt-trait coverage $s_{\text{non-Bt}}$ | 0.10 – 0.20 | linear |
| Current terbufos use on non-Bt area $u_{\text{terbufos}}$ | 0.30 – 0.70 | linear |
| Substitute yield gap $\delta_{\text{substitute}}$ | 0.00 – 0.10 | linear |
| **Combined upper bound** | — | **~1.4 %** |
| **Combined working value** | — | **~0.4 %** |

: Three-factor sensitivity for the SA maize yield impact of a terbufos ban. Even the joint upper bound is well within the brief's "<5 %" headline. {.striped .hover}

# 7. Limitations

- $\varphi_\text{retired}$ is the **biggest single uncertainty**. Substitution to other class I OPs would shrink this dramatically. Restricting the whole class (not just terbufos) is the obvious mitigation.
- The model assumes the **detection signal exists**: without sentinel toxicology ([Appendix D](posts/surveillance_concept_note.qmd)) and a working coordination loop ([Appendix B](technical_appendix_coordination.qmd)), the regulatory action can't fire on time.
- No discounting applied — the 200–400/yr estimate is a steady-state annual figure.

---

::: {.footer-cta}
**Brief:** [Terbufos policy brief](posts/terbufos_policy.qmd) · **Sister appendices:** [A](analysis_report_v2.qmd) · [B](technical_appendix_coordination.qmd) · [D](posts/surveillance_concept_note.qmd)
:::
