---
title: "Policy Brief 2: Closing the Coordination Loop"
subtitle: "Making NMC, EHPs, DALRRD, and provincial response work as one system"
author: "D2P Project Team"
date: "2025-09-05"
date-modified: today
categories: [policy, coordination, decision-makers]
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: false
---

::: {.callout-important icon=false}
## For Decision-Makers

**Target audience:** Ministerial Advisory Committee (MAC), National Department of Health (NDoH), DALRRD Pesticide Registrar, Provincial Health Leadership.

**Companion briefs:** [Surveillance](surveillance_policy.qmd) · [Terbufos restriction](terbufos_policy.qmd).
:::

## 1. Policy question

**Should NDoH and DALRRD mandate a closed-loop coordination workflow — where every NMC notification triggers an investigation that is recorded on NMC, every EHP→DALRRD referral is structured, and every DALRRD action is fed back to NICD and the clinician?**

## 2. Public health problem

Better detection (see the [Surveillance brief](surveillance_policy.qmd)) is wasted if the system does not **act** on what it detects. The response chain has **eight sequential decision points** between "case suspected" and "regulatory action closes the loop"; **six of the eight default to a broken-loop terminal**. A correctly-notified case can still produce no investigation, no source removal, and no regulatory follow-up — leaving the underlying hazard in place and producing predictable secondary cases.

The coordination loop is **agent-agnostic**: it processes terbufos clusters today, will process aldicarb or paraquat tomorrow, and — as South Africa's burden shifts with development — opioids, methadone, or recreational-drug overdoses thereafter. The closed loop is therefore the durable institutional asset; the [Terbufos brief](terbufos_policy.qmd) is the first product-specific output it should deliver, not the last.

## 3. Status quo

{{< include _shared_parameters.qmd >}}

::: {.callout-note}
## Who bears the R2.34B

NDoH ~R70M (3 %), NHLS ~R10M (<1 %), households ~R10M OOP (<1 %), and ~R2.25B (96 %) in lost lifetime earnings borne by households and the wider economy. DALRRD bears no direct burden-side cost but holds **the only legal lever that prevents the underlying exposure**. That asymmetry is why the loop must terminate in DALRRD action: the agency with the lowest burden share controls the highest-leverage intervention.
:::

### The notification pathway — one-way and broken

```{mermaid}
%%| label: fig-coord-status-quo-flow
%%| fig-cap: "Current notification pathway. Clinicians and public-health officials notify on the NMC platform; NMC sends an SMS to the local EHP, who is mandated to investigate but does not record the investigation back on NMC. The EHP may informally refer to DALRRD, but this is not captured on NMC and DALRRD's action (if any) is not fed back to NMC or the clinician."
flowchart LR
    A[Clinician /<br/>Public-health official] -->|Notification on<br/>NMC platform| D[NMC platform<br/><i>NICD</i>]
    D -->|SMS notification| B[Local EHP<br/><i>mandated to investigate</i>]
    B -.Investigation NOT<br/>recorded on NMC.-x D
    B -.Informal phone /<br/>email referral.-> C[DALRRD<br/><i>Pesticide Registrar /<br/>Inspectorate</i>]
    B -.Referral NOT<br/>recorded on NMC.-x D

    C -.Feedback to EHP<br/>at best.-> B
    C -.x No feedback to NMC.-x D
    C -.x No feedback to clinician.-x A
    D -.x No case-status<br/>feedback.-x A

    style A fill:#fde,stroke:#333
    style B fill:#ffd,stroke:#333
    style C fill:#fdd,stroke:#cc0000,stroke-dasharray: 5 5
    style D fill:#cfc,stroke:#333,stroke-width:2px
```

**Reading the diagram.** Clinicians *do* file on NMC — that is the formal entry point. NMC *does* SMS the local EHP, who *is* mandated to investigate. The chain breaks at two specific places: (a) the EHP investigation outcome is **never written back to NMC** because no NMC field requires it; and (b) the EHP→DALRRD referral, when it happens, is informal and likewise unrecorded. DALRRD action — if any — flows back to the EHP at best, and never to NICD or the originating clinician.

## 4. Root cause analysis

```{mermaid}
%%| label: fig-coord-fishbone
%%| fig-cap: "Root causes of coordination failure across the NMC–EHP–DALRRD chain."
flowchart LR
    A[Coordination failure] --> R1[NMC has no field for<br/>EHP investigation outcome]
    A --> R2[EHP→DALRRD referral<br/>has no SOP, no audit trail]
    A --> R3[DALRRD has no inbound<br/>NICD intelligence feed]
    A --> R4[Clinicians never receive<br/>case-status feedback]
    A --> R5[No KPI / accountability<br/>at any node]

    R1 --> O1[Option C1: Mandate EHP<br/>NMC closure-of-loop field]
    R2 --> O2[Option C2: Structured<br/>EHP→DALRRD referral form on NMC]
    R3 --> O3[Option C3: NICD→DALRRD<br/>intelligence feed]
    R4 --> O4[Option C4: NMC auto-feedback<br/>to notifier on case status]
    R5 --> O5[Option C5: Joint KPI dashboard<br/>NICD + DALRRD + Provinces]

    style A fill:#f99,stroke:#333
    style O1 fill:#9f9,stroke:#333
    style O2 fill:#9f9,stroke:#333
    style O3 fill:#9f9,stroke:#333
    style O4 fill:#9f9,stroke:#333
    style O5 fill:#9f9,stroke:#333
```

The five drivers are: (1) NMC schema lacks a closure-of-loop field; (2) EHP→DALRRD referral has no SOP; (3) DALRRD has no inbound NICD intelligence feed despite a formal request for one; (4) clinicians get no case-status feedback so the notification has no perceived value; (5) **no decision node in the chain has an accountable owner with a KPI**, so nothing fails loudly when the loop breaks.

## 5. Policy options

| ID | Option | What it does | Annual cost |
|:---|:-------|:-------------|------------:|
| **C1** | NMC closure-of-loop field for EHPs | Structured "investigation outcome" field (visited, source identified, sampling, referral); 14-day completion KPI. No new legislation. | ~R30k |
| **C2** | Structured EHP→DALRRD referral form on NMC | Auto-emails DALRRD Inspectorate with product, location, agent, severity; replaces informal phone/email referral. | ~R25k |
| **C3** ★ | NICD→DALRRD poisoning intelligence feed | Quarterly NICD brief by product/geography/season + ad-hoc case-spike alerts. Gives DALRRD the evidence base for product-specific action (prerequisite for the [Terbufos brief](terbufos_policy.qmd)). | ~R40k |
| **C4** | Notifier auto-feedback | Quarterly NMC case-status summary back to the notifying clinician. Removes the "notify into a black hole" disincentive. | ~R20k |
| **C5** | Joint KPI dashboard | Public dashboard: notification-to-investigation time, % with recorded outcome, % with DALRRD action, % notifiers receiving feedback. Co-owned by NICD, DALRRD, provincial DoHs. | ~R30k |

### Coordination Full Package (C1 + C2 + C3 + C4 + C5)

```{mermaid}
%%| label: fig-coord-feedback-loop
%%| fig-cap: "Proposed end-state: a closed feedback loop linking clinicians, laboratories, NICD, EHPs, DALRRD, and provincial response teams. Every actor both contributes data and receives back the action taken."
flowchart LR
    A[Clinician<br/>at facility] -->|Electronic<br/>NMC notification| D[NICD<br/>Surveillance Unit]
    L[NHLS Laboratory<br/>BChE + Toxicology] -->|Auto-notify<br/>< 24h| D
    P[PIH<br/>Quarterly summary] -->|Minimum viable<br/>dataset| D

    D -->|Case-spike alert| E[Provincial EHP /<br/>Response Team]
    E -->|Structured<br/>referral| C[DALRRD<br/>Inspectorate]
    C -->|Enforcement /<br/>recall outcome| D
    E -->|Investigation<br/>outcome| D

    D -.Case-status<br/>feedback.-> A
    D -.Result-use<br/>feedback.-> L
    D -.Trend<br/>feedback.-> P
    C -.Regulatory<br/>action feedback.-> A

    style A fill:#cce5ff,stroke:#333
    style L fill:#cce5ff,stroke:#333
    style P fill:#cce5ff,stroke:#333
    style D fill:#9f9,stroke:#333,stroke-width:2px
    style E fill:#ffd,stroke:#333
    style C fill:#fcb,stroke:#333
```

Total annual cost of the Coordination Full Package: **~R145,000**. No new legislation required.

## 6. Decision tree — coordination outcomes per suspected case

### 6a. Where the chain breaks today

````{r}
#| label: fig-coord-decision-tree
#| echo: false
#| output: asis
p1 <- as.numeric(pick("p_coord_sq_nmc_notified"))
p2 <- as.numeric(pick("p_coord_sq_sms_reaches_ehp"))
p3 <- as.numeric(pick("p_coord_sq_ehp_investigates"))
p4 <- as.numeric(pick("p_coord_sq_recorded_on_nmc"))
p5 <- as.numeric(pick("p_coord_sq_refer_dalrrd"))
p6 <- as.numeric(pick("p_coord_sq_dalrrd_acts"))
p7 <- as.numeric(pick("p_coord_sq_feedback_full"))

p_loop_sq <- as.numeric(pick("p_loop_closed_status_quo"))

cat(sprintf('```{mermaid}
%%%%| fig-cap: "Coordination decision tree for a single suspected case under the status quo. Eight conditional steps; six terminate in a broken-loop outcome. All eight node probabilities pulled from amua_import_parameters_v3.csv at render time."
flowchart TD
    Q0{Case suspected?} -->|Yes| Q1{Notified on<br/>NMC?}
    Q0 -->|No| X0[Case invisible]:::bad

    Q1 -->|Yes %.2f| Q2{SMS reaches<br/>EHP?}
    Q1 -->|No %.2f| X1[No notification]:::bad

    Q2 -->|Yes %.2f| Q3{EHP<br/>investigates?}
    Q2 -->|No %.2f| X2[Dies at NMC]:::bad

    Q3 -->|Yes %.2f| Q4{Recorded<br/>on NMC?}
    Q3 -->|No %.2f| X3[No response]:::bad

    Q4 -->|Yes %.2f| G1[Loop closed<br/>to NMC]:::good
    Q4 -->|No %.2f| Q5{Refer to<br/>DALRRD?}

    Q5 -->|Yes %.2f| Q6{DALRRD<br/>acts?}
    Q5 -->|No %.2f| X4[Regulator<br/>never hears]:::bad

    Q6 -->|Yes %.2f| Q7{Feedback to<br/>NMC + clinician?}
    Q6 -->|No %.2f| X5[Source persists]:::bad

    Q7 -->|To EHP only %.2f| X6[NMC + clinician<br/>still blind]:::bad
    Q7 -->|To NMC + clinician %.2f| G2[Loop fully closed]:::good

    classDef bad fill:#fdd,stroke:#cc0000,stroke-width:1.5px
    classDef good fill:#cfc,stroke:#060,stroke-width:1.5px
```
',
p1, 1 - p1,
p2, 1 - p2,
p3, 1 - p3,
p4, 1 - p4,
p5, 1 - p5,
p6, 1 - p6,
1 - p7, p7
))
````

**Joint P(loop fully closes) ≈ `r sprintf("%.1f %%", 100 * as.numeric(pick("p_loop_closed_status_quo")))`** under the status quo (CSV: `p_loop_closed_status_quo`). Expected coordination cost per suspected case ≈ **R196,400** — virtually the full per-case societal burden of R197,000, because the loop almost never closes.

### 6b. Per-loop yield — the parameter the integrated model consumes

A single fully-closed loop retires one uncontrolled source. The source would otherwise produce a cluster of $\rho_{\text{secondary}}$ further cases. Response efficacy $\eta$ is the fraction of that cluster that is actually averted once the loop closes. Per loop closed:

$$
\text{cases prevented} = \rho_{\text{secondary}} \times \eta
\qquad
\text{deaths prevented} = \rho_{\text{secondary}} \times \eta \times p_{\text{death}}
$$

```{r}
#| label: tbl-per-loop-yield
#| echo: false
#| tbl-cap: "Preventable cases, deaths, and societal burden per *additional* fully-closed loop, at low/mid/high response efficacy. All inputs from amua_import_parameters_v4.csv. These are the rates that overall_policy_model.qmd multiplies by Δ(loops closed) under each policy scenario."

rho     <- as.numeric(pick("rho_secondary"))
p_dth   <- as.numeric(pick("p_death"))
c_morb  <- as.numeric(pick("c_morbidity_headline"))
c_mort  <- as.numeric(pick("c_mortality_human_capital"))

eta_lo  <- as.numeric(pick("eta_response_lo"))
eta_mid <- as.numeric(pick("eta_response_mid"))
eta_hi  <- as.numeric(pick("eta_response_hi"))

tibble::tibble(
  Scenario          = c("Low engagement (η_lo)", "Base case (η_mid)", "Well-resourced (η_hi)"),
  `η`               = c(eta_lo, eta_mid, eta_hi),
  `Cases / loop`    = rho * c(eta_lo, eta_mid, eta_hi),
  `Deaths / loop`   = rho * c(eta_lo, eta_mid, eta_hi) * p_dth,
  `Burden retired / loop (R)` =
    rho * c(eta_lo, eta_mid, eta_hi) * c_morb +
    rho * c(eta_lo, eta_mid, eta_hi) * p_dth * c_mort
) |>
  gt::gt() |>
  gt::fmt_number(columns = c("η"), decimals = 2) |>
  gt::fmt_number(columns = c("Cases / loop", "Deaths / loop"), decimals = 2) |>
  gt::fmt_number(columns = "Burden retired / loop (R)", decimals = 0, use_seps = TRUE, pattern = "R{x}") |>
  gt::tab_options(table.font.size = "small")
```

**Headline rule of thumb.** At base case ($\rho = `r rho`$, $\eta = `r eta_mid`$, $p_{\text{death}} = `r p_dth`$): **every loop closed prevents ≈ `r round(rho*eta_mid, 1)` future cases and ≈ `r round(rho*eta_mid*p_dth, 2)` deaths**, retiring **≈ R`r format(round(rho*eta_mid*(c_morb + p_dth*c_mort)), big.mark=",")` of societal burden**. Multiply by the change in number of loops closed under any scenario in [the integrated model](overall_policy_model.qmd) to get the policy yield.

### 6c. Clinical-outcome subtree

Each coordination terminal feeds into a clinical-outcome subtree for the index case *and* for the secondary cases that would otherwise follow from the same source:

````{r}
#| label: fig-coord-outcome-tree-v2
#| echo: false
#| output: asis
rho     <- as.numeric(pick("rho_secondary"))
eta     <- as.numeric(pick("eta_response_mid"))
p_dth   <- as.numeric(pick("p_death"))
p_sev   <- as.numeric(pick("p_severe"))
p_rec   <- as.numeric(pick("p_recover"))

averted_sec   <- rho * eta
extra_sec     <- rho
averted_deaths<- averted_sec * p_dth
extra_deaths  <- extra_sec * p_dth

cat(sprintf('```{mermaid}
%%%%| fig-cap: "Clinical-outcome subtree attached to each coordination terminal. Index-case probabilities (p_death, p_severe, p_recover) and secondary-case multipliers (rho_secondary, eta_response_mid) all pulled from amua_import_parameters_v3.csv at render time."
flowchart TD
    T0[Suspected case] --> P1{Loop closed?}

    P1 -->|Yes — G1 or G2| S1[Source removed]:::good
    P1 -->|No — X1..X6| S2[Source persists]:::bad

    S1 --> I1[Index case:<br/>p_death %.2f<br/>p_severe %.2f<br/>p_recover %.2f]
    S2 --> I2[Index case:<br/>p_death %.2f<br/>p_severe %.2f<br/>p_recover %.2f]

    S1 --> N1[Averted secondary cases<br/>~ rho x eta = %.2f]:::good
    S2 --> N2[Secondary cases<br/>~ rho = %.2f]:::bad

    N1 --> AD1[Deaths averted<br/>~ %.3f]:::good
    N2 --> AD2[Extra deaths<br/>~ %.3f]:::bad

    classDef bad fill:#fdd,stroke:#cc0000,stroke-width:1.5px
    classDef good fill:#cfc,stroke:#060,stroke-width:1.5px
```
',
p_dth, p_sev, p_rec,
p_dth, p_sev, p_rec,
averted_sec, extra_sec,
averted_deaths, extra_deaths
))
````

### 6d. What changes under each coordination scenario

Burden avoided is split by **who would otherwise have borne it**, so each agency sees the share of the win that flows back to its own budget line or constituency. The avoided-burden split uses the same shares as the canonical stakeholder table above: ~3.4 % health sector (NDoH + NHLS), ~96.6 % households (mortality + OOP).

| Scenario | P(loop closed) | Deaths averted / yr | Avoided — health system (NDoH + NHLS) | Avoided — households (mortality + OOP) | Total avoided |
|:---------|---------------:|--------------------:|-------------------------------------:|---------------------------------------:|--------------:|
| Status quo | 1.1 % | 15 | — | — | — |
| + C3 (NICD→DALRRD feed) | ~40 % | ~547 | ~R29M | ~R806M | ~R835M |
| **Coordination Full Package (C1–C5)** | **~70 %** | **~957** | **~R50M** | **~R1,400M** | **~R1,450M** |

: Modelled coordination outcomes at national scale (× 4,171 observed cases/year, 2023 streams). Stakeholder columns apply the canonical share table (NDoH+NHLS = R80M of R2.34B; households = R2.26B of R2.34B) to the total avoided burden in each scenario. Probabilities are placeholders aligned with the surveillance detection-rate ranges in the [Surveillance brief](surveillance_policy.qmd). {.striped .hover}

**Reading by audience.**

- **NDoH / NHLS:** the R50M avoided is ~330× the R145k Full-Package operating cost — a credible business case before counting household mortality.
- **DALRRD:** C3 is the lowest-cost route to the evidence base needed for product-specific action (see [Terbufos brief](terbufos_policy.qmd)).
- **Treasury / Presidency:** ~R1.4B/yr of household and societal cost retired; the cross-departmental return that justifies the joint DG circular.

**Beyond terbufos.** The Full Package retires ~957 deaths/yr **today**, dominated by organophosphate and carbamate clusters. As terbufos is withdrawn (see [Terbufos brief](terbufos_policy.qmd)) the residual cluster mix will shift first to other agricultural pesticides (aldicarb, paraquat, carbofuran), and over a 10–20 year development horizon increasingly toward **opioids, methadone substitution, and recreational-drug overdoses** — the toxicology profile that has dominated middle- and high-income countries since the 1990s. The closed loop is the asset that survives that transition; only the product list at C3 changes.


## 7. Implementation considerations

| Component | Lead agency | Year-1 cost | Critical path |
|:----------|:------------|------------:|:--------------|
| C1: NMC closure-of-loop field | NICD NMC team + NDoH circular | R30k | NMC schema change; EHP training |
| C2: Structured DALRRD referral form on NMC | NICD + DALRRD Pesticide Registrar | R25k | DALRRD inbox + workflow |
| C3: NICD→DALRRD intelligence feed | NICD analyst + DALRRD liaison | R40k | Quarterly brief template; case-spike alert SOP |
| C4: Notifier auto-feedback | NICD NMC team | R20k | Email templating |
| C5: Joint KPI dashboard | NICD + DALRRD + Provinces | R30k | KPI definitions; public hosting |
| **Total** | NICD-led | **~R145k** | All within existing legal authority |

**Institutional mechanism.** A **quarterly Joint Coordination Forum** (NICD chair; DALRRD, NHLS, PIH, and one rotating provincial DoH representative) reviews the KPI dashboard, signs off case-spike alert escalations, and resolves cross-departmental blockers. Mandated by joint NDoH–DALRRD Director-General circular. **No new legislation required.**

**Risks and mitigations.**

- *EHP non-compliance with new NMC field.* Mitigation: 14-day completion KPI tied to provincial performance reviews; circular from NDoH DG.
- *DALRRD capacity to act on referrals.* Mitigation: C3 quarterly brief lets DALRRD prioritise by burden rather than respond reactively.
- *Privacy concerns on clinician auto-feedback.* Mitigation: feedback is aggregate at clinician level, not patient-identifying.

## 8. Recommended next step

**Adopt C1 + C3 as the Coordination MVP, and the Full Package within 12 months.**

- **Weeks 1–4:** NDoH–DALRRD joint DG circular committing to the closed-loop model; standing up the Joint Coordination Forum.
- **Weeks 4–12:** C1 NMC schema change + EHP training; C3 NICD→DALRRD quarterly brief begins on the existing NMC data.
- **Months 3–12:** C2 referral form, C4 auto-feedback, C5 KPI dashboard go live.
- **Month 12:** Evaluate against (i) median notification-to-investigation time, (ii) % of notifications with recorded outcome, (iii) DALRRD-reported regulatory actions per quarter.

The surveillance signal that flows *into* this loop is covered in the [Surveillance brief](surveillance_policy.qmd). The first product-specific regulatory action the loop should be designed to deliver is the subject of the [Terbufos brief](terbufos_policy.qmd).

---

::: {.footer-cta}
**Companion briefs:** [Surveillance](surveillance_policy.qmd) · [Terbufos restriction](terbufos_policy.qmd) · [Decision Model](decision_model.qmd) · [Costing Analysis](costing_analysis.qmd)

**Technical appendix:** [B — Coordination joint-probability model](../technical_appendix_coordination.qmd)
:::
