---
title: "Costing Analysis: Improved Pesticide Surveillance"
subtitle: "Detailed Cost Estimates for Implementation and Operations"
author: "D2P Project Team"
date: "2025-09-05"
date-modified: today
categories: [costing, budget, implementation]
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
---

## Executive Summary

This document provides **detailed cost estimates** for implementing and operating improved pesticide poisoning surveillance in South Africa. All estimates are based on:

- Current government salary scales (2025/2026)
- NHLS IT infrastructure costs (confirmed with NHLS IT management)
- Provincial health department operating budgets
- Expert elicitation for liaison and coordination activities

::: {.callout-important}
## Headline Findings — Two Cost Lenses

**Narrow government-provider cost (NICD/NHLS/PIH/Provincial DoH coordination only):** **~R237,000/year**.

**Pressure-tested whole-system cost** (including downstream hospital toxicology workload, lab test volume, Environmental Health Practitioner (EHP) investigations, and agricultural enforcement): **~R1.55 million/year** (range: R0.9–2.7 million).

**Why the gap matters:** Improved surveillance is not free at the point of detection. Every BChE auto-notification triggers a clinical work-up, a confirmatory test, an EHP follow-up, and — for hotspots — DALRRD enforcement. The narrow figure understates true budget impact by **~6×**. Decision-makers should plan for the comprehensive figure to avoid stranding the programme without operational capacity.

**For context, the comprehensive figure still represents:**

- **0.0028%** of the National Department of Health budget (~R56 billion/year)
- **R172 per severe pesticide poisoning case** (based on ~9,000 cases/year)
- **Less than the salary of 3 mid-level epidemiologists**
:::

## Costing Methodology

### Two Costing Perspectives

This analysis uses **two nested perspectives** to make the trade-offs transparent:

1. **Narrow government-provider perspective** (the original estimate, ~R237k/year). Captures only the costs borne by:
   - National Institute for Communicable Diseases (NICD)
   - National Health Laboratory Service (NHLS) — coordination only
   - Poisons Information Helpline (PIH)
   - Provincial Departments of Health — coordination only

2. **Pressure-tested whole-system perspective** (Section [Pressure-Testing the Costs](#sec-pressure-test)). Adds the **induced demand** that improved surveillance creates downstream:
   - **Hospitals:** clinician time for exposure histories, sample collection, NMC notification completion, and data collation by records clerks
   - **Laboratories:** higher BChE test volumes, confirmatory GC-MS toxicology, sample transport and storage
   - **Environmental Health Practitioners (EHPs):** field investigation of clusters, environmental sampling, source tracing
   - **Agricultural enforcement:** DALRRD compliance officers responding to illegal-pesticide signals, retailer audits, registration follow-up

**Still excluded (for now):**

- Patient out-of-pocket costs and productivity losses
- Wider societal costs (community engagement, media)
- Capital depreciation on existing NHLS lab equipment

::: {.callout-warning}
## Why This Matters

The narrow perspective is appropriate for **"what new line items appear in the surveillance programme budget?"** — but it understates the real fiscal impact on the health and agriculture system. A surveillance system that detects more cases necessarily generates more downstream work. Failing to budget for it leads to: (a) overworked frontline staff, (b) backlogs in lab confirmation, (c) un-investigated hotspots, and (d) erosion of the system within 12–18 months.
:::

### Time Horizon

- **Pilot phase:** 6 months (Q3–Q4 2026)
- **Ongoing operations:** Annual costs for steady-state implementation

### Currency and Inflation

- All costs in **South African Rand (ZAR)**
- Based on **2025/2026 fiscal year** salary scales and price levels
- No inflation adjustment (costs represent current purchasing power)

## Option-by-Option Cost Breakdown

### Status Quo (Baseline)

**Description:** Continue current NMC surveillance approach

```{r}
#| label: tbl-status-quo
#| tbl-cap: "Status quo costs"

library(gt)
library(tidyverse)

status_quo <- tribble(
  ~Cost_Category, ~Annual_Cost, ~Notes,
  "Personnel", "R0", "No additional staff required",
  "IT infrastructure", "R0", "No changes to existing systems",
  "Other operating costs", "R0", "No new expenses",
  "**Total**", "**R0**", "**Baseline (no change)**"
)

status_quo %>%
  gt() %>%
  tab_header(
    title = "Status Quo Costs",
    subtitle = "Current NMC surveillance (baseline)"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 4)
  )
```

---

### Option 1: NHLS Data Sharing Agreement

**Description:** Formalize MOU between NICD and NHLS for monthly aggregated BChE data sharing

#### Personnel Costs

```{r}
#| label: tbl-option1-personnel
#| tbl-cap: "Option 1: Personnel costs"

option1_personnel <- tribble(
  ~Role, ~FTE, ~Annual_Salary, ~Pct_Time_on_Surveillance, ~Annual_Cost,
  "NHLS Data Analyst", 0.05, "R450,000", "5%", "R22,500",
  "NICD Surveillance Officer", 0.02, "R380,000", "2%", "R7,600",
  "**Subtotal Personnel**", NA, NA, NA, "**R30,100**"
)

option1_personnel %>%
  gt() %>%
  tab_header(
    title = "Option 1: Personnel Costs"
  ) %>%
  cols_label(
    Pct_Time_on_Surveillance = "% Time on Surveillance"
  ) %>%
  cols_align(
    align = "right",
    columns = c(FTE, Annual_Salary, Annual_Cost)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 3)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—")
```

#### Other Costs

```{r}
#| label: tbl-option1-other
#| tbl-cap: "Option 1: Other costs"

option1_other <- tribble(
  ~Item, ~One_Time_Cost, ~Annual_Cost, ~Notes,
  "Legal review (MOU drafting)", "R15,000", "R0", "One-time expense in Year 1",
  "Data governance framework", "R10,000", "R5,000", "Initial development + annual review",
  "Liaison meetings (travel, venue)", "R0", "R5,000", "4 meetings/year @ R1,250 each",
  "**Subtotal Other**", "**R25,000**", "**R10,000**", ""
)

option1_other %>%
  gt() %>%
  tab_header(
    title = "Option 1: Other Costs"
  ) %>%
  cols_align(
    align = "right",
    columns = c(One_Time_Cost, Annual_Cost)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 4)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—")
```

#### Total Cost Summary

```{r}
#| label: tbl-option1-total
#| tbl-cap: "Option 1: Total cost summary"

option1_total <- tribble(
  ~Phase, ~Cost,
  "Year 1 (pilot + setup)", "R65,100",
  "Ongoing annual cost", "R40,100"
)

option1_total %>%
  gt() %>%
  tab_header(
    title = "Option 1: Total Costs",
    subtitle = "NHLS Data Sharing Agreement"
  ) %>%
  cols_align(align = "right", columns = Cost) %>%
  tab_style(
    style = cell_fill(color = "lightblue"),
    locations = cells_body(rows = 2)
  )
```

**Key driver:** Personnel time for data extraction and monthly reporting workflows.

---

### Option 2: DALRRD Pesticide Registration Data Sharing

**Description:** Link pesticide poisoning cases to pesticide product registration database

#### Personnel Costs

```{r}
#| label: tbl-option2-personnel
#| tbl-cap: "Option 2: Personnel costs"

option2_personnel <- tribble(
  ~Role, ~FTE, ~Annual_Salary, ~Pct_Time, ~Annual_Cost,
  "NICD Surveillance Officer", 0.04, "R380,000", "4%", "R15,200",
  "DALRRD Liaison Officer", 0.03, "R350,000", "3%", "R10,500",
  "Data cleaning/linkage analyst", 0.01, "R400,000", "1%", "R4,000",
  "**Subtotal Personnel**", NA, NA, NA, "**R29,700**"
)

option2_personnel %>%
  gt() %>%
  tab_header(title = "Option 2: Personnel Costs") %>%
  cols_label(Pct_Time = "% Time") %>%
  cols_align(align = "right", columns = c(FTE, Annual_Salary, Annual_Cost)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 4)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—")
```

#### Other Costs

```{r}
#| label: tbl-option2-other
#| tbl-cap: "Option 2: Other costs"

option2_other <- tribble(
  ~Item, ~One_Time_Cost, ~Annual_Cost,
  "MOU development", "R12,000", "R0",
  "Inter-departmental liaison meetings", "R0", "R8,000",
  "Data quality audits", "R0", "R2,000",
  "**Subtotal Other**", "**R12,000**", "**R10,000**"
)

option2_other %>%
  gt() %>%
  tab_header(title = "Option 2: Other Costs") %>%
  cols_align(align = "right", columns = c(One_Time_Cost, Annual_Cost)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 4)
  )
```

#### Total Cost Summary

```{r}
#| label: tbl-option2-total
#| tbl-cap: "Option 2: Total cost summary"

option2_total <- tribble(
  ~Phase, ~Cost,
  "Year 1 (pilot + setup)", "R51,700",
  "Ongoing annual cost", "R39,700"
)

option2_total %>%
  gt() %>%
  tab_header(
    title = "Option 2: Total Costs",
    subtitle = "DALRRD Data Sharing"
  ) %>%
  cols_align(align = "right", columns = Cost) %>%
  tab_style(
    style = cell_fill(color = "lightblue"),
    locations = cells_body(rows = 2)
  )
```

---

### Option 3: BChE Auto-Notification

**Description:** Configure NHLS Laboratory Information System to auto-notify NICD when BChE results indicate severe poisoning

#### Personnel Costs

```{r}
#| label: tbl-option3-personnel
#| tbl-cap: "Option 3: Personnel costs"

option3_personnel <- tribble(
  ~Role, ~FTE, ~Annual_Salary, ~Pct_Time, ~Annual_Cost,
  "NHLS IT Developer", 0.08, "R550,000", "8%", "R44,000",
  "NICD Surveillance Officer", 0.02, "R380,000", "2%", "R7,600",
  "**Subtotal Personnel**", NA, NA, NA, "**R51,600**"
)

option3_personnel %>%
  gt() %>%
  tab_header(title = "Option 3: Personnel Costs") %>%
  cols_label(Pct_Time = "% Time") %>%
  cols_align(align = "right", columns = c(FTE, Annual_Salary, Annual_Cost)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 3)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—")
```

#### IT Infrastructure Costs

```{r}
#| label: tbl-option3-it
#| tbl-cap: "Option 3: IT infrastructure costs"

option3_it <- tribble(
  ~Item, ~One_Time_Cost, ~Annual_Maintenance,
  "LIS rule configuration", "R25,000", "R5,000",
  "HL7 messaging middleware", "R0", "R8,000",
  "Testing and validation", "R12,000", "R2,000",
  "**Subtotal IT**", "**R37,000**", "**R15,000**"
)

option3_it %>%
  gt() %>%
  tab_header(title = "Option 3: IT Infrastructure Costs") %>%
  cols_align(align = "right", columns = c(One_Time_Cost, Annual_Maintenance)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 4)
  )
```

#### Total Cost Summary

```{r}
#| label: tbl-option3-total
#| tbl-cap: "Option 3: Total cost summary"

option3_total <- tribble(
  ~Phase, ~Cost,
  "Year 1 (development + pilot)", "R103,600",
  "Ongoing annual cost", "R66,600"
)

option3_total %>%
  gt() %>%
  tab_header(
    title = "Option 3: Total Costs",
    subtitle = "BChE Auto-Notification"
  ) %>%
  cols_align(align = "right", columns = Cost) %>%
  tab_style(
    style = cell_fill(color = "lightgreen"),
    locations = cells_body(rows = 2)
  ) %>%
  tab_footnote(
    footnote = "Best value for money — highest effectiveness gain per rand spent",
    locations = cells_column_labels(columns = Cost)
  )
```

**Key driver:** IT development for LIS configuration (one-time) + ongoing middleware maintenance.

---

### Option 4: PIH Quarterly Reports

**Description:** Poisons Information Helpline generates quarterly summary reports for NICD

#### Personnel Costs

```{r}
#| label: tbl-option4-personnel
#| tbl-cap: "Option 4: Personnel costs"

option4_personnel <- tribble(
  ~Role, ~FTE, ~Annual_Salary, ~Pct_Time, ~Annual_Cost,
  "PIH Data Analyst", 0.15, "R420,000", "15%", "R63,000",
  "NICD Surveillance Officer (review)", 0.01, "R380,000", "1%", "R3,800",
  "**Subtotal Personnel**", NA, NA, NA, "**R66,800**"
)

option4_personnel %>%
  gt() %>%
  tab_header(title = "Option 4: Personnel Costs") %>%
  cols_label(Pct_Time = "% Time") %>%
  cols_align(align = "right", columns = c(FTE, Annual_Salary, Annual_Cost)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 3)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—")
```

#### Other Costs

```{r}
#| label: tbl-option4-other
#| tbl-cap: "Option 4: Other costs"

option4_other <- tribble(
  ~Item, ~One_Time_Cost, ~Annual_Cost,
  "Report template development", "R8,000", "R0",
  "Data extraction scripts", "R0", "R3,000",
  "Quality assurance reviews", "R0", "R2,000",
  "**Subtotal Other**", "**R8,000**", "**R5,000**"
)

option4_other %>%
  gt() %>%
  tab_header(title = "Option 4: Other Costs") %>%
  cols_align(align = "right", columns = c(One_Time_Cost, Annual_Cost)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 4)
  )
```

#### Total Cost Summary

```{r}
#| label: tbl-option4-total
#| tbl-cap: "Option 4: Total cost summary"

option4_total <- tribble(
  ~Phase, ~Cost,
  "Year 1 (setup + 4 reports)", "R79,800",
  "Ongoing annual cost", "R71,800"
)

option4_total %>%
  gt() %>%
  tab_header(
    title = "Option 4: Total Costs",
    subtitle = "PIH Quarterly Reports"
  ) %>%
  cols_align(align = "right", columns = Cost) %>%
  tab_style(
    style = cell_fill(color = "lightgreen"),
    locations = cells_body(rows = 2)
  ) %>%
  tab_footnote(
    footnote = "Fastest to implement — can start within 1-2 weeks",
    locations = cells_column_labels(columns = Cost)
  )
```

**Key driver:** PIH analyst time (15% FTE) for quarterly data extraction and report generation.

---

### Option 5: Provincial Hotspot Response Coordination

**Description:** Establish provincial capacity for rapid field investigation of clusters

#### Personnel Costs

```{r}
#| label: tbl-option5-personnel
#| tbl-cap: "Option 5: Personnel costs"

option5_personnel <- tribble(
  ~Role, ~FTE, ~Annual_Salary, ~Pct_Time, ~Annual_Cost,
  "Provincial Surveillance Officers (3 provinces)", 0.05, "R360,000", "5% × 3", "R54,000",
  "NICD National Coordinator", 0.02, "R450,000", "2%", "R9,000",
  "**Subtotal Personnel**", NA, NA, NA, "**R63,000**"
)

option5_personnel %>%
  gt() %>%
  tab_header(title = "Option 5: Personnel Costs") %>%
  cols_label(Pct_Time = "% Time") %>%
  cols_align(align = "right", columns = c(FTE, Annual_Salary, Annual_Cost)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 3)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—")
```

#### Other Costs

```{r}
#| label: tbl-option5-other
#| tbl-cap: "Option 5: Other costs"

option5_other <- tribble(
  ~Item, ~One_Time_Cost, ~Annual_Cost,
  "Training workshops (3 provinces)", "R25,000", "R0",
  "SOP development", "R15,000", "R0",
  "Field investigation kits (sampling equipment)", "R30,000", "R10,000",
  "Travel for field investigations", "R0", "R20,000",
  "Quarterly coordination meetings", "R0", "R8,000",
  "**Subtotal Other**", "**R70,000**", "**R38,000**"
)

option5_other %>%
  gt() %>%
  tab_header(title = "Option 5: Other Costs") %>%
  cols_align(align = "right", columns = c(One_Time_Cost, Annual_Cost)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 6)
  )
```

#### Total Cost Summary

```{r}
#| label: tbl-option5-total
#| tbl-cap: "Option 5: Total cost summary"

option5_total <- tribble(
  ~Phase, ~Cost,
  "Year 1 (training + equipment)", "R171,000",
  "Ongoing annual cost", "R101,000"
)

option5_total %>%
  gt() %>%
  tab_header(
    title = "Option 5: Total Costs",
    subtitle = "Provincial Hotspot Response Coordination"
  ) %>%
  cols_align(align = "right", columns = Cost) %>%
  tab_footnote(
    footnote = "Requires Options 3 or 4 to generate timely alerts",
    locations = cells_column_labels(columns = Cost)
  )
```

**Key driver:** Provincial staff time + field investigation consumables (sampling kits, travel).

---

### Full Package: All Options Combined

**Description:** Implement Options 1–5 simultaneously for comprehensive integrated surveillance

#### Cost Summary

```{r}
#| label: tbl-full-package
#| tbl-cap: "Full Package: Total cost summary"

full_package <- tribble(
  ~Component, ~Annual_Cost, ~Pct_of_Total,
  "Option 1: NHLS MOU", "R40,100", "17%",
  "Option 2: DALRRD data", "R39,700", "17%",
  "Option 3: BChE auto-notify", "R66,600", "28%",
  "Option 4: PIH reports", "R71,800", "30%",
  "Option 5: Provincial coordination", "R101,000", "43%",
  "**Total (with adjustments)**", "**R237,000**", "**100%**"
)

full_package %>%
  gt() %>%
  tab_header(
    title = "Full Package: Annual Cost Summary",
    subtitle = "All options implemented together"
  ) %>%
  cols_label(Pct_of_Total = "% of Total") %>%
  cols_align(align = "right", columns = c(Annual_Cost, Pct_of_Total)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 6)
  ) %>%
  tab_style(
    style = cell_fill(color = "lightyellow"),
    locations = cells_body(rows = 6)
  ) %>%
  tab_footnote(
    footnote = "Total is less than sum of individual options due to shared personnel (no double-counting)",
    locations = cells_body(columns = Component, rows = 6)
  )
```

#### Year 1 vs. Ongoing Costs

```{r}
#| label: fig-year1-vs-ongoing
#| fig-cap: "Year 1 setup costs vs. ongoing annual costs"
#| fig-width: 8
#| fig-height: 5

cost_comparison <- tribble(
  ~Option, ~Year_1, ~Ongoing,
  "Option 1: NHLS MOU", 65100, 40100,
  "Option 2: DALRRD data", 51700, 39700,
  "Option 3: BChE auto-notify", 103600, 66600,
  "Option 4: PIH reports", 79800, 71800,
  "Option 5: Provincial coord", 171000, 101000,
  "Full Package", 320000, 237000
) %>%
  pivot_longer(cols = c(Year_1, Ongoing), names_to = "Phase", values_to = "Cost")

ggplot(cost_comparison, aes(x = reorder(Option, -Cost), y = Cost, fill = Phase)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2", labels = c("Ongoing (annual)", "Year 1 (setup)")) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "R", scale = 1)) +
  coord_flip() +
  labs(
    title = "Year 1 Setup Costs vs. Ongoing Annual Costs",
    subtitle = "All options have higher first-year costs due to setup/training",
    x = "Policy Option",
    y = "Cost (ZAR)",
    fill = "Phase"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## Pressure-Testing the Costs: Downstream Demand on the Whole System {#sec-pressure-test}

The **narrow estimates** above (~R237k/year) capture only the **surveillance coordination** layer — NICD, NHLS liaison, PIH reporting, provincial coordinators. But improved surveillance is not cost-neutral to the wider health and agriculture system. Every additional BChE notification triggers:

1. **Hospital response:** Clinician time to complete exposure histories, order confirmatory tests, fill out NMC forms, liaise with EHPs. Records clerks collate the data.
2. **Laboratory testing:** Higher volume of BChE screening tests (as awareness spreads), confirmatory GC-MS toxicology for hotspots, sample transport logistics.
3. **Environmental Health Practitioner (EHP) investigations:** Field visits to suspected exposure sites, environmental sampling (water, soil, food), source tracing, retailer inspections.
4. **Agricultural enforcement:** DALRRD compliance officers investigate illegal pesticide sales, conduct product recalls, audit registration status.

This section quantifies the **induced demand** from the Full Package (Options 1–5 combined).

---

### Downstream Cost Component 1: Hospital Toxicology Workload

**Scenario:** Full Package detects 85% of hotspots (vs. 45% today). This means ~40% more severe cases are **actively followed up** by hospitals (not just passively recorded). Each follow-up requires:

#### Hospital Personnel Time

```{r}
#| label: tbl-hospital-workload
#| tbl-cap: "Hospital workload induced by improved case detection"

# Assumptions
base_cases_detected <- 900  # Current NMC notifications/year (45% of ~2000 severe cases)
full_package_cases_detected <- 1700  # 85% of ~2000 severe cases
incremental_cases <- full_package_cases_detected - base_cases_detected  # 800 additional cases/year

# Time per case
clinician_time_per_case_hours <- 0.75  # Exposure history, NMC form completion, toxicology consult
records_clerk_time_per_case_hours <- 0.25  # Data entry, filing, liaison with NICD
sample_collection_time_hours <- 0.1  # Phlebotomy for confirmatory toxicology

# Hourly rates (based on government salary scales, loaded with benefits)
clinician_hourly_rate <- 380000 / 1800  # ~R211/hour (assuming R380k salary, 1800 billable hours/year)
records_clerk_hourly_rate <- 200000 / 1800  # ~R111/hour
phlebotomist_hourly_rate <- 250000 / 1800  # ~R139/hour

# Cost calculations
clinician_annual_cost <- incremental_cases * clinician_time_per_case_hours * clinician_hourly_rate
records_clerk_annual_cost <- incremental_cases * records_clerk_time_per_case_hours * records_clerk_hourly_rate
phlebotomy_annual_cost <- incremental_cases * sample_collection_time_hours * phlebotomist_hourly_rate

hospital_personnel_total <- clinician_annual_cost + records_clerk_annual_cost + phlebotomy_annual_cost

hospital_workload <- tribble(
  ~Role, ~Time_per_Case_Hours, ~Incremental_Cases_per_Year, ~Hourly_Rate, ~Annual_Cost,
  "Clinician (exposure history, NMC form)", 0.75, 800, "R211", paste0("R", format(round(clinician_annual_cost), big.mark = ",")),
  "Records clerk (data collation)", 0.25, 800, "R111", paste0("R", format(round(records_clerk_annual_cost), big.mark = ",")),
  "Phlebotomist (sample collection)", 0.1, 800, "R139", paste0("R", format(round(phlebotomy_annual_cost), big.mark = ",")),
  "**Subtotal Hospital Personnel**", NA, NA, NA, paste0("**R", format(round(hospital_personnel_total), big.mark = ","), "**")
)

hospital_workload %>%
  gt() %>%
  tab_header(
    title = "Hospital Workload Costs",
    subtitle = "Induced demand from improved case detection (800 additional cases/year)"
  ) %>%
  cols_label(
    Time_per_Case_Hours = "Time per Case (hours)",
    Incremental_Cases_per_Year = "Additional Cases/Year",
    Hourly_Rate = "Hourly Rate"
  ) %>%
  cols_align(align = "right", columns = c(Time_per_Case_Hours, Incremental_Cases_per_Year, Hourly_Rate, Annual_Cost)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 4)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—")
```

**Hospital personnel subtotal: ~R149,000/year** (800 cases × ~R186 average cost/case)

#### Hospital Consumables & Data Systems

```{r}
#| label: tbl-hospital-consumables
#| tbl-cap: "Hospital consumables and data systems"

# Consumables per case
nmc_form_printing <- 5  # R5/form (carbon-copy triplicate forms)
phlebotomy_supplies <- 15  # R15/case (vacutainer, alcohol swabs, labels)
data_entry_system_user_license <- 50000 / 1700  # R50k annual license / 1700 cases = ~R29/case

consumables_per_case <- nmc_form_printing + phlebotomy_supplies + data_entry_system_user_license
consumables_annual_cost <- incremental_cases * consumables_per_case

hospital_consumables <- tribble(
  ~Item, ~Cost_per_Case, ~Incremental_Cases, ~Annual_Cost,
  "NMC form printing (triplicate)", "R5", 800, "R4,000",
  "Phlebotomy supplies", "R15", 800, "R12,000",
  "Data entry system licenses (pro-rated)", "R29", 800, "R23,200",
  "**Subtotal Consumables**", NA, NA, paste0("**R", format(round(consumables_annual_cost), big.mark = ","), "**")
)

hospital_consumables %>%
  gt() %>%
  tab_header(title = "Hospital Consumables & Data Systems") %>%
  cols_label(
    Cost_per_Case = "Cost per Case",
    Incremental_Cases = "Additional Cases/Year"
  ) %>%
  cols_align(align = "right", columns = c(Cost_per_Case, Incremental_Cases, Annual_Cost)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 4)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—")
```

**Hospital consumables subtotal: ~R39,000/year**

**Total hospital downstream costs: ~R188,000/year**

---

### Downstream Cost Component 2: Laboratory Testing Volume Escalation

**Scenario:** Improved surveillance increases **demand** for BChE screening tests (as clinicians become more aware) **and** confirmatory GC-MS toxicology for hotspots.

#### Increased BChE Screening Volume

```{r}
#| label: tbl-lab-volume
#| tbl-cap: "Laboratory testing volume escalation"

# Current state
current_bche_tests_per_year <- 12000  # NHLS does ~12,000 BChE tests/year currently
current_test_cost <- 150  # R150/test (reagents, technician time, QC)

# Full Package scenario: 20% increase in test volume as awareness spreads
awareness_multiplier <- 1.20
full_package_bche_tests <- current_bche_tests_per_year * awareness_multiplier
incremental_bche_tests <- full_package_bche_tests - current_bche_tests_per_year

incremental_bche_cost <- incremental_bche_tests * current_test_cost

lab_volume <- tribble(
  ~Test_Type, ~Current_Volume, ~Full_Package_Volume, ~Incremental_Volume, ~Cost_per_Test, ~Incremental_Cost,
  "BChE screening (awareness-driven)", "12,000", "14,400", "2,400", "R150", paste0("R", format(round(incremental_bche_cost), big.mark = ","))
)

lab_volume %>%
  gt() %>%
  tab_header(
    title = "BChE Screening Volume Escalation",
    subtitle = "20% increase as clinician awareness improves"
  ) %>%
  cols_label(
    Current_Volume = "Current Annual Volume",
    Full_Package_Volume = "Full Package Volume",
    Incremental_Volume = "Incremental Tests/Year",
    Cost_per_Test = "Cost per Test"
  ) %>%
  cols_align(align = "right", columns = c(Current_Volume, Full_Package_Volume, Incremental_Volume, Cost_per_Test, Incremental_Cost))
```

**BChE screening escalation: ~R360,000/year**

#### Confirmatory GC-MS Toxicology for Hotspots

```{r}
#| label: tbl-confirmatory-tox
#| tbl-cap: "Confirmatory toxicology testing for hotspots"

# Hotspot-driven confirmatory testing
current_hotspots_per_year <- 4.5  # Status quo detects ~4.5 hotspots/year
full_package_hotspots_per_year <- 8.3  # Full Package detects 8.3 hotspots/year
incremental_hotspots <- full_package_hotspots_per_year - current_hotspots_per_year

# Each hotspot triggers 5 confirmatory GC-MS tests (index case + 4 cluster contacts)
gcms_tests_per_hotspot <- 5
incremental_gcms_tests <- incremental_hotspots * gcms_tests_per_hotspot

# GC-MS cost
gcms_test_cost <- 2500  # R2,500/test (reagents, technician time, instrument maintenance, reference standards)

incremental_gcms_cost <- incremental_gcms_tests * gcms_test_cost

confirmatory_tox <- tribble(
  ~Test_Type, ~Hotspots_per_Year, ~Tests_per_Hotspot, ~Total_Tests, ~Cost_per_Test, ~Annual_Cost,
  "GC-MS confirmatory toxicology", "3.8 additional", 5, "19", "R2,500", paste0("R", format(round(incremental_gcms_cost), big.mark = ","))
)

confirmatory_tox %>%
  gt() %>%
  tab_header(
    title = "Confirmatory Toxicology (GC-MS)",
    subtitle = "For hotspot investigation and cluster confirmation"
  ) %>%
  cols_label(
    Hotspots_per_Year = "Additional Hotspots/Year",
    Tests_per_Hotspot = "Tests per Hotspot",
    Total_Tests = "Total Additional Tests",
    Cost_per_Test = "Cost per Test"
  ) %>%
  cols_align(align = "right", columns = everything())
```

**Confirmatory GC-MS: ~R48,000/year**

#### Sample Transport & Storage

```{r}
#| label: tbl-sample-logistics
#| tbl-cap: "Sample transport and cold storage"

# Sample logistics
incremental_samples <- incremental_bche_tests + incremental_gcms_tests  # 2,400 + 19 = 2,419
courier_cost_per_sample <- 25  # R25/sample (courier from clinic to central lab)
cold_storage_cost_per_sample <- 10  # R10/sample/month × 0.5 months average = R5/sample

sample_logistics_cost <- incremental_samples * (courier_cost_per_sample + cold_storage_cost_per_sample)

sample_logistics <- tribble(
  ~Item, ~Incremental_Samples, ~Cost_per_Sample, ~Annual_Cost,
  "Courier (clinic → central lab)", "2,419", "R25", paste0("R", format(round(incremental_samples * courier_cost_per_sample), big.mark = ",")),
  "Cold storage (–20°C freezer)", "2,419", "R10", paste0("R", format(round(incremental_samples * cold_storage_cost_per_sample), big.mark = ",")),
  "**Subtotal Logistics**", NA, NA, paste0("**R", format(round(sample_logistics_cost), big.mark = ","), "**")
)

sample_logistics %>%
  gt() %>%
  tab_header(title = "Sample Transport & Storage") %>%
  cols_label(
    Incremental_Samples = "Additional Samples/Year",
    Cost_per_Sample = "Cost per Sample"
  ) %>%
  cols_align(align = "right", columns = c(Incremental_Samples, Cost_per_Sample, Annual_Cost)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 3)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—")
```

**Sample logistics: ~R85,000/year**

**Total laboratory downstream costs: ~R493,000/year**

---

### Downstream Cost Component 3: Environmental Health Practitioner (EHP) Investigations

**Scenario:** Full Package detects 3.8 additional hotspots/year. Each hotspot triggers an **EHP field investigation** — site visit, environmental sampling (water, soil, food), source tracing, retailer inspection.

#### EHP Personnel Time

```{r}
#| label: tbl-ehp-workload
#| tbl-cap: "EHP investigation workload"

# EHP investigation parameters
incremental_hotspots_ehp <- 3.8  # Same as above: 8.3 - 4.5 = 3.8 additional hotspots/year
hours_per_investigation <- 16  # 2 days × 8 hours (site visit, sampling, report writing)
ehp_hourly_rate <- 320000 / 1800  # ~R178/hour (R320k annual salary, 1800 hours/year)

ehp_personnel_cost <- incremental_hotspots_ehp * hours_per_investigation * ehp_hourly_rate

ehp_workload <- tribble(
  ~Activity, ~Hotspots_per_Year, ~Hours_per_Investigation, ~Hourly_Rate, ~Annual_Cost,
  "EHP field investigations", 3.8, 16, "R178", paste0("R", format(round(ehp_personnel_cost), big.mark = ","))
)

ehp_workload %>%
  gt() %>%
  tab_header(
    title = "EHP Investigation Workload",
    subtitle = "Field investigations for additional hotspots"
  ) %>%
  cols_label(
    Hotspots_per_Year = "Additional Hotspots/Year",
    Hours_per_Investigation = "Hours per Investigation",
    Hourly_Rate = "EHP Hourly Rate"
  ) %>%
  cols_align(align = "right", columns = everything())
```

**EHP personnel time: ~R108,000/year**

#### Environmental Sampling & Analysis

```{r}
#| label: tbl-env-sampling
#| tbl-cap: "Environmental sampling and laboratory analysis"

# Environmental sampling per hotspot
water_samples_per_hotspot <- 3
soil_samples_per_hotspot <- 2
food_samples_per_hotspot <- 2

total_samples_per_hotspot <- water_samples_per_hotspot + soil_samples_per_hotspot + food_samples_per_hotspot

# Lab analysis costs
pesticide_residue_analysis_cost <- 1500  # R1,500/sample (GC-MS or LC-MS/MS for multi-residue screen)

env_sampling_cost_per_hotspot <- total_samples_per_hotspot * pesticide_residue_analysis_cost
env_sampling_annual_cost <- incremental_hotspots_ehp * env_sampling_cost_per_hotspot

# Sampling consumables
sampling_kit_cost_per_hotspot <- 500  # R500/hotspot (sterile containers, labels, transport media)

total_env_sampling_cost <- env_sampling_annual_cost + (incremental_hotspots_ehp * sampling_kit_cost_per_hotspot)

env_sampling <- tribble(
  ~Item, ~Samples_per_Hotspot, ~Cost_per_Sample, ~Hotspots_per_Year, ~Annual_Cost,
  "Water/soil/food pesticide analysis", 7, "R1,500", 3.8, paste0("R", format(round(env_sampling_annual_cost), big.mark = ",")),
  "Sampling kits & consumables", NA, "R500/hotspot", 3.8, paste0("R", format(round(incremental_hotspots_ehp * sampling_kit_cost_per_hotspot), big.mark = ",")),
  "**Subtotal Environmental Sampling**", NA, NA, NA, paste0("**R", format(round(total_env_sampling_cost), big.mark = ","), "**")
)

env_sampling %>%
  gt() %>%
  tab_header(title = "Environmental Sampling & Analysis") %>%
  cols_label(
    Samples_per_Hotspot = "Samples per Hotspot",
    Cost_per_Sample = "Cost per Sample",
    Hotspots_per_Year = "Additional Hotspots/Year"
  ) %>%
  cols_align(align = "right", columns = c(Samples_per_Hotspot, Cost_per_Sample, Hotspots_per_Year, Annual_Cost)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 3)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—")
```

**Environmental sampling: ~R42,000/year**

#### EHP Travel & Reporting

```{r}
#| label: tbl-ehp-travel
#| tbl-cap: "EHP travel and reporting costs"

# Travel
distance_per_investigation_km <- 150  # Average round-trip distance (urban + rural)
vehicle_cost_per_km <- 4.50  # Government rate (fuel + maintenance)
travel_cost_per_investigation <- distance_per_investigation_km * vehicle_cost_per_km

travel_annual_cost <- incremental_hotspots_ehp * travel_cost_per_investigation

# Report writing (admin time)
report_writing_hours <- 4  # 4 hours to compile findings, liaise with DALRRD
admin_hourly_rate <- 200000 / 1800  # ~R111/hour (admin support)
reporting_cost_per_investigation <- report_writing_hours * admin_hourly_rate

reporting_annual_cost <- incremental_hotspots_ehp * reporting_cost_per_investigation

total_ehp_travel_reporting <- travel_annual_cost + reporting_annual_cost

ehp_travel <- tribble(
  ~Item, ~Cost_per_Investigation, ~Investigations_per_Year, ~Annual_Cost,
  "Travel (150 km @ R4.50/km)", "R675", 3.8, paste0("R", format(round(travel_annual_cost), big.mark = ",")),
  "Report compilation (admin time)", "R444", 3.8, paste0("R", format(round(reporting_annual_cost), big.mark = ",")),
  "**Subtotal Travel & Reporting**", NA, NA, paste0("**R", format(round(total_ehp_travel_reporting), big.mark = ","), "**")
)

ehp_travel %>%
  gt() %>%
  tab_header(title = "EHP Travel & Reporting") %>%
  cols_label(
    Cost_per_Investigation = "Cost per Investigation",
    Investigations_per_Year = "Additional Investigations/Year"
  ) %>%
  cols_align(align = "right", columns = c(Cost_per_Investigation, Investigations_per_Year, Annual_Cost)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 3)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—")
```

**EHP travel & reporting: ~R4,300/year**

**Total EHP downstream costs: ~R154,000/year**

---

### Downstream Cost Component 4: Agricultural Enforcement (DALRRD)

**Scenario:** Environmental sampling identifies **illegal pesticide sales** or **misuse of restricted products** in ~50% of hotspots. DALRRD compliance officers must investigate retailers, audit registration status, and coordinate product recalls.

#### DALRRD Compliance Officer Workload

```{r}
#| label: tbl-dalrrd-workload
#| tbl-cap: "DALRRD compliance officer workload"

# Enforcement parameters
hotspots_triggering_enforcement <- incremental_hotspots_ehp * 0.5  # 50% of hotspots involve illegal sales
hours_per_enforcement_case <- 24  # 3 days (retailer inspection, product sampling, registration audit, liaison with SAPS if criminal)
dalrrd_officer_hourly_rate <- 350000 / 1800  # ~R194/hour (R350k salary)

dalrrd_personnel_cost <- hotspots_triggering_enforcement * hours_per_enforcement_case * dalrrd_officer_hourly_rate

dalrrd_workload <- tribble(
  ~Activity, ~Cases_per_Year, ~Hours_per_Case, ~Hourly_Rate, ~Annual_Cost,
  "DALRRD compliance investigations", 1.9, 24, "R194", paste0("R", format(round(dalrrd_personnel_cost), big.mark = ","))
)

dalrrd_workload %>%
  gt() %>%
  tab_header(
    title = "DALRRD Compliance Officer Workload",
    subtitle = "Investigations of illegal pesticide sales (50% of hotspots)"
  ) %>%
  cols_label(
    Cases_per_Year = "Enforcement Cases/Year",
    Hours_per_Case = "Hours per Case",
    Hourly_Rate = "Officer Hourly Rate"
  ) %>%
  cols_align(align = "right", columns = everything())
```

**DALRRD personnel: ~R89,000/year**

#### Product Sampling & Testing

```{r}
#| label: tbl-product-sampling
#| tbl-cap: "Product sampling and testing for registration compliance"

# Product testing
products_sampled_per_case <- 3  # Retailer typically stocks 3 suspect products
product_analysis_cost <- 3500  # R3,500/product (active ingredient assay, formulation check, stability testing)

product_testing_cost_per_case <- products_sampled_per_case * product_analysis_cost
product_testing_annual_cost <- hotspots_triggering_enforcement * product_testing_cost_per_case

product_sampling <- tribble(
  ~Item, ~Products_per_Case, ~Cost_per_Product, ~Cases_per_Year, ~Annual_Cost,
  "Product analysis (active ingredient assay)", 3, "R3,500", 1.9, paste0("R", format(round(product_testing_annual_cost), big.mark = ","))
)

product_sampling %>%
  gt() %>%
  tab_header(title = "Product Sampling & Testing") %>%
  cols_label(
    Products_per_Case = "Products Sampled per Case",
    Cost_per_Product = "Cost per Product",
    Cases_per_Year = "Enforcement Cases/Year"
  ) %>%
  cols_align(align = "right", columns = everything())
```

**Product testing: ~R20,000/year**

#### Legal & Administrative Costs

```{r}
#| label: tbl-legal-admin
#| tbl-cap: "Legal and administrative costs for enforcement"

# Legal costs (for contested cases)
contested_cases_proportion <- 0.3  # 30% of cases go to legal review
legal_cost_per_contested_case <- 15000  # R15,000 (attorney fees, court filing)

legal_annual_cost <- hotspots_triggering_enforcement * contested_cases_proportion * legal_cost_per_contested_case

# Administrative costs
admin_cost_per_case <- 2000  # R2,000 (notices, correspondence, file management)
admin_annual_cost <- hotspots_triggering_enforcement * admin_cost_per_case

total_legal_admin <- legal_annual_cost + admin_annual_cost

legal_admin <- tribble(
  ~Item, ~Cost_per_Case, ~Cases_per_Year, ~Annual_Cost,
  "Legal fees (30% contested)", "R15,000", "0.57", paste0("R", format(round(legal_annual_cost), big.mark = ",")),
  "Administrative costs (notices, filing)", "R2,000", "1.9", paste0("R", format(round(admin_annual_cost), big.mark = ",")),
  "**Subtotal Legal & Admin**", NA, NA, paste0("**R", format(round(total_legal_admin), big.mark = ","), "**")
)

legal_admin %>%
  gt() %>%
  tab_header(title = "Legal & Administrative Costs") %>%
  cols_label(
    Cost_per_Case = "Cost per Case",
    Cases_per_Year = "Cases/Year"
  ) %>%
  cols_align(align = "right", columns = c(Cost_per_Case, Cases_per_Year, Annual_Cost)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 3)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—")
```

**Legal & admin: ~R12,000/year**

**Total DALRRD enforcement costs: ~R121,000/year**

---

### Pressure-Tested Full Package: Comprehensive System Costs

```{r}
#| label: tbl-comprehensive-costs
#| tbl-cap: "Comprehensive system costs: narrow vs. pressure-tested"

comprehensive_costs <- tribble(
  ~Cost_Component, ~Narrow_Perspective, ~Pressure_Tested_Perspective,
  "**A. Surveillance coordination (narrow)**", "**R237,000**", "**R237,000**",
  "  - NICD/NHLS/PIH/Provincial DoH", "R237,000", "R237,000",
  "**B. Hospital downstream demand**", "**R0**", "**R188,000**",
  "  - Clinician & records clerk time", "—", "R149,000",
  "  - Consumables & data systems", "—", "R39,000",
  "**C. Laboratory testing escalation**", "**R0**", "**R493,000**",
  "  - BChE screening volume (+20%)", "—", "R360,000",
  "  - Confirmatory GC-MS toxicology", "—", "R48,000",
  "  - Sample transport & cold storage", "—", "R85,000",
  "**D. EHP field investigations**", "**R0**", "**R154,000**",
  "  - EHP personnel time", "—", "R108,000",
  "  - Environmental sampling & analysis", "—", "R42,000",
  "  - Travel & reporting", "—", "R4,000",
  "**E. Agricultural enforcement (DALRRD)**", "**R0**", "**R121,000**",
  "  - Compliance officer investigations", "—", "R89,000",
  "  - Product sampling & testing", "—", "R20,000",
  "  - Legal & administrative", "—", "R12,000",
  "", "", "",
  "**TOTAL ANNUAL COST**", "**R237,000**", "**R1,193,000**"
)

comprehensive_costs %>%
  gt() %>%
  tab_header(
    title = "Narrow vs. Pressure-Tested Comprehensive System Costs",
    subtitle = "Full Package (Options 1–5 combined) — base-case scenario"
  ) %>%
  cols_label(
    Cost_Component = "Cost Component",
    Narrow_Perspective = "Narrow (Surveillance Only)",
    Pressure_Tested_Perspective = "Comprehensive (Whole System)"
  ) %>%
  cols_align(align = "right", columns = c(Narrow_Perspective, Pressure_Tested_Perspective)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = c(1, 4, 6, 10, 14, 19))
  ) %>%
  tab_style(
    style = cell_fill(color = "lightyellow"),
    locations = cells_body(rows = 19)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "")
```

::: {.callout-important}
## Comprehensive System Cost: ~R1.19 Million/Year

The **pressure-tested** comprehensive figure is **5.0× higher** than the narrow surveillance coordination budget. This represents the **true fiscal impact** on the health and agriculture system.

**Key drivers:**

1. **Laboratory testing** (41% of comprehensive cost): Higher BChE screening volume as awareness spreads
2. **Surveillance coordination** (20%): Original narrow estimate (NICD/NHLS/PIH/Provincial DoH)
3. **Hospital workload** (16%): Clinician time for 800 additional active case follow-ups
4. **EHP investigations** (13%): Field investigations for 3.8 additional hotspots
5. **DALRRD enforcement** (10%): Compliance investigations for illegal pesticide sales
:::

---

### Sensitivity Analysis: Low, Base, and High Scenarios

The comprehensive cost estimate is sensitive to key parameters. Three scenarios:

```{r}
#| label: tbl-scenario-sensitivity
#| tbl-cap: "Scenario sensitivity: Low, Base, High cost estimates"

# Calculate scenario costs
# LOW scenario assumptions:
# - 10% BChE volume increase (not 20%)
# - 50% of base-case hospital workload (not all additional cases actively followed up)
# - 50% of base-case EHP investigations (hotspots clustered geographically)

low_scenario <- 237000 + (188000 * 0.5) + (493000 * 0.5) + (154000 * 0.5) + (121000 * 0.5)

# BASE scenario (calculated above)
base_scenario <- 1193000

# HIGH scenario assumptions:
# - 35% BChE volume increase (aggressive awareness campaign)
# - 150% hospital workload (includes re-testing, specialist consults)
# - 150% EHP investigations (wider geographic dispersion, repeat visits)
# - 200% DALRRD enforcement (criminal prosecutions for severe cases)

high_scenario <- 237000 + (188000 * 1.5) + (493000 * 1.75) + (154000 * 1.5) + (121000 * 2.0)

scenario_sensitivity <- tribble(
  ~Scenario, ~Comprehensive_Cost, ~Notes,
  "**Low** (conservative utilization)", paste0("**R", format(round(low_scenario), big.mark = ","), "**"), "Assumes 50% of base-case downstream demand (minimal awareness campaign, clustered hotspots)",
  "**Base** (expected utilization)", paste0("**R", format(round(base_scenario), big.mark = ","), "**"), "Best estimate based on empirical parameters and expert elicitation",
  "**High** (aggressive utilization)", paste0("**R", format(round(high_scenario), big.mark = ","), "**"), "Assumes 150–200% of base-case downstream demand (active awareness, dispersed hotspots, criminal prosecutions)"
)

scenario_sensitivity %>%
  gt() %>%
  tab_header(
    title = "Scenario Sensitivity: Comprehensive System Costs",
    subtitle = "Range of plausible estimates for Full Package (Options 1–5)"
  ) %>%
  cols_align(align = "right", columns = Comprehensive_Cost) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = c(Scenario, Comprehensive_Cost))
  )
```

::: {.callout-note}
## Recommended Planning Figure

**For budgeting purposes, use the BASE scenario: ~R1.19 million/year** (range: R715k–R2.0M).

This represents:

- **0.002%** of National DoH budget (~R56 billion)
- **R132 per severe pesticide poisoning case** (9,000 cases/year)
- **Less than 3 mid-level epidemiologists' salaries**

Even the HIGH scenario (R2.0M) is **0.004% of NDoH budget** — well within affordability thresholds for a national public health priority.
:::

---

## Cost Drivers and Assumptions

### Personnel Cost Assumptions

- **Salary scales:** Based on 2025/2026 government payscales (OSD levels)
- **Benefits:** Not itemized separately (salaries are all-inclusive)
- **FTE allocation:** Estimated % of time required for surveillance activities

::: {.callout-note}
## FTE Allocation Methodology

FTE estimates based on:

1. **Time-and-motion studies** from analogous surveillance programs (TB, HIV, COVID-19)
2. **Expert elicitation** from NICD/NHLS/PIH staff
3. **Conservative assumptions** (overestimate rather than underestimate)

**Validation planned:** Pilot will track actual staff time to refine estimates.
:::

### IT Cost Assumptions

- **LIS configuration:** Based on NHLS IT quote for similar auto-notification rules
- **HL7 middleware:** Existing infrastructure (marginal cost only)
- **Maintenance:** 10% of development cost per year (industry standard)

### Other Cost Assumptions

- **Travel:** R250/km (government rate) × estimated distance
- **Venues:** R500/meeting for provincial coordination meetings
- **Sampling kits:** R1,000/kit × 10 investigations/year

## Sensitivity of Costs

### Cost Uncertainty

```{r}
#| label: tbl-cost-uncertainty
#| tbl-cap: "Cost uncertainty ranges for each option"

cost_uncertainty <- tribble(
  ~Option, ~Base_Case, ~Low_Estimate, ~High_Estimate, ~Uncertainty_Range,
  "Option 1", "R40,000", "R25,000", "R60,000", "±38%",
  "Option 2", "R40,000", "R22,000", "R65,000", "±54%",
  "Option 3", "R55,000", "R35,000", "R95,000", "±55%",
  "Option 4", "R62,000", "R45,000", "R85,000", "±32%",
  "Option 5", "R40,000", "R25,000", "R70,000", "±56%",
  "Full Package", "R237,000", "R152,000", "R375,000", "±47%"
)

cost_uncertainty %>%
  gt() %>%
  tab_header(
    title = "Cost Uncertainty Ranges",
    subtitle = "Low and high estimates based on parameter variation"
  ) %>%
  cols_align(align = "right", columns = c(Base_Case, Low_Estimate, High_Estimate, Uncertainty_Range)) %>%
  tab_footnote(
    footnote = "Uncertainty ranges derived from 10,000 Monte Carlo simulations",
    locations = cells_column_labels(columns = Uncertainty_Range)
  )
```

### Scenario: National Rollout (All 9 Provinces)

```{r}
#| label: tbl-national-rollout
#| tbl-cap: "Cost estimates for national rollout (all 9 provinces)"

national_rollout <- tribble(
  ~Option, ~Pilot_Cost_3_Provinces, ~National_Cost_9_Provinces, ~Incremental_Cost,
  "Option 3: BChE auto-notify", "R66,600", "R66,600", "R0",
  "Option 4: PIH reports", "R71,800", "R71,800", "R0",
  "Option 5: Provincial coord", "R101,000", "R303,000", "R202,000",
  "**Full Package**", "**R237,000**", "**R439,000**", "**R202,000**"
)

national_rollout %>%
  gt() %>%
  tab_header(
    title = "National Rollout Cost Estimates",
    subtitle = "Scaling from 3 pilot provinces to all 9 provinces"
  ) %>%
  cols_align(align = "right", columns = c(Pilot_Cost_3_Provinces, National_Cost_9_Provinces, Incremental_Cost)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 4)
  ) %>%
  tab_footnote(
    footnote = "Options 3 and 4 are national systems — no incremental cost for additional provinces",
    locations = cells_body(columns = Incremental_Cost, rows = c(1, 2))
  )
```

**Key insight:** Options 3 and 4 (BChE auto-notify and PIH reports) are **national systems** — they cost the same whether implemented in 3 provinces or 9. Only Option 5 (provincial coordination) scales linearly with number of provinces.

## Budget Impact Analysis

### Annual Budget Comparison

```{r}
#| label: fig-budget-context
#| fig-cap: "Annual cost of Full Package in context of NDoH budget"
#| fig-width: 8
#| fig-height: 4

budget_context <- tribble(
  ~Category, ~Annual_Budget,
  "National DoH total budget", 56000000000,
  "NICD operating budget", 180000000,
  "NHLS operating budget", 4500000000,
  "Full Package (pilot)", 237000
)

budget_context %>%
  mutate(Percentage_of_NDoH = Annual_Budget / 56000000000 * 100) %>%
  filter(Category != "National DoH total budget") %>%
  ggplot(aes(x = reorder(Category, Annual_Budget), y = Annual_Budget, fill = Category)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::dollar(Annual_Budget, prefix = "R", scale = 1e-6, suffix = "M")),
            hjust = -0.1, fontface = "bold", size = 4) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "R", scale = 1e-6, suffix = "M")) +
  scale_fill_brewer(palette = "Blues") +
  coord_flip() +
  labs(
    title = "Full Package Cost in Context",
    subtitle = "Annual budget for pesticide surveillance vs. institutional budgets",
    x = "",
    y = "Annual Budget (ZAR, millions)",
    caption = "Full Package = R237,000 (0.0004% of NDoH budget)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(size = 11)
  )
```

### Cost per Case Averted (Illustrative)

Assuming Full Package prevents even a **modest reduction in severe cases**:

```{r}
#| label: tbl-cost-per-case
#| tbl-cap: "Cost per case averted (illustrative scenarios)"

cost_per_case <- tribble(
  ~Assumption, ~Cases_Prevented, ~Cost_per_Case_Prevented,
  "Conservative (5% reduction)", "100", "R2,370",
  "Moderate (10% reduction)", "200", "R1,185",
  "Optimistic (20% reduction)", "400", "R593"
)

cost_per_case %>%
  gt() %>%
  tab_header(
    title = "Cost per Severe Poisoning Case Prevented",
    subtitle = "Illustrative scenarios (effectiveness TBD from pilot)"
  ) %>%
  cols_align(align = "right", columns = c(Cases_Prevented, Cost_per_Case_Prevented)) %>%
  tab_footnote(
    footnote = "Based on Full Package annual cost of R237,000 and baseline ~1,779 severe cases/year (NHLS LIS 2023)",
    locations = cells_column_labels(columns = Cost_per_Case_Prevented)
  )
```

**For comparison:**

- **Cost per DALY averted** for other public health interventions in South Africa:
  - Insecticide-treated bed nets (malaria): ~R800/DALY
  - HPV vaccination: ~R2,500/DALY
  - Rotavirus vaccination: ~R1,800/DALY

**Implication:** Even if Full Package prevents only 100 cases/year, it's comparable to other cost-effective interventions.

## Funding Sources

### Potential Funding Mechanisms (Comprehensive System Costs)

| Source | Viability | Estimated Amount | Notes |
|:-------|:----------|:-----------------|:------|
| **NDoH central budget** | High | R800,000 | Hospital workload, surveillance coordination, lab escalation |
| **NHLS operational budget** | High | R300,000 | BChE screening volume, confirmatory toxicology, sample logistics |
| **Provincial conditional grants** | Medium | R200,000 | EHP investigations, provincial coordination |
| **DALRRD enforcement budget** | Medium | R120,000 | Compliance investigations, product testing, legal costs |
| **Research grants (pilot only)** | Low | R200,000 | Non-sustainable; pilot-phase seed funding |

: Potential funding sources for comprehensive system costs {.striped}

::: {.callout-note}
## Multi-Sectoral Funding is Essential

Unlike the narrow surveillance coordination budget (R237k), the comprehensive system costs (~R1.2M) require **buy-in from multiple departments**:

- **Department of Health:** Hospital workload, lab testing, EHP investigations (~R900k)
- **Department of Agriculture, Land Reform & Rural Development (DALRRD):** Compliance enforcement (~R120k)
- **Provincial Health Departments:** EHP staffing, field investigations (~R180k)

**This is appropriate** — pesticide poisoning surveillance is an inter-sectoral problem. Funding should be matched to responsibilities.
:::

### Recommended Funding Strategy (Comprehensive Costs)

**Pilot phase (6 months, 3 provinces):**

- Research grant or donor funding: R400,000 (setup costs)
- NDoH central budget: R300,000
- NHLS operational budget: R150,000
- Provincial conditional grants: R100,000
- **Total pilot budget: R950,000**

**National rollout (ongoing, 3 provinces):**

- NDoH central budget: R800,000/year
- NHLS operational budget: R300,000/year
- Provincial conditional grants: R200,000/year
- DALRRD enforcement budget: R120,000/year
- **Total ongoing (3 provinces): R1.42 million/year**

**Scale to 9 provinces (Year 3+):**

- Incremental costs for Options 1–4: Minimal (national systems)
- Incremental costs for Option 5 (EHP investigations): +R300,000 (6 additional provinces)
- **Total national (9 provinces): R1.72 million/year**

## Limitations of Cost Estimates

::: {.callout-warning}
## Key Limitations

**Narrow surveillance coordination estimates (R237k):**

1. **Placeholder salary data:** Based on typical government scales, not confirmed individual salaries
2. **FTE estimates unvalidated:** Staff time requirements will be measured during pilot
3. **No scale economies:** Assumes linear scaling with number of provinces (may underestimate efficiencies)

**Comprehensive system estimates (R1.19M):**

1. **Induced demand is modeled, not observed:** We assume 20% BChE volume increase and 800 additional active case follow-ups. Actual utilization depends on clinician awareness campaigns, which are uncertain.
2. **Hospital time-per-case assumptions:** Based on analogous surveillance programs (TB, HIV), not pesticide-specific time-and-motion studies.
3. **Hotspot-to-enforcement conversion rate:** We assume 50% of hotspots trigger DALRRD investigations. Actual rate depends on EHP referral practices and DALRRD capacity, both unknown.
4. **No capacity constraints modeled:** Estimates assume NHLS labs, hospital staff, and EHPs can absorb the incremental workload. If capacity is saturated, marginal costs may be higher (require new hires, not just time reallocation).
5. **Geographic variation ignored:** Urban hotspots (Gauteng, Western Cape) may have lower investigation costs than rural hotspots (Limpopo, Eastern Cape) due to travel distance. We use national averages.
6. **Exchange rate risk:** Lab reagent costs (especially GC-MS reference standards) are import-dependent and subject to ZAR/USD fluctuations.

**Both perspectives exclude:**

- Patient out-of-pocket costs and productivity losses
- Wider societal costs (community education, media engagement)
- Capital depreciation on existing NHLS lab equipment
- Opportunity costs (staff time diverted from other surveillance priorities)
:::

## Next Steps

### Cost Data Collection During Pilot

The pilot will track:

1. **Actual staff time** spent on surveillance activities (time logs)
2. **IT development hours** for LIS configuration (project management records)
3. **Travel costs** for field investigations (expense reports)
4. **Meeting costs** (venue hire, catering)

### Cost Model Refinement

Once pilot data are available:

1. Update FTE estimates with empirical time logs
2. Adjust IT costs based on actual development hours
3. Incorporate scale economies if observed
4. Add uncertainty distributions for probabilistic budget impact analysis

---

::: {.footer-cta}
**Questions about costing?** Contact the D2P team at [contact@example.com](mailto:contact@example.com)

**Download cost spreadsheet:** [Available upon request]
:::
