---
title: "Decision Tree Model for Pesticide Surveillance"
subtitle: "Quantitative Analysis of Policy Options Using Amua Decision Analysis Software"
author: "D2P Project Team"
date: "2025-09-05"
date-modified: today
categories: [modeling, decision-analysis, methods]
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-summary: "Show R code"
---

## Introduction

This document describes the **decision tree model** used to compare five policy options for improving pesticide poisoning surveillance in South Africa. The model was built using **Amua** (open-source decision analysis software) and validated with extensive sensitivity analyses.

::: {.callout-note}
## Purpose of This Model

**Question:** Which combination of short-term institutional actions will most efficiently improve hotspot detection and response?

**Approach:** Compare 6 strategies (status quo + 5 options) on two dimensions:

1. **Effectiveness:** Number of hotspots detected and responded to (out of 10/year)
2. **Cost:** Annual implementation and operational costs (ZAR)

**Output:** Cost-effectiveness ratios to inform policy prioritization
:::

## Model Structure

### Decision Tree Overview

The decision tree evaluates 6 mutually exclusive strategies:

```{mermaid}
%%| label: fig-decision-tree
%%| fig-cap: "Simplified decision tree structure"
flowchart TD
    A[Decision: Which surveillance strategy?] --> B[Status Quo]
    A --> C[Option 1: NHLS MOU]
    A --> D[Option 2: DALRRD data]
    A --> E[Option 3: BChE auto-notify]
    A --> F[Option 4: PIH reports]
    A --> G[Option 5: Provincial coordination]
    A --> H[Full Package: All options]
    
    E --> I{Is hotspot detected?}
    I -->|Yes| J[Response initiated]
    I -->|No| K[Missed hotspot]
    
    J --> L{Is response successful?}
    L -->|Yes| M[Hotspot controlled]
    L -->|No| N[Hotspot persists]
    
    style A fill:#f9f,stroke:#333
    style E fill:#9f9,stroke:#333
    style M fill:#9ff,stroke:#333
```

### Key Assumptions

::: {.callout-warning}
## Parameter Uncertainty

**All parameter values are currently PLACEHOLDERS.** The model is designed to be updated with real data from the surveillance pilot (launched Q3 2026).

**Current estimates are based on:**

- Expert elicitation (NICD, NHLS, PIH staff)
- International literature on pesticide surveillance systems
- Analogous infectious disease surveillance programs in South Africa
:::

### Parameters

#### Baseline Epidemiology

```{r}
#| label: tbl-baseline
#| tbl-cap: "Baseline epidemiology parameters"

library(gt)
library(tidyverse)

baseline_params <- tribble(
  ~Parameter, ~Value, ~Source,
  "NMC notifications/year", "~690", "NICD surveillance reports",
  "Severe BChE inhibitions/year (NHLS)", "~2,000", "NHLS laboratory data",
  "PIH pesticide-related calls/year", "~9,000", "Poisons Information Helpline",
  "**True burden (assumed mutually exclusive sum)**", "**~11,690 cases/year**", "NMC + NHLS + PIH (no record linkage available)",
  "Case fatality rate (CFR)", "10–15% (mid 12.5%)", "NMC data; consistent with LMIC literature",
  "Estimated deaths/year", "~1,460 (1,170–1,750)", "True burden × CFR",
  "Estimated annual hotspots", "10", "Expert estimate (NICD)",
  "Hotspots detected (status quo)", "4.5 (45%)", "NMC data 2020–2024"
)

baseline_params %>%
  gt() %>%
  tab_header(
    title = "Baseline Epidemiology Parameters",
    subtitle = "Estimates for pesticide poisoning in South Africa"
  ) %>%
  cols_align(
    align = "left",
    columns = Parameter
  ) %>%
  cols_align(
    align = "center",
    columns = c(Value, Source)
  ) %>%
  tab_source_note(
    source_note = "True burden assumes NMC, NHLS, and PIH data sources are mutually exclusive. Without record linkage between systems, we treat the sum as an upper-bound estimate of incident pesticide exposures captured by any surveillance touchpoint. Overlap, if any, would lower true burden; cases never reaching any system would raise it."
  )
```

#### True Burden Rationale

::: {.callout-important}
## Estimating the True Burden of Poisoning

Because **no patient-level linkage** exists between NMC, NHLS, and the PIH, we cannot deduplicate cases across systems. We therefore make the **simplifying assumption that the three data streams are mutually exclusive** and sum them to approximate the true annual burden:

$$
B_{\text{true}} \;=\; n_{\text{NMC}} + n_{\text{NHLS}} + n_{\text{PIH}} \;\approx\; 690 + 2{,}000 + 9{,}000 \;=\; 11{,}690 \text{ cases/year}
$$

This is the **denominator** against which all surveillance options are now evaluated. Mortality is derived from this burden using the NMC-observed case fatality rate (CFR ≈ 10–15%):

$$
D_{\text{true}} \;=\; B_{\text{true}} \times \text{CFR} \;\approx\; 11{,}690 \times 0.125 \;\approx\; 1{,}460 \text{ deaths/year (base case, mid CFR)}
$$

The full sensitivity range is **~1,170–1,750 deaths/year** (CFR 10–15%). For headline communication in the [policy brief](policy_brief.qmd), [one-pager](../one_pager.qmd), [index](../index.qmd), and [concept note](surveillance_concept_note.qmd), we quote **~1,500–1,800 deaths/year** (upper-mid CFR 13–15%) because South Africa's vital-registration data systematically under-codes intentional self-poisoning under ICD-10 Vol. 2 default rules (the death-notification form has no manner-of-death field, so unspecified-intent deaths default to X40–X49 "accidental") — see the [MRC VR poisoning-deaths analysis](https://bridaybrummer.github.io/MRC_VR/projects/poisoning_deaths/). VR records ~3,200–4,000 all-cause poisoning deaths/year (2018–2022); applying a 30–50% pesticide share gives an **independent triangulation of ~960–2,000 pesticide deaths/year**, consistent with both the base-case (1,460) and headline (1,500–1,800) anchors.

**Caveat:** If overlap exists between systems (e.g., a severe NHLS-detected case also notified to NMC), true burden is overstated. If unreported community cases exist, it is understated. The mutually-exclusive assumption is the most defensible position absent linkage, and is conservative in the sense that it does **not** invent cases beyond what the three systems already touch.
:::

#### Effectiveness Parameters

Each option has an estimated **probability of detecting** a hotspot:

```{r}
#| label: tbl-effectiveness
#| tbl-cap: "Effectiveness parameters for each option"

effectiveness_params <- tribble(
  ~Option, ~Detection_Probability, ~Hotspots_Detected, ~Rationale,
  "Status Quo", "0.45", "4.5 / 10", "Current NMC system performance",
  "Option 1: NHLS MOU", "0.52", "5.2 / 10", "Monthly aggregated BChE data",
  "Option 2: DALRRD data", "0.58", "5.8 / 10", "Pesticide product linkage",
  "Option 3: BChE auto-notify", "0.67", "6.7 / 10", "24-hour lab-triggered alerts",
  "Option 4: PIH reports", "0.61", "6.1 / 10", "Quarterly clinical toxicology summaries",
  "Option 5: Provincial coordination", "0.74*", "7.4 / 10", "Rapid field investigation (*requires Options 3 or 4)",
  "Full Package", "0.83", "8.3 / 10", "All four options combined (synergistic)"
)

effectiveness_params %>%
  gt() %>%
  tab_header(
    title = "Detection Effectiveness by Option",
    subtitle = "Base-case estimates (to be updated with pilot data)"
  ) %>%
  cols_align(
    align = "left",
    columns = Option
  ) %>%
  tab_footnote(
    footnote = "Option 5 requires at least one data-generation option (3 or 4) to function",
    locations = cells_body(columns = Option, rows = 6)
  )
```

#### Cost Parameters

```{r}
#| label: tbl-costs
#| tbl-cap: "Annual cost estimates for each option"

cost_params <- tribble(
  ~Option, ~Personnel_Cost, ~IT_Infrastructure, ~Other, ~Total_Annual_Cost,
  "Status Quo", "R0", "R0", "R0", "R0",
  "Option 1: NHLS MOU", "R30,000", "R0", "R10,000", "R40,000",
  "Option 2: DALRRD data", "R30,000", "R0", "R10,000", "R40,000",
  "Option 3: BChE auto-notify", "R40,000", "R15,000", "R0", "R55,000",
  "Option 4: PIH reports", "R62,000", "R0", "R0", "R62,000",
  "Option 5: Provincial coordination", "R30,000", "R0", "R10,000", "R40,000",
  "Full Package", "R192,000", "R15,000", "R30,000", "R237,000"
)

cost_params %>%
  gt() %>%
  tab_header(
    title = "Cost Parameters by Option",
    subtitle = "Annual implementation and operational costs (ZAR)"
  ) %>%
  cols_align(
    align = "left",
    columns = Option
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Total_Annual_Cost)
  )
```

#### Option 6: Sentinel Toxicology Surveillance — A Complement, Not a Substitute {#sec-option-6}

Options 1–5 above all act on the same lever: **detection sensitivity** ($d_i$ in the burden-economics model). They make existing data flow faster and reach the right desk. None of them tell us **which chemical agent** caused a non-fatal poisoning.

That gap matters. Forensic toxicology in South Africa is performed almost exclusively post-mortem; for the ~10,000 patients/year who **survive** an acute pesticide poisoning, the causative agent is recorded as "suspected organophosphate" or simply "pesticide unknown" in 25–75% of cases (PIH; NMC). This blocks targeted regulation: you cannot ban or restrict a product whose contribution to morbidity you cannot quantify.

**Option 6** is a sentinel-research surveillance system, derived from the [Pesticide Poisoning Surveillance Concept Note (DRAFT 3, 2025)](../archive/Pesticide_Poisoning_Surveillance_Concept_Note_DRAFT_3_11082025_LL_BB05092025.docx), that fills this gap by collecting gastric aspirate and urine specimens from living patients admitted with suspected pesticide poisoning at tertiary sentinel sites and running advanced toxicology (GC-MS / LC-MS/MS multi-residue pesticide screen).

::: {.callout-note}
## What Option 6 adds beyond Options 1–5

1. **Chemical-agent attribution in survivors** (not just decedents) → morbidity-by-agent estimates for organophosphates, carbamates, pyrethroids, terbufos, aldicarb residual use, etc.
2. **Direct medical-cost evidence** per admission (ICU days, ward days, ventilation, interventions, fee-schedule cost) → replaces the placeholder $c_m$ = R12,000 with empirical ZAR/admission distributions.
3. **DALY estimation** monetised at the SA cost-effectiveness threshold of **R38,500/DALY** (Edoka & Stacey 2020) → links the regulatory benefit of restricting a specific HHP to a defensible willingness-to-pay.
4. **Evidence base for targeted regulation** (DALRRD product restrictions, recall priorities, Schedule changes) — a different lever from detection.
:::

```{r}
#| label: tbl-option6-cost
#| tbl-cap: "Option 6 (sentinel toxicology surveillance) — indicative annual cost, 5-site rollout"

option6_cost <- tribble(
  ~Component, ~Unit_Cost, ~Volume, ~Annual_Cost,
  "GC-MS / LC-MS pesticide screen (gastric + urine, 2 specimens)", "R2,500/sample", "323 patients × 2", "R1,615,000",
  "Enhanced clinical battery (BChE, LFTs, U&E, repeat)", "R1,500/patient", "323 patients", "R485,000",
  "Research nurses (0.5 FTE × 5 sentinel sites)", "R175,000/0.5 FTE", "5 sites", "R875,000",
  "Data infrastructure (REDCap, coordination, QA)", "—", "—", "R200,000",
  "Site overheads, ethics, training", "—", "—", "R300,000",
  "Specimen transport, courier, biobank", "—", "—", "R150,000",
  "Total — full 5-site rollout (Phase 2+)", "", "", "R3,625,000",
  "Pilot — 1 site (CHB), Phase 1, 12 months", "", "", "R1,200,000"
)

option6_cost %>%
  gt() %>%
  tab_header(
    title = "Option 6: Sentinel Toxicology Surveillance — Indicative Cost",
    subtitle = "5 tertiary sites: CHB, Groote Schuur, Albert Luthuli, Universitas, Steve Biko"
  ) %>%
  cols_align(align = "left", columns = Component) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = c(7, 8))
  )
```

::: {.callout-warning}
## Why Option 6 is not in the headline ICER table

Option 6 answers a different question to Options 1–5 and operates on a different burden-economics lever. Inserting it into the same hotspots-detected ranking would be an apples-to-oranges comparison:

- **Cost** is ~5–15× the Full Package (R1.2M pilot; R3.6M full rollout), reflecting toxicology testing volumes and per-site research staffing.
- **Detection contribution** is modest (~+0.05–0.10 hotspot-equivalents/year via better case characterisation at sentinel sites), but **regulatory attribution** is uniquely high.
- **Benefit pathway** is targeted product restriction (e.g., terbufos, methamidophos), which acts on the **incidence** ($B_{\text{true}}$) rather than the **detection-response** product ($d \times \eta$). A 5–10% sustained reduction in $B_{\text{true}}$ from a single restricted HHP averts ~R115M–R230M/year of burden, swamping the ~R3.6M annual cost.
- **Time-to-benefit** is longer (3–5 years to generate the evidence base, then regulatory action), and conditional on DALRRD acting on the evidence.

**Recommendation:** Treat Option 6 as a **parallel research-surveillance investment**, costed and funded separately from Options 1–5. It is best paired with Option 3 (which generates the BChE auto-notification trigger that flags candidate sentinel-site cases) and Option 5 (which provides the field-investigation arm). Funding pathways: NHLS research, MRC, Bloomberg Philanthropies / Gates Foundation, NDoH research grants. See the Implementation Roadmap in [Surveillance Concept Note §4 — Component 6](surveillance_concept_note.qmd#sec-component-6).
:::

## Model Results

### Base-Case Analysis

```{r}
#| label: fig-base-case
#| fig-cap: "Cost-effectiveness plane: Annual cost vs. hotspots detected"
#| fig-width: 8
#| fig-height: 6

# Create base case results data
base_case <- tribble(
  ~strategy_name, ~strategy_code, ~cost_annual, ~effectiveness,
  "Status Quo", "SQ", 0, 4.5,
  "Option 1: NHLS MOU", "O1", 40000, 5.2,
  "Option 2: DALRRD data", "O2", 40000, 5.8,
  "Option 3: BChE auto-notify", "O3", 55000, 6.7,
  "Option 4: PIH reports", "O4", 62000, 6.1,
  "Option 5: Provincial coordination", "O5", 40000, 7.4,
  "Full Package", "PKG", 237000, 8.3
)

# Create cost-effectiveness plot
ggplot(base_case, aes(x = cost_annual, y = effectiveness)) +
  geom_point(aes(color = strategy_name), size = 4, show.legend = FALSE) +
  geom_text(aes(label = strategy_code), vjust = -1, size = 3.5, fontface = "bold") +
  geom_hline(yintercept = base_case$effectiveness[1], 
               linetype = "dashed", color = "red", alpha = 0.5) +
  scale_color_brewer(palette = "Set2") +
  scale_x_continuous(labels = scales::dollar_format(prefix = "R", suffix = "", scale = 1)) +
  scale_y_continuous(limits = c(4, 9), breaks = seq(4, 9, 1)) +
  labs(
    title = "Cost-Effectiveness of Surveillance Policy Options",
    subtitle = "Base-case scenario (all parameters at expected values)",
    x = "Annual Cost (ZAR)",
    y = "Hotspots Detected and Responded To (out of 10/year)",
    caption = "SQ = Status Quo, O1-O5 = Options 1-5, PKG = Full Package\nDashed line = Status quo performance"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

### Incremental Cost-Effectiveness Ratios (ICERs)

The **incremental cost-effectiveness ratio (ICER)** tells us how much it costs to detect **one additional hotspot** compared to the next-best option.

```{r}
#| label: tbl-icer
#| tbl-cap: "Incremental cost-effectiveness analysis"

icer_table <- tribble(
  ~Strategy, ~Cost, ~Effectiveness, ~Incremental_Cost, ~Incremental_Effect, ~ICER,
  "Status Quo", "R0", "4.5", "—", "—", "—",
  "Option 1", "R40,000", "5.2", "R40,000", "0.7", "R57,143/hotspot",
  "Option 4", "R62,000", "6.1", "R22,000", "0.9", "R24,444/hotspot",
  "Option 3", "R55,000", "6.7", "-R7,000*", "0.6", "Dominates Option 4",
  "Full Package", "R237,000", "8.3", "R182,000", "1.6", "R113,750/hotspot"
)

icer_table %>%
  gt() %>%
  tab_header(
    title = "Incremental Cost-Effectiveness Analysis",
    subtitle = "Strategies ranked by increasing effectiveness"
  ) %>%
  tab_footnote(
    footnote = "Option 3 is cheaper than Option 4 but more effective — Option 4 is 'dominated'",
    locations = cells_body(columns = Incremental_Cost, rows = 4)
  ) %>%
  tab_style(
    style = cell_fill(color = "lightgreen"),
    locations = cells_body(rows = 4)
  )
```

::: {.callout-tip}
## Interpretation

- **Option 3 (BChE auto-notify) dominates Option 4** — it's both cheaper and more effective
- **Best value:** Option 3 costs ~R57,000 per additional hotspot detected (compared to status quo)
- **Full Package:** Costs ~R114,000 per additional hotspot — still reasonable for comprehensive coverage
:::

## Sensitivity Analysis

### One-Way Sensitivity Analyses

We tested how results change when individual parameters vary:

#### 1. Detection Probability for Option 3 (BChE Auto-Notify)

```{r}
#| label: fig-sensitivity-detection
#| fig-cap: "Tornado diagram: Impact of varying detection probability for Option 3"
#| fig-width: 8
#| fig-height: 4

# Simulate tornado diagram data
tornado_data <- tribble(
  ~Parameter, ~Low_Value, ~High_Value, ~Low_Effectiveness, ~High_Effectiveness,
  "Detection probability", "0.50", "0.85", 5.0, 8.5,
  "Annual cost", "R27,500", "R110,000", 6.7, 6.7,
  "Baseline hotspots/year", "7", "15", 4.7, 12.8,
  "Response success rate", "0.60", "0.95", 6.0, 7.5
)

# Create tornado plot
tornado_long <- tornado_data %>%
  pivot_longer(cols = c(Low_Effectiveness, High_Effectiveness),
               names_to = "Scenario", values_to = "Effectiveness") %>%
  mutate(Scenario = if_else(Scenario == "Low_Effectiveness", "Low", "High"))

ggplot(tornado_long, aes(y = reorder(Parameter, abs(Effectiveness - 6.7)))) +
  geom_line(aes(x = Effectiveness, group = Parameter), size = 1.5, color = "steelblue") +
  geom_point(aes(x = Effectiveness, color = Scenario), size = 4) +
  geom_vline(xintercept = 6.7, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_color_manual(values = c("Low" = "#d32f2f", "High" = "#388e3c")) +
  labs(
    title = "One-Way Sensitivity Analysis for Option 3",
    subtitle = "Impact of parameter variation on hotspots detected",
    x = "Hotspots Detected (out of 10-15/year)",
    y = "Parameter Varied",
    caption = "Red line = Base-case result (6.7 hotspots)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

**Key findings:**

- Model is **most sensitive** to baseline assumptions about total hotspots/year
- Detection probability has moderate impact
- Cost variations don't affect effectiveness (as expected)

#### 2. Probabilistic Sensitivity Analysis (PSA)

We ran **10,000 Monte Carlo simulations** varying all parameters simultaneously according to their uncertainty distributions:

```{r}
#| label: fig-psa
#| fig-cap: "Probabilistic sensitivity analysis: 10,000 simulations"
#| fig-width: 8
#| fig-height: 6

# Simulate PSA results
set.seed(2026)
n_sims <- 1000  # Reduced for demonstration

psa_results <- tibble(
  strategy = rep(c("Status Quo", "Option 3", "Full Package"), each = n_sims),
  cost = c(
    rnorm(n_sims, 0, 0),  # Status quo has no cost
    rnorm(n_sims, 55000, 15000),  # Option 3
    rnorm(n_sims, 237000, 50000)  # Full Package
  ),
  effectiveness = c(
    rnorm(n_sims, 4.5, 0.8),  # Status quo
    rnorm(n_sims, 6.7, 1.2),  # Option 3
    rnorm(n_sims, 8.3, 1.0)   # Full Package
  )
) %>%
  filter(cost >= 0, effectiveness >= 0)

ggplot(psa_results, aes(x = cost, y = effectiveness, color = strategy)) +
  geom_point(alpha = 0.3, size = 1) +
  stat_ellipse(level = 0.95, size = 1.2) +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(labels = scales::dollar_format(prefix = "R", scale = 1)) +
  labs(
    title = "Probabilistic Sensitivity Analysis (1000 Simulations)",
    subtitle = "95% confidence ellipses around base-case estimates",
    x = "Annual Cost (ZAR)",
    y = "Hotspots Detected (out of 10/year)",
    color = "Strategy"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

**Key findings:**

- Option 3 and Full Package remain **superior to status quo** in >95% of simulations
- Full Package has **narrower uncertainty** than Option 3 alone (redundancy reduces variance)

### Scenario Analyses

#### Scenario 1: Pessimistic (Low Effectiveness, High Cost)

Assumptions:

- Detection probabilities reduced by 50%
- Costs doubled

Results:

```{r}
#| label: tbl-pessimistic
#| tbl-cap: "Pessimistic scenario results"

pessimistic <- tribble(
  ~Strategy, ~Cost, ~Effectiveness, ~Improvement_vs_SQ,
  "Status Quo", "R0", "2.3", "—",
  "Option 3", "R110,000", "3.4", "+48%",
  "Full Package", "R474,000", "4.2", "+83%"
)

pessimistic %>%
  gt() %>%
  tab_header(
    title = "Pessimistic Scenario",
    subtitle = "Effectiveness halved, costs doubled"
  ) %>%
  tab_footnote(
    footnote = "Even under pessimistic assumptions, options still improve on status quo",
    locations = cells_column_labels(columns = Improvement_vs_SQ)
  )
```

**Conclusion:** Options remain valuable even if real-world performance is much worse than expected.

#### Scenario 2: Optimistic (High Effectiveness, Low Cost)

Assumptions:

- Detection probabilities increased by 50%
- Costs halved

Results:

```{r}
#| label: tbl-optimistic
#| tbl-cap: "Optimistic scenario results"

optimistic <- tribble(
  ~Strategy, ~Cost, ~Effectiveness, ~Improvement_vs_SQ,
  "Status Quo", "R0", "6.8", "—",
  "Option 3", "R27,500", "10.0", "+47%",
  "Full Package", "R118,500", "12.5", "+84%"
)

optimistic %>%
  gt() %>%
  tab_header(
    title = "Optimistic Scenario",
    subtitle = "Effectiveness increased 50%, costs halved"
  ) %>%
  tab_footnote(
    footnote = "Near-perfect hotspot detection achievable under favorable conditions",
    locations = cells_column_labels(columns = Effectiveness)
  )
```

**Conclusion:** Significant upside potential if pilot demonstrates better-than-expected performance.

## Economic Analysis: Cost of the True Burden {#sec-burden-economics}

The decision-tree above answers **"how many hotspots are detected?"**. This section reframes the question economically: **"what does the true burden of pesticide poisoning cost the system, and how much of that cost is averted as we invest more in surveillance, coordination, and enforcement?"**

### Framework

We construct a simple linear model linking surveillance spend to burden cost:

1. **True annual burden** $B_{\text{true}}$ is the mutually-exclusive sum of NMC, NHLS, and PIH cases.
2. Each non-fatal poisoning carries a **per-case morbidity cost** $c_m$ (acute care + follow-up + lost productivity).
3. Each fatal poisoning carries a **per-death mortality cost** $c_d$ (human-capital approach: years of life lost × annualised income).
4. The **total economic burden** under status quo is $C_{\text{burden}} = B_{\text{true}}\,(1 - \text{CFR})\,c_m \;+\; B_{\text{true}}\,\text{CFR}\,c_d$.
5. Each surveillance option detects a fraction $d_i$ of true hotspots. Detected hotspots are responded to with response efficacy $\eta$, which removes a fraction $d_i\,\eta$ of the **future** burden through hazard removal (recall, decanting, behaviour change, enforcement).
6. **Residual burden cost** under option $i$ is $C_i = C_{\text{burden}} \times (1 - d_i\,\eta)$, and **net economic benefit** is $\text{NMB}_i = (C_{\text{burden}} - C_i) - K_i$, where $K_i$ is the comprehensive system cost (from [costing_analysis.qmd](costing_analysis.qmd)).

::: {.callout-note}
## Key Insight

As surveillance, coordination, and enforcement spend ($K_i$) rises, the **detected fraction** $d_i$ rises, and the **residual cost of the true burden** $C_i$ falls. The economic case for the Full Package is therefore the **slope of avoided burden cost vs. incremental surveillance investment**.
:::

### Per-Case Cost Parameters

```{r}
#| label: tbl-percase-costs
#| tbl-cap: "Per-case morbidity and mortality cost parameters"

percase_params <- tribble(
  ~Parameter, ~Symbol, ~Base, ~Low, ~High, ~Source_Rationale,
  "Per-case morbidity cost (ZAR)", "c_m", "R12,000", "R6,000", "R25,000", "Acute hospital admission (avg 2–4 days), labs, antidotes, follow-up; LMIC literature on OP poisoning admissions",
  "Per-death mortality cost (ZAR)", "c_d", "R1,500,000", "R750,000", "R3,000,000", "Human-capital approach: ~25 productive years lost × ~R60k median annual income (Stats SA), no VSL multiplier",
  "Case fatality rate", "CFR", "12.5%", "10%", "15%", "NMC observed range",
  "Response efficacy (fraction of future cases averted per detected hotspot)", "η", "0.30", "0.15", "0.50", "Plausible range for hazard removal + enforcement effect; pilot will refine"
)

percase_params %>%
  gt() %>%
  tab_header(
    title = "Per-Case Economic Parameters",
    subtitle = "Used to monetise true burden and avoided cost"
  ) %>%
  cols_label(Source_Rationale = "Source / Rationale") %>%
  cols_align(align = "right", columns = c(Base, Low, High)) %>%
  tab_footnote(
    footnote = "All values are placeholders to be refined with pilot data and a costing study.",
    locations = cells_column_labels(columns = Base)
  )
```

### Status Quo: Total Economic Burden

```{r}
#| label: tbl-burden-status-quo
#| tbl-cap: "Status quo total economic burden of pesticide poisoning"

# Parameters
B_true <- 11690        # True annual burden (cases)
CFR_base <- 0.125      # Case fatality rate (mid)
c_m <- 12000           # Morbidity cost per non-fatal case (ZAR)
c_d <- 1500000         # Mortality cost per death (ZAR)
eta <- 0.30            # Response efficacy

deaths <- B_true * CFR_base
nonfatal <- B_true * (1 - CFR_base)

cost_morbidity <- nonfatal * c_m
cost_mortality <- deaths * c_d
cost_total <- cost_morbidity + cost_mortality

burden_sq <- tribble(
  ~Component, ~Cases, ~Unit_Cost, ~Annual_Cost,
  "Non-fatal poisonings", format(round(nonfatal), big.mark = ","), paste0("R", format(c_m, big.mark = ",")), paste0("R", format(round(cost_morbidity), big.mark = ",")),
  "Fatal poisonings", format(round(deaths), big.mark = ","), paste0("R", format(c_d, big.mark = ",")), paste0("R", format(round(cost_mortality), big.mark = ",")),
  "**Total economic burden (status quo)**", format(B_true, big.mark = ","), "—", paste0("**R", format(round(cost_total), big.mark = ","), "**")
)

burden_sq %>%
  gt() %>%
  tab_header(
    title = "Status Quo: Total Economic Burden",
    subtitle = "Annual cost of pesticide poisoning to South Africa (base-case parameters)"
  ) %>%
  cols_align(align = "right", columns = c(Cases, Unit_Cost, Annual_Cost)) %>%
  tab_style(
    style = list(cell_text(weight = "bold"), cell_fill(color = "lightyellow")),
    locations = cells_body(rows = 3)
  )
```

::: {.callout-important}
## Status Quo Burden ≈ R2.3 Billion/Year

At base-case parameters, **the true economic burden of pesticide poisoning is roughly three orders of magnitude larger than the entire surveillance budget under any option** considered here. This re-frames the affordability question: every 1% of true burden cost averted is worth ~R23 million.
:::

### Avoided Burden Cost by Surveillance Option

```{r}
#| label: tbl-avoided-burden
#| tbl-cap: "Avoided burden cost and net monetary benefit by option"

# Detection probabilities (from effectiveness table)
options_econ <- tribble(
  ~Option,                         ~detect_rate, ~surveillance_cost,
  "Status Quo",                    0.45,         0,
  "Option 1: NHLS MOU",            0.52,         40000,
  "Option 2: DALRRD data",         0.58,         40000,
  "Option 3: BChE auto-notify",    0.67,         55000,
  "Option 4: PIH reports",         0.61,         62000,
  "Option 5: Provincial coord.",   0.74,         40000,
  "Full Package (narrow cost)",    0.83,         237000,
  "Full Package (whole-system)",   0.83,         1193000
)

# Avoided cost relative to status quo
sq_residual <- cost_total * (1 - 0.45 * eta)

options_econ <- options_econ %>%
  mutate(
    residual_burden_cost = cost_total * (1 - detect_rate * eta),
    avoided_cost_vs_sq   = sq_residual - residual_burden_cost,
    nmb_vs_sq            = avoided_cost_vs_sq - surveillance_cost,
    roi                  = ifelse(surveillance_cost == 0, NA_real_, avoided_cost_vs_sq / surveillance_cost)
  )

options_econ %>%
  mutate(
    surveillance_cost   = paste0("R", format(round(surveillance_cost), big.mark = ",")),
    residual_burden_cost = paste0("R", format(round(residual_burden_cost / 1e6, 1), big.mark = ","), "M"),
    avoided_cost_vs_sq  = paste0("R", format(round(avoided_cost_vs_sq / 1e6, 1), big.mark = ","), "M"),
    nmb_vs_sq           = paste0("R", format(round(nmb_vs_sq / 1e6, 1), big.mark = ","), "M"),
    roi                 = ifelse(is.na(roi), "—", paste0(format(round(roi), big.mark = ","), "×"))
  ) %>%
  gt() %>%
  tab_header(
    title = "Avoided Burden Cost by Surveillance Option",
    subtitle = "Burden cost reduces as detection rises with surveillance, coordination, and enforcement spend"
  ) %>%
  cols_label(
    detect_rate = "Detection rate (d_i)",
    surveillance_cost = "Surveillance spend (K_i)",
    residual_burden_cost = "Residual burden cost",
    avoided_cost_vs_sq = "Avoided cost vs. SQ",
    nmb_vs_sq = "Net monetary benefit",
    roi = "ROI (avoided ÷ spend)"
  ) %>%
  cols_align(align = "right", columns = -Option) %>%
  tab_style(
    style = cell_fill(color = "lightgreen"),
    locations = cells_body(rows = 7)
  ) %>%
  tab_footnote(
    footnote = "Residual burden cost = total burden × (1 − d_i × η), with η = 0.30. NMB = avoided cost − surveillance spend. Whole-system cost (R1.19M) includes induced demand on hospitals, labs, EHPs, and DALRRD enforcement.",
    locations = cells_column_labels(columns = nmb_vs_sq)
  )
```

### Spend–Burden Curve

```{r}
#| label: fig-spend-burden
#| fig-cap: "Cost of the true burden falls as surveillance, coordination, and enforcement spend rises"
#| fig-width: 8
#| fig-height: 5

ggplot(options_econ, aes(x = surveillance_cost, y = residual_burden_cost / 1e6)) +
  geom_line(color = "grey60", linetype = "dashed") +
  geom_point(aes(color = Option), size = 4) +
  geom_text(aes(label = Option), vjust = -1.2, size = 3) +
  scale_x_continuous(labels = scales::label_dollar(prefix = "R", scale = 1e-3, suffix = "k")) +
  scale_y_continuous(labels = scales::label_dollar(prefix = "R", suffix = "M")) +
  labs(
    title = "Surveillance Spend vs. Residual Cost of True Burden",
    subtitle = "Each rand spent on surveillance, coordination, and enforcement reduces the cost of true poisoning burden",
    x = "Annual surveillance / coordination / enforcement spend (K_i)",
    y = "Residual cost of true burden (C_i)",
    caption = "Base case: η = 0.30, CFR = 12.5%, true burden = 11,690 cases/year"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )
```

### Sensitivity: CFR and Response Efficacy

The economic case is most sensitive to **CFR** (which scales mortality cost) and **response efficacy** $\eta$ (which scales avoided cost per detected hotspot).

```{r}
#| label: tbl-econ-sensitivity
#| tbl-cap: "Sensitivity of Full Package net monetary benefit to CFR and response efficacy"

cfr_grid <- c(0.10, 0.125, 0.15)
eta_grid <- c(0.15, 0.30, 0.50)
K_full <- 1193000   # whole-system cost
d_full <- 0.83
d_sq   <- 0.45

sens <- expand_grid(CFR = cfr_grid, eta_val = eta_grid) %>%
  mutate(
    burden_total = (B_true * (1 - CFR) * c_m) + (B_true * CFR * c_d),
    avoided      = burden_total * (d_full - d_sq) * eta_val,
    nmb          = avoided - K_full
  ) %>%
  mutate(
    `Total burden` = paste0("R", format(round(burden_total / 1e9, 2), big.mark = ","), "B"),
    `Avoided cost (Full Pkg)` = paste0("R", format(round(avoided / 1e6, 1), big.mark = ","), "M"),
    `NMB (Full Pkg)` = paste0("R", format(round(nmb / 1e6, 1), big.mark = ","), "M")
  ) %>%
  select(CFR, eta_val, `Total burden`, `Avoided cost (Full Pkg)`, `NMB (Full Pkg)`)

sens %>%
  gt() %>%
  tab_header(
    title = "Sensitivity: CFR × Response Efficacy",
    subtitle = "Full Package net monetary benefit (whole-system cost = R1.19M)"
  ) %>%
  cols_label(eta_val = "η (response efficacy)") %>%
  fmt_percent(columns = CFR, decimals = 1) %>%
  cols_align(align = "right", columns = everything())
```

::: {.callout-tip}
## Bottom Line

Across the full plausible CFR × $\eta$ grid, the Full Package generates a **net monetary benefit measured in hundreds of millions of rand**, dwarfing its R0.24M (narrow) or R1.19M (whole-system) annual cost. Even at the **least favourable corner** (CFR = 10%, η = 0.15), avoided burden cost exceeds programme cost by orders of magnitude.

**Policy implication:** The economic case for surveillance investment does not hinge on optimistic assumptions. It holds because the true burden is large, and even modest improvements in detection translate into substantial avoided costs.
:::

### Limitations of the Economic Model

::: {.callout-warning}
## Caveats

1. **Mutually-exclusive assumption** likely overstates true burden if NMC/NHLS/PIH overlap; understates it if community cases never reach any system. Sensitivity to a 30% deduplication: total burden falls ~30% but Full Package NMB remains strongly positive.
2. **Per-case morbidity cost (R12,000)** is a placeholder. A South-African micro-costing study of pesticide-poisoning admissions is needed.
3. **Per-death cost (R1.5M)** uses the human-capital approach, which is conservative relative to value-of-statistical-life (VSL) estimates (typically R5–15M for SA). Switching to VSL would magnify NMB by ~3–10×.
4. **Response efficacy η = 0.30** is the largest single uncertainty. Efficacy depends on enforcement capacity, retailer compliance, and community uptake — all measurable in pilot.
5. **Linear scaling** assumes each detected hotspot averts the same fraction of future cases. In reality, marginal hotspots may be smaller or harder to act on (diminishing returns).
6. **No discounting** applied (short time horizon).
:::

## Model Validation

### Face Validity

Model reviewed by:

- NICD surveillance epidemiologists
- NHLS laboratory managers
- PIH clinical toxicologists
- Academic health economists (UCT School of Public Health)

**Feedback incorporated:**

- Adjusted detection probabilities based on expert elicitation
- Added response success rate as probabilistic node
- Included scenario for "Option 5 requires Option 3 or 4"

### Internal Validity

- **Parameter consistency:** All probabilities sum to 1.0
- **Cost additivity:** Full Package cost = sum of individual option costs
- **Effectiveness synergy:** Full Package > sum of individual effects (captures coordination benefits)

### External Validity (Pending)

Model will be validated against **pilot data** (Q3 2026 – Q1 2027):

1. Compare predicted vs. observed detection rates
2. Update cost parameters with actual implementation expenses
3. Refine effectiveness estimates based on measured time-to-detection
4. Re-run PSA with empirical parameter distributions

## Model Limitations

::: {.callout-warning}
## Key Limitations

1. **Placeholder parameters:** All values subject to revision based on pilot data
2. **Linear effectiveness assumption:** Model assumes options combine additively (may underestimate synergies)
3. **No discounting:** Costs and effects not discounted over time (justified for short time horizon)
4. **No health outcomes:** Model tracks hotspots detected, not deaths/disabilities averted (future extension planned)
5. **No equity analysis:** Model doesn't differentiate by age, sex, or socioeconomic status
:::

## Software and Reproducibility

### Tools Used

- **Amua v1.0.3:** Decision tree modeling and Monte Carlo simulation
- **R v4.3.1:** Data wrangling, visualization, and reporting
- **Quarto v1.3:** Reproducible document generation

### Model Files

All model files and code are available in the project repository:

- `amua_import_parameters_v3.csv` — Parameter definitions for Amua import (see [Parameter Appraisal](../parameter_appraisal.qmd) for the v2 → v3 critique)
- `wrangling_v2.r` — Data processing script
- `methods_v2.r` — Custom functions for analysis
- `analysis_report_v2.qmd` — Full analysis report with embedded code

### Reproducing Results

To reproduce this analysis:

1. Download Amua from [https://github.com/zward/Amua](https://github.com/zward/Amua)
2. Import `amua_import_parameters_v3.csv`
3. Run decision tree model (10,000 PSA iterations)
4. Export results to CSV
5. Run `wrangling_v2.r` to process outputs
6. Render `analysis_report_v2.qmd`

## Next Steps

### Model Updates Planned

Once pilot data become available:

1. **Update effectiveness parameters** with empirical detection rates
2. **Refine cost estimates** based on actual implementation expenses
3. **Extend model** to include health outcomes (deaths averted, DALYs gained)
4. **Add equity dimensions** (stratify by age, sex, province)
5. **Multi-year projection** with discounting

### Decision Analysis Extensions

Future analyses could include:

- **Value of information analysis:** How much is it worth to reduce parameter uncertainty?
- **Budget impact analysis:** What if we scale up to all 9 provinces?
- **Threshold analysis:** At what cost would options no longer be cost-effective?

---

::: {.footer-cta}
**Questions about the model?** Contact the D2P analytical team at [contact@example.com](mailto:contact@example.com)

**Download model files:** [amua_import_parameters_v3.csv](../amua_import_parameters_v3.csv) — see [Parameter Critical Appraisal](../parameter_appraisal.qmd) for the methodology and v2 critique.
:::
