---
title: "Overall Policy Model"
subtitle: "Integrated steady-state model of surveillance, coordination, and terbufos phase-out"
author: "D2P Project Team"
date: "2026-05-13"
date-modified: today
categories: [policy, integrated-model, decision-makers, auditable]
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: true
    code-tools: true
---

## Scope

A single steady-state model of acute pesticide poisoning in South Africa in which the three companion briefs ([Surveillance](surveillance_policy.qmd), [Coordination](coordination_policy.qmd), [Terbufos](terbufos_policy.qmd)) are encoded as toggleable options. Every parameter is either a named row in [`amua_import_parameters_v4.csv`](../amua_import_parameters_v4.csv) or a declared constant at the top of the model chunk.

The model is designed to demonstrate two structural results:

1. **Detection–workload coupling.** Induced EHP investigation and DALRRD enforcement workload scales with the false-positive-inflated alert volume, producing diminishing marginal returns once community-tier surveillance is added.
2. **Source-control coupling.** A 12–24 month terbufos phase-out reduces true cases and deaths proportionally, lowering health/YLL costs and induced surveillance workload by the same fraction.

## Burden

StatsSA MACOD 2022 records 1,014 pesticide-specific poisoning deaths/yr (X48+X68+X87+Y18; interim anchor pending 2023 release). Applied to a mid case-fatality of 12.5 %, this implies ~8,112 true cases/yr. NMC currently notifies ~12.5 % of true cases. Terbufos (WHO Class Ia organophosphate) is the dominant agent in fatal paediatric events (Davies 2023) with an attributable fraction of 20–35 %.

## Model

```{r}
#| label: overall-model
#| echo: true
#| message: false
#| warning: false

library(tidyverse)
library(gt)
library(scales)

# ----------------------------------------------------------------------
# 1. PULL CANONICAL PARAMETERS FROM THE CSV
# ----------------------------------------------------------------------
raw <- readr::read_csv("../amua_import_parameters_v4.csv", show_col_types = FALSE)
pick <- function(name) {
  v <- raw |> filter(Name == name) |> pull(Expression)
  if (length(v) == 0) NA_real_ else suppressWarnings(as.numeric(v[1]))
}

# Burden anchors
N_true    <- pick("n_true_burden_from_statssa_mid")  # 8,112 true cases/yr
N_deaths  <- pick("n_deaths_headline")               # 1,014 deaths/yr
c_morb    <- pick("c_morbidity_headline")            # R/morbidity case
c_mort    <- pick("c_mortality_human_capital")       # R/death (YLL human-capital)

# Detection probabilities along the surveillance ladder
d_SQ      <- pick("d_status_quo")           # NMC only
d_MVP     <- pick("d_surveillance_mvp")     # MVP: NHLS feed + BChE auto + PIH MVD
d_S1b_mid <- pick("d_S1b_event_community_mid")  # adds community/EHP layer

# Surveillance operating costs (annual, regulator-paid)
c_surv_MVP <- pick("c_surv_MVP_op")          # R137k
c_S1b_op   <- pick("c_S1b_op_mid")           # R1.5M
p_fp_S1b   <- pick("p_fp_S1b_mid")           # 35 % false-positive alerts

# Coordination Full Package operating
c_coord_FP <- pick("c_full_package_narrow_operating") - c_surv_MVP
# (≈ R100k/yr incremental for C1–C5 closure-of-loop machinery)

# Coordination outcome — deaths averted at scale
n_deaths_av_coord_FP <- pick("n_deaths_averted_coord_FP")  # 957

# Terbufos phase-out
f_terb_mid <- pick("frac_terbufos_attributable_mid")  # 0.25
phi_terb   <- pick("frac_T3_burden_retired")          # 0.70
c_T2_enf   <- pick("c_T2_RUP_enforcement_mid")        # R2M

# ----------------------------------------------------------------------
# 2. ASSUMPTIONS DECLARED INLINE (NOT in the CSV — edit here)
# ----------------------------------------------------------------------
# These are the unit costs that translate detected alerts into induced
# downstream workload. They are kept here, named, so a reviewer can
# change them and re-render.

c_per_investigation  <- 4000     # R/case: EHP environmental investigation + travel + write-up
c_per_dalrrd_action  <- 12000    # R/case: DALRRD inspectorate site visit + sampling + enforcement letter
p_dalrrd_action      <- 0.30     # P(DALRRD action | confirmed case worth referring)

# Terbufos phase-out ramp: linear from 0 to phi_terb over RAMP_MONTHS
RAMP_MONTHS          <- 18       # 12–24 month policy window; user-editable

# ----------------------------------------------------------------------
# 3. THE MODEL — one function, four scenarios
# ----------------------------------------------------------------------
# A scenario is fully specified by:
#   d            : per-case detection probability
#   c_surv       : annual surveillance operating cost
#   c_coord      : annual coordination operating cost
#   p_fp         : false-positive rate inflating alert volume
#   terb_phase   : 0..1, share of terbufos-attributable burden retired this year
#   c_regulatory : DALRRD enforcement spend (terbufos restriction)

run_scenario <- function(d, c_surv, c_coord, p_fp,
                         terb_phase = 0, c_regulatory = 0) {

  # (A) Effective true burden after terbufos retirement this year
  N_true_eff   <- N_true   * (1 - f_terb_mid * terb_phase)
  N_deaths_eff <- N_deaths * (1 - f_terb_mid * terb_phase)

  # (B) Detected (true) cases and gross alerts (incl. false positives)
  detected_true  <- d * N_true_eff
  gross_alerts   <- detected_true / (1 - p_fp)   # false-positive inflation
  false_alerts   <- gross_alerts - detected_true

  # (C) Induced workload costs (HCW/EHP/DALRRD) — scale with GROSS alerts
  c_investigation <- gross_alerts * c_per_investigation
  c_dalrrd_load   <- gross_alerts * p_dalrrd_action * c_per_dalrrd_action

  # (D) Health & YLL cost on the *true* burden (only true cases incur it)
  c_health <- N_true_eff   * c_morb
  c_yll    <- N_deaths_eff * c_mort

  # (E) Regulator-paid total this year
  c_regulator <- c_surv + c_coord + c_investigation + c_dalrrd_load + c_regulatory

  # (F) Societal total this year
  c_society   <- c_regulator + c_health + c_yll

  # (G) Deaths averted relative to status quo (no terbufos retirement, NMC-only)
  deaths_sq   <- N_deaths
  deaths_now  <- N_deaths_eff * (1 - (d - d_SQ))  # crude: detection→response→aversion
  # Use a single response efficacy: eta = 0.30 (mid) applied to incremental detected cases
  eta <- pick("eta_response_mid")
  deaths_av_surv <- eta * (d - d_SQ) * N_true_eff * (N_deaths_eff / N_true_eff)
  deaths_av_terb <- N_deaths * f_terb_mid * terb_phase
  deaths_av      <- deaths_av_surv + deaths_av_terb

  tibble(
    detected_true   = detected_true,
    gross_alerts    = gross_alerts,
    false_alerts    = false_alerts,
    c_surv          = c_surv,
    c_coord         = c_coord,
    c_investigation = c_investigation,
    c_dalrrd_load   = c_dalrrd_load,
    c_regulatory    = c_regulatory,
    c_health        = c_health,
    c_yll           = c_yll,
    c_regulator_tot = c_regulator,
    c_society_tot   = c_society,
    deaths_averted  = deaths_av,
    cost_per_death_averted =
      if_else(deaths_av > 0, c_regulator / deaths_av, NA_real_)
  )
}

# ----------------------------------------------------------------------
# 4. FOUR LAYERED SCENARIOS
# ----------------------------------------------------------------------
phase_now <- min(1, 12 / RAMP_MONTHS)   # year-1 ramp fraction (12 months elapsed)

scenarios <- bind_rows(
  `0 · Status quo (NMC only)`                = run_scenario(d_SQ,      0,        0,         p_fp = 0,        terb_phase = 0,         c_regulatory = 0),
  `1 · Surveillance MVP + Coordination FP`   = run_scenario(d_MVP,     c_surv_MVP, c_coord_FP, p_fp = 0,        terb_phase = 0,         c_regulatory = 0),
  `2 · + Community surveillance (S1b)`       = run_scenario(d_S1b_mid, c_surv_MVP + c_S1b_op, c_coord_FP, p_fp = p_fp_S1b, terb_phase = 0,         c_regulatory = 0),
  `3 · + Terbufos phase-out (yr 1, ramp ${RAMP_MONTHS}m)` =
                                              run_scenario(d_S1b_mid, c_surv_MVP + c_S1b_op, c_coord_FP, p_fp = p_fp_S1b, terb_phase = phase_now, c_regulatory = c_T2_enf),
  .id = "Scenario"
)

# Rename the dynamic label
scenarios$Scenario[4] <- sprintf("3 · + Terbufos phase-out (yr 1, ramp %dm)", RAMP_MONTHS)
```

## Scenario comparison

```{r}
#| label: tbl-overall
#| tbl-cap: "Annual steady-state outputs under each layered policy option."

scenarios |>
  mutate(
    detected_true = round(detected_true),
    gross_alerts  = round(gross_alerts),
    false_alerts  = round(false_alerts),
    `Regulator (R/yr)`   = scales::comma(round(c_regulator_tot)),
    `Health + YLL (R/yr)`= scales::comma(round(c_health + c_yll)),
    `Society (R/yr)`     = scales::comma(round(c_society_tot)),
    `Deaths averted`     = round(deaths_averted),
    `R / death averted`  = if_else(deaths_averted > 0,
                                   scales::comma(round(c_regulator_tot / deaths_averted)),
                                   "—")
  ) |>
  select(Scenario,
         `Detected (true)` = detected_true,
         `Gross alerts`    = gross_alerts,
         `False alerts`    = false_alerts,
         `Regulator (R/yr)`,
         `Health + YLL (R/yr)`,
         `Society (R/yr)`,
         `Deaths averted`,
         `R / death averted`) |>
  gt() |>
  tab_header(title = "Layered policy options — annual steady-state view") |>
  opt_row_striping() |>
  tab_options(table.font.size = "small")
```

Rows 1→2 raise detection from 5 % to 19 % at R237k/yr incremental regulator spend, with no false-positive inflation (administrative feeds only). Row 3 adds community-tier surveillance, which lifts detection further but introduces a 35 % false-positive rate; induced workload scales with gross (true + false) alert volume. Row 4 layers a `r RAMP_MONTHS`-month terbufos phase-out, which reduces true cases and deaths proportionally to the attributable fraction × ramp progress.

## Detection–workload coupling

```{r}
#| label: fig-diminishing-returns
#| fig-cap: "Marginal cost per death averted rises sharply with community-surveillance false-positive rate. The terbufos phase-out shifts the curve down because the system is now processing fewer true cases."

fp_grid <- seq(0.10, 0.60, by = 0.025)

curve_df <- bind_rows(
  tibble(
    fp = fp_grid,
    label = "Surveillance only (no terbufos action)",
    res = map(fp, ~ run_scenario(d_S1b_mid, c_surv_MVP + c_S1b_op, c_coord_FP,
                                 p_fp = .x, terb_phase = 0, c_regulatory = 0))
  ),
  tibble(
    fp = fp_grid,
    label = sprintf("Surveillance + terbufos phase-out (yr 1, %dm ramp)", RAMP_MONTHS),
    res = map(fp, ~ run_scenario(d_S1b_mid, c_surv_MVP + c_S1b_op, c_coord_FP,
                                 p_fp = .x, terb_phase = phase_now, c_regulatory = c_T2_enf))
  )
) |>
  mutate(cost_per_death = map_dbl(res, ~ .x$cost_per_death_averted)) |>
  select(-res)

ggplot(curve_df, aes(x = fp, y = cost_per_death, colour = label)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.5) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::comma_format(prefix = "R")) +
  labs(
    x = "Community-surveillance false-positive rate (p_fp_S1b)",
    y = "Regulator cost per death averted (R/yr)",
    colour = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## Terbufos phase-out trajectory

```{r}
#| label: fig-terbufos-ramp
#| fig-cap: "As terbufos is phased out (linear ramp), deaths and true-case burden fall, dragging *both* the health/YLL bill *and* the induced surveillance workload down with them."

months <- 0:24
ramp_frac <- pmin(1, months / RAMP_MONTHS)

ramp_df <- map_dfr(seq_along(months), function(i) {
  r <- run_scenario(d_S1b_mid, c_surv_MVP + c_S1b_op, c_coord_FP,
                    p_fp = p_fp_S1b,
                    terb_phase = ramp_frac[i],
                    c_regulatory = c_T2_enf)
  tibble(
    month                   = months[i],
    `True cases (residual)` = N_true   * (1 - f_terb_mid * ramp_frac[i]),
    `Deaths (residual)`     = N_deaths * (1 - f_terb_mid * ramp_frac[i]),
    `Health + YLL bill`     = r$c_health + r$c_yll,
    `Surveillance + induced workload` =
      r$c_surv + r$c_coord + r$c_investigation + r$c_dalrrd_load
  )
})

ramp_long <- ramp_df |>
  pivot_longer(-month, names_to = "metric", values_to = "value") |>
  group_by(metric) |>
  mutate(value_idx = value / first(value) * 100) |>
  ungroup()

ggplot(ramp_long, aes(x = month, y = value_idx, colour = metric)) +
  geom_line(linewidth = 1.1) +
  geom_vline(xintercept = RAMP_MONTHS, linetype = "dashed", alpha = 0.4) +
  annotate("text", x = RAMP_MONTHS, y = 102, label = "Ramp complete",
           hjust = 1.05, size = 3.2, alpha = 0.6) +
  labs(
    x = "Months after policy commencement",
    y = "Indexed to month 0 (= 100)",
    colour = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## Sensitivity analyses

Three two-way sensitivity tables are reported. Each varies two model inputs over their CSV-anchored ranges and holds all others at working values. Cells are colour-shaded from low (green) to high (red) on the reported outcome.

### S1. False-positive rate × detection probability

Outcome: regulator cost per death averted (R/yr), under the full surveillance + coordination stack with no terbufos action.

```{r}
#| label: tbl-sens-fp-detection
#| tbl-cap: "Regulator cost per death averted (R/yr) as community false-positive rate and per-case detection are varied jointly. All other parameters held at working values; terbufos phase-out off."

fp_levels    <- c(0.20, 0.35, 0.50)
d_levels     <- c(d_MVP, 0.30, d_S1b_mid, 0.50)
d_labels     <- c(sprintf("MVP (%.0f%%)", d_MVP*100),
                  "30%",
                  sprintf("S1b mid (%.0f%%)", d_S1b_mid*100),
                  "50%")

sens1 <- expand_grid(fp = fp_levels, d = d_levels) |>
  mutate(
    res = map2(d, fp, ~ run_scenario(.x, c_surv_MVP + c_S1b_op, c_coord_FP,
                                     p_fp = .y, terb_phase = 0, c_regulatory = 0)),
    cost_per_death = map_dbl(res, ~ .x$cost_per_death_averted),
    d_lab = factor(d, levels = d_levels, labels = d_labels)
  ) |>
  select(fp, d_lab, cost_per_death) |>
  pivot_wider(names_from = d_lab, values_from = cost_per_death) |>
  mutate(`FP rate` = scales::percent(fp, accuracy = 1), .before = 1) |>
  select(-fp)

sens1 |>
  gt(rowname_col = "FP rate") |>
  fmt_number(columns = -1, decimals = 0, use_seps = TRUE, pattern = "R{x}") |>
  data_color(columns = -1, palette = c("#1a9850", "#fee08b", "#d73027")) |>
  tab_stubhead(label = "FP rate \\ Detection") |>
  tab_options(table.font.size = "small")
```

Cost per death averted rises with both the false-positive rate (more induced workload per detected true case) and, at high FP, with detection itself (the additional true cases averted do not offset the alert-processing cost).

### S2. Terbufos attributable fraction × ramp length

Outcome: deaths averted in year 1 attributable to the phase-out, holding surveillance + coordination at the S1b mid scenario.

```{r}
#| label: tbl-sens-terbufos
#| tbl-cap: "Year-1 deaths averted from terbufos phase-out, varying attributable fraction (rows) and ramp length (columns). At month 12 the ramp is complete for the 12-month option and 50–67 % complete for the 18- and 24-month options respectively."

f_levels    <- c(pick("frac_terbufos_attributable_lo"),
                 pick("frac_terbufos_attributable_mid"),
                 pick("frac_terbufos_attributable_hi"))
ramp_levels <- c(12, 18, 24)

sens2 <- expand_grid(f_attr = f_levels, ramp = ramp_levels) |>
  mutate(
    phase_yr1 = pmin(1, 12 / ramp),
    deaths_av = f_attr * phi_terb * N_deaths * phase_yr1
  ) |>
  select(f_attr, ramp, deaths_av) |>
  pivot_wider(names_from = ramp, values_from = deaths_av,
              names_glue = "{ramp}-month ramp") |>
  mutate(`Attributable fraction` = scales::percent(f_attr, accuracy = 1),
         .before = 1) |>
  select(-f_attr)

sens2 |>
  gt(rowname_col = "Attributable fraction") |>
  fmt_number(columns = -1, decimals = 0) |>
  data_color(columns = -1, palette = c("#d73027", "#fee08b", "#1a9850")) |>
  tab_stubhead(label = "f_attr \\ Ramp") |>
  tab_options(table.font.size = "small")
```

The 12-month ramp captures the full attributable retirement in year 1; the 24-month ramp halves it. At the working attributable fraction (25 %), this is a difference of ~`r round(0.25 * phi_terb * N_deaths * (1 - 0.5))` deaths in year 1.

### S3. Response efficacy × terbufos phase-out completion

Outcome: total annual deaths averted (surveillance response + source control), under the S1b stack at mid FP.

```{r}
#| label: tbl-sens-eta-terbufos
#| tbl-cap: "Total annual deaths averted (surveillance + terbufos retirement) as response efficacy η and phase-out completion are varied jointly. Negative complementarity is absent: the two levers add."

eta_levels  <- c(pick("eta_response_lo"),
                 pick("eta_response_mid"),
                 pick("eta_response_hi"))
phase_levels <- c(0, 0.5, 1)

sens3 <- expand_grid(eta = eta_levels, phase = phase_levels) |>
  mutate(
    deaths_av_surv = eta * (d_S1b_mid - d_SQ) * N_true *
                     (1 - f_terb_mid * phase) *
                     (N_deaths / N_true),
    deaths_av_terb = N_deaths * f_terb_mid * phase * phi_terb,
    deaths_av_tot  = deaths_av_surv + deaths_av_terb
  ) |>
  select(eta, phase, deaths_av_tot) |>
  pivot_wider(names_from = phase, values_from = deaths_av_tot,
              names_glue = "{scales::percent(phase, accuracy = 1)} retired") |>
  mutate(`Response efficacy η` = scales::percent(eta, accuracy = 1),
         .before = 1) |>
  select(-eta)

sens3 |>
  gt(rowname_col = "Response efficacy η") |>
  fmt_number(columns = -1, decimals = 0) |>
  data_color(columns = -1, palette = c("#d73027", "#fee08b", "#1a9850")) |>
  tab_stubhead(label = "η \\ Terbufos retired") |>
  tab_options(table.font.size = "small")
```

Surveillance contribution scales with η and is partially eroded as terbufos retirement reduces the residual true-case pool. Source-control contribution scales linearly with retirement. The two levers are additive at all parameter values in the supported ranges.

## Results

**Detection–workload coupling.** Administrative-tier surveillance (NHLS feed, BChE auto-notify, PIH MVD) and coordination (C1–C5) raise per-case detection from 5 % to 19 % without false-positive inflation, because alerts originate from confirmed laboratory or clinical events. Community-tier surveillance (S1b) raises detection further but at a benchmarked false-positive rate of 20–50 %; gross alert volume is 1.25–2 × the true-case count and marginal regulator cost per death averted rises monotonically with `p_fp_S1b`. The model implies S1b should be piloted with direct measurement of `p_fp_S1b` before national rollout.

**Source-control coupling.** A `r RAMP_MONTHS`-month linear terbufos phase-out retires `r round(N_deaths * f_terb_mid * phi_terb)` deaths/yr and `r round(N_true * f_terb_mid * phi_terb)` cases/yr at full ramp (`frac_terbufos_attributable_mid = 0.25`, `frac_T3_burden_retired = 0.70`). Because induced workload and health/YLL costs both scale with true cases, all four output streams (cases, deaths, health/YLL, induced workload) fall by the same fraction. Source control is complementary to surveillance: it is the only lever that reduces the denominator the system has to process.

## Editing the model

Inputs are either CSV rows in [`amua_import_parameters_v4.csv`](../amua_import_parameters_v4.csv) (registry: [Parameters Hub](../parameters_hub.html)) or declared constants at the top of the model chunk (`c_per_investigation`, `c_per_dalrrd_action`, `p_dalrrd_action`, `RAMP_MONTHS`). Sensitivity ranges (`_lo`, `_hi`) are provided in the CSV for every working value. Typical edits:

- `RAMP_MONTHS <- 12` or `24` — tighter or slower phase-out.
- Replace `p_fp_S1b_mid` with `p_fp_S1b_hi` in the S1b scenario — stress-test community FP.
- Drop `c_mort` from `c_society_tot` — health-system perspective only.

## Relation to companion briefs

```{mermaid}
%%| label: fig-overall-fit
%%| fig-cap: "Each companion brief specifies one component; this model evaluates them jointly."
flowchart LR
    A[1 · Surveillance<br/>MVP + S1b ladder] --> M[Integrated model<br/>this page]
    B[2 · Coordination<br/>NMC→EHP→DALRRD closure] --> M
    C[3 · Terbufos<br/>T2 → T3 escalation] --> M
    style M fill:#fff3b0,stroke:#cc8800,stroke-width:2px
```

---

Companion briefs: [Surveillance](surveillance_policy.qmd) · [Coordination](coordination_policy.qmd) · [Terbufos](terbufos_policy.qmd) · [Decision Model](decision_model.qmd) · [Costing](costing_analysis.qmd).
