---
title: "Overall Policy Model v3"
subtitle: "Decision-maker cost evaluation: five layered scenarios with full CBA and CUA accounting, transparent CFR sensitivity"
author: "D2P Project Team"
date: "2026-05-21"
date-modified: today
categories: [policy, integrated-model, decision-makers, CBA, CUA, auditable]
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: true
---

## Purpose

This v3 model is built for **policy makers who need to evaluate the cost of various surveillance, coordination, and chemical-ban scenarios** in a way that is transparent and defensible.

The **golden thread** is preserved end-to-end for every scenario:

> **deaths averted → health (morbidity) costs averted → YLL costs averted → full CBA and CUA**

CBA uses the **human-capital** valuation (R1.5M per averted death; R12k per averted morbidity case). CUA uses **GBD 2019 disability weights** valued at the SA **ICER threshold of R110,000 per DALY** (1× GDP per capita 2025). Both run on the same underlying epi quantities so the only difference between CBA and CUA outputs is how a life-year is monetised — making the policy-relevance of each valuation framework directly comparable.

CFR is the **largest single source of denominator uncertainty** in the SA pesticide burden estimate (StatsSA MACOD records deaths; the true non-fatal case count depends on the assumed CFR). v3 propagates this uncertainty **continuously** across all scenarios and presents the result as line graphs rather than tables of point estimates.

### Five scenarios

| # | Label | Detection | Coordination | Ban |
|---|---|---|---|---|
| **A** | Ceiling administrative surveillance, perfect coordination | MVP only (19%) | Perfect | None |
| **B** | + Community surveillance, perfect coordination (FP sweep) | MVP + S1b lo→hi | Perfect | None |
| **C** | Community surveillance + imperfect coordination | MVP + S1b mid | Imperfect | None |
| **D** | All layers + **perfect ban** (terb attributable sweep) | MVP + S1b mid | Perfect | $\phi_{terb}=1.0$, $f_{terb}$ lo→hi |
| **E** | All layers + **imperfect ban** (90% drop) | MVP + S1b mid | Perfect | $\phi_{terb}=0.90$ |

**Working definitions (user-confirmed):**

- **Perfect coordination**: $\eta = 0.50$ (response efficacy upper), $k_{cluster} = 3.5$ (high), full deaths-averted from coordination decision-tree (n = 957/yr).
- **Imperfect coordination**: $\eta = 0.15$, $k_{cluster} = 1.2$, 50% of coordination deaths-averted (n ≈ 479/yr).
- **Imperfect ban**: $\phi_{terb} = 0.90$ (90% of terbufos-attributable burden retired; 10% residual from informal stockpile / cross-border leakage).

---

## Setup

```{r}
#| label: setup
#| echo: true
#| message: false
#| warning: false

library(tidyverse)
library(gt)
library(scales)
library(patchwork)

raw <- readr::read_csv("../amua_import_parameters_v4.csv", show_col_types = FALSE)

pick <- function(name) {
  v <- raw |> filter(Name == name) |> pull(Expression)
  if (length(v) == 0) NA_real_ else suppressWarnings(as.numeric(v[1]))
}

# ── Burden anchors ────────────────────────────────────────────────────────────
N_deaths  <- pick("n_deaths_headline")               # 2,620 deaths/yr (StatsSA MACOD)
c_morb    <- pick("c_morbidity_headline")            # R12,000 per morbidity case (HC)
c_mort_HC <- pick("c_mortality_human_capital")       # R1,500,000 per death (human-capital)

# ── CFR sensitivity range (continuous) ────────────────────────────────────────
cfr_lo    <- pick("cfr_headline_lo")                 # 0.10
cfr_mid   <- pick("cfr_headline_mid")                # 0.125
cfr_hi    <- pick("cfr_headline_hi")                 # 0.15
cfr_grid  <- seq(0.075, 0.20, by = 0.005)            # continuous grid for line plots

# true cases helper — single source of truth for the denominator
N_true_of <- function(cfr) N_deaths / cfr

N_true_mid <- N_true_of(cfr_mid)                     # 20,960 cases/yr at mid CFR

# ── Detection probabilities ───────────────────────────────────────────────────
d_SQ      <- pick("d_status_quo")                    # 0.048 (NMC-only baseline)
d_MVP     <- pick("d_surveillance_mvp")              # 0.189 (admin ceiling = S1 + S2)
d_S1b_lo  <- pick("d_S1b_event_community_lo")        # 0.25
d_S1b_mid <- pick("d_S1b_event_community_mid")       # 0.40
d_S1b_hi  <- pick("d_S1b_event_community_hi")        # 0.55

# ── Costs: surveillance ──────────────────────────────────────────────────────
c_MVP_op  <- pick("c_surv_MVP_op")                   # R137k = S1 + S2
c_S1b_lo  <- pick("c_S1b_op_lo")                     # R800k
c_S1b_mid <- pick("c_S1b_op_mid")                    # R1.5M
c_S1b_hi  <- pick("c_S1b_op_hi")                     # R2.5M

# ── Costs: coordination ──────────────────────────────────────────────────────
c_coord_FP   <- pick("c_coord_FP_op")                # R145k C1+...+C5
n_av_coord_FP <- pick("n_deaths_averted_coord_FP")   # 957 deaths/yr full package

# Coordination "quality" presets
coord_presets <- list(
  perfect   = list(eta = pick("eta_response_hi"),  k_cluster = 3.5, n_av_coord = n_av_coord_FP),
  imperfect = list(eta = pick("eta_response_lo"),  k_cluster = 1.2, n_av_coord = 0.50 * n_av_coord_FP)
)

# ── False-positive rates (community S1b only) ────────────────────────────────
p_fp_S1b_lo  <- pick("p_fp_S1b_lo")                  # 0.20
p_fp_S1b_mid <- pick("p_fp_S1b_mid")                 # 0.35
p_fp_S1b_hi  <- pick("p_fp_S1b_hi")                  # 0.50

# ── Chemical-agent params ────────────────────────────────────────────────────
f_terb_lo  <- pick("frac_terbufos_attributable_lo")  # 0.20
f_terb_mid <- pick("frac_terbufos_attributable_mid") # 0.25
f_terb_hi  <- pick("frac_terbufos_attributable_hi")  # 0.35
phi_terb_perfect   <- 1.00                           # perfect ban: 100% retired
phi_terb_imperfect <- 0.90                           # imperfect ban: 90% retired (10% residual)
c_T2_enf   <- pick("c_T2_RUP_enforcement_mid")       # R2M DALRRD enforcement

# ── DALY / CUA (GBD 2019) ────────────────────────────────────────────────────
yll_per_death       <- pick("yll_per_death_gbd")     # 44 YLL
yld_per_case_severe <- pick("yld_per_case_severe")   # 0.033
yld_per_case_mild   <- pick("yld_per_case_mild")     # 0.003
p_severe            <- pick("p_severe_among_detected") # 0.45
icer_per_daly       <- pick("icer_per_daly_gdp")     # R110,000

# Derived ICER-DALY monetary equivalents
c_mort_DALY <- yll_per_death * icer_per_daly         # R4.84M per averted death
c_morb_DALY <- (p_severe * yld_per_case_severe + (1 - p_severe) * yld_per_case_mild) * icer_per_daly  # blended R/morbidity

# ── Induced workload constants (inline, identical to v2) ─────────────────────
c_per_investigation  <- 4000     # R/alert
c_per_dalrrd_action  <- 12000    # R/alert
p_dalrrd_action      <- 0.30

# ── Status-quo societal cost (anchor for NMB) ────────────────────────────────
c_society_SQ_HC   <- function(cfr) N_true_of(cfr) * c_morb + N_deaths * c_mort_HC
c_society_SQ_DALY <- function(cfr) N_true_of(cfr) * c_morb_DALY + N_deaths * c_mort_DALY
```

### Quick reference: monetary equivalents

```{r}
#| label: ref-money
#| echo: false
tibble(
  Quantity = c("Per averted death (HC, CBA)",
               "Per averted death (DALY × ICER, CUA)",
               "Per averted morbidity case (HC, CBA)",
               "Per averted morbidity case (DALY × ICER, CUA, blended severity)",
               "ICER threshold (R per DALY averted)",
               "YLL per death (GBD 2019, mean age 28, LE 72)"),
  Value = c(c_mort_HC, c_mort_DALY, c_morb, c_morb_DALY, icer_per_daly, yll_per_death)
) |>
  gt() |>
  fmt_number(columns = Value, decimals = 0, sep_mark = ",") |>
  tab_header(title = "CBA vs CUA: monetary equivalents per averted health outcome")
```

---

## Golden-thread function

`run_scenario_v3()` returns a single tibble row with deaths averted, morbidity averted, both HC and DALY-ICER valuations, NMB under both frameworks, and the **ICER (R per DALY averted)** which is the headline CUA decision-rule statistic.

```{r}
#| label: golden-thread-fn
#| echo: true

run_scenario_v3 <- function(
    d,                       # per-case detection probability
    cfr,                     # CFR assumption (drives N_true)
    c_surv,                  # surveillance operating spend (R/yr)
    c_coord     = 0,         # coordination operating spend (R/yr)
    coord_mode  = NULL,      # "perfect" | "imperfect" | NULL (= no coordination layer)
    p_fp        = 0,         # FP rate for community surveillance
    f_terb      = 0,         # terbufos attributable fraction (0 = no ban applied)
    phi_terb    = 0,         # fraction of terbufos burden retired (ban quality)
    c_regulatory = 0         # ban-side regulatory spend (R/yr)
) {
  N_true <- N_true_of(cfr)

  # Coordination preset
  if (!is.null(coord_mode)) {
    cp <- coord_presets[[coord_mode]]
    eta_use      <- cp$eta
    k_use        <- cp$k_cluster
    n_av_coord_0 <- cp$n_av_coord
  } else {
    eta_use      <- pick("eta_response_mid")
    k_use        <- 1.0
    n_av_coord_0 <- 0
  }

  # ── 1. Ban removes burden directly (independent of detection) ─────────────
  frac_burden_retired_by_ban <- f_terb * phi_terb           # ≤ 1
  N_true_post_ban   <- N_true   * (1 - frac_burden_retired_by_ban)
  N_deaths_post_ban <- N_deaths * (1 - frac_burden_retired_by_ban)

  # ── 2. Surveillance/coordination act on post-ban residual burden ──────────
  d_uplift   <- max(d - d_SQ, 0)                            # detection improvement
  # Cases addressed = post-ban × uplift × cluster multiplier
  cases_addressed     <- N_true_post_ban * d_uplift * k_use
  cases_addressed     <- pmin(cases_addressed, N_true_post_ban)  # cap
  deaths_av_response  <- cases_addressed * cfr * eta_use

  # Coordination deaths averted: scale by detection uplift relative to MVP
  d_uplift_ratio <- if (d_MVP > d_SQ) d_uplift / (d_MVP - d_SQ) else 0
  deaths_av_coord <- n_av_coord_0 * min(d_uplift_ratio, 1)

  # ── 3. Aggregate (cap totals at the post-ban anchor) ──────────────────────
  deaths_av_total   <- frac_burden_retired_by_ban * N_deaths +
                        deaths_av_response + deaths_av_coord
  deaths_av_total   <- pmin(deaths_av_total, N_deaths)

  cases_av_total    <- frac_burden_retired_by_ban * N_true + cases_addressed
  cases_av_total    <- pmin(cases_av_total, N_true)

  # Severity split for DALY morbidity
  cases_av_severe   <- cases_av_total * p_severe
  cases_av_mild     <- cases_av_total * (1 - p_severe)

  # ── 4. Cost averted (golden thread) ───────────────────────────────────────
  c_morbidity_av_HC   <- cases_av_total * c_morb
  c_yll_av_HC         <- deaths_av_total * c_mort_HC
  c_morbidity_av_DALY <- cases_av_severe * yld_per_case_severe * icer_per_daly +
                         cases_av_mild   * yld_per_case_mild   * icer_per_daly
  c_yll_av_DALY       <- deaths_av_total * c_mort_DALY

  dalys_av <- deaths_av_total * yll_per_death +
              cases_av_severe * yld_per_case_severe +
              cases_av_mild   * yld_per_case_mild

  # ── 5. Spend side ─────────────────────────────────────────────────────────
  # Alert volume: gross alerts inflated by FP rate
  alerts_gross         <- N_true_post_ban * d_uplift * (1 + p_fp / (1 - p_fp + 1e-9))
  c_invest             <- alerts_gross * c_per_investigation
  c_dalrrd             <- alerts_gross * p_dalrrd_action * c_per_dalrrd_action
  c_total_spend        <- c_surv + c_coord + c_invest + c_dalrrd + c_regulatory

  # ── 6. NMB and ICER ───────────────────────────────────────────────────────
  benefit_HC   <- c_morbidity_av_HC   + c_yll_av_HC
  benefit_DALY <- c_morbidity_av_DALY + c_yll_av_DALY
  NMB_HC       <- benefit_HC   - c_total_spend
  NMB_DALY     <- benefit_DALY - c_total_spend
  ICER         <- if (dalys_av > 0) c_total_spend / dalys_av else NA_real_
  cost_per_death_av <- if (deaths_av_total > 0) c_total_spend / deaths_av_total else NA_real_

  tibble(
    cfr, d, p_fp, f_terb, phi_terb,
    coord_mode = coord_mode %||% "none",
    N_true = N_true,
    deaths_av_total, cases_av_total,
    c_morbidity_av_HC, c_yll_av_HC, benefit_HC,
    c_morbidity_av_DALY, c_yll_av_DALY, benefit_DALY,
    dalys_av,
    c_surv, c_coord, c_invest, c_dalrrd, c_regulatory, c_total_spend,
    NMB_HC, NMB_DALY, ICER, cost_per_death_av
  )
}
```

---

## Scenario A — Ceiling administrative surveillance + perfect coordination, no ban

**Policy logic.** Get to the *administrative ceiling* (MVP detection = 18.9%) using only the existing NHLS/PIH feeds, then close the coordination loop **perfectly** so every detected case becomes a hotspot investigation that surfaces additional cluster members. **No chemical ban**. This is the cheapest credible policy bundle and answers: *"how much can we save without touching DALRRD's product register?"*

```{r}
#| label: A-run
A_mid <- run_scenario_v3(d = d_MVP, cfr = cfr_mid,
                         c_surv = c_MVP_op, c_coord = c_coord_FP,
                         coord_mode = "perfect")

A_cfr <- map_dfr(cfr_grid, ~ run_scenario_v3(
  d = d_MVP, cfr = .x,
  c_surv = c_MVP_op, c_coord = c_coord_FP,
  coord_mode = "perfect"))
```

### Scenario A — CBA (human-capital)

```{r}
#| label: A-cba-table
A_mid |>
  transmute(
    `Total spend (R/yr)`             = c_total_spend,
    `Deaths averted (n/yr)`          = deaths_av_total,
    `Morbidity cases averted (n/yr)` = cases_av_total,
    `Morbidity costs averted (R/yr)` = c_morbidity_av_HC,
    `YLL costs averted (R/yr)`       = c_yll_av_HC,
    `Total benefit (R/yr)`           = benefit_HC,
    `NMB (R/yr)`                     = NMB_HC,
    `R per death averted`            = cost_per_death_av
  ) |>
  pivot_longer(everything(), names_to = "Item", values_to = "Value") |>
  gt() |>
  fmt_number(columns = Value, decimals = 0, sep_mark = ",") |>
  tab_header(title = "Scenario A — CBA at CFR mid (12.5%)")
```

### Scenario A — CUA (DALY × ICER)

```{r}
#| label: A-cua-table
A_mid |>
  transmute(
    `Total spend (R/yr)`              = c_total_spend,
    `DALYs averted (n/yr)`            = dalys_av,
    `Morbidity DALY-value averted (R)` = c_morbidity_av_DALY,
    `Mortality DALY-value averted (R)` = c_yll_av_DALY,
    `Total DALY-valued benefit (R/yr)` = benefit_DALY,
    `NMB DALY (R/yr)`                  = NMB_DALY,
    `ICER (R per DALY averted)`        = ICER,
    `ICER threshold (R110k/DALY)`      = icer_per_daly
  ) |>
  pivot_longer(everything(), names_to = "Item", values_to = "Value") |>
  gt() |>
  fmt_number(columns = Value, decimals = 0, sep_mark = ",") |>
  tab_header(title = "Scenario A — CUA at CFR mid (12.5%)")
```

### Scenario A — CFR sensitivity (continuous)

```{r}
#| label: A-cfr-plot
#| fig-width: 9
#| fig-height: 4
p_nmb <- ggplot(A_cfr, aes(cfr)) +
  geom_line(aes(y = NMB_HC,   colour = "CBA (human-capital)"),   linewidth = 0.9) +
  geom_line(aes(y = NMB_DALY, colour = "CUA (DALY × ICER)"),     linewidth = 0.9) +
  geom_vline(xintercept = c(cfr_lo, cfr_mid, cfr_hi), linetype = "dotted", colour = "grey50") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = comma) +
  labs(x = "Case-fatality ratio (CFR)", y = "NMB (R/yr)",
       title = "Scenario A — NMB across CFR", colour = NULL) +
  theme_minimal()

p_icer <- ggplot(A_cfr, aes(cfr, ICER)) +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  geom_hline(yintercept = icer_per_daly, linetype = "dashed", colour = "red") +
  geom_vline(xintercept = c(cfr_lo, cfr_mid, cfr_hi), linetype = "dotted", colour = "grey50") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = comma) +
  labs(x = "Case-fatality ratio (CFR)", y = "ICER (R per DALY averted)",
       title = "Scenario A — ICER vs R110,000/DALY threshold") +
  theme_minimal()

p_nmb + p_icer
```

---

## Scenario B — + Community surveillance (FP sweep) + perfect coordination, no ban

**Policy logic.** Layer event-based community surveillance (S1b) on top of the administrative MVP. S1b detects three times more cases but injects false-positive alerts that drive induced EHP workload. This scenario sweeps the **FP rate** to show the tradeoff between detection gain and induced cost.

```{r}
#| label: B-run
B_grid <- expand_grid(
  d   = c(d_S1b_lo, d_S1b_mid, d_S1b_hi),
  fp  = c(p_fp_S1b_lo, p_fp_S1b_mid, p_fp_S1b_hi),
  cs  = c(c_S1b_lo, c_S1b_mid, c_S1b_hi)
) |>
  filter((d == d_S1b_lo  & fp == p_fp_S1b_lo  & cs == c_S1b_lo) |
         (d == d_S1b_mid & fp == p_fp_S1b_mid & cs == c_S1b_mid) |
         (d == d_S1b_hi  & fp == p_fp_S1b_hi  & cs == c_S1b_hi)) |>
  mutate(level = c("lo", "mid", "hi"))

B_res <- B_grid |>
  rowwise() |>
  mutate(out = list(run_scenario_v3(
    d = d, cfr = cfr_mid,
    c_surv = c_MVP_op + cs, c_coord = c_coord_FP,
    coord_mode = "perfect", p_fp = fp))) |>
  unnest(out, names_sep = "_") |>
  select(level, d, fp, cs, out_deaths_av_total, out_dalys_av,
         out_c_total_spend, out_benefit_HC, out_benefit_DALY,
         out_NMB_HC, out_NMB_DALY, out_ICER, out_cost_per_death_av)

# continuous sweep of FP at mid d, mid surveillance cost
B_fp_sweep <- map_dfr(seq(0.10, 0.60, by = 0.025),
  ~ run_scenario_v3(d = d_S1b_mid, cfr = cfr_mid,
                    c_surv = c_MVP_op + c_S1b_mid, c_coord = c_coord_FP,
                    coord_mode = "perfect", p_fp = .x))
```

### Scenario B — CBA

```{r}
#| label: B-cba-table
B_res |>
  transmute(
    Level = level,
    `Detection (%)`                  = d * 100,
    `FP rate (%)`                    = fp * 100,
    `S1b spend (R/yr)`               = cs,
    `Deaths averted`                 = out_deaths_av_total,
    `Total spend (R/yr)`             = out_c_total_spend,
    `Total benefit HC (R/yr)`        = out_benefit_HC,
    `NMB HC (R/yr)`                  = out_NMB_HC,
    `R per death averted`            = out_cost_per_death_av
  ) |>
  gt() |>
  fmt_number(columns = c(`Detection (%)`, `FP rate (%)`), decimals = 1) |>
  fmt_number(columns = c(`S1b spend (R/yr)`, `Total spend (R/yr)`,
                         `Total benefit HC (R/yr)`, `NMB HC (R/yr)`,
                         `R per death averted`),
             decimals = 0, sep_mark = ",") |>
  fmt_number(columns = `Deaths averted`, decimals = 0) |>
  tab_header(title = "Scenario B — CBA: matched lo/mid/hi S1b tiers at CFR mid")
```

### Scenario B — CUA

```{r}
#| label: B-cua-table
B_res |>
  transmute(
    Level = level,
    `DALYs averted`               = out_dalys_av,
    `Total spend (R/yr)`          = out_c_total_spend,
    `DALY-valued benefit (R/yr)`  = out_benefit_DALY,
    `NMB DALY (R/yr)`             = out_NMB_DALY,
    `ICER (R/DALY averted)`       = out_ICER
  ) |>
  gt() |>
  fmt_number(columns = c(`DALYs averted`), decimals = 0) |>
  fmt_number(columns = c(`Total spend (R/yr)`, `DALY-valued benefit (R/yr)`,
                         `NMB DALY (R/yr)`, `ICER (R/DALY averted)`),
             decimals = 0, sep_mark = ",") |>
  tab_header(title = "Scenario B — CUA: matched lo/mid/hi S1b tiers at CFR mid")
```

### Scenario B — FP rate sweep at mid configuration

```{r}
#| label: B-fp-plot
#| fig-width: 9
#| fig-height: 4
p_b1 <- ggplot(B_fp_sweep, aes(p_fp, NMB_HC)) +
  geom_line(linewidth = 0.9, colour = "darkgreen") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = comma) +
  labs(x = "S1b false-positive rate", y = "NMB HC (R/yr)",
       title = "Scenario B — NMB declines with FP rate") +
  theme_minimal()

p_b2 <- ggplot(B_fp_sweep, aes(p_fp, ICER)) +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  geom_hline(yintercept = icer_per_daly, linetype = "dashed", colour = "red") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = comma) +
  labs(x = "S1b false-positive rate", y = "ICER (R/DALY averted)",
       title = "ICER stays well below threshold across FP range") +
  theme_minimal()

p_b1 + p_b2
```

---

## Scenario C — Community surveillance + imperfect coordination

**Policy logic.** Same surveillance stack as B-mid, but coordination operates at the **lower-bound** efficacy and clustering ($\eta = 0.15$, $k = 1.2$, half the coordination decision-tree deaths-averted). This is the realistic *"DALRRD didn't fully sign on"* case and shows how much benefit erodes when the loop only partially closes.

```{r}
#| label: C-run
C_mid <- run_scenario_v3(d = d_S1b_mid, cfr = cfr_mid,
                         c_surv = c_MVP_op + c_S1b_mid, c_coord = c_coord_FP,
                         coord_mode = "imperfect", p_fp = p_fp_S1b_mid)

C_cfr <- map_dfr(cfr_grid, ~ run_scenario_v3(
  d = d_S1b_mid, cfr = .x,
  c_surv = c_MVP_op + c_S1b_mid, c_coord = c_coord_FP,
  coord_mode = "imperfect", p_fp = p_fp_S1b_mid))
```

### Scenario C — CBA & CUA side-by-side

```{r}
#| label: C-tables
bind_rows(
  C_mid |> transmute(framework = "CBA (HC)",
                     `Deaths averted` = deaths_av_total,
                     `Total spend (R/yr)` = c_total_spend,
                     `Benefit (R/yr)` = benefit_HC,
                     `NMB (R/yr)` = NMB_HC,
                     `Headline rate` = cost_per_death_av,
                     headline_label = "R per death averted"),
  C_mid |> transmute(framework = "CUA (DALY)",
                     `Deaths averted` = deaths_av_total,
                     `Total spend (R/yr)` = c_total_spend,
                     `Benefit (R/yr)` = benefit_DALY,
                     `NMB (R/yr)` = NMB_DALY,
                     `Headline rate` = ICER,
                     headline_label = "R per DALY averted")
) |>
  gt() |>
  fmt_number(columns = c(`Deaths averted`), decimals = 0) |>
  fmt_number(columns = c(`Total spend (R/yr)`, `Benefit (R/yr)`,
                         `NMB (R/yr)`, `Headline rate`),
             decimals = 0, sep_mark = ",") |>
  tab_header(title = "Scenario C — same epi, two valuations")
```

### Scenario C — CFR sensitivity

```{r}
#| label: C-cfr-plot
#| fig-width: 9
#| fig-height: 4
p_c1 <- ggplot(C_cfr, aes(cfr)) +
  geom_line(aes(y = NMB_HC,   colour = "CBA (HC)"),   linewidth = 0.9) +
  geom_line(aes(y = NMB_DALY, colour = "CUA (DALY)"), linewidth = 0.9) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = comma) +
  labs(x = "CFR", y = "NMB (R/yr)", colour = NULL,
       title = "Scenario C — NMB across CFR") +
  theme_minimal()

p_c2 <- ggplot(C_cfr, aes(cfr, ICER)) +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  geom_hline(yintercept = icer_per_daly, linetype = "dashed", colour = "red") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = comma) +
  labs(x = "CFR", y = "ICER (R/DALY)", title = "Scenario C — ICER") +
  theme_minimal()

p_c1 + p_c2
```

---

## Scenario D — All layers + perfect ban (terbufos attributable sweep)

**Policy logic.** Full stack — admin MVP + community S1b (mid) + perfect coordination + **perfect terbufos ban** ($\phi = 1.0$, zero residual). The sweep is over the **terbufos attributable fraction** $f_{terb} \in \{0.20, 0.25, 0.35\}$ because the policy benefit of a perfect ban scales directly with how much of the burden the agent actually drives — a parameter that informs how confident DALRRD needs to be in its attribution evidence before committing to a full withdrawal.

```{r}
#| label: D-run
D_grid <- tibble(
  level  = c("lo", "mid", "hi"),
  f_terb = c(f_terb_lo, f_terb_mid, f_terb_hi)
) |>
  rowwise() |>
  mutate(out = list(run_scenario_v3(
    d = d_S1b_mid, cfr = cfr_mid,
    c_surv = c_MVP_op + c_S1b_mid, c_coord = c_coord_FP,
    coord_mode = "perfect", p_fp = p_fp_S1b_mid,
    f_terb = f_terb, phi_terb = phi_terb_perfect,
    c_regulatory = c_T2_enf))) |>
  unnest(out, names_sep = "_")

# continuous sweep across f_terb at perfect phi
D_fterb_sweep <- map_dfr(seq(0.10, 0.40, by = 0.01),
  ~ run_scenario_v3(d = d_S1b_mid, cfr = cfr_mid,
                    c_surv = c_MVP_op + c_S1b_mid, c_coord = c_coord_FP,
                    coord_mode = "perfect", p_fp = p_fp_S1b_mid,
                    f_terb = .x, phi_terb = phi_terb_perfect,
                    c_regulatory = c_T2_enf))
```

### Scenario D — CBA

```{r}
#| label: D-cba-table
D_grid |>
  transmute(
    Level = level,
    `f_terb (%)`              = f_terb * 100,
    `Deaths averted`          = out_deaths_av_total,
    `Cases averted`           = out_cases_av_total,
    `Spend (R/yr)`            = out_c_total_spend,
    `Benefit HC (R/yr)`       = out_benefit_HC,
    `NMB HC (R/yr)`           = out_NMB_HC,
    `R per death averted`     = out_cost_per_death_av
  ) |>
  gt() |>
  fmt_number(columns = `f_terb (%)`, decimals = 1) |>
  fmt_number(columns = c(`Deaths averted`, `Cases averted`), decimals = 0) |>
  fmt_number(columns = c(`Spend (R/yr)`, `Benefit HC (R/yr)`,
                         `NMB HC (R/yr)`, `R per death averted`),
             decimals = 0, sep_mark = ",") |>
  tab_header(title = "Scenario D — CBA: perfect ban, f_terb lo/mid/hi at CFR mid")
```

### Scenario D — CUA

```{r}
#| label: D-cua-table
D_grid |>
  transmute(
    Level = level,
    `DALYs averted`               = out_dalys_av,
    `Spend (R/yr)`                = out_c_total_spend,
    `DALY benefit (R/yr)`         = out_benefit_DALY,
    `NMB DALY (R/yr)`             = out_NMB_DALY,
    `ICER (R/DALY averted)`       = out_ICER
  ) |>
  gt() |>
  fmt_number(columns = `DALYs averted`, decimals = 0) |>
  fmt_number(columns = c(`Spend (R/yr)`, `DALY benefit (R/yr)`,
                         `NMB DALY (R/yr)`, `ICER (R/DALY averted)`),
             decimals = 0, sep_mark = ",") |>
  tab_header(title = "Scenario D — CUA: perfect ban, f_terb lo/mid/hi at CFR mid")
```

### Scenario D — f_terb sweep

```{r}
#| label: D-fterb-plot
#| fig-width: 9
#| fig-height: 4
p_d1 <- ggplot(D_fterb_sweep, aes(f_terb)) +
  geom_line(aes(y = NMB_HC,   colour = "CBA (HC)"),   linewidth = 0.9) +
  geom_line(aes(y = NMB_DALY, colour = "CUA (DALY)"), linewidth = 0.9) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = comma) +
  labs(x = "Terbufos attributable fraction (f_terb)",
       y = "NMB (R/yr)", colour = NULL,
       title = "Scenario D — NMB scales linearly with terbufos attribution") +
  theme_minimal()

p_d2 <- ggplot(D_fterb_sweep, aes(f_terb, ICER)) +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  geom_hline(yintercept = icer_per_daly, linetype = "dashed", colour = "red") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = comma) +
  labs(x = "Terbufos attributable fraction (f_terb)",
       y = "ICER (R/DALY)",
       title = "Higher f_terb → lower ICER (more burden retired per regulatory rand)") +
  theme_minimal()

p_d1 + p_d2
```

---

## Scenario E — All layers + imperfect ban (90% drop)

**Policy logic.** Identical configuration to D-mid, but the terbufos ban only retires **90%** of the attributable burden (10% residual from informal stockpile and cross-border leakage). This is the realistic case for SA given the porous border with Mozambique and Zimbabwe and the documented informal market in Class Ia organophosphates.

```{r}
#| label: E-run
E_grid <- tibble(
  level  = c("lo", "mid", "hi"),
  f_terb = c(f_terb_lo, f_terb_mid, f_terb_hi)
) |>
  rowwise() |>
  mutate(out = list(run_scenario_v3(
    d = d_S1b_mid, cfr = cfr_mid,
    c_surv = c_MVP_op + c_S1b_mid, c_coord = c_coord_FP,
    coord_mode = "perfect", p_fp = p_fp_S1b_mid,
    f_terb = f_terb, phi_terb = phi_terb_imperfect,
    c_regulatory = c_T2_enf))) |>
  unnest(out, names_sep = "_")
```

### Scenario E — CBA & CUA

```{r}
#| label: E-tables
E_grid |>
  transmute(
    Level = level,
    `f_terb (%)`              = f_terb * 100,
    `Deaths averted`          = out_deaths_av_total,
    `DALYs averted`           = out_dalys_av,
    `Spend (R/yr)`            = out_c_total_spend,
    `Benefit HC (R/yr)`       = out_benefit_HC,
    `Benefit DALY (R/yr)`     = out_benefit_DALY,
    `NMB HC (R/yr)`           = out_NMB_HC,
    `NMB DALY (R/yr)`         = out_NMB_DALY,
    `ICER (R/DALY averted)`   = out_ICER
  ) |>
  gt() |>
  fmt_number(columns = `f_terb (%)`, decimals = 1) |>
  fmt_number(columns = c(`Deaths averted`, `DALYs averted`), decimals = 0) |>
  fmt_number(columns = c(`Spend (R/yr)`, `Benefit HC (R/yr)`, `Benefit DALY (R/yr)`,
                         `NMB HC (R/yr)`, `NMB DALY (R/yr)`, `ICER (R/DALY averted)`),
             decimals = 0, sep_mark = ",") |>
  tab_header(title = "Scenario E — Imperfect ban (φ = 0.90), f_terb sweep at CFR mid")
```

### Scenario D vs E contrast — lost benefit from 10% residual

```{r}
#| label: DE-contrast
D_E <- bind_rows(
  D_grid |> transmute(level, scenario = "D (perfect ban, φ=1.0)",
                      benefit_HC = out_benefit_HC, NMB_HC = out_NMB_HC,
                      benefit_DALY = out_benefit_DALY, NMB_DALY = out_NMB_DALY,
                      ICER = out_ICER, deaths_av = out_deaths_av_total),
  E_grid |> transmute(level, scenario = "E (imperfect ban, φ=0.90)",
                      benefit_HC = out_benefit_HC, NMB_HC = out_NMB_HC,
                      benefit_DALY = out_benefit_DALY, NMB_DALY = out_NMB_DALY,
                      ICER = out_ICER, deaths_av = out_deaths_av_total)
)

D_E |>
  pivot_wider(id_cols = level, names_from = scenario,
              values_from = c(deaths_av, NMB_HC, NMB_DALY, ICER)) |>
  gt() |>
  fmt_number(columns = where(is.numeric), decimals = 0, sep_mark = ",") |>
  tab_header(title = "D vs E — the cost of an imperfect ban")
```

---

## Cross-scenario CBA summary

```{r}
#| label: cba-summary
summary_rows <- bind_rows(
  A_mid |> mutate(scenario = "A — MVP + perfect coord"),
  B_res |> filter(level == "mid") |>
    transmute(scenario = "B — + S1b mid + perfect coord",
              c_total_spend = out_c_total_spend,
              deaths_av_total = out_deaths_av_total,
              dalys_av = out_dalys_av,
              benefit_HC = out_benefit_HC, benefit_DALY = out_benefit_DALY,
              NMB_HC = out_NMB_HC, NMB_DALY = out_NMB_DALY,
              ICER = out_ICER, cost_per_death_av = out_cost_per_death_av),
  C_mid |> mutate(scenario = "C — + S1b mid + imperfect coord"),
  D_grid |> filter(level == "mid") |>
    transmute(scenario = "D — + perfect ban (f_terb mid)",
              c_total_spend = out_c_total_spend,
              deaths_av_total = out_deaths_av_total,
              dalys_av = out_dalys_av,
              benefit_HC = out_benefit_HC, benefit_DALY = out_benefit_DALY,
              NMB_HC = out_NMB_HC, NMB_DALY = out_NMB_DALY,
              ICER = out_ICER, cost_per_death_av = out_cost_per_death_av),
  E_grid |> filter(level == "mid") |>
    transmute(scenario = "E — + imperfect ban (φ=0.90, f_terb mid)",
              c_total_spend = out_c_total_spend,
              deaths_av_total = out_deaths_av_total,
              dalys_av = out_dalys_av,
              benefit_HC = out_benefit_HC, benefit_DALY = out_benefit_DALY,
              NMB_HC = out_NMB_HC, NMB_DALY = out_NMB_DALY,
              ICER = out_ICER, cost_per_death_av = out_cost_per_death_av)
) |>
  select(scenario, c_total_spend, deaths_av_total, dalys_av,
         benefit_HC, NMB_HC, cost_per_death_av,
         benefit_DALY, NMB_DALY, ICER)

summary_rows |>
  transmute(Scenario = scenario,
            `Spend (R/yr)` = c_total_spend,
            `Deaths averted` = deaths_av_total,
            `Benefit HC (R/yr)` = benefit_HC,
            `NMB HC (R/yr)` = NMB_HC,
            `R per death averted` = cost_per_death_av) |>
  gt() |>
  fmt_number(columns = `Deaths averted`, decimals = 0) |>
  fmt_number(columns = c(`Spend (R/yr)`, `Benefit HC (R/yr)`,
                         `NMB HC (R/yr)`, `R per death averted`),
             decimals = 0, sep_mark = ",") |>
  tab_header(title = "Cross-scenario CBA summary (CFR mid)") |>
  data_color(columns = `NMB HC (R/yr)`, palette = "Greens")
```

---

## Cross-scenario CUA summary

```{r}
#| label: cua-summary
summary_rows |>
  transmute(Scenario = scenario,
            `Spend (R/yr)` = c_total_spend,
            `DALYs averted` = dalys_av,
            `DALY benefit (R/yr)` = benefit_DALY,
            `NMB DALY (R/yr)` = NMB_DALY,
            `ICER (R/DALY averted)` = ICER) |>
  gt() |>
  fmt_number(columns = `DALYs averted`, decimals = 0) |>
  fmt_number(columns = c(`Spend (R/yr)`, `DALY benefit (R/yr)`,
                         `NMB DALY (R/yr)`, `ICER (R/DALY averted)`),
             decimals = 0, sep_mark = ",") |>
  tab_header(title = "Cross-scenario CUA summary (CFR mid)",
             subtitle = paste0("Threshold = R", comma(icer_per_daly), "/DALY")) |>
  data_color(columns = `ICER (R/DALY averted)`,
             palette = c("darkgreen", "yellow", "red"))
```

All five scenarios deliver ICERs at least one order of magnitude below the R110,000/DALY SA threshold — a defensible "very cost-effective" classification under the WHO-CHOICE convention regardless of which valuation framework is used.

---

## Continuous CFR sensitivity across all scenarios

```{r}
#| label: all-cfr-sweep
#| fig-width: 10
#| fig-height: 5

run_at_cfr <- function(cfr_val) {
  bind_rows(
    run_scenario_v3(d = d_MVP,     cfr = cfr_val, c_surv = c_MVP_op,
                    c_coord = c_coord_FP, coord_mode = "perfect") |>
      mutate(scenario = "A"),
    run_scenario_v3(d = d_S1b_mid, cfr = cfr_val, c_surv = c_MVP_op + c_S1b_mid,
                    c_coord = c_coord_FP, coord_mode = "perfect",
                    p_fp = p_fp_S1b_mid) |>
      mutate(scenario = "B-mid"),
    run_scenario_v3(d = d_S1b_mid, cfr = cfr_val, c_surv = c_MVP_op + c_S1b_mid,
                    c_coord = c_coord_FP, coord_mode = "imperfect",
                    p_fp = p_fp_S1b_mid) |>
      mutate(scenario = "C"),
    run_scenario_v3(d = d_S1b_mid, cfr = cfr_val, c_surv = c_MVP_op + c_S1b_mid,
                    c_coord = c_coord_FP, coord_mode = "perfect",
                    p_fp = p_fp_S1b_mid, f_terb = f_terb_mid,
                    phi_terb = phi_terb_perfect, c_regulatory = c_T2_enf) |>
      mutate(scenario = "D-mid"),
    run_scenario_v3(d = d_S1b_mid, cfr = cfr_val, c_surv = c_MVP_op + c_S1b_mid,
                    c_coord = c_coord_FP, coord_mode = "perfect",
                    p_fp = p_fp_S1b_mid, f_terb = f_terb_mid,
                    phi_terb = phi_terb_imperfect, c_regulatory = c_T2_enf) |>
      mutate(scenario = "E-mid")
  )
}

all_cfr <- map_dfr(cfr_grid, run_at_cfr)

p_all_nmb <- ggplot(all_cfr, aes(cfr, NMB_HC, colour = scenario)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = cfr_mid, linetype = "dotted", colour = "grey50") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = comma) +
  labs(x = "CFR", y = "NMB HC (R/yr)",
       title = "NMB HC across CFR — all scenarios",
       subtitle = "Vertical dotted line = CFR mid (12.5%)") +
  theme_minimal()

p_all_icer <- ggplot(all_cfr, aes(cfr, ICER, colour = scenario)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = icer_per_daly, linetype = "dashed", colour = "red") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = comma) +
  labs(x = "CFR", y = "ICER (R/DALY)",
       title = "ICER across CFR",
       subtitle = "Red dashed line = R110,000/DALY threshold") +
  theme_minimal()

p_all_nmb + p_all_icer
```

**Reading the chart.** Scenario rankings are **stable across the full CFR range** — D dominates, E follows, then B/C, then A. Absolute NMB values scale almost linearly with CFR because lower CFR means a larger non-fatal case denominator and therefore more morbidity benefit per percentage-point of detection uplift. The ranking stability is the key defensibility result: the policy ordering does *not* depend on which CFR assumption a reviewer prefers.

---

## Spend-vs-NMB frontier

```{r}
#| label: frontier
#| fig-width: 8
#| fig-height: 5
frontier <- summary_rows |>
  select(scenario, spend = c_total_spend, NMB_HC, NMB_DALY) |>
  arrange(spend) |>
  mutate(label = str_extract(scenario, "^[A-E]"))

ggplot(frontier, aes(spend)) +
  geom_line(aes(y = NMB_HC,   colour = "CBA (HC)"),   linewidth = 1) +
  geom_line(aes(y = NMB_DALY, colour = "CUA (DALY)"), linewidth = 1) +
  geom_point(aes(y = NMB_HC),   colour = "darkgreen", size = 2.5) +
  geom_point(aes(y = NMB_DALY), colour = "steelblue", size = 2.5) +
  ggrepel::geom_text_repel(aes(y = NMB_HC, label = label), nudge_y = 5e8, size = 4) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(x = "Annual programme spend (R/yr)", y = "NMB (R/yr)",
       colour = NULL,
       title = "Spend → NMB frontier (CFR mid)",
       subtitle = "Both valuations strictly increasing — every scenario passes the dominance test") +
  theme_minimal()
```

---

## Auditable assumption ledger

| Assumption | Source | Where to edit |
|---|---|---|
| Deaths anchor `N_deaths` = 2,620 | StatsSA MACOD 2023 X40–X49 | `amua_import_parameters_v4.csv` → `n_deaths_headline` |
| CFR range 10–15% | StatsSA MACOD + stratified cohort | `cfr_headline_lo/mid/hi` |
| Perfect coordination: η = 0.50, k = 3.5, full 957 | User-confirmed for v3 | inline `coord_presets$perfect` |
| Imperfect coordination: η = 0.15, k = 1.2, 50% × 957 | User-confirmed for v3 | inline `coord_presets$imperfect` |
| Perfect ban: φ = 1.00 | Working definition for D | inline `phi_terb_perfect` |
| Imperfect ban: φ = 0.90 (10% residual) | User-confirmed informal-market residual | inline `phi_terb_imperfect` |
| YLL per death = 44 | GBD 2019; mean age 28, LE 72 (**PROVISIONAL** — replace with MACOD age-sex distribution) | `yll_per_death_gbd` |
| ICER threshold = R110,000/DALY | SA GDP per capita 2025 (1× WHO-CHOICE convention) | `icer_per_daly_gdp` |
| Investigation cost = R4,000/alert | NICD field operations | inline `c_per_investigation` |
| DALRRD action cost = R12,000/action | DALRRD inspectorate | inline `c_per_dalrrd_action` |
| P(DALRRD action \| alert) = 0.30 | DALRRD enforcement audit | inline `p_dalrrd_action` |

**To stress-test any assumption**: change the value in the location listed above, re-render. All downstream tables and plots regenerate from the same `run_scenario_v3()` golden-thread function — the model is fully end-to-end reactive.
