---
title: "Policy Brief 1: Strengthening Pesticide Poisoning Surveillance"
subtitle: "Closing the detection gap between NMC, NHLS, and PIH"
author: "D2P Project Team"
date: "2025-09-05"
date-modified: today
categories: [policy, surveillance, decision-makers]
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: false
---

::: {.callout-important icon=false}
## For Decision-Makers

**Target audience:** Ministerial Advisory Committee (MAC), National Department of Health (NDoH), Provincial Health Leadership, NICD/NHLS leadership.

**Companion briefs:** [Coordination](coordination_policy.qmd) · [Terbufos restriction](terbufos_policy.qmd).
:::

## 1. Policy question

**Should NDoH invest in automated, multi-source pesticide-poisoning surveillance to close the detection gap between the NMC, NHLS laboratory, and PIH data streams?**

## 2. Public health problem

South Africa under-detects pesticide poisoning. Three data streams each see a fraction of the problem; none sees the whole:

- Most cases never appear in NMC.
- Severe laboratory results do not trigger a public-health alert.
- The country's largest single source of clinical signal — the Poisons Information Helpline — is not used for surveillance at all.

## 3. Status quo

{{< include _shared_parameters.qmd >}}

::: {.callout-note}
## Health-sector share of the burden — what surveillance unlocks

The public health system bears **~`r sprintf("R%.0fM", as.numeric(pick("share_health_system_total"))/1e6)` of the `r sprintf("R%.2fB", as.numeric(pick("C_burden_total_headline"))/1e9)` annual burden directly on its own budget lines** — ~`r sprintf("R%.0fM", as.numeric(pick("share_doh_morbidity"))/1e6)` to NDoH (UPFS ward + ICU days, ventilation, atropine + PAM) and ~`r sprintf("R%.0fM", as.numeric(pick("share_nhls_morbidity"))/1e6)` to NHLS (BChE, GC-MS confirmatory volume). Better surveillance is a **necessary but not sufficient** input to reducing that burden: surveillance changes only what is *seen*; the burden reduction itself comes from the response side — source removal, regulatory action, clinical management — covered in the [Coordination brief](coordination_policy.qmd) and the [Terbufos brief](terbufos_policy.qmd). **This brief therefore quantifies P_detect uplift only and does not claim cases or deaths averted.**
:::

**Architecturally, the three streams are unlinked:**

```{mermaid}
%%| label: fig-surv-current-system
%%| fig-cap: "Current surveillance architecture. NMC is NDoH-owned and is the source of truth. NHLS BChE results now route to NMC via NHLS CDW → NICD SDW (MOU feed, now operational). Two active gaps remain: PIH consultations have no structured NMC feed, and forensic chemical toxicology is post-mortem only and fully de-identified (not fit for surveillance linkage in its current form; MLDI system work is a potential pathway)."
flowchart LR
    A[Healthcare Facility<br/>Clinician / PHO] -->|NMC notification| E[("NDoH / NMC<br/><b>NDoH-owned</b><br/>Source of truth")]
    A --> NHLS

    subgraph NHLS[NHLS]
        direction TB
        C1[BChE Assay<br/>1,779 severe/year<br/>10,626 tests 2023<br/><i>Live patients</i>]
        C2[Chemical Toxicology<br/><i>Post-mortem only —<br/>not yet done on<br/>live patients</i>]
        CDW[(NHLS CDW)]
        C1 -->|monthly extract| CDW
    end

    SDW[(NICD SDW)]
    CDW -->|"✓ MOU feed<br/>operational"| SDW
    SDW -->|BChE monthly| E

    A --> D[PIH Consultation<br/>~1,158 clinician calls/yr<br/><i>No minimum dataset</i>]

    C2 -.x Not available for<br/>live-patient surveillance.-x E
    D -.No minimum viable<br/>dataset defined.-> E

    style E fill:#9f9,stroke:#060,stroke-width:2px
    style A fill:#fde,stroke:#333
    style C1 fill:#bbf,stroke:#333
    style C2 fill:#ffd9b3,stroke:#cc8800,stroke-dasharray: 5 5
    style D fill:#bfb,stroke:#333
    style CDW fill:#e6f0ff,stroke:#3366cc
    style SDW fill:#e6f0ff,stroke:#3366cc
```

**What this means in plain numbers.**

- NMC notifications: `r scales::comma(as.numeric(pick("n_nmc_annual")))` ([`n_nmc_annual`](../parameters_hub.qmd#tbl-burden))
- NHLS severe BChE inhibitions: `r scales::comma(as.numeric(pick("n_nhls_severe_annual")))` ([`n_nhls_severe_annual`](../parameters_hub.qmd#tbl-burden)) out of `r scales::comma(as.numeric(pick("n_nhls_total_annual")))` BChE tests ([`n_nhls_total_annual`](../parameters_hub.qmd#tbl-burden))
- PIH clinician-initiated calls: `r scales::comma(as.numeric(pick("n_pih_annual")))` ([`n_pih_annual`](../parameters_hub.qmd#tbl-burden))
- Implied true annual burden: $N_\text{true} = \frac{2,620}{0.125} = 20,960$ ([`n_true_burden_from_statssa_mid`](../parameters_hub.qmd#tbl-burden))

NMC alone captures $P_\text{detect} = \frac{1,013}{20,960} = 0.048$ ([`d_status_quo`](../parameters_hub.qmd#tbl-detection)), i.e. **< 5 %**.

::: {.callout-caution appearance="minimal" collapse="true"}
## Caveat: this denominator rests on a CFR assumption
The 20,960 denominator applies a 12.5% CFR to the StatsSA MACOD death count — but that CFR is from NMC-*notified* severe cases (the sicker end). The undetected majority are likely milder; a blended CFR of 3–5% implies 52,000–87,000 true cases and a status-quo P_detect of 1–2%. Relative option rankings (S1, S2, MVP) are stable across this range; absolute gap size is not. Mid-burden is used as the working denominator throughout.
:::

### Detection vs. true burden — a critical distinction

Better surveillance has **two opposite numerical effects** that decision-makers must not confuse:

| Effect | Direction | Driver | Interpretation |
|:-------|:---------:|:-------|:---------------|
| **Observed** burden | ↑ | Better case-finding (auto-notify, MVD ingest) | The headline "case count" will rise as detection improves — this is success, not failure. |
| **True** burden | ↓ | Faster source removal, targeted regulation, prevention | Number of actual poisonings falls as cases are detected and acted on earlier. |

Post-pilot case counts will rise — this is a detection effect, not a true burden increase. Interpret against completeness-adjusted true incidence (laboratory + PIH denominators as proxy); agree this framing with NDoH and MAC before pilot launch.

## 4. Root cause analysis

```{mermaid}
%%| label: fig-surv-fishbone
%%| fig-cap: "Root causes of surveillance under-detection and the surveillance-improvement options that address each one."
flowchart LR
    A[Surveillance under-detection] --> B[NHLS–NMC MOU not<br/>IT-implemented]
    A --> C[PIH data not used<br/>for surveillance]
    A --> D[Delayed manual reporting]
    A --> F[No live-patient<br/>chemical toxicology]

    B --> G["✓ NHLS→NMC IT feed<br/><i>Completed</i>"]
    C --> H[Option S1: PIH MVD<br/>+ dashboard]
    D --> I[Option S2: BChE<br/>auto-notify <i>capability exists</i>]
    F --> J[Option S3: Sentinel<br/>toxicology + auto-notify]

    style A fill:#f99,stroke:#333
    style G fill:#e0e0e0,stroke:#999,stroke-dasharray: 5 5
    style H fill:#9f9,stroke:#333
    style I fill:#9f9,stroke:#333
    style J fill:#9f9,stroke:#333
```

Three active root causes remain **executional, not legislative**, and are each already legally permissible under Section 90 of the National Health Act (2003) and the existing NMC regulations. The fourth root cause (NHLS–NMC IT connectivity) has been resolved — the LIS-to-NMC feed is now operational.

### 4b. System-wide root cause analysis — what is in scope, and what is not

The fishbone above is deliberately narrow to surveillance under-detection. The figure below widens the lens to the **full set of root causes of pesticide-poisoning harm in South Africa** — surveillance, coordination, regulatory, and contextual — and maps each to a policy option in the D2P portfolio (this brief, the [Coordination brief](coordination_policy.qmd), and the [Terbufos brief](terbufos_policy.qmd)). **Drivers that are not connected to a policy option are explicitly out of scope** for the current D2P workstream: they are real problems, but addressing them sits outside this project's mandate, funding window, or institutional authority.

This chunk is intentionally self-contained (Mermaid only, no R dependencies) so it can be copied verbatim into the coordination or terbufos briefs, the analysis report, or a standalone slide deck.


```{mermaid}
%%| label: fig-system-rca
%%| fig-cap: "System-wide root cause analysis of pesticide-poisoning harm in South Africa, with each driver mapped to a D2P policy option. Green = addressed by a D2P option; grey-dashed = explicitly out of scope for D2P (real problem, but not the mandate of this workstream)."
flowchart LR
    HARM[Pesticide-poisoning<br/>harm in SA<br/>~2,620 deaths/yr<br/>~R2.34B/yr burden]:::harm

    %% ===== Four root-cause domains =====
    HARM --> SURV[Surveillance gaps]:::dom
    HARM --> COORD[Coordination &<br/>response gaps]:::dom
    HARM --> BAN[Regulatory /<br/>banning gaps]:::dom
    HARM --> CTX[Contextual<br/>upstream drivers]:::dom

    %% ===== Surveillance drivers =====
    SURV --> SV1[No integrated 3-stream<br/>system NMC + NHLS + PIH]
    SURV --> SV2[No live-patient<br/>chemical toxicology]
    SURV --> SV3[Hospital-only<br/>surveillance reach]
    SURV --> SV4[Limited EBS / community<br/>signal capture]
    SURV --> SV5[Poor data standardisation<br/>NMC does not follow<br/>AfriTox agent naming]
    SURV --> SV6[Forensic chemistry<br/>fully de-identified —<br/>no demographics]:::oos
    SURV --> SV7[MLDI system limited<br/>for surveillance utility]:::oos
    SURV --> SV8[CRVS / StatsSA MACOD<br/>3–4 year publication lag]:::oos
    SURV --> SV9[CRVS does not capture<br/>manner of death]:::oos

    SV1 --> OS_MVP[Surveillance MVP<br/>S1 + S2]:::opt
    SV2 --> OS3[Surveillance S3<br/>sentinel toxicology]:::opt
    SV3 --> OS1b[Surveillance S1b<br/>community / EBS layer<br/><i>Phase 2, conditional</i>]:::opt
    SV4 --> OS1b
    SV5 --> OS1[Surveillance S1<br/>AfriTox harmonisation<br/>+ NMC agent drop-down]:::opt

    %% ===== Coordination drivers =====
    COORD --> CO1[EHP investigation<br/>outcome not recorded<br/>back on NMC]
    COORD --> CO2[EHP → DALRRD referral<br/>has no SOP, no audit trail]
    COORD --> CO3[Clinicians get no<br/>case-status feedback]
    COORD --> CO4[No M&E of investigation,<br/>source removal, or<br/>regulatory action]
    COORD --> CO5[Unknown chemical agent<br/>registries — storage / use<br/>locations not shared<br/>with health sector]:::oos

    CO1 --> OC1[Coordination C1<br/>NMC closure-of-loop<br/>EHP field]:::opt
    CO2 --> OC2[Coordination C2<br/>structured EHP → DALRRD<br/>referral form]:::opt
    CO3 --> OC4[Coordination C4<br/>notifier auto-feedback]:::opt
    CO4 --> OC5[Coordination C5<br/>joint KPI dashboard]:::opt
    CO4 --> OC3[Coordination C3<br/>NICD → DALRRD<br/>intelligence feed]:::opt

    %% ===== Banning / regulatory drivers =====
    BAN --> BN1[Most toxic agent in recent<br/>mass-poisonings still legally<br/>registered Terbufos<br/>contrary to international std]
    BAN --> BN2[Informally traded<br/>street-pesticide market]
    BAN --> BN3[Repackaged into<br/>unlabelled containers]

    BN1 --> OT2[Terbufos T2<br/>cancel registration<br/>12–24 mo phase-out]:::opt
    BN2 --> OT2
    BN3 --> OT2

    %% ===== Contextual / upstream drivers — explicitly out of scope =====
    CTX --> CX1[HHPs used as<br/>vermin control<br/>in informal settlements]:::oos
    CTX --> CX2[Industry / trade pushback<br/>to protect crop yield<br/>despite safer substitutes]:::oos
    CTX --> CX3[Mental health burden<br/>and suicide pathway]:::oos
    CTX --> CX4[Low public awareness<br/>of informal-pesticide toxicity]:::oos
    CTX --> CX5[Low / unknown occupational<br/>health & safety on farms]:::oos

    %% ===== Styling =====
    classDef harm  fill:#f99,stroke:#900,stroke-width:2px,color:#000
    classDef dom   fill:#fde,stroke:#333,stroke-width:1.5px,color:#000
    classDef opt   fill:#cfc,stroke:#060,stroke-width:1.5px,color:#000
    classDef oos   fill:#eee,stroke:#999,stroke-dasharray: 5 5,color:#555
```


**How to read.** Each domain (surveillance / coordination / banning / contextual) lists every root cause raised in the D2P scoping work. **Green nodes** flow to a concrete policy option in this portfolio. **Grey-dashed nodes** are explicitly *not* addressed: forensic-chemistry de-identification and MLDI redesign are separate NHLS / NICD workstreams; CRVS lag and manner-of-death capture sit with StatsSA and Home Affairs; the chemical-agent storage registry sits with DALRRD and DTIC; and the contextual drivers (HHPs-as-rodenticide, industry pushback, mental health, public awareness, farm OHS) are real upstream determinants but are owned by departments and disciplines outside the D2P remit. Calling them out as out-of-scope here is the honest alternative to either pretending the portfolio addresses them or pretending they do not matter.

## 5. Policy options

::: {.callout-tip icon=false}
## ✓ NHLS–NMC IT feed — now operational

The NHLS–NICD LIS-to-NMC connector is operational; monthly BChE data flows to NICD under the existing MOU. The three active options below build on this.
:::

::: {.callout-note collapse="true"}
## Audit trail — mini-budget backing the S0 operating cost (R60,000/yr, now in baseline) {#sec-s0-mini-budget}

`c_S0_NHLS_NMC_feed_op` = R60,000/yr (gamma(25, 2400)); `c_S0_NHLS_NMC_feed_setup` = R80,000 one-off (already incurred). This feed is **in the status-quo baseline** — it is not a marginal cost of adopting the Surveillance MVP, but its operating cost is relevant context for decision-makers auditing total program spend.

| # | Line item | Driver | Annual (R) | Source / anchor |
|:-:|:----------|:-------|-----------:|:----------------|
| 1 | NICD analyst data extraction + cleaning | 0.05 FTE × R650k DPSA L10-11 | 32,500 | DPSA L10-11 scale 2024 (Tab 1 NMC analyst CTE) |
| 2 | NHLS CDW data extract QA (monthly BChE pull validation) | 0.02 FTE × R900k NHLS medical scientist | 18,000 | NHLS Patterson grade D 2024 (Tab 1 NHLS scientist CTE) |
| 3 | NICD SDW hosting + data-feed monitoring (shared cloud, pesticide slice) | AWS af-south-1 shared allocation | 6,000 | NICD data platform budget 2024 (Tab 2 "Data warehouse / ETL") |
| 4 | Annual MOU compliance review + governance documentation | Inter-agency MOU review + admin | 3,500 | NICD legal + NHLS MOU management |
|   | **Total** |   | **60,000** | matches `c_S0_NHLS_NMC_feed_op` in [`amua_import_parameters_v4.csv`](../amua_import_parameters_v4.csv) |

**Distribution.** gamma(shape = 25, scale = 2,400) → mean R60,000, SD ≈ R12,000. Items 1 and 2 are time-allocated personnel cost at standard CTE rates; items 3 and 4 are direct cash outlays. The same analyst CTE anchor (R650k DPSA L10-11) underpins the NMC baseline in §7 Tab 1, so S0 is consistent with and additive to the status-quo build-up.

**Setup.** `c_S0_NHLS_NMC_feed_setup` = R80,000 (one-off) covered: legal/MOU negotiation, data-mapping and schema agreement, NICD SDW ingestion script development, UAT. Already incurred; not a cost of adopting the MVP.
:::

### Option S1 — PIH live dashboard + Minimum Viable Dataset *(★ recommended)*

Two layers:

- **Short term (1–2 weeks):** PIH publishes a live dashboard of pesticide-related calls (counts, geography, agent, severity, demographics, **AfriTox agent name**). No new data; just a daily extract from existing call records.
- **Medium term (3–9 months):** PIH and NICD agree a **Minimum Viable Dataset (MVD)** so the ~1,158 clinician-initiated calls/year can be ingested into NMC as a structured surveillance stream. **Agent-name capture is a mandatory MVD field** — both PIH and NMC must record the AfriTox product name for every case. To make this reciprocal, **NMC should add a structured agent drop-down** (pesticide class → active ingredient → AfriTox product name) to its NDoH-hosted notification form: PIH records what the caller *reports*; NMC records what the clinician *confirms*. Combined: agent-attribution signal system-wide at near-zero marginal cost.

Detection rises to ~`r sprintf("%.0f", as.numeric(pick("d_S1_PIH_MVD")) * 100)` % of true cases (up from ~`r sprintf("%.0f", as.numeric(pick("d_status_quo")) * 100)` %) for a one-off setup of **`r zar(as.numeric(pick("c_S1_PIH_MVD_setup")))`**, then **`r zar(as.numeric(pick("c_S1_PIH_MVD_op")))`/year** from Year 2 onward (Year 1 total: **`r zar(as.numeric(pick("c_S1_PIH_MVD_setup")) + as.numeric(pick("c_S1_PIH_MVD_op")))`**; includes AfriTox agent-name harmonization). The dashboard is the lowest-hanging fruit in the whole portfolio: it ships in days.

::: {.callout-note collapse="true"}
## Audit trail — mini-budget backing the R82,000/yr operating cost {#sec-s1-mini-budget}

The `c_S1_PIH_MVD_op` parameter (mean R82,000/yr; gamma(25, 3280)) is **not a single quoted price** — it is the sum of seven small recurring line items required to keep the PIH dashboard live, the MVD pipeline running, and the AfriTox agent drop-down synchronised between PIH and NMC. All figures are 2024–2025 ZAR, marginal over the status-quo PIH operation (which is already costed in §7 Tab 1–2 from [`data/surveillance_costing.xlsx`](../data/surveillance_costing.xlsx)).

| # | Line item | Driver | Annual (R) | Source / anchor |
|:-:|:----------|:-------|-----------:|:----------------|
| 1 | Dashboard hosting (PIH Tygerberg cloud VM) | AWS af-south-1 t3.small + RDS + backup, 12 mo | 18,000 | AWS af-south-1 pricing 2024 (also Tab 2 of [`surveillance_costing.xlsx`](../data/surveillance_costing.xlsx), row "Dashboard hosting (Tygerberg cloud VM)", pesticide share) |
| 2 | Daily PIH→dashboard extract pipeline (script + on-call) | 0.02 FTE × R780k poisons specialist | 15,600 | DoH pharmacy OSD scale 2024 (PIH specialist CTE used in Tab 1) |
| 3 | AfriTox agent-name harmonisation maintenance (PIH ↔ NMC drop-down sync, quarterly refresh) | 0.02 FTE × R650k NICD epidemiologist | 13,000 | DPSA L10-11 scale 2024 (Tab 1 NMC analyst CTE) |
| 4 | MVD ingest pipeline operations (ETL run, error handling, schema drift) | NICD data engineer 0.015 FTE × R750k | 11,250 | SITA / state IT contracting 2024 (Tab 1 Cross-cutting) |
| 5 | QA + monthly NMC reconciliation (PIH-ingested rows vs NMC) | 1 h/wk × R650k NICD analyst (0.01 FTE) | 6,500 | DPSA L10-11 scale 2024 |
| 6 | Annual data-dictionary review + minor MVD releases | PIH + NICD workshop, 2 d/yr; consumables | 10,000 | PIH admin estimate (consistent with `c_S1_PIH_MVD_setup` R40,000 one-off) |
| 7 | Licences + alert SMS (Power BI workspace seats, ~5,000 outbound SMS/yr @ R0.20) | Microsoft ZA EA + Clickatell rate | 7,650 | Microsoft ZA EA 2024; Clickatell / BulkSMS ZA 2024 (same anchors as §7 Tab 2) |
|   | **Total** |   | **82,000** | matches `c_S1_PIH_MVD_op` in [`amua_import_parameters_v4.csv`](../amua_import_parameters_v4.csv) |

**How to interpret.** Items 1, 4 and 7 are direct cash outlays (cloud, ETL, licences, SMS). Items 2, 3, 5 and 6 are **time-allocated personnel cost** at standard public-sector cost-to-employer rates — i.e. they do not require new hires; they ring-fence small fractions of existing PIH and NICD posts. The same CTE anchors and unit-cost sources are used in the §7 status-quo build-up, so the marginal S1 line is consistent with — and additive to — the costed baseline.

**Distribution.** The CSV models this as gamma(shape = 25, scale = 3,280) → mean R82,000, SD ≈ R16,400 (95% CI ≈ R53k–R116k). This range absorbs reasonable variation across all seven items (e.g. cloud-VM sizing, AfriTox refresh frequency, FTE %).

**Sensitivity — what would push it materially higher.** (i) Hosting the dashboard on Power BI Premium dedicated capacity rather than a small VM (+R30–50k/yr); (ii) replacing the 0.02 FTE pharmacist time on the daily extract with a dedicated data engineer FTE (+R60k/yr if 0.1 FTE); (iii) extending alert-SMS volume 5× during outbreak periods (+R3–5k/yr). None of these move the option out of the "lowest-hanging fruit" tier vs S2 (R55,000/yr) or S1b (R-millions/yr).
:::

### Option S1b — Phase 2 community layer *(commission after MVP pilot if rural/informal gap confirms)*

S1 and S2 improve detection within the **formal healthcare system** — via PIH, NHLS BChE, and NMC-notified facilities. They structurally cannot reach cases that never arrive at a formal facility: the child taken to a traditional healer after ingesting street-sold granules, the farm worker who doesn't present at all, the informal-settlement cluster whose care pathway bypasses formal services entirely.

**The MVP ceiling (~`r sprintf("%.0f", as.numeric(pick("d_surveillance_mvp")) * 100)`% P_detect) is real and bounded** by the reach of formal administrative streams. S1b is the mechanism for going beyond it — but only after the MVP has established the baseline and confirmed that the remaining gap is structural rather than simply a compliance issue within the formal system.

**Trigger condition.** Commission S1b if, after 6 months of MVP operation, the geographic distribution of PIH + NHLS + NMC cases shows ≥40% of severe cases originating in districts with low formal-facility access (defined as <1 public hospital per 50km radius). This is a data-driven go/no-go, not a discretionary expansion. CSV parameters: `d_S1b_event_community_*`, `c_S1b_setup_*`, `c_S1b_op_*`, `p_fp_S1b_*`.

**False positives are a design parameter, not a disqualifier.** The gross false-positive rate is ~`r sprintf("%.0f%%", as.numeric(pick("p_fp_S1b_mid")) * 100)` (range `r sprintf("%.0f–%.0f%%", as.numeric(pick("p_fp_S1b_lo")) * 100, as.numeric(pick("p_fp_S1b_hi")) * 100)`). At P_detect = `r pick("d_S1b_event_community_mid")` on the mid-burden of `r scales::comma(as.numeric(pick("n_true_burden_from_statssa_mid")))` true cases, the gross notification volume would be ~`r scales::comma(round(as.numeric(pick("n_true_burden_from_statssa_mid")) * as.numeric(pick("d_S1b_event_community_mid")) / (1 - as.numeric(pick("p_fp_S1b_mid")))))` events/yr — of which ~`r scales::comma(round(as.numeric(pick("n_true_burden_from_statssa_mid")) * as.numeric(pick("d_S1b_event_community_mid"))))` are true cases and ~`r scales::comma(round(as.numeric(pick("n_true_burden_from_statssa_mid")) * as.numeric(pick("d_S1b_event_community_mid")) * as.numeric(pick("p_fp_S1b_mid")) / (1 - as.numeric(pick("p_fp_S1b_mid")))))` are false alerts *before filtering*. This is solvable with a **two-tier triage architecture** — the same design used by WHO-EWARN, APSED, and comparable LMIC systems:

1. **Tier 1 — CHW structured checklist**: Community health workers report via a mobile checklist requiring ≥2 of: acute onset, ≥2 affected persons, known or suspected product contact. Events not meeting criteria are logged but no dispatch is triggered.
2. **Tier 2 — EHP call-back within 24 hours**: A provincial EHP calls back to confirm before committing to on-site investigation.

With a two-tier filter, the effective net FP rate drops to approximately 8–12% — within the investigation capacity modelled in the [Coordination brief](coordination_policy.qmd).

**Two-stage rollout — Free State and Gauteng first.** These two provinces account for an estimated **~50% of the annual pesticide-poisoning burden** (working assumption; to be confirmed against the MVP pilot's geographic distribution from NMC, NHLS, and PIH data). Comparable LMIC systems have achieved P_detect in the range `r pick("d_S1b_event_community_lo")`–`r pick("d_S1b_event_community_hi")` — `r sprintf("%.1f–%.1f×", as.numeric(pick("d_S1b_event_community_lo")) / as.numeric(pick("d_surveillance_mvp")), as.numeric(pick("d_S1b_event_community_hi")) / as.numeric(pick("d_surveillance_mvp")))` the MVP ceiling.

| Stage | Provinces | Est. burden share | Setup cost | Operating cost/yr | Go/no-go criterion |
|:------|:----------|------------------:|-----------:|------------------:|:-------------------|
| **S1b pilot** | Free State + Gauteng | ~50% | ~`r sprintf("R%.0fM", as.numeric(pick("c_S1b_setup_lo"))/1e6)` | ~`r sprintf("R%.1fM/yr", as.numeric(pick("c_S1b_op_lo"))/1e6)` | Month 12: net P_detect >35% and net FP <15% |
| **S1b national** | All 9 provinces | 100% | ~`r sprintf("R%.1fM", as.numeric(pick("c_S1b_setup_hi"))/1e6)` | ~`r sprintf("R%.1fM/yr", as.numeric(pick("c_S1b_op_hi"))/1e6)` | Conditional on pilot outcome |

The cost range (`r sprintf("R%.0fM–R%.1fM setup; R%.1fM–R%.1fM/yr operating", as.numeric(pick("c_S1b_setup_lo"))/1e6, as.numeric(pick("c_S1b_setup_hi"))/1e6, as.numeric(pick("c_S1b_op_lo"))/1e6, as.numeric(pick("c_S1b_op_hi"))/1e6)` — approximately `r sprintf("%.0f–%.0f×", as.numeric(pick("c_S1b_op_lo")) / as.numeric(pick("c_surv_MVP_op")), as.numeric(pick("c_S1b_op_hi")) / as.numeric(pick("c_surv_MVP_op")))` the MVP operating cost) directly reflects the two-stage choice.

::: {.callout-note collapse="true"}
## Audit trail — mini-budget backing the S1b operating cost range {#sec-s1b-mini-budget}

The three `c_S1b_op_*` parameters reflect **distinct deployment footprints**, not sensitivity bounds around a single estimate. All figures are 2024–2025 ZAR, marginal over the Surveillance MVP baseline.

**Lo scenario — pilot (2 provinces: Free State + Gauteng) | `c_S1b_op_lo` = R800,000/yr**

| # | Line item | Driver | Annual (R) | Source / anchor |
|:-:|:----------|:-------|-----------:|:----------------|
| 1 | Provincial coordinators ×4 (FS + Gauteng) | 4 × R120,000 DPSA L8 pesticide-surveillance allocation | 480,000 | DPSA L8 mid-step CTE ~R600k–R750k; R120k = ~17% FTE allocation |
| 2 | Platform hosting + mobile checklist app | AWS af-south-1 + off-the-shelf EBS mobile app, 2-province scale | 120,000 | AWS af-south-1 pricing 2024 |
| 3 | CHW structured training + annual refresher | 2 provinces × 2 training cohorts + consumables | 120,000 | LMIC community surveillance analogue (IDSR pilot) |
| 4 | NICD data analyst + monthly reporting | 0.1 FTE × R650k DPSA L10-11 | 65,000 | DPSA L10-11 scale 2024 |
| 5 | EHP call-back protocol management + QA | 0.015 FTE × R650k NICD analyst | 15,000 | DPSA L10-11 scale 2024 |
|   | **Total (lo / pilot)** |   | **800,000** | matches `c_S1b_op_lo` |

**Mid scenario — working value (8 provinces) | `c_S1b_op_mid` = R1,500,000/yr**

From CSV note: *"~8 provincial coordinators (DPSA L8, R120k each) + national analyst + platform hosting + annual CHW refresher."*

| # | Line item | Driver | Annual (R) | Source / anchor |
|:-:|:----------|:-------|-----------:|:----------------|
| 1 | Provincial coordinators ×8 | 8 × R120,000 DPSA L8 | 960,000 | DPSA L8 scale 2024 (same unit cost as lo) |
| 2 | National NICD analyst | 0.2 FTE × R650k DPSA L10-11 | 130,000 | DPSA L10-11 scale 2024 |
| 3 | Platform hosting + app maintenance | AWS multi-province scale + maintenance | 180,000 | AWS af-south-1 pricing 2024 |
| 4 | Annual CHW refresher training | 8 provinces × cohort costs | 230,000 | EWARN/IDSR analogues |
|   | **Total (mid / working)** |   | **1,500,000** | matches `c_S1b_op_mid` |

**Hi scenario — full 9-province + dedicated EHP nodes | `c_S1b_op_hi` = R2,500,000/yr**

From CSV note: *"full 9-province deployment with dedicated EHP nodes and enhanced data cleaning."*

| # | Line item | Driver | Annual (R) | Source / anchor |
|:-:|:----------|:-------|-----------:|:----------------|
| 1 | Provincial coordinators + dedicated EHP nodes ×9 | 9 × R160,000 DPSA L8-9 (dedicated EHP grade) | 1,440,000 | DPSA L8-L9 scale 2024 |
| 2 | National NICD analyst (enhanced share) | 0.3 FTE × R650k | 195,000 | DPSA L10-11 scale 2024 |
| 3 | Platform hosting + enhanced data-cleaning infrastructure | 9-province scale + QA pipeline | 350,000 | AWS af-south-1 + QA infrastructure estimate |
| 4 | Annual CHW refresher training (9 provinces) | 9 provinces × full cohort | 515,000 | WHO-EWARN mature deployment analogue |
|   | **Total (hi / full)** |   | **2,500,000** | matches `c_S1b_op_hi` |

**DPSA L8 anchor.** R120,000 represents ~16–20% FTE of a DPSA L8 mid-step CTE (~R600k–R750k/yr). In practice this is an existing EHP or health-promotion officer whose pesticide-poisoning community-surveillance duties are ring-fenced at that fraction — not a new hire. The same unit cost applies across all three scenarios; the scenarios differ in the number of provinces covered and whether dedicated EHP nodes (requiring L8-9 grade) are included.

**Setup parameters.** `c_S1b_setup_lo/mid/hi` = R1M / R2M / R3.5M (one-off) cover mobile-app development, protocol design, and initial CHW training. They are separately parameterised and not included in the operating cost rows above.
:::

### Option S2 — BChE auto-notification *(★ recommended — capability exists)*

Configure the NHLS LIS to fire an automatic NICD notification when a BChE result shows severe inhibition (< 50 % of normal), within 24 hours of validation. **No new test, no new reagent contract** — only an IT configuration on top of the ~`r scales::comma(as.numeric(pick("n_nhls_total_annual")))` BChE tests/year already performed (`r scales::comma(as.numeric(pick("n_nhls_severe_annual")))` severe in 2023). Detection rises to ~`r sprintf("%.0f", as.numeric(pick("d_S2_BChE_autonotify")) * 100)` % of true cases (up from ~`r sprintf("%.0f", as.numeric(pick("d_status_quo")) * 100)` %) for a one-off setup of **`r zar(as.numeric(pick("c_S2_BChE_autonotify_setup")))`**, then **`r zar(as.numeric(pick("c_S2_BChE_autonotify_op")))`/year** from Year 2 onward (Year 1 total: **`r zar(as.numeric(pick("c_S2_BChE_autonotify_setup")) + as.numeric(pick("c_S2_BChE_autonotify_op")))`**) in 2–4 weeks. This is the highest-value, fastest option on the menu.

::: {.callout-note collapse="true"}
## Audit trail — mini-budget backing the R55,000/yr operating cost {#sec-s2-mini-budget}

`c_S2_BChE_autonotify_op` = R55,000/yr (gamma(25, 2200)) covers the pure IT infrastructure required to keep the HL7 auto-notification middleware running — **no new laboratory tests, no new reagent contract, no new hires.** The one-off LIS reconfiguration and UAT cost (`c_S2_BChE_autonotify_setup` = R200,000) is separately parameterised. All figures are 2024–2025 ZAR, marginal over the status-quo NHLS BChE baseline.

| # | Line item | Driver | Annual (R) | Source / anchor |
|:-:|:----------|:-------|-----------:|:----------------|
| 1 | HL7 middleware hosting (Mirth Connect; af-south-1 t3.small VM) | VM + storage, 12 mo | 18,000 | AWS af-south-1 pricing 2024 (same tier as §7 Tab 2 "LIS rule engine middleware") |
| 2 | NHLS LIS IT engineer maintenance | 0.02 FTE × R750k SITA/state IT rate | 15,000 | SITA / state IT contracting 2024 (Tab 1 Cross-cutting data engineer CTE) |
| 3 | NHLS medical scientist alert QA review | 0.005 FTE × R900k NHLS Patterson grade D | 4,500 | NHLS Patterson grade D 2024 (Tab 1 NHLS scientist CTE) |
| 4 | NICD epidemiologist alert triage + NMC routing | 0.01 FTE × R650k DPSA L10-11 | 6,500 | DPSA L10-11 scale 2024 (Tab 1 NMC analyst CTE) |
| 5 | Mirth Connect enterprise support licence / maintenance agreement | Annual contract | 7,000 | NHLS IT estimate (cited in `c_S2_BChE_autonotify_op` source note) |
| 6 | NICD analyst QA reconciliation (NHLS→NMC matching spot-checks) | 0.005 FTE × R650k | 3,250 | DPSA L10-11 scale 2024 |
| 7 | On-call incident-response allowance (middleware outage cover) | Annual allowance | 750 | NHLS IT estimate |
|   | **Total** |   | **55,000** | matches `c_S2_BChE_autonotify_op` in [`amua_import_parameters_v4.csv`](../amua_import_parameters_v4.csv) |

**How to interpret.** Items 1 and 5 are direct cash outlays (cloud hosting + licence). Items 2, 3, 4 and 6 are time-allocated personnel cost at standard public-sector CTE rates — the same anchors used in §7 Tab 1. The NHLS medical scientist and data-engineer CTE rows already appear in the status-quo build-up; the S2 marginal allocation is additive to the existing pesticide share of those posts.

**Distribution.** gamma(shape = 25, scale = 2,200) → mean R55,000, SD ≈ R11,000 (95% CI ≈ R36k–R78k).

**Sensitivity.** Dominant driver: middleware hosting tier (upgrading t3.small → t3.medium adds ~R12,000/yr). Doubling the NHLS IT maintenance FTE share adds ~R15,000/yr. Neither moves S2 above S1 in operating cost (R82,000/yr).

**Setup.** `c_S2_BChE_autonotify_setup` = R200,000 (one-off) covers NHLS LIS HL7 rule development, configuration of the severity threshold (< 50% inhibition), UAT across NHLS network sites, and go-live sign-off. Spent once; not part of the recurring R55,000/yr.
:::

### Option S3 — Sentinel live-patient toxicology *(★ recommended, medium term — distinct purpose)*

NHLS currently performs chemical toxicology (qualitative/quantitative pesticide assays) **post-mortem only**. There is no live-patient toxicology capability. Option S3 funds a sentinel tertiary site (e.g. CHB) to perform live-patient chemical toxicology, scaling to 4–5 sites in phase 2–3.

::: {.callout-caution appearance="minimal"}
**Forensic chemistry is fully de-identified and not fit for surveillance linkage.** No patient, facility, or case-linkage identifiers: cannot be matched to NMC or PIH records without structural re-design of specimen accessioning. The **MLDI system** is the longer-term pathway for surveillance-ready post-mortem toxicology — a separate workstream from S3.
:::

**S3 has a different purpose from S1 and S2** and is *not* a detection-coverage uplift. S1 and S2 increase the share of true cases that enter the surveillance system. **S3 identifies the proportion of chemical agents being found in severely poisoned living patients at sentinel sites** — the missing evidence base for product-specific regulation (see [Terbufos brief](terbufos_policy.qmd)). Without S3, ~10,000 survivors/year are recorded with agent "unknown / suspected" and targeted product restriction has no quantitative footing.

Cost: ~`r sprintf("R%.1fM", round(as.numeric(pick("c_option6_pilot_year1"))/1e6, 1))` pilot site → ~`r sprintf("R%.1fM", round(as.numeric(pick("c_option6_full_5site_annual"))/1e6, 1))` for 5-site rollout, 6–12 months. Funded from research-surveillance streams (SAMRC, Bloomberg, Gates), not the surveillance operating budget. Full quantitative concept in the [sentinel toxicology appendix](surveillance_concept_note.qmd).

::: {.callout-note collapse="true"}
## Audit trail — mini-budget backing the S3 sentinel toxicology costs {#sec-s3-mini-budget}

Four component-level parameters in [`amua_import_parameters_v4.csv`](../amua_import_parameters_v4.csv) underpin both S3 cost rows. The pilot and 5-site budgets run **below the maximum specimen capacity** (`n_specimens_per_site_per_year` = 646): the pilot targets ~150 GC-MS screens and ~200 clinical batteries (ramp-up year); the 5-site rollout targets ~100 screens and ~133 batteries per site per year (steady-state research enrolment).

**Pilot — 1 sentinel site (CHB) | `c_option6_pilot_year1` = R1,200,000/yr**

| # | Line item | Driver | Annual (R) | CSV parameter / source |
|:-:|:----------|:-------|-----------:|:----------------------|
| 1 | Research nurse 0.5 FTE (TASK research band) | Per-site personnel anchor | 175,000 | `c_research_nurse_05FTE_per_site` |
| 2 | GC-MS / LC-MS multi-residue screens (~150 specimens × R2,500) | Pilot ramp-up volume | 375,000 | `c_gcms_screen_per_specimen` × pilot volume |
| 3 | Enhanced clinical batteries (~200 patients × R1,500; BChE + LFT + U&E + repeat) | All enrolled patients | 300,000 | `c_clinical_battery_per_patient` × pilot volume |
| 4 | Chemical pathologist co-investigator time (0.07 FTE × R1,500k NHLS Patterson grade E) | Sign-out + supervision | 105,000 | NHLS Patterson grade E 2024 (Tab 1 NHLS chemical pathologist CTE) |
| 5 | REDCap data management + statistical reporting | 1-site platform + annual report | 80,000 | Concept Note DRAFT 3 |
| 6 | Site consumables, specimen transport + cold chain, overheads | ~150 specimens × ~R1,100 blended overhead | 165,000 | NHLS logistics tariff 2024 + consumables estimate |
|   | **Total (pilot, 1 site)** |   | **1,200,000** | matches `c_option6_pilot_year1` |

**Full rollout — 5 sentinel sites | `c_option6_full_5site_annual` = R3,625,000/yr**

| # | Line item | Driver | Annual (R) | CSV parameter / source |
|:-:|:----------|:-------|-----------:|:----------------------|
| 1 | Research nurses ×5 (0.5 FTE each) | 5 × `c_research_nurse_05FTE_per_site` | 875,000 | `c_research_nurse_05FTE_per_site` × 5 |
| 2 | GC-MS screens (5 sites × ~100 specimens × R2,500) | Steady-state research volume | 1,250,000 | `c_gcms_screen_per_specimen` × 5-site volume |
| 3 | Enhanced clinical batteries (5 sites × ~133 patients × R1,500) | BChE + LFT + U&E + repeat | 997,500 | `c_clinical_battery_per_patient` × 5-site volume |
| 4 | Multi-site coordination (NHLS + NICD; 0.2 FTE × R750k data engineer) | Network coordination | 150,000 | SITA / state IT contracting 2024 (Tab 1 Cross-cutting) |
| 5 | Data management + reporting (multi-site REDCap + analysis) | 5-site QA + annual report | 202,500 | Concept Note DRAFT 3 |
| 6 | Site consumables, transport, cold chain (5 sites × R30k) | Per-site recurring overhead | 150,000 | NHLS logistics tariff 2024 |
|   | **Total (5-site)** |   | **3,625,000** | matches `c_option6_full_5site_annual` |

**Why S3 is funded from research streams, not the surveillance operating budget.** S3 is a **research investment** — it generates the agent-attribution evidence base for targeted regulatory action. It does not increase P_detect (the share of true cases entering the surveillance system) and therefore does not appear in the S1/S2/MVP operating-cost parameters. Its value is quantified in the [Terbufos brief](terbufos_policy.qmd) where regulatory restriction converts agent-attribution evidence into deaths averted.

**Why specimens run below the per-site capacity of 646.** `n_specimens_per_site_per_year` (= 646) is the *anatomically derived maximum* (2 specimens per patient × 323 patients — the sample-size calculation for the agent-attribution study at p = 0.3, d = 0.05). Actual enrolment ramps up over the pilot year (~150 screens) and reaches a steady-state of ~100–133 screens per site in a funded 5-site network constrained by the research budget, not by specimen throughput.
:::

### Summary comparison


| Option | Setup (one-off) | Annual operating (yr 2+) | Year 1 total | P_detect (per-case) | Δ vs. status quo | Cost / pp | Time | Parameter |
|:-------|----------------:|-------------------------:|-------------:|:-------------------:|:----------------:|----------:|:-----|:----------|
| Status quo | — | R0 | R0 | $\frac{1,013}{20,960} = 0.048$ | — | — | — | [`d_status_quo`](../parameters_hub.qmd#tbl-detection) |
| S1: PIH dashboard + MVD (+AfriTox agent name) | `r zar(as.numeric(pick("c_S1_PIH_MVD_setup")))` | `r zar(as.numeric(pick("c_S1_PIH_MVD_op")))` | `r zar(as.numeric(pick("c_S1_PIH_MVD_setup")) + as.numeric(pick("c_S1_PIH_MVD_op")))` | $\frac{1,013 + 1,158}{20,960} = 0.104$ | `r sprintf("+%.1f pp", (as.numeric(pick("d_S1_PIH_MVD")) - as.numeric(pick("d_status_quo"))) * 100)` | `r zar(round(as.numeric(pick("c_S1_PIH_MVD_op")) / ((as.numeric(pick("d_S1_PIH_MVD")) - as.numeric(pick("d_status_quo"))) * 100)))` | 1–2 wks dash / 3–9 mo MVD | [`d_S1_PIH_MVD`](../parameters_hub.qmd#tbl-detection) |
| S2: BChE auto-notify | `r zar(as.numeric(pick("c_S2_BChE_autonotify_setup")))` | `r zar(as.numeric(pick("c_S2_BChE_autonotify_op")))` | `r zar(as.numeric(pick("c_S2_BChE_autonotify_setup")) + as.numeric(pick("c_S2_BChE_autonotify_op")))` | $\frac{1,013 + 1,779}{20,960} = 0.133$ | `r sprintf("+%.1f pp", (as.numeric(pick("d_S2_BChE_autonotify")) - as.numeric(pick("d_status_quo"))) * 100)` | `r zar(round(as.numeric(pick("c_S2_BChE_autonotify_op")) / ((as.numeric(pick("d_S2_BChE_autonotify")) - as.numeric(pick("d_status_quo"))) * 100)))` | 2–4 wks | [`d_S2_BChE_autonotify`](../parameters_hub.qmd#tbl-detection) |
| **S1 + S2 (Surveillance MVP)** | **`r zar(as.numeric(pick("c_surv_MVP_setup")))`** | **`r zar(as.numeric(pick("c_surv_MVP_op")))`** | **`r zar(as.numeric(pick("c_surv_MVP_setup")) + as.numeric(pick("c_surv_MVP_op")))`** | $\frac{1,013 + 1,779 + 1,158}{20,960} = 0.189$ | **`r sprintf("+%.1f pp", (as.numeric(pick("d_surveillance_mvp")) - as.numeric(pick("d_status_quo"))) * 100)`** | **`r zar(round(as.numeric(pick("c_surv_MVP_op")) / ((as.numeric(pick("d_surveillance_mvp")) - as.numeric(pick("d_status_quo"))) * 100)))`** | **12 weeks** | [`d_surveillance_mvp`](../parameters_hub.qmd#tbl-detection) |
| S1b: Phase 2 community layer *(pilot: FS + Gauteng; national conditional on pilot)* | `r sprintf("R%.0fM–R%.1fM", as.numeric(pick("c_S1b_setup_lo"))/1e6, as.numeric(pick("c_S1b_setup_hi"))/1e6)` | Pilot `r sprintf("R%.1fM/yr", as.numeric(pick("c_S1b_op_lo"))/1e6)`; national `r sprintf("R%.1fM/yr", as.numeric(pick("c_S1b_op_hi"))/1e6)` | `r sprintf("R%.1fM–R%.1fM", (as.numeric(pick("c_S1b_setup_lo")) + as.numeric(pick("c_S1b_op_lo")))/1e6, (as.numeric(pick("c_S1b_setup_hi")) + as.numeric(pick("c_S1b_op_hi")))/1e6)` | `r pick("d_S1b_event_community_lo")`–`r pick("d_S1b_event_community_hi")` (working: `r pick("d_S1b_event_community_mid")`) | `r sprintf("+%.0f to +%.0f pp", (as.numeric(pick("d_S1b_event_community_lo")) - as.numeric(pick("d_status_quo"))) * 100, (as.numeric(pick("d_S1b_event_community_hi")) - as.numeric(pick("d_status_quo"))) * 100)` | `r sprintf("%s–%s", zar(round(as.numeric(pick("c_S1b_op_hi")) / ((as.numeric(pick("d_S1b_event_community_hi")) - as.numeric(pick("d_status_quo"))) * 100))), zar(round(as.numeric(pick("c_S1b_op_lo")) / ((as.numeric(pick("d_S1b_event_community_lo")) - as.numeric(pick("d_status_quo"))) * 100))))` | 12 mo pilot → national | [`d_S1b_event_community_mid`](../parameters_hub.qmd) |
| S3: Sentinel toxicology (agent attribution — distinct purpose) | — (incl. in yr 1) | `r sprintf("R%.1fM/yr (5-site)", round(as.numeric(pick("c_option6_full_5site_annual"))/1e6,1))` | `r sprintf("R%.1fM (1-site pilot)", round(as.numeric(pick("c_option6_pilot_year1"))/1e6,1))` | n/a (not a coverage uplift) | feeds product-banning evidence base | n/a | 6–12 mo pilot | — |

: **Intermediate-outcome costing analysis — surveillance options, base-case.** Costs are 2024–25 ZAR; P_detect is the per-case detection probability anchored to the StatsSA MACOD 2023 mid-burden of `r scales::comma(as.numeric(pick("n_true_burden_from_statssa_mid")))` true cases/year. **Cost per pp** is the annual operating cost divided by the percentage-point gain in P_detect over status quo. S1 and S2 expand coverage within the formal healthcare system; **S3 is a parallel agent-attribution investment**, not a coverage uplift — see option text above. **S1b is a Phase 2 community layer**, commissioned conditional on MVP pilot data confirming a structural rural/informal gap (trigger: ≥40% of severe cases in low-facility-access districts after 6 months of MVP operation). {.striped .hover}

::: {.callout-note appearance="simple"}
**Scope of this brief.** Surveillance changes only what is *seen*. The status quo P_detect = `r pick("d_status_quo")` (§6 below) is a **measurement** of the current detection gap, not a counterfactual against which deaths are averted. Cases-averted and deaths-averted figures live in the [Coordination brief](coordination_policy.qmd) and [Terbufos brief](terbufos_policy.qmd) respectively.
:::

::: {.callout-note collapse="true"}
## Audit trail — mini-budget backing the R137,000/yr Surveillance MVP operating cost {#sec-mvp-mini-budget}

`c_surv_MVP_op` = R137,000/yr is a **deterministic sum** of two independent components. No distributional assumption is placed on the total; uncertainty propagates through the component gamma distributions in any PSA.

| Component | CSV parameter | Annual (R) | Mini-budget |
|:----------|:-------------|----------:|:-----------|
| S1 PIH dashboard + MVD (+ AfriTox harmonisation) | `c_S1_PIH_MVD_op` | 82,000 | [S1 audit trail](#sec-s1-mini-budget) |
| S2 BChE auto-notification | `c_S2_BChE_autonotify_op` | 55,000 | [S2 audit trail](#sec-s2-mini-budget) |
| **Surveillance MVP total** | `c_surv_MVP_op` | **137,000** | |

**No double-counting.** The S1 and S2 cost items sit on entirely separate operational pathways:

- **S1 items** (PIH cloud hosting, MVD ETL pipeline, AfriTox harmonisation, QA, licences/SMS) are driven by PIH call-record flows and the NICD/PIH data-dictionary agreement.
- **S2 items** (NHLS HL7 middleware, LIS IT maintenance, NHLS scientist QA, NICD alert triage) are driven by the NHLS LIS auto-notification rule.

The two NICD analyst FTE allocations sum to ~0.05 FTE total (S1: ~0.035 FTE; S2: ~0.015 FTE) — within the 0.2 FTE analyst post already costed in the §7 Tab 1 NMC baseline. **The MVP does not require a new NICD headcount.**

**Setup.** `c_surv_MVP_setup` = R240,000 (one-off) = `c_S1_PIH_MVD_setup` (R40,000) + `c_S2_BChE_autonotify_setup` (R200,000). S0 is already operational and its setup cost was separately incurred; it is not double-counted here. The S2 LIS HL7 reconfiguration (R200,000) dominates.
:::

::: {.callout-note}
## Audit: parameter names for costing each policy option

Every annual cost and setup cost in this brief is directly linked to a parameter in [`amua_import_parameters_v4.csv`](../amua_import_parameters_v4.csv). These are the exact parameter names used in the model and costing tables:

| Option | Setup cost parameter | Operating cost parameter |
|:-------|:--------------------|:------------------------|
| S1 — PIH dashboard + MVD (+ AfriTox harmonisation) | `c_S1_PIH_MVD_setup` | `c_S1_PIH_MVD_op` |
| S2 — BChE auto-notification | `c_S2_BChE_autonotify_setup` | `c_S2_BChE_autonotify_op` |
| Surveillance MVP (S1 + S2) | `c_surv_MVP_setup` | `c_surv_MVP_op` |
| S1b — event-based / community *(sensitivity)* | `c_S1b_setup_mid` (lo/hi also exist) | `c_S1b_op_mid` (lo/hi also exist) |
| S0 — NHLS-NMC IT feed (now operational) | `c_S0_NHLS_NMC_feed_setup` | `c_S0_NHLS_NMC_feed_op` |
| Status quo (NICD program) | — | `c_status_quo_program` |

**How to audit:**
- Each parameter above is a row in the CSV, with value, distribution, and source citation.
- The same parameter names are used in the decision tree (see §6) and the marginal cost table (§7, Tab 5).
- Every option has a **collapsible mini-budget audit trail** directly below its option text showing how each bottom-line figure is built up from unit-cost and personnel-time components: [S0](#sec-s0-mini-budget) · [S1](#sec-s1-mini-budget) · [S1b](#sec-s1b-mini-budget) · [S2](#sec-s2-mini-budget) · [S3](#sec-s3-mini-budget) · [MVP](#sec-mvp-mini-budget).
- All mini-budgets use the same CTE anchors as the §7 status-quo build-up (DPSA salary scales, NHLS Patterson grades, SITA contracting benchmarks, AWS af-south-1 pricing), so the marginal option costs are consistent with and additive to the costed baseline.
- To update a cost, edit the value in the CSV and re-render — all outputs update automatically.
:::

## 6. Decision tree — from surveillance investment to detection uplift {#sec-decision-tree}

::: {.callout-note collapse="true"}
## Audit trail — how every P_detect value in this brief was derived

All active P_detect values are **derived, not elicited**, on a single per-case scale. The arithmetic is auditable from three published anchors.

### Step 1 — the implied true annual burden (denominator)

| Anchor | Value | Source |
|--|--:|--|
| Annual NMC pesticide notifications | `r scales::comma(as.numeric(pick("n_nmc_annual")))` | `n_nmc_annual` — NICD NMC Annual Report 2023 |
| NMC-hospitalised case-fatality rate | `r sprintf("%.1f %%", as.numeric(pick("cfr_headline_mid")) * 100)` | `cfr_headline_mid` (range `r sprintf("%.0f–15 %%", as.numeric(pick("cfr_headline_lo")) * 100)`) — Free State 2018-19 + NMC observed |
| StatsSA MACOD pesticide-poisoning deaths/yr | `r scales::comma(as.numeric(pick("n_deaths_statssa_macod")))` | `n_deaths_statssa_macod` — StatsSA Causes of Death, 2023: Table 4.12: Accidental poisoning by and exposure to noxious substance (X40-X49) |

If notified cases have CFR ≈ 12.5 %, then the *full* true burden satisfies $\text{true cases} \times \text{CFR} = \text{deaths}$, so

$$
N_\text{true} \;=\; \frac{2{,}620}{0.125} \;=\; \mathbf{20{,}960}\ \text{cases/year} \quad (\texttt{n\_true\_burden\_from\_statssa\_mid})
$$

This is the **single denominator** used for every per-case P_detect below.

### Step 2 — the per-stream numerators (2023 actuals)

| Stream | Annual count | CSV row | Source |
|--|--:|--|--|
| NMC notifications | `r scales::comma(as.numeric(pick("n_nmc_annual")))` | `n_nmc_annual` | NICD NMC Annual Report 2023 |
| NHLS severe BChE inhibitions (<50 % normal) | `r scales::comma(as.numeric(pick("n_nhls_severe_annual")))` | `n_nhls_severe_annual` | NHLS LIS 2023 (of `r scales::comma(as.numeric(pick("n_nhls_total_annual")))` total tests) |
| PIH clinician-initiated calls | `r scales::comma(as.numeric(pick("n_pih_annual")))` | `n_pih_annual` | PIH (Tygerberg) 2023 |

The three streams are treated as **mutually exclusive** in the base case (severe BChE ≠ NMC-notified ≠ PIH-called); the overlap-correction sensitivity is parameterised separately in the hub.

### Step 3 — P_detect for each option

Each option adds one or more streams to the NMC base. The numerator is the sum of streams reaching the surveillance system; the denominator is always 20,960.

| Option | Streams summed | Arithmetic | P_detect | CSV row |
|--|--|--:|--:|--|
| Status quo | NMC only | $1{,}013 / 20{,}960$ | **0.048** | `d_status_quo` |
| S1 — PIH MVD + dashboard | NMC + PIH | $(1{,}013 + 1{,}158) / 20{,}960 = 2{,}171 / 20{,}960$ | **0.104** | `d_S1_PIH_MVD` |
| S2 — BChE auto-notify | NMC + NHLS severe | $(1{,}013 + 1{,}779) / 20{,}960 = 2{,}792 / 20{,}960$ | **0.133** | `d_S2_BChE_autonotify` |
| **Surveillance MVP (S1 + S2)** | NMC + NHLS severe + PIH | $(1{,}013 + 1{,}779 + 1{,}158) / 20{,}960 = 3{,}950 / 20{,}960$ | **0.189** | `d_surveillance_mvp` |

### Step 4 — CFR sensitivity bracket on the status-quo anchor

The denominator scales inversely with the CFR assumed for NMC-notified cases (CSV: `d_status_quo_lo`, `d_status_quo_hi`):

- Lower bound (CFR = `r pick("cfr_headline_lo")`): $P_\text{detect}^{SQ} = 1{,}013 \times 0.10 / 2{,}620 = 0.039$ (implied burden `r scales::comma(as.numeric(pick("n_true_burden_from_statssa_hi")))`)
- Mid (CFR = `r pick("cfr_headline_mid")`): $P_\text{detect}^{SQ} = `r pick("d_status_quo")`$ (implied burden `r scales::comma(as.numeric(pick("n_true_burden_from_statssa_mid")))`)
- Upper (CFR = `r pick("cfr_headline_hi")`): $P_\text{detect}^{SQ} = 1{,}013 \times 0.15 / 2{,}620 = 0.058$ (implied burden `r scales::comma(as.numeric(pick("n_true_burden_from_statssa_lo")))`)

The same CFR bracket applies proportionally to S1/S2/MVP — e.g. the MVP varies from `r sprintf("%.3f", as.numeric(pick("d_surveillance_mvp")) * as.numeric(pick("n_true_burden_from_statssa_lo")) / as.numeric(pick("n_true_burden_from_statssa_mid")))` (CFR `r pick("cfr_headline_hi")`) to `r sprintf("%.3f", as.numeric(pick("d_surveillance_mvp")) * as.numeric(pick("n_true_burden_from_statssa_hi")) / as.numeric(pick("n_true_burden_from_statssa_mid")))` (CFR `r pick("cfr_headline_lo")`).

### Step 5 — why this replaces the v1 value of 0.45

The legacy v1 figure was an expert-elicited estimate on a different scale, derived from a model of spatial event sensitivity rather than per-case detection. That quantity answered a different question and is not comparable to a per-case detection probability. This brief now uses the **per-case** framing throughout — every value above is directly auditable from NICD + NHLS + PIH + StatsSA published figures and the canonical [`amua_import_parameters_v4.csv`](../amua_import_parameters_v4.csv).
:::

````{r}
#| label: fig-surv-decision-tree
#| echo: false
#| output: asis

# Source the decision-tree engine and the surveillance tree builder.
# Both scripts auto-resolve their paths whether rendered from the project
# root or from posts/. All numbers below come from the rollback of the tree
# object built off amua_import_parameters_v4.csv -- no inline arithmetic.
source("../R/surveillance_tree.r")

params   <- load_params()                          # named list from CSV
tree     <- build_surveillance_tree(params)        # dt_node object
res      <- evs_to_df(eval_tree(tree))             # per-case rollback (tibble)
annual   <- scale_surveillance_to_annual(res, params$n_true_burden_from_statssa_mid)

# Pull arm-level values out of the rolled-back tibble. The arm names match
# those defined in build_surveillance_tree() in R/surveillance_tree.r.
arm <- function(a) annual %>% filter(arm == a)
d_sq  <- arm("status_quo")$detected
d_s1  <- arm("S1_PIH_MVD")$detected
d_s2  <- arm("S2_BChE_autonotify")$detected
d_mvp <- arm("MVP_S1_plus_S2")$detected

c_s1   <- arm("S1_PIH_MVD")$annual_cost_zar
c_s2   <- arm("S2_BChE_autonotify")$annual_cost_zar
c_mvp  <- arm("MVP_S1_plus_S2")$annual_cost_zar

pp <- function(d) sprintf("%+.1f pp", (d - d_sq) * 100)

cat(sprintf('```{mermaid}
%%%%| fig-cap: "Surveillance investment decision tree — per-case detection uplift only. Each option raises P_detect, the share of true annual cases (~20,960 StatsSA-anchored mid-burden) that enter the surveillance system. The absolute burden impact of these uplifts depends on response-side parameters covered in the coordination and terbufos briefs. S3 (sentinel toxicology) is shown separately because it is an agent-attribution investment, not a coverage uplift. All numerical values pulled from amua_import_parameters_v4.csv at render time."
flowchart TD
    INV{Surveillance<br/>investment?} -->|None| SQ[Status quo<br/>P_detect = %.3f<br/>~5%% of true cases]:::bad
    INV -->|S1 only<br/>R%s/yr| O1[+ PIH MVD<br/>P_detect = %.3f<br/>%s vs SQ]:::warn
    INV -->|S2 only<br/>R%s/yr| O2[+ BChE auto-notify<br/>P_detect = %.3f<br/>%s vs SQ]:::warn
    INV -->|S1 + S2<br/>R%s/yr| MVP[Surveillance MVP<br/>P_detect = %.3f<br/>%s vs SQ]:::good
    S3[S3 Sentinel toxicology<br/>agent attribution<br/><i>parallel investment —<br/>see Terbufos brief</i>]:::aux

    classDef bad fill:#fdd,stroke:#cc0000,stroke-width:1.5px
    classDef warn fill:#ffe9b3,stroke:#cc8800,stroke-width:1.5px
    classDef good fill:#cfc,stroke:#060,stroke-width:1.5px
    classDef aux fill:#e6f0ff,stroke:#3366cc,stroke-width:1.5px,stroke-dasharray: 5 5
```
',
d_sq,
format(c_s1,  big.mark = ","), d_s1,  pp(d_s1),
format(c_s2,  big.mark = ","), d_s2,  pp(d_s2),
format(c_mvp, big.mark = ","), d_mvp, pp(d_mvp)
))
````

The tree reports **per-case detection uplift in percentage points**. The status-quo node is a **measurement** of the current detection gap (1,013 NMC notifications ÷ ~20,960 implied true cases), not a baseline from which deaths are counted as averted — surveillance has no response attached. Four detection inputs (`d_status_quo`, `d_S2_BChE_autonotify`, `d_S1_PIH_MVD`, `d_surveillance_mvp`) and three operating-cost rows (`c_S1_PIH_MVD_op`, `c_S2_BChE_autonotify_op`, `c_surv_MVP_op`) live in [`amua_import_parameters_v4.csv`](../amua_import_parameters_v4.csv) — edit the CSV and re-render. Option S3 sits outside this tree: it is an agent-attribution investment whose value is parameterised in the [Terbufos brief](terbufos_policy.qmd) and the [sentinel toxicology appendix](surveillance_concept_note.qmd).

::: {.callout-note appearance="minimal"}
**S1b (Phase 2 community layer) is not modelled in this decision tree.** The MVP arm represents the ceiling achievable by linking existing administrative streams. S1b would require a new arm with P_detect drawn from `d_S1b_event_community_mid` = `r pick("d_S1b_event_community_mid")` (range `r pick("d_S1b_event_community_lo")`–`r pick("d_S1b_event_community_hi")`) and annual operating cost drawn from `c_S1b_op_mid` = `r scales::comma(as.numeric(pick("c_S1b_op_mid")))`. It is excluded from the current tree because it is a Phase 2 option commissioned conditional on MVP pilot results, not a parallel near-term decision; and because its coordination cost depends on the net false-positive rate after two-tier triage (`p_fp_S1b_mid` = `r pick("p_fp_S1b_mid")` gross; ~8–12% net) modelled in the [Coordination brief](coordination_policy.qmd).
:::

**For burden impact, see the next brief.** The same P_detect figures feed into the [Coordination brief](coordination_policy.qmd) decision tree, where response efficacy η converts detection into cases averted, and into the [Terbufos brief](terbufos_policy.qmd), where regulatory restriction converts cases into deaths averted.

## 7. Status-quo costing — what surveillance costs today {#sec-costing}

::: {.callout-note icon=false}
## How to read & edit this section

The tables below are the **average-cost economic build-up of running surveillance for pesticide poisoning today** (status quo, 2024–25 ZAR). They are sourced from an editable workbook — [`data/surveillance_costing.xlsx`](../data/surveillance_costing.xlsx) — with three sheets:

- **`personnel`** — staff time (CTE × FTE %) by stream
- **`system`** — infrastructure, hosting, SMS, reagents by stream
- **`allocation`** — what share of each stream's total volume is pesticide-related (e.g. ~100 pesticide notifications/month out of ~10,000 total NMC notifications ≈ **0.84 %** of NMC platform OPEX is attributable to pesticide work)

To revise a unit cost or FTE, **edit a cell in the .xlsx and re-render** — every number below updates. The CSV (`amua_import_parameters_v4.csv`) remains the source of truth for *model* parameters and *marginal* policy-option costs (S1 / S2 / MVP). Salary anchors come from the DPSA scales (L10–11), NHLS Patterson grades (D/E), DoH OSD scales (MO, pharmacy), NICD recruitment notices and SITA contracting benchmarks; system anchors come from NHLS State Tender RT-262, AWS af-south-1 pricing, Clickatell/BulkSMS SA, and the NICD ICT budget — full citations are in each row's `source` column.
:::

```{r}
#| label: load-costing
#| message: false
#| warning: false

library(readxl)
library(scales)
library(gt)
library(dplyr)

cost_path <- "../data/surveillance_costing.xlsx"
personnel  <- read_excel(cost_path, sheet = "personnel")
system_dat <- read_excel(cost_path, sheet = "system")
allocation <- read_excel(cost_path, sheet = "allocation")

# Allocation factor: share of stream activity that is pesticide-related.
alloc_factor <- function(stream_name) {
  a <- allocation %>% filter(stream == stream_name)
  if (nrow(a) == 0) return(1)
  a$allocation_pct[1] / 100
}

# Apply pesticide-share allocation to platform-level OPEX rows (rationale
# contains the literal phrase "pesticide share").
system_alloc <- system_dat %>%
  mutate(
    is_platform = grepl("pesticide share", rationale, ignore.case = TRUE),
    pesticide_share_pct = if_else(
      is_platform,
      vapply(stream, function(s) alloc_factor(s) * 100, numeric(1)),
      100
    ),
    pesticide_cost_zar = round(annual_cost_zar * pesticide_share_pct / 100)
  )

zar <- function(x) paste0("R", scales::comma(round(x)))
```

::: {.panel-tabset}

### 1. Personnel

Annual cost-to-employer × FTE share attributable to pesticide surveillance, by stream.

```{r}
#| label: tbl-cost-personnel

personnel %>%
  arrange(stream, desc(annual_cost_zar)) %>%
  select(stream, role, annual_cte_zar, fte_pct, annual_cost_zar, rationale, source) %>%
  gt(rowname_col = "role", groupname_col = "stream") %>%
  fmt_number(columns = c(annual_cte_zar, annual_cost_zar), decimals = 0) %>%
  fmt_number(columns = fte_pct, decimals = 1, suffixing = FALSE, pattern = "{x}%") %>%
  cols_label(
    annual_cte_zar  = "Annual CTE (R)",
    fte_pct         = "FTE %",
    annual_cost_zar = "Allocated cost (R/yr)",
    rationale       = "Rationale",
    source          = "Source"
  ) %>%
  grand_summary_rows(
    columns = annual_cost_zar,
    fns = list("Total personnel" = ~ sum(.)),
    fmt = list(~ fmt_number(., decimals = 0))
  ) %>%
  tab_options(table.font.size = 11)
```

### 2. System

Infrastructure, hosting, SMS, reagents and licences — unit cost × annual volume, then allocated to the pesticide share of stream activity where applicable.

```{r}
#| label: tbl-cost-system

system_alloc %>%
  arrange(stream, desc(pesticide_cost_zar)) %>%
  select(stream, item, unit_cost_zar, units_per_yr, annual_cost_zar,
         pesticide_share_pct, pesticide_cost_zar, rationale, source) %>%
  gt(rowname_col = "item", groupname_col = "stream") %>%
  fmt_number(columns = c(unit_cost_zar, units_per_yr,
                         annual_cost_zar, pesticide_cost_zar),
             decimals = 0) %>%
  fmt_number(columns = pesticide_share_pct, decimals = 1, pattern = "{x}%") %>%
  cols_label(
    unit_cost_zar       = "Unit cost (R)",
    units_per_yr        = "Units/yr",
    annual_cost_zar     = "Raw cost (R/yr)",
    pesticide_share_pct = "Pesticide share",
    pesticide_cost_zar  = "Allocated cost (R/yr)",
    rationale           = "Rationale",
    source              = "Source"
  ) %>%
  grand_summary_rows(
    columns = pesticide_cost_zar,
    fns = list("Total system (pesticide-allocated)" = ~ sum(.)),
    fmt = list(~ fmt_number(., decimals = 0))
  ) %>%
  tab_options(table.font.size = 11)
```

### 3. Allocation factors

Why these proportions matter: platform-level OPEX (NMC hosting, PIH telephony, AfriTox licence) covers *all* notifiable / call activity, of which pesticide poisoning is only a slice. The denominator drives the allocated cost.

```{r}
#| label: tbl-cost-allocation

allocation %>%
  select(stream, basis, pesticide_count_yr, total_count_yr,
         allocation_pct, rationale, source) %>%
  gt(rowname_col = "basis", groupname_col = "stream") %>%
  fmt_number(columns = c(pesticide_count_yr, total_count_yr), decimals = 0) %>%
  fmt_number(columns = allocation_pct, decimals = 2, pattern = "{x}%") %>%
  cols_label(
    pesticide_count_yr = "Pesticide (n/yr)",
    total_count_yr     = "Total (n/yr)",
    allocation_pct     = "Share",
    rationale          = "Rationale",
    source             = "Source"
  ) %>%
  tab_options(table.font.size = 11)
```

### 4. Status-quo total

Status-quo annual cost of running pesticide-poisoning surveillance across the three streams, summed from the personnel and system sheets.

```{r}
#| label: tbl-cost-total

pers_by_stream <- personnel %>%
  group_by(stream) %>%
  summarise(personnel_zar = sum(annual_cost_zar), .groups = "drop")

sys_by_stream <- system_alloc %>%
  group_by(stream) %>%
  summarise(system_zar = sum(pesticide_cost_zar), .groups = "drop")

totals <- full_join(pers_by_stream, sys_by_stream, by = "stream") %>%
  mutate(across(c(personnel_zar, system_zar), ~ replace_na(., 0)),
         total_zar = personnel_zar + system_zar) %>%
  arrange(desc(total_zar))

totals %>%
  gt(rowname_col = "stream") %>%
  fmt_number(columns = c(personnel_zar, system_zar, total_zar), decimals = 0) %>%
  cols_label(
    personnel_zar = "Personnel (R/yr)",
    system_zar    = "System (R/yr)",
    total_zar     = "Total (R/yr)"
  ) %>%
  grand_summary_rows(
    columns = c(personnel_zar, system_zar, total_zar),
    fns = list("Status-quo total" = ~ sum(.)),
    fmt = list(~ fmt_number(., decimals = 0))
  ) %>%
  tab_options(table.font.size = 11)
```

### 5. Marginal cost of policy options

For comparison: the **additional** annual operating cost of each surveillance investment, pulled directly from [`amua_import_parameters_v4.csv`](../amua_import_parameters_v4.csv) (these are the same numbers driving the decision tree in §6).

```{r}
#| label: tbl-cost-marginal

marginal <- tibble::tribble(
  ~option, ~setup_param, ~op_param,
  "S1 — PIH dashboard + MVD (+ AfriTox harmonisation)", "c_S1_PIH_MVD_setup",           "c_S1_PIH_MVD_op",
  "S2 — BChE auto-notification",                         "c_S2_BChE_autonotify_setup",   "c_S2_BChE_autonotify_op",
  "Surveillance MVP (S1 + S2)",                          "c_surv_MVP_setup",             "c_surv_MVP_op",
  "S0 — NHLS-NMC IT feed (now operational)",             "c_S0_NHLS_NMC_feed_setup",     "c_S0_NHLS_NMC_feed_op",
  "S1b — event-based community (sensitivity, lo)",       "c_S1b_setup_lo",               "c_S1b_op_lo",
  "S1b — event-based community (sensitivity, mid)",      "c_S1b_setup_mid",              "c_S1b_op_mid",
  "S1b — event-based community (sensitivity, hi)",       "c_S1b_setup_hi",               "c_S1b_op_hi"
) %>%
  mutate(
    setup_zar = vapply(setup_param, function(n) as.numeric(pick(n)), numeric(1)),
    op_zar    = vapply(op_param,    function(n) as.numeric(pick(n)), numeric(1))
  ) %>%
  select(option, setup_zar, op_zar, setup_param, op_param)

marginal %>%
  gt(rowname_col = "option") %>%
  fmt_number(columns = c(setup_zar, op_zar), decimals = 0) %>%
  cols_label(
    setup_zar   = "Setup (R, one-off)",
    op_zar      = "Operating (R/yr)",
    setup_param = "CSV: setup",
    op_param    = "CSV: operating"
  ) %>%
  tab_options(table.font.size = 11)
```

:::

**Headline read.** The status-quo surveillance system already costs roughly **`r zar(sum(totals$total_zar))`/yr** in personnel + pesticide-allocated system spend across NMC, PIH, and NHLS BChE — borne almost entirely by NDoH, NHLS, and Stellenbosch FoMHS rather than appearing as a line item. Against that base, the **Surveillance MVP (S1 + S2) adds only `r zar(as.numeric(pick("c_surv_MVP_op")))`/yr** — a **`r sprintf("%.1f%%", 100 * as.numeric(pick("c_surv_MVP_op")) / sum(totals$total_zar))` uplift on current spend** for a **`r sprintf("+%.1f pp", (as.numeric(pick("d_surveillance_mvp")) - as.numeric(pick("d_status_quo"))) * 100)`** jump in P_detect.

## 8. Implementation considerations

| Component | Lead agency | Year-1 cost | Critical path |
|:----------|:------------|------------:|:--------------|
| S1 PIH dashboard | PIH (Tygerberg) + NICD | `r scales::comma(as.numeric(pick("c_S1_PIH_MVD_op")))` | Daily extract + hosting; MVD workshop |
| S2 BChE auto-notify | NHLS LIS team | `r scales::comma(as.numeric(pick("c_S2_BChE_autonotify_op")))` | LIS rule configuration; QA review |
| S3 Sentinel toxicology | NHLS clinical chemistry + sentinel hospital | `r zar(as.numeric(pick("c_option6_pilot_year1")))` pilot | Lab method validation; sample logistics |
| S1b Phase 2 community layer *(pilot: FS + Gauteng; commission after Month 6 MVP review)* | NICD + provincial DoH EHPs | Pilot: `r sprintf("R%.0fM setup + R%.1fM/yr", as.numeric(pick("c_S1b_setup_lo"))/1e6, as.numeric(pick("c_S1b_op_lo"))/1e6)`; national: `r sprintf("R%.1fM setup + R%.1fM/yr", as.numeric(pick("c_S1b_setup_hi"))/1e6, as.numeric(pick("c_S1b_op_hi"))/1e6)` | CHW two-tier triage training, mobile checklist app, EHP call-back protocol, QA pipeline |

**Privacy & legal.** The PIH-to-NMC MVD feed transmits personal health information (age-band, sex, geography, clinical data) and must use a **secure, authenticated HTTPS/TLS API** — no plaintext or email transmission. De-identification before transmission is mandatory (age-band not exact age; province/district not street address; no name or phone number; see Appendix A2). The NHLS BChE auto-notify feed transmits result-level data with patient identifiers to NMC, consistent with the existing NMC notification pathway, and must use the same secure HL7-over-TLS channel. Both flows are POPIA-compliant under existing Section 90 NHA authority. **No new legislation required.**

**Risks.** (i) Low BChE testing uptake by clinicians — S2 fires only on tests ordered. (ii) Data-quality drift if facility metadata is incomplete. (iii) Capacity bottleneck at provincial response level — the [Coordination brief](coordination_policy.qmd) addresses this.

## 9. Recommended next step

**Adopt the Surveillance MVP (S1 + S2) and pilot in 3 provinces (Gauteng, Western Cape, KZN) for 6 months.** Approximate pilot budget: **~`r zar(as.numeric(pick("c_surv_MVP_op")))` operating** (NHLS–NMC IT feed is now operational). Lead agency: NICD as national coordinator.

- **Weeks 1–2:** S1 PIH dashboard live; S2 BChE LIS configuration begins.
- **Months 1–3:** S1 MVD workshop with PIH finalised; S2 BChE auto-notify QA review completed.
- **Month 6:** Evaluate against (a) completeness of NHLS severe results in NMC and (b) signal-to-investigation latency.
- **Month 9:** Add S3 sentinel toxicology site, funded from a separate research-surveillance stream.
- **Month 12:** Review MVP geographic distribution. If ≥40% of severe cases originate in low-facility-access districts, commission the S1b Phase 2 pilot (Free State + Gauteng; ~`r sprintf("R%.0fM", as.numeric(pick("c_S1b_setup_lo"))/1e6)` setup + `r sprintf("R%.1fM/yr", as.numeric(pick("c_S1b_op_lo"))/1e6)` operating).

The coordination side of the same system — how NICD, EHPs, DALRRD, and provincial response teams act on the improved signal — is covered in the [Coordination brief](coordination_policy.qmd). The product-specific regulatory action that S3 makes possible is the subject of the [Terbufos brief](terbufos_policy.qmd).

---

::: {.footer-cta}
**Companion briefs:** [Coordination](coordination_policy.qmd) · [Terbufos restriction](terbufos_policy.qmd) · [Decision Model](decision_model.qmd) · [Costing Analysis](costing_analysis.qmd)

**Technical appendices:** [A — Surveillance decision analysis](../analysis_report_v2.qmd) · [D — Sentinel toxicology (S3 deep dive)](surveillance_concept_note.qmd)

**Appendices (this brief):** [Appendix A — MVD specifications (NMC + PIH)](#sec-mvd)
:::

---

## Appendix A — Minimum Viable Dataset (MVD) specifications {#sec-mvd}

### A1. NMC MVD — required fields for PIH-to-NMC structured feed

The table below specifies the **Minimum Viable Dataset** for ingesting PIH clinician-initiated calls into the National Microbiology and Communicable Disease Notification (NMC) system as a structured surveillance stream. These fields must be captured at the point of PIH call (what the *caller reports*) and confirmed/updated by the notifying clinician (what the clinician *confirms*). The pesticide-type and circumstances classification is drawn from the existing NMC notification form; the MVD mandates their structured capture for every pesticide-poisoning case.

::: {.callout-note appearance="minimal"}
**Agent-name capture is mandatory.** Both PIH and NMC must record the **AfriTox product name** for every case. NMC should add a structured agent drop-down (pesticide class → active ingredient → AfriTox product name) to the NDoH-hosted notification form to make this reciprocal with PIH. Cases with agent "unknown" are acceptable (do not exclude), but the field must be present.
:::

**Pesticide type (structured drop-down — mandatory)**

| Code | Pesticide type | Example active ingredients |
|:-----|:--------------|:--------------------------|
| OPC  | Organophosphates | Chlorpyrifos, dimethoate, profenofos |
| CAR  | Carbamates | Aldicarb, methomyl, carbofuran |
| PYR  | Pyrethroids | Cypermethrin, deltamethrin, permethrin |
| PAR  | Paraquat & Diquat | Paraquat dichloride, diquat dibromide |
| PHO  | Phosphine & Phosphides | Aluminium phosphide, zinc phosphide |
| NEO  | Neonicotinoid Insecticides | Imidacloprid, thiamethoxam |
| ORG  | Organochlorines | DDT, endosulfan, lindane |
| CPH  | Chlorophenoxy Herbicides | 2,4-D, MCPA, triclopyr |
| GLY  | Glyphosates | Glyphosate-IPA, glyphosate-ammonium |
| COU  | Coumarin Rodenticides | Brodifacoum, bromadiolone, chlorophacinone |
| FOR  | Formamidines | Amitraz, chlordimeform |
| FUN  | Fungicides — Unspecified | — |
| HRB  | Herbicides — Unspecified | — |
| INS  | Insecticides — Unspecified | — |
| DEE  | Diethyl-m-toluamide (DEET) | DEET |
| AFX  | Aflatoxin | Aflatoxin B1/B2/G1/G2 |
| NAP  | Naphthalene (marbles) | Naphthalene |
| OTH  | Other | (free text required) |
| UNK  | Unknown | — |

: NMC pesticide-type classification. Source: existing NMC notification form. AfriTox product-name lookup linked to each class at data entry. {#tbl-mvd-pesticide-type .striped .hover}

**Circumstances / intent (structured drop-down — mandatory)**

| Code | Circumstance | ICD-10 intent mapping |
|:-----|:------------|:----------------------|
| UI-OCC | Unintentional — Occupational | X48 (farm, pest control, industrial) |
| UI-DOM | Unintentional — Domestic | X40–X44 |
| INT-SH | Intentional — Self-harm | X60–X69 |
| INT-AH | Intentional — Assault/Homicide | X85–X90 |
| ENV-DR | Environmental / Drift | X48 (environmental context) |
| ACC-CN | Accidental contamination (food/water) | X49 |
| UNK-UI | Unknown / Under investigation | — |

: NMC circumstances/intent classification. Source: existing NMC notification form. Must be captured at both PIH call (caller-reported) and NMC notification (clinician-confirmed); these may differ and both values should be retained. {#tbl-mvd-circumstances .striped .hover}

**Remaining mandatory MVD fields for NMC notification**

| Field | Format | Notes |
|:------|:-------|:------|
| Case date (date of exposure / presentation) | ISO 8601 (YYYY-MM-DD) | Date of poisoning event or first clinical presentation |
| Facility / point of care | Facility NHIS code | Or "PIH referral" if never attended facility |
| Province | 2-letter code (e.g. WC, GP, FS) | Province of exposure, not of facility |
| District | LIM-code or free text | For geographic clustering analysis |
| Outcome at notification | Survived / Died / Unknown | MUST be updated if outcome changes |
| AfriTox product name | Free text + drop-down lookup | Mandatory field; "unknown" is valid if product unidentified |
| Notification source | NMC-direct / PIH-MVD / NHLS-autonotify | Tracks which stream generated the case |

: MVD — additional mandatory fields. These are the minimum set required to (a) link across streams, (b) support geographic clustering, and (c) support product-specific attribution. Fields marked "unknown" are acceptable; empty/null is not. {#tbl-mvd-core .striped .hover}

---

### A2. PIH MVD — required fields including demographic variables

PIH captures call data in its existing Toxicology Management System (TMS). The following fields must be structured (coded, not free-text) in the PIH-to-NMC MVD feed to ensure the data are surveillance-ready. **Demographic variables are mandatory** — they are the primary tool for detecting vulnerable sub-populations (occupational exposure in farm workers, paediatric accidental ingestion, gender asymmetry in intentional self-harm) and for geographic clustering.

| Domain | Field | Format | Rationale |
|:-------|:------|:-------|:---------|
| **Demographics** | Age | Integer (years) or age-band (0–4, 5–14, 15–24, 25–34, 35–49, 50–64, ≥65) | Paediatric vs. adult exposure patterns differ; required for burden modelling |
| **Demographics** | Sex | M / F / Unknown | Gender asymmetry in intentional self-harm is a known signal |
| **Demographics** | Caller type | Patient / Clinician / Caregiver / Other | Distinguishes clinician-initiated (surveillance-relevant) from lay calls |
| **Demographics** | Occupation (if stated) | Free text + coded: farm worker / pesticide applicator / domestic worker / child / other / unknown | Occupational exposure is systematically under-identified |
| **Geography** | Province of caller | 2-letter code | Derived from caller area code or stated by caller |
| **Geography** | District (if known) | LIM-code or free text | Optional but captures rural/urban signal |
| **Exposure** | Pesticide type | NMC drop-down (see A1) | Must align with NMC classification for linkage |
| **Exposure** | AfriTox product name | Free text + drop-down lookup | Mandatory; "unknown" valid |
| **Exposure** | Route | Ingestion / Dermal / Inhalation / Ocular / Unknown | Route influences clinical management and surveillance signal |
| **Circumstances** | Intent | NMC circumstance codes (see A1) | Caller-reported; may be updated by treating clinician |
| **Clinical** | Severity (caller-assessed) | Mild / Moderate / Severe / Unknown | PSS (Poisoning Severity Score) if TMS supports it |
| **Clinical** | Referred to facility | Yes / No / Unknown | Links call record to potential NMC notification |
| **Linkage** | NMC notification number | Free text (if referred and notified) | Enables de-duplicated case linking across PIH + NMC streams |
| **System** | Call date and time | ISO 8601 (YYYY-MM-DDTHH:MM) | Required for temporal clustering and latency audit |
| **System** | TMS call ID | Alphanumeric | Internal identifier; retained for audit; not transmitted externally |

: PIH MVD — full field specification including demographic variables. Demographic fields (age, sex, caller type, occupation) are mandatory for all calls; a structured unknown code is required if data are not available — blank fields will be rejected in the NMC ingest pipeline. {#tbl-pih-mvd .striped .hover}

::: {.callout-note appearance="minimal"}
**POPIA compliance and secure transmission.** All PIH-to-NMC transmissions must be de-identified before transmission: no name, phone number, or exact address. Age-band (not exact age), province/district (not street address), and anonymised TMS call ID are sufficient for surveillance linkage and comply with Section 90 of the National Health Act (2003) and POPIA. Exact age may be retained internally within PIH TMS for clinical purposes and transmitted as age-band only.

De-identified health data still constitutes personal information under POPIA. The PIH-to-NMC feed **must be implemented as a secure HTTPS/TLS API with mutual authentication** (client certificate or token-based) between the PIH TMS and the NMC ingest endpoint. Transmission over email, FTP, or unencrypted HTTP is not acceptable. The MVD ingest pipeline (line item 4 in the S1 operating cost) includes the cost of maintaining this secure channel; the one-off setup cost (`c_S1_PIH_MVD_setup` = R40,000) must include API security configuration and penetration testing as a go-live requirement.
:::

::: {.callout-tip appearance="minimal"}
**AfriTox agent-name harmonisation.** Both PIH (TMS) and NMC (notification form) must use the same AfriTox product-name controlled vocabulary so that product-level attribution is consistent across streams. The harmonisation workshop (S1 MVD workshop, Months 1–3) should produce a shared AfriTox lookup table embedded in both systems. This is the lowest-cost, highest-value interoperability intervention in the MVD: it costs a single workshop and yields cross-stream agent attribution at scale.
:::
