---
title: "Policy Brief 1: Strengthening Pesticide Poisoning Surveillance"
subtitle: "Closing the detection gap between NMC, NHLS, and PIH"
author: "D2P Project Team"
date: "2025-09-05"
date-modified: today
categories: [policy, surveillance, decision-makers]
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: false
---

::: {.callout-important icon=false}
## For Decision-Makers

**Target audience:** Ministerial Advisory Committee (MAC), National Department of Health (NDoH), Provincial Health Leadership, NICD/NHLS leadership.

**Companion briefs:** [Coordination](coordination_policy.qmd) · [Terbufos restriction](terbufos_policy.qmd).
:::

## 1. Policy question

**Should NDoH invest in automated, multi-source pesticide-poisoning surveillance to close the detection gap between the NMC, NHLS laboratory, and PIH data streams?**

## 2. Public health problem

South Africa under-detects pesticide poisoning. Three data streams each see a fraction of the problem; none sees the whole:

- Most cases never appear in NMC.
- Severe laboratory results do not trigger a public-health alert.
- The country's largest single source of clinical signal — the Poisons Information Helpline — is not used for surveillance at all.

Cases from contaminated food, illegal sales, or unsafe occupational exposure are detected late or not at all; agencies have no real-time view of which products are causing harm; and the country cannot target regulatory or public-health response to where harm is concentrating.

## 3. Status quo

{{< include _shared_parameters.qmd >}}

::: {.callout-note}
## Health-sector share of the burden — what surveillance unlocks

The public health system bears **~R80M of the R2.34B annual burden directly on its own budget lines** — ~R70M to NDoH (UPFS ward + ICU days, ventilation, atropine + PAM) and ~R10M to NHLS (BChE, GC-MS confirmatory volume). Better surveillance is a **necessary but not sufficient** input to reducing that burden: surveillance changes only what is *seen*; the burden reduction itself comes from the response side — source removal, regulatory action, clinical management — covered in the [Coordination brief](coordination_policy.qmd) and the [Terbufos brief](terbufos_policy.qmd). **This brief therefore quantifies P_detect uplift only and does not claim cases or deaths averted.**
:::

**Architecturally, the three streams are unlinked:**

```{mermaid}
%%| label: fig-surv-current-system
%%| fig-cap: "Current surveillance architecture. NMC is NDoH-owned and is the source of truth. NHLS BChE results now route to NMC via NHLS CDW → NICD SDW (MOU feed, now operational). Two active gaps remain: PIH consultations have no structured NMC feed, and chemical toxicology is post-mortem only."
flowchart LR
    A[Healthcare Facility<br/>Clinician / PHO] -->|NMC notification| E[("NDoH / NMC<br/><b>NDoH-owned</b><br/>Source of truth")]
    A --> NHLS

    subgraph NHLS[NHLS]
        direction TB
        C1[BChE Assay<br/>1,779 severe/year<br/>10,626 tests 2023<br/><i>Live patients</i>]
        C2[Chemical Toxicology<br/><i>Post-mortem only —<br/>not yet done on<br/>live patients</i>]
        CDW[(NHLS CDW)]
        C1 -->|monthly extract| CDW
    end

    SDW[(NICD SDW)]
    CDW -->|"✓ MOU feed<br/>operational"| SDW
    SDW -->|BChE monthly| E

    A --> D[PIH Consultation<br/>~1,158 clinician calls/yr<br/><i>No minimum dataset</i>]

    C2 -.x Not available for<br/>live-patient surveillance.-x E
    D -.No minimum viable<br/>dataset defined.-> E

    style E fill:#9f9,stroke:#060,stroke-width:2px
    style A fill:#fde,stroke:#333
    style C1 fill:#bbf,stroke:#333
    style C2 fill:#ffd9b3,stroke:#cc8800,stroke-dasharray: 5 5
    style D fill:#bfb,stroke:#333
    style CDW fill:#e6f0ff,stroke:#3366cc
    style SDW fill:#e6f0ff,stroke:#3366cc
```

**What this means in plain numbers.**

- NMC notifications: 1,013 ([`n_nmc_annual`](../parameters_hub.qmd#tbl-burden))
- NHLS severe BChE inhibitions: 1,779 ([`n_nhls_severe_annual`](../parameters_hub.qmd#tbl-burden)) out of 10,626 BChE tests ([`n_nhls_total_annual`](../parameters_hub.qmd#tbl-burden))
- PIH clinician-initiated calls: 1,158 ([`n_pih_annual`](../parameters_hub.qmd#tbl-burden))
- Implied true annual burden: $N_\text{true} = \frac{2,620}{0.125} = 20,960$ ([`n_true_burden_from_statssa_mid`](../parameters_hub.qmd#tbl-burden))

NMC alone captures $P_\text{detect} = \frac{1,013}{20,960} = 0.048$ ([`d_status_quo`](../parameters_hub.qmd#tbl-detection)), i.e. **< 5 %**.

::: {.callout-caution appearance="minimal" collapse="true"}
## Caveat: this denominator rests on a CFR assumption
The 20,960 figure is derived by applying a 12.5% case-fatality rate to the StatsSA MACOD death count — but that CFR is drawn from NMC-*notified*, clinically severe cases, which are by definition the sicker end of the spectrum. The undetected ~95% are likely milder, with a lower blended CFR. If the true blended CFR is 3–5%, the implied burden is 52,000–87,000 cases and the status-quo P_detect is 1–2%, not 4.8%. The detection gap would be simultaneously **larger** and harder to fully close. This caveat does not change the **relative rankings** of S1, S2, and MVP — those are stable across the plausible CFR range — but it does affect the absolute scale of the gap. The mid-burden estimate is used as the working denominator throughout this brief for option comparison; it should not be read as a precise point estimate of true annual incidence.
:::

### Detection vs. true burden — a critical distinction

Better surveillance has **two opposite numerical effects** that decision-makers must not confuse:

| Effect | Direction | Driver | Interpretation |
|:-------|:---------:|:-------|:---------------|
| **Observed** burden | ↑ | Better case-finding (auto-notify, MVD ingest) | The headline "case count" will rise as detection improves — this is success, not failure. |
| **True** burden | ↓ | Faster source removal, targeted regulation, prevention | Number of actual poisonings falls as cases are detected and acted on earlier. |

A naive reading of post-pilot data will show "more cases" and conclude the system is getting worse. The correct reading is to compare **completeness-adjusted true incidence** against the baseline, using laboratory and PIH denominators as the closest proxy. This needs to be agreed up front with NDoH and the MAC.

## 4. Root cause analysis

```{mermaid}
%%| label: fig-surv-fishbone
%%| fig-cap: "Root causes of surveillance under-detection and the surveillance-improvement options that address each one."
flowchart LR
    A[Surveillance under-detection] --> B[NHLS–NMC MOU not<br/>IT-implemented]
    A --> C[PIH data not used<br/>for surveillance]
    A --> D[Delayed manual reporting]
    A --> F[No live-patient<br/>chemical toxicology]

    B --> G["✓ NHLS→NMC IT feed<br/><i>Completed</i>"]
    C --> H[Option S1: PIH MVD<br/>+ dashboard]
    D --> I[Option S2: BChE<br/>auto-notify <i>capability exists</i>]
    F --> J[Option S3: Sentinel<br/>toxicology + auto-notify]

    style A fill:#f99,stroke:#333
    style G fill:#e0e0e0,stroke:#999,stroke-dasharray: 5 5
    style H fill:#9f9,stroke:#333
    style I fill:#9f9,stroke:#333
    style J fill:#9f9,stroke:#333
```

Three active root causes remain **executional, not legislative**, and are each already legally permissible under Section 90 of the National Health Act (2003) and the existing NMC regulations. The fourth root cause (NHLS–NMC IT connectivity) has been resolved — the LIS-to-NMC feed is now operational.

## 5. Policy options

::: {.callout-tip icon=false}
## ✓ NHLS–NMC IT feed — now operational

The NHLS–NICD LIS-to-NMC connector is operational; monthly BChE data flows to NICD under the existing MOU. This root-cause is addressed. The three active options below build on that foundation.
:::

### Option S1 — PIH live dashboard + Minimum Viable Dataset *(★ recommended)*

Two layers:

- **Short term (1–2 weeks):** PIH publishes a live dashboard of pesticide-related calls (counts, geography, agent, severity, demographics, **AfriTox agent name**). No new data; just a daily extract from existing call records.
- **Medium term (3–9 months):** PIH and NICD agree a **Minimum Viable Dataset (MVD)** so the ~1,158 clinician-initiated calls/year can be ingested into NMC as a structured surveillance stream. **Agent-name capture is a mandatory MVD field** — both PIH and NMC must record the AfriTox product name for every case. To make this reciprocal, **NMC should add a structured agent drop-down** (pesticide class → active ingredient → AfriTox product name) to its NDoH-hosted notification form: PIH records what the caller *reports*; NMC records what the clinician *confirms*. Combined, they provide agent-attribution signal across severity strata — the same signal S3 sentinel toxicology provides at tertiary sites, but available system-wide at near-zero marginal cost.

Detection rises to ~`r sprintf("%.0f", as.numeric(pick("d_S1_PIH_MVD")) * 100)` % of true cases (up from ~`r sprintf("%.0f", as.numeric(pick("d_status_quo")) * 100)` %) for **`r zar(as.numeric(pick("c_S1_PIH_MVD_op")))`/year** (includes AfriTox agent-name harmonization). The dashboard is the lowest-hanging fruit in the whole portfolio: it ships in days.

### Option S1b — Phase 2 community layer *(commission after MVP pilot if rural/informal gap confirms)*

S1 and S2 improve detection within the **formal healthcare system** — via PIH, NHLS BChE, and NMC-notified facilities. They structurally cannot reach cases that never arrive at a formal facility: the child taken to a traditional healer after ingesting street-sold granules, the farm worker who doesn't present at all, the informal-settlement cluster whose care pathway bypasses formal services entirely. These cases are invisible to S1 and S2 regardless of how well those options are implemented.

**The MVP ceiling (~`r sprintf("%.0f", as.numeric(pick("d_surveillance_mvp")) * 100)`% P_detect) is real and bounded** by the reach of formal administrative streams. S1b is the mechanism for going beyond it — but only after the MVP has established the baseline and confirmed that the remaining gap is structural rather than simply a compliance issue within the formal system.

**Trigger condition.** Commission S1b if, after 6 months of MVP operation, the geographic distribution of PIH + NHLS + NMC cases shows ≥40% of severe cases originating in districts with low formal-facility access (defined as <1 public hospital per 50km radius). This is a data-driven go/no-go, not a discretionary expansion. CSV parameters: `d_S1b_event_community_*`, `c_S1b_setup_*`, `c_S1b_op_*`, `p_fp_S1b_*`.

**False positives are a design parameter, not a disqualifier.** Community reporters without clinical training will flag events that turn out not to be pesticide poisonings. The gross false-positive rate is ~`r sprintf("%.0f%%", as.numeric(pick("p_fp_S1b_mid")) * 100)` (range `r sprintf("%.0f–%.0f%%", as.numeric(pick("p_fp_S1b_lo")) * 100, as.numeric(pick("p_fp_S1b_hi")) * 100)`). At P_detect = `r pick("d_S1b_event_community_mid")` on the mid-burden of `r scales::comma(as.numeric(pick("n_true_burden_from_statssa_mid")))` true cases, the gross notification volume would be ~`r scales::comma(round(as.numeric(pick("n_true_burden_from_statssa_mid")) * as.numeric(pick("d_S1b_event_community_mid")) / (1 - as.numeric(pick("p_fp_S1b_mid")))))` events/yr — of which ~`r scales::comma(round(as.numeric(pick("n_true_burden_from_statssa_mid")) * as.numeric(pick("d_S1b_event_community_mid"))))` are true cases and ~`r scales::comma(round(as.numeric(pick("n_true_burden_from_statssa_mid")) * as.numeric(pick("d_S1b_event_community_mid")) * as.numeric(pick("p_fp_S1b_mid")) / (1 - as.numeric(pick("p_fp_S1b_mid")))))` are false alerts *before filtering*. This is solvable with a **two-tier triage architecture** — the same design used by WHO-EWARN, APSED, and comparable LMIC systems:

1. **Tier 1 — CHW structured checklist**: Community health workers report via a mobile checklist requiring ≥2 of: acute onset, ≥2 affected persons, known or suspected product contact. Events not meeting criteria are logged but no dispatch is triggered.
2. **Tier 2 — EHP call-back within 24 hours**: A provincial EHP calls back to confirm before committing to on-site investigation.

With a two-tier filter, the effective net FP rate drops to approximately 8–12% — within the investigation capacity modelled in the [Coordination brief](coordination_policy.qmd). The coordination cost implication is therefore driven by the *net* rate after triage, not the gross `r sprintf("%.0f%%", as.numeric(pick("p_fp_S1b_mid")) * 100)`.

**Two-stage rollout — Free State and Gauteng first.** These two provinces account for an estimated **~50% of the annual pesticide-poisoning burden** (working assumption; to be confirmed against the MVP pilot's geographic distribution from NMC, NHLS, and PIH data). Starting there maximises the return on the pilot investment and generates the go/no-go evidence base for national rollout. Comparable LMIC systems have achieved P_detect in the range `r pick("d_S1b_event_community_lo")`–`r pick("d_S1b_event_community_hi")` — `r sprintf("%.1f–%.1f×", as.numeric(pick("d_S1b_event_community_lo")) / as.numeric(pick("d_surveillance_mvp")), as.numeric(pick("d_S1b_event_community_hi")) / as.numeric(pick("d_surveillance_mvp")))` the MVP ceiling.

| Stage | Provinces | Est. burden share | Setup cost | Operating cost/yr | Go/no-go criterion |
|:------|:----------|------------------:|-----------:|------------------:|:-------------------|
| **S1b pilot** | Free State + Gauteng | ~50% | ~`r sprintf("R%.0fM", as.numeric(pick("c_S1b_setup_lo"))/1e6)` | ~`r sprintf("R%.1fM/yr", as.numeric(pick("c_S1b_op_lo"))/1e6)` | Month 12: net P_detect >35% and net FP <15% |
| **S1b national** | All 9 provinces | 100% | ~`r sprintf("R%.1fM", as.numeric(pick("c_S1b_setup_hi"))/1e6)` | ~`r sprintf("R%.1fM/yr", as.numeric(pick("c_S1b_op_hi"))/1e6)` | Conditional on pilot outcome |

The cost range (`r sprintf("R%.0fM–R%.1fM setup; R%.1fM–R%.1fM/yr operating", as.numeric(pick("c_S1b_setup_lo"))/1e6, as.numeric(pick("c_S1b_setup_hi"))/1e6, as.numeric(pick("c_S1b_op_lo"))/1e6, as.numeric(pick("c_S1b_op_hi"))/1e6)` — approximately `r sprintf("%.0f–%.0f×", as.numeric(pick("c_S1b_op_lo")) / as.numeric(pick("c_surv_MVP_op")), as.numeric(pick("c_S1b_op_hi")) / as.numeric(pick("c_surv_MVP_op")))` the MVP operating cost) directly reflects the two-stage choice.

### Option S2 — BChE auto-notification *(★ recommended — capability exists)*

Configure the NHLS LIS to fire an automatic NICD notification when a BChE result shows severe inhibition (< 50 % of normal), within 24 hours of validation. **No new test, no new reagent contract** — only an IT configuration on top of the ~`r scales::comma(as.numeric(pick("n_nhls_total_annual")))` BChE tests/year already performed (`r scales::comma(as.numeric(pick("n_nhls_severe_annual")))` severe in 2023). Detection rises to ~`r sprintf("%.0f", as.numeric(pick("d_S2_BChE_autonotify")) * 100)` % of true cases (up from ~`r sprintf("%.0f", as.numeric(pick("d_status_quo")) * 100)` %) for **`r zar(as.numeric(pick("c_S2_BChE_autonotify_op")))`/year** in 2–4 weeks. This is the highest-value, fastest option on the menu.

### Option S3 — Sentinel live-patient toxicology *(★ recommended, medium term — distinct purpose)*

NHLS currently performs chemical toxicology (qualitative/quantitative pesticide assays) **post-mortem only**. There is no live-patient toxicology capability. Option S3 funds a sentinel tertiary site (e.g. CHB) to perform live-patient chemical toxicology, scaling to 4–5 sites in phase 2–3.

**S3 has a different purpose from S1 and S2** and is *not* a detection-coverage uplift. S1 and S2 increase the share of true cases that enter the surveillance system. **S3 identifies the proportion of chemical agents being found in severely poisoned living patients at sentinel sites** — the missing evidence base for product-specific regulation (see [Terbufos brief](terbufos_policy.qmd)). Without S3, ~10,000 survivors/year are recorded with agent "unknown / suspected" and targeted product restriction has no quantitative footing.

Cost: ~`r sprintf("R%.1fM", round(as.numeric(pick("c_option6_pilot_year1"))/1e6, 1))` pilot site → ~`r sprintf("R%.1fM", round(as.numeric(pick("c_option6_full_5site_annual"))/1e6, 1))` for 5-site rollout, 6–12 months. Funded from research-surveillance streams (SAMRC, Bloomberg, Gates), not the surveillance operating budget. Full quantitative concept in the [sentinel toxicology appendix](surveillance_concept_note.qmd).

### Summary comparison


| Option | Annual cost | P_detect (per-case) | Δ vs. status quo | Time | Parameter |
|:-------|------------:|---------------------:|----------------:|:-----|:----------|
| Status quo | R0 | $\frac{1,013}{20,960} = 0.048$ | — | — | [`d_status_quo`](../parameters_hub.qmd#tbl-detection) |
| S1: PIH dashboard + MVD (+AfriTox agent name) | R82k | $\frac{1,013 + 1,158}{20,960} = 0.104$ | +5.6 pp | 1–2 wks dash / 3–9 mo MVD | [`d_S1_PIH_MVD`](../parameters_hub.qmd#tbl-detection) |
| S2: BChE auto-notify | R55k | $\frac{1,013 + 1,779}{20,960} = 0.133$ | +8.5 pp | 2–4 wks | [`d_S2_BChE_autonotify`](../parameters_hub.qmd#tbl-detection) |
| **S1 + S2 (Surveillance MVP)** | **~R137k** | $\frac{1,013 + 1,779 + 1,158}{20,960} = 0.189$ | **+14.1 pp** | **12 weeks** | [`d_surveillance_mvp`](../parameters_hub.qmd#tbl-detection) |
| S1b: Phase 2 community layer *(pilot: FS + Gauteng; national conditional on pilot)* | Pilot `r sprintf("R%.1fM/yr", as.numeric(pick("c_S1b_op_lo"))/1e6)`; national `r sprintf("R%.1fM/yr", as.numeric(pick("c_S1b_op_hi"))/1e6)` | `r pick("d_S1b_event_community_lo")`–`r pick("d_S1b_event_community_hi")` (working: `r pick("d_S1b_event_community_mid")`) | `r sprintf("+%.0f to +%.0f pp", (as.numeric(pick("d_S1b_event_community_lo")) - as.numeric(pick("d_status_quo"))) * 100, (as.numeric(pick("d_S1b_event_community_hi")) - as.numeric(pick("d_status_quo"))) * 100)` | 12 mo pilot → national | [`d_S1b_event_community_mid`](../parameters_hub.qmd) |
| S3: Sentinel toxicology (agent attribution — distinct purpose) | R1.2M pilot → R3.6M | n/a (not a coverage uplift) | feeds product-banning evidence base | 6–12 mo pilot | — |

: Surveillance options — base-case. P_detect is the per-case detection probability, anchored to the StatsSA MACOD 2023 mid-burden of 20,960 true cases/year. S1 and S2 expand coverage within the formal healthcare system; **S3 is a parallel agent-attribution investment**, not a coverage uplift — see option text above. **S1b is a Phase 2 community layer**, commissioned conditional on MVP pilot data confirming a structural rural/informal gap (trigger: ≥40% of severe cases in low-facility-access districts after 6 months of MVP operation). Translating P_detect into cases or deaths averted requires the response-side parameters (η, agency capacity, regulatory action) covered in the [Coordination](coordination_policy.qmd) and [Terbufos](terbufos_policy.qmd) briefs. {.striped .hover}

::: {.callout-note appearance="simple"}
**Scope of this brief.** Surveillance changes only what is *seen*. The status quo P_detect = 0.048 (§6 below) is a **measurement** of the current detection gap, not a counterfactual against which deaths are averted. Cases-averted and deaths-averted figures therefore live in the [Coordination brief](coordination_policy.qmd) (where the response is parameterised) and the [Terbufos brief](terbufos_policy.qmd) (where regulatory action is parameterised). This separation keeps the surveillance pitch honest: better surveillance is **necessary but not sufficient**.
:::

::: {.callout-note}
## Audit: parameter names for costing each policy option

Every annual cost and setup cost in this brief is directly linked to a parameter in [`amua_import_parameters_v4.csv`](../amua_import_parameters_v4.csv). These are the exact parameter names used in the model and costing tables:

| Option | Setup cost parameter | Operating cost parameter |
|:-------|:--------------------|:------------------------|
| S1 — PIH dashboard + MVD (+ AfriTox harmonisation) | `c_S1_PIH_MVD_setup` | `c_S1_PIH_MVD_op` |
| S2 — BChE auto-notification | `c_S2_BChE_autonotify_setup` | `c_S2_BChE_autonotify_op` |
| Surveillance MVP (S1 + S2) | `c_surv_MVP_setup` | `c_surv_MVP_op` |
| S1b — event-based / community *(sensitivity)* | `c_S1b_setup_mid` (lo/hi also exist) | `c_S1b_op_mid` (lo/hi also exist) |
| S0 — NHLS-NMC IT feed (now operational) | `c_S0_NHLS_NMC_feed_setup` | `c_S0_NHLS_NMC_feed_op` |
| Status quo (NICD program) | — | `c_status_quo_program` |

**How to audit:**
- Each parameter above is a row in the CSV, with value, distribution, and source citation.
- The same parameter names are used in the decision tree (see §6) and the marginal cost table (§7, Tab 5).
- To update a cost, edit the value in the CSV and re-render — all outputs update automatically.
:::

## 6. Decision tree — from surveillance investment to detection uplift {#sec-decision-tree}

::: {.callout-note collapse="true"}
## Audit trail — how every P_detect value in this brief was derived

All active P_detect values are **derived, not elicited**, on a single per-case scale. The arithmetic is auditable from three published anchors.

### Step 1 — the implied true annual burden (denominator)

| Anchor | Value | Source |
|--|--:|--|
| Annual NMC pesticide notifications | 1,013 | `n_nmc_annual` — NICD NMC Annual Report 2023 |
| NMC-hospitalised case-fatality rate | 12.5 % | `cfr_headline_mid` (range 10–15 %) — Free State 2018-19 + NMC observed |
| StatsSA MACOD pesticide-poisoning deaths/yr | 2,620 | `n_deaths_statssa_macod` — StatsSA Causes of Death, 2023: Table 4.12: Accidental poisoning by and exposure to noxious substance (X40-X49) |

If notified cases have CFR ≈ 12.5 %, then the *full* true burden satisfies $\text{true cases} \times \text{CFR} = \text{deaths}$, so

$$
N_\text{true} \;=\; \frac{2{,}620}{0.125} \;=\; \mathbf{20{,}960}\ \text{cases/year} \quad (\texttt{n\_true\_burden\_from\_statssa\_mid})
$$

This is the **single denominator** used for every per-case P_detect below.

### Step 2 — the per-stream numerators (2023 actuals)

| Stream | Annual count | CSV row | Source |
|--|--:|--|--|
| NMC notifications | 1,013 | `n_nmc_annual` | NICD NMC Annual Report 2023 |
| NHLS severe BChE inhibitions (<50 % normal) | 1,779 | `n_nhls_severe_annual` | NHLS LIS 2023 (of 10,626 total tests) |
| PIH clinician-initiated calls | 1,158 | `n_pih_annual` | PIH (Tygerberg) 2023 |

The three streams are treated as **mutually exclusive** in the base case (severe BChE ≠ NMC-notified ≠ PIH-called); the overlap-correction sensitivity is parameterised separately in the hub.

### Step 3 — P_detect for each option

Each option adds one or more streams to the NMC base. The numerator is the sum of streams reaching the surveillance system; the denominator is always 20,960.

| Option | Streams summed | Arithmetic | P_detect | CSV row |
|--|--|--:|--:|--|
| Status quo | NMC only | $1{,}013 / 20{,}960$ | **0.048** | `d_status_quo` |
| S1 — PIH MVD + dashboard | NMC + PIH | $(1{,}013 + 1{,}158) / 20{,}960 = 2{,}171 / 20{,}960$ | **0.104** | `d_S1_PIH_MVD` |
| S2 — BChE auto-notify | NMC + NHLS severe | $(1{,}013 + 1{,}779) / 20{,}960 = 2{,}792 / 20{,}960$ | **0.133** | `d_S2_BChE_autonotify` |
| **Surveillance MVP (S1 + S2)** | NMC + NHLS severe + PIH | $(1{,}013 + 1{,}779 + 1{,}158) / 20{,}960 = 3{,}950 / 20{,}960$ | **0.189** | `d_surveillance_mvp` |

### Step 4 — CFR sensitivity bracket on the status-quo anchor

The denominator scales inversely with the CFR assumed for NMC-notified cases (CSV: `d_status_quo_lo`, `d_status_quo_hi`):

- Lower bound (CFR = 0.10): $P_\text{detect}^{SQ} = 1{,}013 \times 0.10 / 2{,}620 = 0.039$ (implied burden 26,200)
- Mid (CFR = 0.125): $P_\text{detect}^{SQ} = 0.048$ (implied burden 20,960)
- Upper (CFR = 0.15): $P_\text{detect}^{SQ} = 1{,}013 \times 0.15 / 2{,}620 = 0.058$ (implied burden 17,467)

The same CFR bracket applies proportionally to S1/S2/MVP — e.g. the MVP varies from 0.151 (CFR 0.10) to 0.226 (CFR 0.15).

### Step 5 — why this replaces the v1 value of 0.45

The legacy v1 figure was an expert-elicited estimate on a different scale, derived from a model of spatial event sensitivity rather than per-case detection. That quantity answered a different question and is not comparable to a per-case detection probability. This brief now uses the **per-case** framing throughout — every value above is directly auditable from NICD + NHLS + PIH + StatsSA published figures and the canonical [`amua_import_parameters_v4.csv`](../amua_import_parameters_v4.csv).
:::

````{r}
#| label: fig-surv-decision-tree
#| echo: false
#| output: asis

# Source the decision-tree engine and the surveillance tree builder.
# Both scripts auto-resolve their paths whether rendered from the project
# root or from posts/. All numbers below come from the rollback of the tree
# object built off amua_import_parameters_v4.csv -- no inline arithmetic.
source("../R/surveillance_tree.r")

params   <- load_params()                          # named list from CSV
tree     <- build_surveillance_tree(params)        # dt_node object
res      <- evs_to_df(eval_tree(tree))             # per-case rollback (tibble)
annual   <- scale_surveillance_to_annual(res, params$n_true_burden_from_statssa_mid)

# Pull arm-level values out of the rolled-back tibble. The arm names match
# those defined in build_surveillance_tree() in R/surveillance_tree.r.
arm <- function(a) annual %>% filter(arm == a)
d_sq  <- arm("status_quo")$detected
d_s1  <- arm("S1_PIH_MVD")$detected
d_s2  <- arm("S2_BChE_autonotify")$detected
d_mvp <- arm("MVP_S1_plus_S2")$detected

c_s1   <- arm("S1_PIH_MVD")$annual_cost_zar
c_s2   <- arm("S2_BChE_autonotify")$annual_cost_zar
c_mvp  <- arm("MVP_S1_plus_S2")$annual_cost_zar

pp <- function(d) sprintf("%+.1f pp", (d - d_sq) * 100)

cat(sprintf('```{mermaid}
%%%%| fig-cap: "Surveillance investment decision tree — per-case detection uplift only. Each option raises P_detect, the share of true annual cases (~20,960 StatsSA-anchored mid-burden) that enter the surveillance system. The absolute burden impact of these uplifts depends on response-side parameters covered in the coordination and terbufos briefs. S3 (sentinel toxicology) is shown separately because it is an agent-attribution investment, not a coverage uplift. All numerical values pulled from amua_import_parameters_v4.csv at render time."
flowchart TD
    INV{Surveillance<br/>investment?} -->|None| SQ[Status quo<br/>P_detect = %.3f<br/>~5%% of true cases]:::bad
    INV -->|S1 only<br/>R%s/yr| O1[+ PIH MVD<br/>P_detect = %.3f<br/>%s vs SQ]:::warn
    INV -->|S2 only<br/>R%s/yr| O2[+ BChE auto-notify<br/>P_detect = %.3f<br/>%s vs SQ]:::warn
    INV -->|S1 + S2<br/>R%s/yr| MVP[Surveillance MVP<br/>P_detect = %.3f<br/>%s vs SQ]:::good
    S3[S3 Sentinel toxicology<br/>agent attribution<br/><i>parallel investment —<br/>see Terbufos brief</i>]:::aux

    classDef bad fill:#fdd,stroke:#cc0000,stroke-width:1.5px
    classDef warn fill:#ffe9b3,stroke:#cc8800,stroke-width:1.5px
    classDef good fill:#cfc,stroke:#060,stroke-width:1.5px
    classDef aux fill:#e6f0ff,stroke:#3366cc,stroke-width:1.5px,stroke-dasharray: 5 5
```
',
d_sq,
format(c_s1,  big.mark = ","), d_s1,  pp(d_s1),
format(c_s2,  big.mark = ","), d_s2,  pp(d_s2),
format(c_mvp, big.mark = ","), d_mvp, pp(d_mvp)
))
````

The tree reports **per-case detection uplift in percentage points**. The status-quo node is a **measurement** of the current detection gap (1,013 NMC notifications ÷ ~20,960 implied true cases), not a baseline from which deaths are counted as averted — surveillance has no response attached. Four detection inputs (`d_status_quo`, `d_S2_BChE_autonotify`, `d_S1_PIH_MVD`, `d_surveillance_mvp`) and three operating-cost rows (`c_S1_PIH_MVD_op`, `c_S2_BChE_autonotify_op`, `c_surv_MVP_op`) live in [`amua_import_parameters_v4.csv`](../amua_import_parameters_v4.csv) — edit the CSV and re-render. Option S3 sits outside this tree: it is an agent-attribution investment whose value is parameterised in the [Terbufos brief](terbufos_policy.qmd) and the [sentinel toxicology appendix](surveillance_concept_note.qmd).

::: {.callout-note appearance="minimal"}
**S1b (Phase 2 community layer) is not modelled in this decision tree.** The MVP arm represents the ceiling achievable by linking existing administrative streams. S1b would require a new arm with P_detect drawn from `d_S1b_event_community_mid` = `r pick("d_S1b_event_community_mid")` (range `r pick("d_S1b_event_community_lo")`–`r pick("d_S1b_event_community_hi")`) and annual operating cost drawn from `c_S1b_op_mid` = `r scales::comma(as.numeric(pick("c_S1b_op_mid")))`. It is excluded from the current tree because it is a Phase 2 option commissioned conditional on MVP pilot results, not a parallel near-term decision; and because its coordination cost depends on the net false-positive rate after two-tier triage (`p_fp_S1b_mid` = `r pick("p_fp_S1b_mid")` gross; ~8–12% net) modelled in the [Coordination brief](coordination_policy.qmd).
:::

**For burden impact, see the next brief.** The same P_detect figures feed into the [Coordination brief](coordination_policy.qmd) decision tree, where response efficacy η converts detection into cases averted, and into the [Terbufos brief](terbufos_policy.qmd), where regulatory restriction converts cases into deaths averted.

## 7. Status-quo costing — what surveillance costs today {#sec-costing}

::: {.callout-note icon=false}
## How to read & edit this section

The tables below are the **average-cost economic build-up of running surveillance for pesticide poisoning today** (status quo, 2024–25 ZAR). They are sourced from an editable workbook — [`data/surveillance_costing.xlsx`](../data/surveillance_costing.xlsx) — with three sheets:

- **`personnel`** — staff time (CTE × FTE %) by stream
- **`system`** — infrastructure, hosting, SMS, reagents by stream
- **`allocation`** — what share of each stream's total volume is pesticide-related (e.g. ~100 pesticide notifications/month out of ~10,000 total NMC notifications ≈ **0.84 %** of NMC platform OPEX is attributable to pesticide work)

To revise a unit cost or FTE, **edit a cell in the .xlsx and re-render** — every number below updates. The CSV (`amua_import_parameters_v4.csv`) remains the source of truth for *model* parameters and *marginal* policy-option costs (S1 / S2 / MVP). Salary anchors come from the DPSA scales (L10–11), NHLS Patterson grades (D/E), DoH OSD scales (MO, pharmacy), NICD recruitment notices and SITA contracting benchmarks; system anchors come from NHLS State Tender RT-262, AWS af-south-1 pricing, Clickatell/BulkSMS SA, and the NICD ICT budget — full citations are in each row's `source` column.
:::

```{r}
#| label: load-costing
#| message: false
#| warning: false

library(readxl)
library(scales)
library(gt)
library(dplyr)

cost_path <- "../data/surveillance_costing.xlsx"
personnel  <- read_excel(cost_path, sheet = "personnel")
system_dat <- read_excel(cost_path, sheet = "system")
allocation <- read_excel(cost_path, sheet = "allocation")

# Allocation factor: share of stream activity that is pesticide-related.
alloc_factor <- function(stream_name) {
  a <- allocation %>% filter(stream == stream_name)
  if (nrow(a) == 0) return(1)
  a$allocation_pct[1] / 100
}

# Apply pesticide-share allocation to platform-level OPEX rows (rationale
# contains the literal phrase "pesticide share").
system_alloc <- system_dat %>%
  mutate(
    is_platform = grepl("pesticide share", rationale, ignore.case = TRUE),
    pesticide_share_pct = if_else(
      is_platform,
      vapply(stream, function(s) alloc_factor(s) * 100, numeric(1)),
      100
    ),
    pesticide_cost_zar = round(annual_cost_zar * pesticide_share_pct / 100)
  )

zar <- function(x) paste0("R", scales::comma(round(x)))
```

::: {.panel-tabset}

### 1. Personnel

Annual cost-to-employer × FTE share attributable to pesticide surveillance, by stream.

```{r}
#| label: tbl-cost-personnel

personnel %>%
  arrange(stream, desc(annual_cost_zar)) %>%
  select(stream, role, annual_cte_zar, fte_pct, annual_cost_zar, rationale, source) %>%
  gt(rowname_col = "role", groupname_col = "stream") %>%
  fmt_number(columns = c(annual_cte_zar, annual_cost_zar), decimals = 0) %>%
  fmt_number(columns = fte_pct, decimals = 1, suffixing = FALSE, pattern = "{x}%") %>%
  cols_label(
    annual_cte_zar  = "Annual CTE (R)",
    fte_pct         = "FTE %",
    annual_cost_zar = "Allocated cost (R/yr)",
    rationale       = "Rationale",
    source          = "Source"
  ) %>%
  grand_summary_rows(
    columns = annual_cost_zar,
    fns = list("Total personnel" = ~ sum(.)),
    fmt = list(~ fmt_number(., decimals = 0))
  ) %>%
  tab_options(table.font.size = 11)
```

### 2. System

Infrastructure, hosting, SMS, reagents and licences — unit cost × annual volume, then allocated to the pesticide share of stream activity where applicable.

```{r}
#| label: tbl-cost-system

system_alloc %>%
  arrange(stream, desc(pesticide_cost_zar)) %>%
  select(stream, item, unit_cost_zar, units_per_yr, annual_cost_zar,
         pesticide_share_pct, pesticide_cost_zar, rationale, source) %>%
  gt(rowname_col = "item", groupname_col = "stream") %>%
  fmt_number(columns = c(unit_cost_zar, units_per_yr,
                         annual_cost_zar, pesticide_cost_zar),
             decimals = 0) %>%
  fmt_number(columns = pesticide_share_pct, decimals = 1, pattern = "{x}%") %>%
  cols_label(
    unit_cost_zar       = "Unit cost (R)",
    units_per_yr        = "Units/yr",
    annual_cost_zar     = "Raw cost (R/yr)",
    pesticide_share_pct = "Pesticide share",
    pesticide_cost_zar  = "Allocated cost (R/yr)",
    rationale           = "Rationale",
    source              = "Source"
  ) %>%
  grand_summary_rows(
    columns = pesticide_cost_zar,
    fns = list("Total system (pesticide-allocated)" = ~ sum(.)),
    fmt = list(~ fmt_number(., decimals = 0))
  ) %>%
  tab_options(table.font.size = 11)
```

### 3. Allocation factors

Why these proportions matter: platform-level OPEX (NMC hosting, PIH telephony, AfriTox licence) covers *all* notifiable / call activity, of which pesticide poisoning is only a slice. The denominator drives the allocated cost.

```{r}
#| label: tbl-cost-allocation

allocation %>%
  select(stream, basis, pesticide_count_yr, total_count_yr,
         allocation_pct, rationale, source) %>%
  gt(rowname_col = "basis", groupname_col = "stream") %>%
  fmt_number(columns = c(pesticide_count_yr, total_count_yr), decimals = 0) %>%
  fmt_number(columns = allocation_pct, decimals = 2, pattern = "{x}%") %>%
  cols_label(
    pesticide_count_yr = "Pesticide (n/yr)",
    total_count_yr     = "Total (n/yr)",
    allocation_pct     = "Share",
    rationale          = "Rationale",
    source             = "Source"
  ) %>%
  tab_options(table.font.size = 11)
```

### 4. Status-quo total

Status-quo annual cost of running pesticide-poisoning surveillance across the three streams, summed from the personnel and system sheets.

```{r}
#| label: tbl-cost-total

pers_by_stream <- personnel %>%
  group_by(stream) %>%
  summarise(personnel_zar = sum(annual_cost_zar), .groups = "drop")

sys_by_stream <- system_alloc %>%
  group_by(stream) %>%
  summarise(system_zar = sum(pesticide_cost_zar), .groups = "drop")

totals <- full_join(pers_by_stream, sys_by_stream, by = "stream") %>%
  mutate(across(c(personnel_zar, system_zar), ~ replace_na(., 0)),
         total_zar = personnel_zar + system_zar) %>%
  arrange(desc(total_zar))

totals %>%
  gt(rowname_col = "stream") %>%
  fmt_number(columns = c(personnel_zar, system_zar, total_zar), decimals = 0) %>%
  cols_label(
    personnel_zar = "Personnel (R/yr)",
    system_zar    = "System (R/yr)",
    total_zar     = "Total (R/yr)"
  ) %>%
  grand_summary_rows(
    columns = c(personnel_zar, system_zar, total_zar),
    fns = list("Status-quo total" = ~ sum(.)),
    fmt = list(~ fmt_number(., decimals = 0))
  ) %>%
  tab_options(table.font.size = 11)
```

### 5. Marginal cost of policy options

For comparison: the **additional** annual operating cost of each surveillance investment, pulled directly from [`amua_import_parameters_v4.csv`](../amua_import_parameters_v4.csv) (these are the same numbers driving the decision tree in §6).

```{r}
#| label: tbl-cost-marginal

marginal <- tibble::tribble(
  ~option, ~setup_param, ~op_param,
  "S1 — PIH dashboard + MVD (+ AfriTox harmonisation)", "c_S1_PIH_MVD_setup",           "c_S1_PIH_MVD_op",
  "S2 — BChE auto-notification",                         "c_S2_BChE_autonotify_setup",   "c_S2_BChE_autonotify_op",
  "Surveillance MVP (S1 + S2)",                          "c_surv_MVP_setup",             "c_surv_MVP_op",
  "S0 — NHLS-NMC IT feed (now operational)",             "c_S0_NHLS_NMC_feed_setup",     "c_S0_NHLS_NMC_feed_op",
  "S1b — event-based community (sensitivity, lo)",       "c_S1b_setup_lo",               "c_S1b_op_lo",
  "S1b — event-based community (sensitivity, mid)",      "c_S1b_setup_mid",              "c_S1b_op_mid",
  "S1b — event-based community (sensitivity, hi)",       "c_S1b_setup_hi",               "c_S1b_op_hi"
) %>%
  mutate(
    setup_zar = vapply(setup_param, function(n) as.numeric(pick(n)), numeric(1)),
    op_zar    = vapply(op_param,    function(n) as.numeric(pick(n)), numeric(1))
  ) %>%
  select(option, setup_zar, op_zar, setup_param, op_param)

marginal %>%
  gt(rowname_col = "option") %>%
  fmt_number(columns = c(setup_zar, op_zar), decimals = 0) %>%
  cols_label(
    setup_zar   = "Setup (R, one-off)",
    op_zar      = "Operating (R/yr)",
    setup_param = "CSV: setup",
    op_param    = "CSV: operating"
  ) %>%
  tab_options(table.font.size = 11)
```

:::

**Headline read.** The status-quo surveillance system already costs roughly **`r zar(sum(totals$total_zar))`/yr** in personnel + pesticide-allocated system spend across NMC, PIH, and NHLS BChE — borne almost entirely by NDoH, NHLS, and Stellenbosch FoMHS rather than appearing as a line item. Against that base, the **Surveillance MVP (S1 + S2) adds only `r zar(as.numeric(pick("c_surv_MVP_op")))`/yr** — a **`r sprintf("%.1f%%", 100 * as.numeric(pick("c_surv_MVP_op")) / sum(totals$total_zar))` uplift on current spend** for a **`r sprintf("+%.1f pp", (as.numeric(pick("d_surveillance_mvp")) - as.numeric(pick("d_status_quo"))) * 100)`** jump in P_detect. The marginal investment is small relative to the in-place baseline.

## 8. Implementation considerations

| Component | Lead agency | Year-1 cost | Critical path |
|:----------|:------------|------------:|:--------------|
| S1 PIH dashboard | PIH (Tygerberg) + NICD | `r scales::comma(as.numeric(pick("c_S1_PIH_MVD_op")))` | Daily extract + hosting; MVD workshop |
| S2 BChE auto-notify | NHLS LIS team | `r scales::comma(as.numeric(pick("c_S2_BChE_autonotify_op")))` | LIS rule configuration; QA review |
| S3 Sentinel toxicology | NHLS clinical chemistry + sentinel hospital | `r zar(as.numeric(pick("c_option6_pilot_year1")))` pilot | Lab method validation; sample logistics |
| S1b Phase 2 community layer *(pilot: FS + Gauteng; commission after Month 6 MVP review)* | NICD + provincial DoH EHPs | Pilot: `r sprintf("R%.0fM setup + R%.1fM/yr", as.numeric(pick("c_S1b_setup_lo"))/1e6, as.numeric(pick("c_S1b_op_lo"))/1e6)`; national: `r sprintf("R%.1fM setup + R%.1fM/yr", as.numeric(pick("c_S1b_setup_hi"))/1e6, as.numeric(pick("c_S1b_op_hi"))/1e6)` | CHW two-tier triage training, mobile checklist app, EHP call-back protocol, QA pipeline |

**Privacy & legal.** All flows are de-identified (age, sex, facility, result, date). POPIA-compliant under existing Section 90 NHA authority. **No new legislation required.**

**Risks.** (i) Low BChE testing uptake by clinicians — S2 fires only on tests ordered. (ii) Data-quality drift if facility metadata is incomplete. (iii) Capacity bottleneck at provincial response level — the [Coordination brief](coordination_policy.qmd) addresses this.

## 9. Recommended next step

**Adopt the Surveillance MVP (S1 + S2) and pilot in 3 provinces (Gauteng, Western Cape, KZN) for 6 months.** Approximate pilot budget: **~`r zar(as.numeric(pick("c_surv_MVP_op")))` operating** (NHLS–NMC IT feed is now operational). Lead agency: NICD as national coordinator.

- **Weeks 1–2:** S1 PIH dashboard live; S2 BChE LIS configuration begins.
- **Months 1–3:** S1 MVD workshop with PIH finalised; S2 BChE auto-notify QA review completed.
- **Month 6:** Evaluate against (a) completeness of NHLS severe results in NMC and (b) signal-to-investigation latency.
- **Month 9:** Add S3 sentinel toxicology site, funded from a separate research-surveillance stream.
- **Month 12:** Review MVP geographic distribution. If ≥40% of severe cases originate in low-facility-access districts, commission the S1b Phase 2 pilot (Free State + Gauteng; ~`r sprintf("R%.0fM", as.numeric(pick("c_S1b_setup_lo"))/1e6)` setup + `r sprintf("R%.1fM/yr", as.numeric(pick("c_S1b_op_lo"))/1e6)` operating).

The coordination side of the same system — how NICD, EHPs, DALRRD, and provincial response teams act on the improved signal — is covered in the [Coordination brief](coordination_policy.qmd). The product-specific regulatory action that S3 makes possible is the subject of the [Terbufos brief](terbufos_policy.qmd).

---

::: {.footer-cta}
**Companion briefs:** [Coordination](coordination_policy.qmd) · [Terbufos restriction](terbufos_policy.qmd) · [Decision Model](decision_model.qmd) · [Costing Analysis](costing_analysis.qmd)

**Technical appendices:** [A — Surveillance decision analysis](../analysis_report_v2.qmd) · [D — Sentinel toxicology (S3 deep dive)](surveillance_concept_note.qmd)
:::
