---
title: "Policy Brief 3: Restricting or Banning Terbufos in South Africa"
subtitle: "A priority-product action under the surveillance + coordination framework"
author: "D2P Project Team"
date: "2025-09-05"
date-modified: today
categories: [policy, regulation, terbufos, decision-makers]
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: false
---

::: {.callout-important icon=false}
## For Decision-Makers

**Target audience:** DALRRD Pesticide Registrar, Ministerial Advisory Committee (MAC), National Department of Health (NDoH), Department of Trade, Industry & Competition (substitution support).

**Companion briefs:** [Surveillance](surveillance_policy.qmd) · [Coordination](coordination_policy.qmd).

**Status:** **DRAFT.** One genuine SA evidence gap remains (the adult **live-patient** attributable fraction, resolved only by Surveillance Option S3). The other quantitative inputs — market volume, yield impact, substitute hazard, substitute pricing, enforcement cost — are **not epistemic gaps**: they are either administrative numbers DALRRD can pull, or values defensibly anchored to the EU/US/Sri Lanka evidence base (see [§2b](#international-natural-experiment-evidence) and [Interim assumptions](#interim-assumptions-anchored-to-international-evidence)).
:::

```{r}
#| label: headline-numbers
#| echo: false
#| message: false
#| warning: false

# Pull all headline inputs from the canonical CSV — no inline magic numbers.
source("../R/terbufos_tree.r")  # also sources decision_tree_engine.r
params <- load_params()

f_lo    <- params$frac_terbufos_attributable_lo
f_mid   <- params$frac_terbufos_attributable_mid
f_hi    <- params$frac_terbufos_attributable_hi
phi     <- params$frac_T3_burden_retired
N_d     <- params$n_deaths_headline
N_cases <- params$n_true_burden_from_statssa_mid
C_tot   <- params$C_burden_total_headline
c_enf   <- params$c_T2_RUP_enforcement_mid

# Headline averted (working / mid)
deaths_av_mid    <- f_mid * phi * N_d
deaths_av_lo     <- f_lo  * phi * N_d
deaths_av_hi     <- f_hi  * phi * N_d
poisonings_av_mid <- f_mid * phi * N_cases
poisonings_av_lo  <- f_lo  * phi * N_cases
poisonings_av_hi  <- f_hi  * phi * N_cases
cost_av_mid      <- f_mid * phi * C_tot
cost_av_lo       <- f_lo  * phi * C_tot
cost_av_hi       <- f_hi  * phi * C_tot
leverage_mid     <- cost_av_mid / c_enf

fnum <- function(x) format(round(x), big.mark = ",")
fbn  <- function(x) sprintf("%.2f", x / 1e9)
fmn  <- function(x) sprintf("%.0f", x / 1e6)
fpct <- function(x) sprintf("%.0f%%", 100 * x)
```

::: {.callout-tip icon=false}
## Headlines — what banning terbufos buys SA per year

> **`r fnum(deaths_av_mid)` deaths averted/yr**  *(range `r fnum(deaths_av_lo)`–`r fnum(deaths_av_hi)`)*<br>
> **`r fnum(poisonings_av_mid)` poisonings averted/yr**  *(range `r fnum(poisonings_av_lo)`–`r fnum(poisonings_av_hi)`)*<br>
> **~R`r fbn(cost_av_mid)`B/yr in retired economic burden**  *(range R`r fbn(cost_av_lo)`–R`r fbn(cost_av_hi)`B)*<br>
> **at a DALRRD enforcement cost of ~R`r fmn(c_enf)`M/yr** — a household-burden-retired : regulator-spend leverage ratio of roughly **`r fnum(leverage_mid)` : 1**.

**Arithmetic (working / mid value).** Every number above is the product of three CSV-driven inputs and a stakeholder share:

| Quantity | Formula | Working value |
|:---|:---|---:|
| Deaths averted | $f_\text{attr} \times \varphi_\text{retired} \times N_\text{deaths}$ | `r fpct(f_mid)` × `r fpct(phi)` × `r fnum(N_d)` = **`r fnum(deaths_av_mid)`/yr** |
| Poisonings averted | $f_\text{attr} \times \varphi_\text{retired} \times N_\text{true cases}$ | `r fpct(f_mid)` × `r fpct(phi)` × `r fnum(N_cases)` = **`r fnum(poisonings_av_mid)`/yr** |
| Burden retired | $f_\text{attr} \times \varphi_\text{retired} \times C_\text{burden total}$ | `r fpct(f_mid)` × `r fpct(phi)` × R`r fbn(C_tot)`B = **R`r fbn(cost_av_mid)`B/yr** |
| Regulator spend | `c_T2_RUP_enforcement_mid` | **R`r fmn(c_enf)`M/yr** |
| Leverage | $C_\text{burden retired} \div C_\text{enforcement}$ | R`r fbn(cost_av_mid)`B ÷ R`r fmn(c_enf)`M ≈ **`r fnum(leverage_mid)` : 1** |

Where: $f_\text{attr}$ = terbufos attributable fraction of fatal SA pesticide events (Davies 2023 + Gunnell 2017); $\varphi_\text{retired}$ = fraction of attributable burden retired post-ban (Sri Lanka substitution-incompleteness, `r fpct(phi)`); $N_\text{deaths}$, $N_\text{true cases}$, $C_\text{burden total}$ = StatsSA MACOD 2023 + implied true burden + total annual economic burden.
:::

## 1. Policy question

**Should South Africa ban terbufos as a registered agricultural pesticide?**

## 2. Public health problem

Terbufos is a **WHO Class Ia ("Extremely Hazardous")** organophosphate insecticide,[^who] with rat oral LD50 **1.6–4.5 mg/kg** — among the most acutely toxic active ingredients in commercial use anywhere.[^epa] Sub-gram doses are lethal in adults; very small quantities are lethal in children.

**The SA-specific evidence pinning the harm to this agent.** Davies et al. (2023) identified **terbufos and methamidophos as the dominant agents in fatal SA childhood pesticide poisoning**, with most deaths following ingestion of granular formulations diverted into the informal "street-pesticide" market.[^davies] NICD/NMC mass-poisoning reports 2022–2024 are independently consistent.[^nicd]

**International natural-experiment anchor.** Sri Lanka's sequential Class Ia withdrawals (1995–2011) retired **20–40 %** of fatal pesticide poisonings per banned agent without compensating method substitution.[^gunnell] EU (2009) withdrew terbufos with **no documented maize-yield collapse** in any producer state. Davies 2023 places SA terbufos in the same 20–40 % bracket on the paediatric data alone.

[^gunnell]: Gunnell D, Knipe D, Chang S-S, Pearson M, Konradsen F, Eddleston M et al. *Prevention of suicide with regulations aimed at restricting access to highly hazardous pesticides: a systematic review of the international evidence.* Lancet Glob Health 2017;5:e1026–e1037. Knipe DW et al. *Suicide prevention through means restriction: impact of the 2008–2011 pesticide restrictions on suicide in Sri Lanka.* PLOS One 2017.

[^who]: World Health Organization. *The WHO Recommended Classification of Pesticides by Hazard and Guidelines to Classification*, current edition. Terbufos is listed in Class Ia (Extremely Hazardous).

[^epa]: US Environmental Protection Agency. *Reregistration Eligibility Decision (RED): Terbufos.* EPA-738-R-06-006, September 2006. Rat acute oral LD50 1.6–4.5 mg/kg; rat acute dermal LD50 1.0–7.4 mg/kg.

[^davies]: Davies J et al. (2023). Paediatric pesticide deaths in South Africa; documents terbufos and methamidophos dominance among fatal cases. Cited in [glossary.qmd](../glossary.qmd#key-sources).

[^nicd]: NICD/NMC clinical case reports and public communications, 2022–2024, on mass-poisoning incidents linked to informally-sold granular organophosphate pesticides. Formal case-series compilation pending (open item TBF-2).

## 3. Status quo

{{< include _shared_parameters.qmd >}}

::: {.callout-note}
## DALRRD holds the lever — and bears almost none of the cost

DALRRD itself bears **~R0 of the R2.34B annual burden**. That cost is borne by NDoH (~R70M hospital), NHLS (~R10M lab), and most heavily by **households (~R2.26B in OOP costs and lost lifetime earnings)**. DALRRD nevertheless holds the *only* lever — registration, restriction, recall — that prevents the underlying exposure. Action borne by DALRRD (~R1–3M/yr enforcement) retires household-borne mortality cost at a leverage ratio of roughly **1 : 500**. This is the single highest-leverage regulatory action on the SA pesticide-poisoning portfolio.
:::

| Item | Status |
|:-----|:-------|
| SA registration | Registered as a soil insecticide (Act 36 of 1947) |
| WHO classification | Class Ia — Extremely Hazardous[^who] |
| Acute oral LD50 (rat) | 1.6–4.5 mg/kg[^epa] |
| EU | Not approved since 2009 |
| US EPA | Restricted-Use; granular formulations voluntarily cancelled[^epa] |
| SA paediatric mortality | Dominant agent in fatal childhood pesticide poisoning[^davies] |

## 4. Root causes of disproportionate terbufos harm

```{mermaid}
%%| label: fig-terb-fishbone
%%| fig-cap: "Root causes that make terbufos a disproportionate public-health hazard in SA. All five drivers funnel to the same regulatory remedy — banning terbufos."
flowchart LR
    A[Disproportionate<br/>terbufos-related harm] --> R1[High intrinsic toxicity<br/>WHO Class Ia]
    A --> R2[Granular formulation<br/>resembles food / seed]
    A --> R3[Diversion into informal<br/>retail / pest-control use]
    A --> R4[Limited live-patient<br/>agent attribution]
    A --> R5[Substitutes available but<br/>switching not mandated]

    R1 --> BAN[Ban terbufos<br/><i>Option T2</i>]:::ban
    R2 --> BAN
    R3 --> BAN
    R5 --> BAN
    R4 -.-> S3[Surveillance S3<br/><i>refines attribution<br/>but not a prerequisite</i>]:::aux

    classDef ban fill:#9f9,stroke:#060,stroke-width:2px
    classDef aux fill:#bbf,stroke:#333
    style A fill:#f99,stroke:#333
```

Four of the five drivers (intrinsic hazard, formulation, informal-market diversion, substitution readiness) are addressed by a single regulatory act — withdrawing terbufos from the SA pesticide register. The fifth driver (live-patient attribution) is sharpened by [Surveillance Option S3](surveillance_policy.qmd), but the existing paediatric attribution (Davies 2023) and the Sri Lanka international evidence are already sufficient to act.

## 5. Policy options

The action space collapses to two options:

### Option T1 — Status quo (no change)

Terbufos remains registered under existing conditions. The system continues to rely on general pesticide-handling regulations and clinician/EHP-led NMC notifications to detect harm *after* it occurs. **This is the implicit decision if no regulatory action is taken.**

### Option T2 — Ban terbufos *(★ recommended)*

Cancel terbufos product registrations under the Fertilizers, Farm Feeds, Agricultural Remedies and Stock Remedies Act, with a **12–24 month phase-out window** to allow substitute uptake and existing stock to clear. Aligns SA with the EU 2009 position. During the phase-out, terbufos is moved to Restricted-Use status (sales only via certified applicators with logged transactions) so the regulatory mechanism is operational from gazette day one.

*Intermediate restriction options (RUP, formulation restriction, distributor audit) are best understood as the **mechanics of the phase-out**, not as alternative end-states. The end-state is withdrawal.*

## 6. Decision tree — deaths, burden, and yield under each option

````{r}
#| label: fig-terb-decision-tree
#| echo: false
#| output: asis

# Source the decision-tree engine and the terbufos tree builder. Numbers
# below come from the rollback of the tree object built off
# amua_import_parameters_v3.csv -- no inline arithmetic.
source("../R/terbufos_tree.r")

params  <- load_params()
tree    <- build_terbufos_tree(params)
res     <- evs_to_df(eval_tree(tree))
averted <- compute_avertedness(res)

# Parameters still pulled directly from the named list -- these drive the
# bracket annotations in the diagram caption and the lo/hi range labels.
f_lo  <- params$frac_terbufos_attributable_lo
f_mid <- params$frac_terbufos_attributable_mid
f_hi  <- params$frac_terbufos_attributable_hi
phi   <- params$frac_T3_burden_retired
N_deaths_year <- params$n_deaths_headline
burden_total  <- params$C_burden_total_headline

# Status-quo (T1) terbufos-attributable burden, pulled from the tibble.
sq <- res %>% filter(arm == "status_quo")
deaths_attrib_mid <- round(sq$deaths_per_yr)
deaths_attrib_lo  <- round(f_lo * N_deaths_year)
deaths_attrib_hi  <- round(f_hi * N_deaths_year)
burden_attrib_mid <- round(sq$burden_zar_per_yr / 1e9, 2)

# Deaths and burden averted under the ban, pulled from compute_avertedness().
da <- averted %>% filter(metric == "deaths/yr")
ba <- averted %>% filter(metric == "burden ZAR/yr")
deaths_avert_mid  <- round(da$averted)
deaths_avert_lo   <- round(f_lo * phi * N_deaths_year)
deaths_avert_hi   <- round(f_hi * phi * N_deaths_year)
burden_avert_mid  <- round(ba$averted / 1e9, 2)
burden_avert_lo   <- round(f_lo * phi * burden_total / 1e9, 2)
burden_avert_hi   <- round(f_hi * phi * burden_total / 1e9, 2)

# Residual deaths after the ban (substitution-incompleteness)
deaths_residual_mid <- deaths_attrib_mid - deaths_avert_mid

cat(sprintf('```{mermaid}
%%%%| fig-cap: "Two-node terbufos decision tree. The split is binary — ban or no ban. Each branch terminates in three parallel outcomes: deaths, monetised burden, and maize yield. Deaths-averted arithmetic: f_attr × φ_retired × N_deaths_total = %.2f × %.2f × %s = %.0f/yr (working). Attributable fraction range %.0f–%.0f%% from Davies 2023 + Gunnell 2017; substitution-incompleteness %.0f%% from Sri Lanka."
flowchart TD
    Q{Ban terbufos?} -->|No — T1<br/>Status quo| T1[Continued registration]:::bad
    Q -->|Yes — T2<br/>Cancel registration| T2[12–24 month phase-out]:::good

    T1 --> D1["<b>Deaths</b><br/>~%.0f/yr terbufos-attributable<br/>~range %.0f–%.0f~<br/>= %.0f%% × %s total"]:::bad
    T1 --> B1["<b>Burden</b><br/>~R%.2fB/yr terbufos-attributable<br/>retained on households + NDoH"]:::bad
    T1 --> Y1["<b>Maize yield</b><br/>No change"]:::neutral

    T2 --> D2["<b>Deaths averted</b><br/>~%.0f/yr<br/>~range %.0f–%.0f~<br/>= %.0f%% × %.0f%% × %s"]:::good
    T2 --> B2["<b>Burden retired</b><br/>~R%.2fB/yr<br/>~range R%.2f–R%.2fB~"]:::good
    T2 --> Y2["<b>Maize yield</b><br/><5%% impact<br/>~see §7~"]:::neutral

    D2 --> R2["Residual ~%.0f deaths/yr<br/>substitution-incompleteness<br/>~Sri Lanka %.0f%%~"]:::warn

    classDef bad fill:#fdd,stroke:#cc0000,stroke-width:1.5px
    classDef good fill:#cfc,stroke:#060,stroke-width:2px
    classDef warn fill:#ffe9b3,stroke:#cc8800,stroke-width:1.5px
    classDef neutral fill:#eee,stroke:#666
```
',
f_mid, phi, format(round(N_deaths_year), big.mark = ","), deaths_avert_mid,
100*f_lo, 100*f_hi, 100*phi,
deaths_attrib_mid, deaths_attrib_lo, deaths_attrib_hi, 100*f_mid, format(round(N_deaths_year), big.mark = ","),
burden_attrib_mid,
deaths_avert_mid, deaths_avert_lo, deaths_avert_hi, 100*f_mid, 100*phi, format(round(N_deaths_year), big.mark = ","),
burden_avert_mid, burden_avert_lo, burden_avert_hi,
deaths_residual_mid, 100*(1 - phi)
))
````

**Reading the tree.** The decision is binary because the upstream driver — high-intrinsic-toxicity granular OP in informal circulation — is binary: either the product is on the register or it isn't. Each branch terminates in three parallel outcomes:

- **Deaths** — the headline public-health metric. Computed as $f_\text{attr} \times \varphi_\text{retired} \times N_\text{deaths}$. The working value pulls from three CSV parameters:
  - `frac_terbufos_attributable_mid` = `r sprintf("%.0f%%", 100*as.numeric(pick("frac_terbufos_attributable_mid")))` (range `r sprintf("%.0f–%.0f%%", 100*as.numeric(pick("frac_terbufos_attributable_lo")), 100*as.numeric(pick("frac_terbufos_attributable_hi")))`) — SA paediatric attribution (Davies 2023) + Sri Lanka single-agent withdrawal bracket (Gunnell 2017).
  - `frac_T3_burden_retired` = `r sprintf("%.0f%%", 100*as.numeric(pick("frac_T3_burden_retired")))` — fraction of attributable burden retired post-ban, anchored to Sri Lanka substitution-incompleteness. The complement is the *residual* node — deaths that persist because some users will switch to other Class I OPs rather than to less-hazardous substitutes.
  - `n_deaths_headline` = `r format(round(as.numeric(pick("n_deaths_headline"))), big.mark = ",")` — StatsSA MACOD 2023 fatal pesticide-poisoning total.
- **Burden** — the same multiplication applied to the R`r sprintf("%.2f", as.numeric(pick("C_burden_total_headline"))/1e9)`B total annual burden. Working value: **~R`r round(as.numeric(pick("frac_terbufos_attributable_mid")) * as.numeric(pick("frac_T3_burden_retired")) * as.numeric(pick("C_burden_total_headline"))/1e9, 2)`B/yr retired**, falling on households + NDoH at the canonical share split.
- **Maize yield** — bounded at <5 % (see §7).

The full closed-form derivation, with the deaths-averted bracket calculated under every $f_\text{attr} \times \varphi_\text{retired}$ combination, lives in [Technical Appendix C](../technical_appendix_terbufos.qmd).

## 6b. Alternative view — per-individual decision tree (Amua-style)

The tree in §6 is a **population-aggregate** model: it takes the StatsSA mortality total as given and asks what share is retired by a ban. An alternative — drafted in Amua — is a **per-individual** tree that walks a single SA resident through one year, splits on whether they experience a pesticide poisoning, then splits on whether the agent is terbufos, and finally splits on clinical outcome (died / hospitalised / asymptomatic). Each terminal carries a per-person cost; the rollback returns expected deaths, hospitalisations and ZAR cost **per individual per year**, which can be multiplied by the SA population to recover annual national totals.

The two trees answer the same policy question from opposite ends. The aggregate tree is more compact and audits cleanly against StatsSA; the per-individual tree exposes the *clinical-pathway logic* (CFR by agent) that drives the effect, and makes the personal stakes of registration policy explicit. The numbers below come from the same CSV-driven parameter set as §6.

````{r}
#| label: fig-terb-tree-personal
#| echo: false
#| output: asis

source("../R/terbufos_tree_personal.r")

params_p <- load_params()
tree_p   <- build_terbufos_tree_personal(params_p)
res_p    <- evs_to_df(eval_tree(tree_p), decision_name = "ban_terbufos_personal")
av_p     <- compute_avertedness_personal(res_p, params_p$sa_population_ref)

# Pull mid values for the diagram labels.
ppe_sq   <- params_p$p_p_exposed_sq
ppe_ban  <- params_p$p_p_exposed_ban
pterb_sq <- params_p$p_terbufos_among_poisoned_sq
pterb_bn <- params_p$p_terbufos_among_poisoned_ban
pd_t     <- params_p$p_die_given_terbufos
ph_t     <- params_p$p_hosp_given_terbufos
pa_t     <- params_p$p_asymp_given_terbufos
pd_o     <- params_p$p_die_given_other
ph_o     <- params_p$p_hosp_given_other
pa_o     <- params_p$p_asymp_given_other

# Averted (national) — for caption.
d_av <- av_p %>% filter(metric == "deaths/yr (national)")
h_av <- av_p %>% filter(metric == "hospitalisations/yr (national)")
c_av <- av_p %>% filter(metric == "cost ZAR/yr (national)")

cat(sprintf('```{mermaid}
%%%%| fig-cap: "Per-individual decision tree (Amua draft). An SA resident is followed for one year. Probabilities are CSV-driven (see <code>terb.personal_tree</code> block). Scaled to a population of %s, this tree gives **%s deaths averted/yr**, **%s hospitalisations averted/yr**, and **~R%sB/yr** in retired cost under a ban. Compare with the aggregate tree in §6: the two trees should give numbers in the same order of magnitude."
flowchart LR
    Q{Ban terbufos?}

    Q -->|No — T1<br/>Status quo| SQ[Status quo]:::bad
    Q -->|Yes — T2<br/>Cancel registration| BN[Ban]:::good

    SQ -->|p=%.6f<br/>Poisoning exposed| SQ_P[Poisoned]
    SQ -->|1 - p<br/>No poisoning| SQ_N[Not poisoned]:::neutral
    SQ_P -->|%.2f<br/>Terbufos| SQ_T[Outcome | terbufos]:::bad
    SQ_P -->|%.2f<br/>Other agent| SQ_O[Outcome | other]
    SQ_T --> SQ_TD["Died %.3f"]:::bad
    SQ_T --> SQ_TH["Hospitalised %.3f"]:::warn
    SQ_T --> SQ_TA["Asymptomatic %.3f"]:::neutral
    SQ_O --> SQ_OD["Died %.3f"]:::warn
    SQ_O --> SQ_OH["Hospitalised %.3f"]:::neutral
    SQ_O --> SQ_OA["Asymptomatic %.3f"]:::neutral

    BN -->|p=%.6f<br/>Poisoning exposed| BN_P[Poisoned]
    BN -->|1 - p<br/>No poisoning| BN_N[Not poisoned]:::neutral
    BN_P -->|%.2f<br/>Terbufos residual| BN_T[Outcome | terbufos]:::warn
    BN_P -->|%.2f<br/>Other agent| BN_O[Outcome | other]:::good
    BN_T --> BN_TD["Died %.3f"]:::bad
    BN_T --> BN_TH["Hospitalised %.3f"]:::warn
    BN_T --> BN_TA["Asymptomatic %.3f"]:::neutral
    BN_O --> BN_OD["Died %.3f"]:::warn
    BN_O --> BN_OH["Hospitalised %.3f"]:::neutral
    BN_O --> BN_OA["Asymptomatic %.3f"]:::neutral

    classDef bad fill:#fdd,stroke:#cc0000,stroke-width:1.5px
    classDef good fill:#cfc,stroke:#060,stroke-width:2px
    classDef warn fill:#ffe9b3,stroke:#cc8800,stroke-width:1.5px
    classDef neutral fill:#eee,stroke:#666
```
',
format(params_p$sa_population_ref, big.mark = ","),
format(round(d_av$averted), big.mark = ","),
format(round(h_av$averted), big.mark = ","),
sprintf("%.2f", c_av$averted / 1e9),
ppe_sq, pterb_sq, 1 - pterb_sq, pd_t, ph_t, pa_t, pd_o, ph_o, pa_o,
ppe_ban, pterb_bn, 1 - pterb_bn, pd_t, ph_t, pa_t, pd_o, ph_o, pa_o
))
````

**Per-individual rollback (expected values per SA resident per year).**

```{r}
#| label: tbl-terb-tree-personal-pp
#| echo: false
knitr::kable(
  res_p %>%
    mutate(across(where(is.numeric), ~signif(.x, 3))),
  caption = "Expected deaths, hospitalisations, asymptomatic events, and ZAR cost per individual per year, by arm. Multiply by the SA population reference to get annual national totals."
)
```

**Scaled to SA population.**

```{r}
#| label: tbl-terb-tree-personal-national
#| echo: false
knitr::kable(
  av_p %>%
    mutate(status_quo    = format(round(status_quo), big.mark = ","),
           ban           = format(round(ban),        big.mark = ","),
           averted       = format(round(averted),    big.mark = ","),
           pct_reduction = sprintf("%.1f%%", 100 * pct_reduction)),
  caption = sprintf("National annual totals under each arm, and the reduction (status quo - ban) delivered by an enforced ban. Compare with the §6 headline of ~%s deaths averted/yr from the aggregate tree.",
                    format(round(f_mid * phi * N_d), big.mark = ","))
)
```

**Reading this against §6.** The two trees use the same upstream data but parameterise the effect differently:

- The **§6 aggregate tree** multiplies the StatsSA total by $f_\text{attr} \times \varphi_\text{retired}$. The effect size is bounded by the attributable-fraction literature (Davies 2023, Gunnell 2017).
- The **per-individual tree** above derives the effect from the *case-fatality differential* between terbufos (`r sprintf("%.1f%%", 100*pd_t)`) and substitute agents (`r sprintf("%.1f%%", 100*pd_o)`), applied to the agent-mix shift (`r sprintf("%.0f%% → %.0f%%", 100*pterb_sq, 100*pterb_bn)`).

If the per-individual tree gives a *much larger* number than §6, the implication is that the personal-tree's CFR-by-agent inputs are more optimistic than the aggregate attributable-fraction literature supports — i.e. the personal tree is implicitly assuming a stronger ban effect. If it gives a *similar* number, the two views corroborate each other. The audience-facing brief should pick ONE — the aggregate view is usually safer for regulator audiences; the per-individual view is more impactful for clinician / parliamentary audiences.

## 7. Crop-yield impact (brief)

A ban affects production primarily through soil-insect / stalk-borer control on maize. The expected SA-specific yield impact is **<5 %**, concentrated in **non-Bt smallholder maize**, for three independent reasons:

1. **SA maize is ~85 % Bt-trait.** The dominant stalk-borer use case is largely retired biologically before any chemical intervention.
2. **Registered substitutes exist** for the residual soil-insect / nematode use case (chlorantraniliprole, thiamethoxam, imidacloprid, biological controls). These are **100–1,000× less acutely toxic** than terbufos on rat oral LD50 — the acute-poisoning burden retired is therefore essentially the full terbufos attributable share, not a redistribution to a similar-hazard substitute.
3. **EU 2009 withdrawal produced no documented maize-yield collapse** in any producer state — a directly relevant natural experiment.

Substitute input cost is roughly **2–5× per hectare** vs commodity granular OPs, but maize gross output is ~R20,000/ha, so the input-cost delta is single-digit-% of gross output. Smallholder transition support (DALRRD grower advisory + targeted subsidy on substitute inputs for the first two seasons) addresses the access constraint.

The detailed yield-impact derivation, with sensitivity ranges on Bt coverage, substitute pricing, and smallholder access, is presented in the supplementary technical note: [Technical Appendix C, §4 — Maize yield impact](../technical_appendix_terbufos.qmd#sec-yield-impact).

## 8. Recommended next step

**Adopt Option T2 (ban) with a 24-month phase-out.**

1. **Month 0–3.** DALRRD gazette intent to cancel registration; immediate RUP designation as the phase-out mechanism (no informal-market sales from day one).
2. **Month 3–6.** Public-comment process; DALRRD–grain-industry transition plan including smallholder substitute subsidy.
3. **Month 6–12.** [Surveillance S3](surveillance_policy.qmd) sentinel toxicology stands up to track attributable fraction during the phase-out (refines the bracket; does not gate the decision).
4. **Month 12–24.** Final cancellation; existing stock cleared; DALRRD inspectorate audit confirms market withdrawal.
5. **Month 24.** Joint NDoH–DALRRD evaluation against pre-agreed reduction in attributable serious-poisoning cases.

This is the public-health action with the highest demonstrated leverage on the SA pesticide-poisoning portfolio, supported by both SA-specific paediatric data (Davies 2023) and a directly applicable international natural experiment (Sri Lanka, EU). The agricultural cost is small, bounded, and addressable through standard substitution support.

---

::: {.footer-cta}
**Companion briefs:** [Surveillance](surveillance_policy.qmd) · [Coordination](coordination_policy.qmd) · [Decision Model](decision_model.qmd) · [Costing Analysis](costing_analysis.qmd)

**Technical appendix:** [C — Terbufos bracketed deaths-averted estimate and yield-impact model](../technical_appendix_terbufos.qmd)
:::
